### 2019-06-19 submission/manuscript.Rmd
Add references and put manuscript into .Rmd

---
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
geometry: margin=1.0in
---

```{r knitr_settings, tidy=TRUE, eval=TRUE, echo=FALSE, cache=FALSE}
options(tidyverse.quiet = TRUE)

library(tidyverse)
library(knitr)

opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("message" = FALSE)
opts_chunk$set("cache" = FALSE)

inline_hook <- function(x){
	print(x)

	if(is.list(x)){
		x <- unlist(x)
	}
	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps^0.5){
			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
		} else {
			paste(format(x,big.mark=',', digits=1, nsmall=1, scientific=FALSE))
		}
	} else {
    	paste(x)      
	}
}
knitr::knit_hooks$set(inline=inline_hook)
```

\vspace{35mm}

# Proton pump inhibitor administration does not promote *Clostridium difficile* colonization in a murine model

\vspace{35mm}

Sarah Tomkovich${^1}$, Nicholas A. Lesniak${^1}$, Yuan Li${^1}$, Lucas Bishop${^1}$, Madison J. Fitzgerald${^1}$, Patrick D. Schloss^1$\dagger$^

\vspace{40mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

$1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109

\newpage
\linenumbers

## Abstract
Proton pump inhibitor (PPI) use has been associated with microbiota alterations and susceptibility to *Clostridium difficile* infections (CDIs) in humans. We assessed how PPI treatment alters the fecal microbiota and whether PPIs promote CDIs using a CDI mouse model. Mice were treated with the PPI omeprazole for 7 days prior to *C. difficile* challenge and were compared to mice that were treated with the antibiotic clindamycin or both omeprazole and clindamycin. We found that omeprazole was not sufficient to promote *C. difficile* colonization. When omeprazole treatment was combined with the antibiotic, there was a mixed level of *C. difficile* colonization, where one cage of mice remained resistant and the other cage was colonized. 16S rRNA sequencing analysis revealed that omeprazole had minimal impact on the murine microbiota throughout the 16 days of PPI exposure. *C. difficile* colonization of clindamycin-treated mice was associated with changes in *Alistipes*, *Barnesiella*, Porphyromonadaceae, and Ruminococcaceae. These results suggest PPI treatment alone is not sufficient to disrupt microbiota resistance to C. difficle infection in mice that are normally resistant in the absence of antibiotic treatment.

## Importance
Antibiotics are a major risk factor for *Clostridium difficile* infections (CDIs), but it is less clear what other factors contribute to CDI incidence, severity, and recurrence. However, other medications besides antibiotics have been shown to alter the microbiota. Proton pump inhibitors (PPIs) have been shown to alter the human intestinal microbiota in observational and interventional stduies and PPI use has also been associated with CDIs. We evaluated the ability of the PPI omeprazole to alter the murine intestinal microbiota and disrupt resistance to *C. difficile*. We found PPI treatment had minimal impact on the murine intestinal microbiota and was not enough of a disruption to promote C. difficile colonization after challenge. Further studies are needed to determine whether other factors such as the composition of the staring bacterial community, comorbidities, and use of additional medications contribute to the association between PPIs and CDIs.

\newpage

## Introduction
Proton pump inhibitors (PPIs) are one of the most widely used drugs and are available over the counter. PPIs are approved to treat a variety of conditions including gastroesophageal reflux disease, ulcers, and prophylaxis for nonsteroidal anti-inflammatory drug (NSAID)-associated gastrointestinal bleeding [@nehra2018proton]. Unfortunately, there have been a number of adverse events associated with their use including small intestinal bacterial overgrowth, decreased calcium absorption, and increased risk of enteric infections [@nehra2018proton]. There’s also been a number of large cohort studies as well as interventional clinical trials that have demonstrated specific alterations in the intestinal microbiome associated with PPI use [@Imhann2017; @Naito2018]. PPI-associated microbiota changes have been attributed to the PPIs ability to increase stomach pH which may promote the survival of oral and pathogenic bacteria. Human fecal microbiota changes with PPI use include increases in Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae with decreases in Ruminococcaceae [@Imhann2017]. An in vitro screen, demonstrated that 3 PPIs inhibit the growth of 7-8 strains of Lachnospiraceae and Ruminococcaceae, which represented ~16% of the 40 intestinal bacteria strains that were screened [@maier2018extensive].
Antibiotics have a large impact on the intestinal microbiome and are a primary risk factor for developing CDIs [@Schubert2015]. It is less clear whether other human medications that impact the microbiota also influence *C. difficile* colonization resistance. Multiple epidemiological studies have suggested an association between PPI use and incidence of Clostridium difficile infections (CDIs) [@tariq2017association; @nehra2018proton; @Naito2018]. However, most of the studies suggesting a link between PPIs and *C. difficile* are retrospective and did not evaluate the microbiome [@tariq2017association; @nehra2018proton]. Thus, it is unclear whether the gastrointestinal microbiome changes associated with PPI use play a role in the association between PPIs and CDI incidence, severity or recurrence. Additionally, epidemiological studies have a limited capacity to address potential confounders and comorbidities in patients that were on PPIs and developed CDI infections. 
Although there have been a few *C. difficile* mouse model studies that have demonstrated PPIs have some effect on *C. difficile* colonization with or without additional antibiotic treatment, none of these studies looked at the interplay with the intestinal microbiome [@hung2015proton; @Kaur2007; @kaur2011effect]. Here, we evaluated the impact of a PPI on the murine microbiome and susceptibility to C. difficile colonization in relation to clindamycin, an antibiotic that perturbs the microbiome enough to allow C. difficile to colonize but is mild enough that C. difficile is cleared within 10 days [@Jenior2018].

## Results

**Murine fecal microbiomes were minimally affected by PPI treatment **
To test if PPI treatment alters the microbiome and promotes susceptibility to CDIs, we gavaged mice daily with a high dose of omeprazole for 7 days before C. difficile challenge (Figure 1A) and compared the PPI-treated mice to a group that received the antibiotic, clindamycin, and a group that received both PPI and clindamycin treatment. A principle coordinates analysis (PCoA) of the Bray-Curtis distances over the initial 7 days of treatment revealed the bacterial communities of PPI-treated mice remained relatively unchanged (Figure 1B). In contrast, the microbiomes from the 2 groups of mice that received clindamycin started to shift away from the untreated and PPI-treated mice the day after treatment (Figure 1B).

**PPI-treatment did not promote susceptibility to Clostridium difficile infection in mice**
After 7 days of PPI treatment or 1 day after clindamycin treatment, mice were challenged with C. difficile 630 spores. While all 4 of the clindamycin-treated mice were colonized with C. difficile, all of the PPI-treated mice were resistant to C. difficile colonization (Figure 2A). Interestingly, only 1 cage of the PPI and clindamycin-treated mice were colonized, while the other cage was not (Figure 2A). PCoAs of the bacterial communities from the 3 groups of mice after C. difficile challenge, showed that the greatest shifts in bacterial communities occurred in the clindamycin-treated mice (Figure 2B). Within 5 days, all of the mice had cleared C. difficile, suggesting there was no difference in rate of clearance for the mice that were initially colonized with C. difficile. Streptococcus, ___, ____ are all taxa that have previously been impacted by PPIs in human studies, but we see no significant fluctuations in the stool of our PPI-treated mice throughout the course of the 16-day experiment (Figure 2C). Together these results suggest that PPI treatment alone had minor effects on microbial community resistance to C. difficile colonization in mice. Instead most of the differences between our 3 groups appear to be driven by clindamycin treatment and included decreased Alistipes, Barnesiella, Porhyromonadaceae, Ruminococcaceae (Schubert), taxa previously found to be altered in clindamycin-treated mice that were challenged with C. difficile.

## Discussion and Conclusions

Our results demonstrate the PPI omeprazole had minimal impact on the stool microbiomes of mice and did not promote susceptibility to C. difficile colonization. In contrast, clindamycin-treated mice were initially colonized, but cleared the infection within 10 days. Combining PPI and clindamycin treatment had a mixed effect, one cage was colonized with C. difficile while the other cage remained resistant.
Our findings that PPI treatment had minimal impact on the fecal microbiome are comparable to several other mouse studies that indicate PPIs have more of an effect on the small intestinal microbiota compared to the fecal microbiota (Yasutomi). Our findings differ from the taxa observed to be significantly impacted by PPI use in human studies, but a closer examinations suggest that the significance of these associations is driven by some individuals more than others and the effect size can be relatively small with overall differences on a PCoA plot difficult to distinguish (Imhann, Freedberg). Also, most of the studies to date compare different individuals (PPI users versus non-users) or the same individuals before and after treatment (Naito 2018, Imhann), unfortunately these limited microbiota snapshots may be skewed by day-to-day microbiota variations (Llorens-Rico & Raes Nature 2019), which become apparent in the longitudinally sampled mice. Enterococcus and Streptococcus are frequently mentioned in the context of human PPI use and human CDIs (Imhann 2017), however we did not find much of either taxa in our mice. 
Previous work that administered the PPI lansoprazole daily for 2 weeks to mice and then challenged with C. difficile demonstrated that PPI treatment alone resulted in detectable C. difficile CFUs in the stool 1 week after challenge, but also showed there was detectable C. difficile in mice not treated with antibiotics. The presence of C. difficile in mice that were not treated with either antibiotics or PPIs could be attributed to the higher dose of vegetative cells (108 CFU) used to challenge the mice and may partly explain their observation of C. difficile in PPI-treated mice. We have previously shown mice from are colony that are not given antibiotics are resistant to C. difficile 630 when challenged with 103 spores (Jenior 2017). Another group also used 106 spores or 107 vegetative cells to challenge antibiotic-treated mice or mice treated with both the PPI esomeprazole for 2 days and antibiotics and demonstrated the antibiotic/PPI-treated mice experience more severe CDI. Our study extended previous work examining PPIs and C. difficile in mice to examine the contribution of the intestinal microbiota. Unfortunately, we found the PPI omeprazole had minimal impact on the murine intestinal microbiota and did not promote C. difficile colonization. Our 16S rRNA sequencing suggests that Streptococcus and Enterococcus are rare taxa in our C57BL/6 mouse colony. If these taxa are important contributors, to the associations between PPIs and C. difficile infection in humans, this may be part of the reason we did not see an effect in our CDI mouse model. There could also be differences at the species level for taxa that are impacted with PPIs in humans compared to mice, which may also contribute to the association. Other murine starting communities or the addition of human-associated strains may be needed to determine whether PPIs can increase Enterococcus and Streptococcus and subsequently promote or worsen CDIs.  
Other factors such as age, body mass index, comorbidities, and use of other medications in human studies may also be contributing to the association between PPIs and CDIs. NSAID treatment can increase CDI severity in mice (Maseda 2019), given that PPIs are sometimes used in combination with NSAIDs, it may be worth examining  This study only looked at the impact of PPIs with or without antibiotics, but future work may want to incorporate age, other comorbidities and bacterial strains that are less common in mice to evaluate whether PPIs in conjunction with other factors can increase the risk of C. difficile.

## Materials and Methods

**Animals** All mouse experiments were performed with 7- to 12-week-old C57BL/6 male and female mice. 2-3 mice were housed per cage and male and female mice were housed separately. All animal experiments were approved by the University of Michigan Animal Care and Use Committee (IACUC) under protocol number PRO00006983.

**Drug treatments**
Omeprazole (Sigma Aldrich) was prepared in a vehicle solution of 40% polyethylene glycol 400 (Sigma-Aldrich) in phosphate buffered saline. Omeprazole was prepared from 20 mg/mL frozen aliquots and diluted to an 8mg/mL prior to gavage. All mice received ~40mg/kg omeprazole or vehicle solution once per day through the duration of the experiment starting 7 days before C. difficile challenge. One day prior to C. difficile challenge, mice received an intraperitoneal injection of 10 mg/kg clindamycin or sterile saline vehicle. All drugs were filter sterilized through a 0.22 micron syringe filter before administration to animals.

**C. difficile infection model**
Mice were challenged with C. difficile 630 seven days after the start of omeprazole treatment and one day after clindamycin treatment. Mice were challenged with 103 spores in ultrapure distilled water. Samples were collected for 16S rRNA sequencing or quantifying C. difficile CFU throughout the duration of the experiments at the indicated timepoints (Figure 1A). Samples for 16S rRNA sequencing were flash frozen in liquid nitrogen and stored at -80oC until DNA extraction, while samples for CFU quantification were transferred into an anaerobic chamber and serially diluted in PBS. Diluted samples were plated on TCCFA (taurocholate, cycloserine, cefoxitin, fructose agar) plates and incubated at 37oC for 24 hours under anaerobic conditions to quantify C. difficile CFU.

**16S rRNA gene sequencing** DNA for 16S rRNA gene sequencing was extracted from 10-50 mg fecal pellets from each mouse using the DNeasy Powersoil HTP 96 Kit (Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf). The V4 hypervariable region of the 16S rRNA gene was amplified with Accuprime Pfx DNA polymerase (Thermo Fisher Scientific)) using previously described custom barcoded primers (Kozich et al 2013). The 16S rRNA amplicon library was sequenced with the MiSeq (Illumina). Amplicons were cleaned up and normalized with the SequalPrep Normalization Plate Kit (ThermoFisher Scientific) and pooled amplicons were quantified with the KAPA library quantification kit (KAPA Biosystems).

**16S rRNA gene sequence analysis**Mothur (v1.40.5) was used for all sequence processing steps (Schloss2009) using the previously published protocol (Kozich2013).  R (v.3.5.1) was used to generate figures and perform statistical analysis.

**Statistical Analysis**

**Code availability** The code for all sequence processing and analysis step as well as an Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Tomkovich_PPI_XXXX_2019.

**Data availability** The 16S rRNA sequencing data have been deposited in the NCBI Sequence Read Archive (Accession no. _________)

\newpage

## Figures

**Figure 1. PPI treatment had minimal impact on the murine fecal microbiota** A. Mouse experiment timeline and logistics.

\newpage

## References

### 2019-06-19 code/merge_labels.R
Rename group names and subtract 7 from day column to match experiment setup where C. difficile is administered on day 0

@@ -33,7 +33,10 @@ metadata <- read_xlsx('data/raw/PPI_metadata.xlsx') %>%
   rename("Plate_label" = "Plate map label") %>% 
   # convert names of seq to match the format of the shared samples
   mutate(shared_names = gsub('^_{1,3}','', Plate_label),
-         shared_names = gsub('_', '+', shared_names)) 
+         shared_names = gsub('_', '+', shared_names)) %>% 
+  mutate(Group = case_when(Group == 'O+' ~ 'PPI', #make group names more readable
+                           Group == 'C+' ~ 'Clindamycin',
+                           Group == 'CO+' ~ 'Clindamycin + PPI')) 
   # select only the samples that are present in the shared data
 matching <- inner_join(metadata, shared_sample_names, by = c('shared_names' = 'Group')) 
   
@@ -58,6 +61,10 @@ ppi_cdiff_cfu <- metadata %>%
 #Join CFU data to metadata file
 output_df <- full_join(metadata,
                        ppi_cdiff_cfu, by = c('Mouse ID', 'day'))
+
+#Change day to match experimental setup where C. difficile challenge day equals day 0. Implement by subtracting 7 from each day.
+output_df <- output_df %>% 
+  mutate(day = day - 7)
 
 write.table(output_df, 'data/process/ppi_metadata.txt', 
   row.names = F, quote = F, sep = '\t')

### 2019-06-19 code/ppi_pcoa.R
Clean up code to account for updates to metadata.txt

@@ -21,10 +21,8 @@ pcoa <- read_tsv('data/mothur/ppi.opti_mcc.braycurtis.0.03.lt.ave.pcoa.axes') %>
   select(group, axis1, axis2) %>% rename(shared_names = group) %>% 
 	right_join(metadata, by = "shared_names") # join with cleaned up metadata for day, mouse, treatment
   # Select O+, C+, and CO+ groups. Convert Group letter labels to more descriptive labels
-pcoa <- pcoa %>% filter(Group != "NA") %>% 
-  mutate(Group = case_when(Group == 'O+' ~ 'PPI', 
-                           Group == 'C+' ~ 'Clindamycin',
-                           Group == 'CO+' ~ 'Clindamycin + PPI'))
+pcoa <- pcoa %>% filter(Group != "NA")
+
 #plot all samples
 pcoa %>%  
   ggplot(aes(x=axis1, y=axis2, color=Group, alpha = day)) +
@@ -67,7 +65,7 @@ pcoa %>% filter(Group == "Clindamycin + PPI") %>%
   theme(plot.title = element_text(hjust = 0.5))
 
 # plot for before abx
-pcoa %>% filter(day < 7) %>% 
+pcoa %>% filter(day < -1) %>% 
   ggplot(aes(x=axis1, y=axis2, color=Group, alpha = day)) +
   geom_point() +
   theme_classic() +
@@ -76,7 +74,7 @@ pcoa %>% filter(day < 7) %>%
   
 
 #plot before abx with D0 removed
-pcoa %>% filter(day < 7) %>% filter(day > 0) %>% 
+pcoa %>% filter(day < -1) %>% filter(day > -7) %>% 
   ggplot(aes(x=axis1, y=axis2, color=Group, alpha = day)) +
   geom_point() +
   theme_classic()+
@@ -84,7 +82,7 @@ pcoa %>% filter(day < 7) %>% filter(day > 0) %>%
   theme(plot.title = element_text(hjust = 0.5))
 
 # plot day after abx
-pcoa %>%	filter(day == 7) %>% 
+pcoa %>%	filter(day == 0) %>% 
   ggplot(aes(x=axis1, y=axis2, color=Group, alpha = day)) +
   geom_point() +
   theme_classic()+
@@ -92,7 +90,7 @@ pcoa %>%	filter(day == 7) %>%
   theme(plot.title = element_text(hjust = 0.5))
 
[…]

### 2019-06-19 code/ppi_taxa.R

@@ -18,12 +18,9 @@
 library(tidyverse)
 library(broom)
 
-# Import metadata into data frame and clean
+# Import metadata into data frame
 metadata <- read.table('data/process/ppi_metadata.txt', header = T, sep = '\t', stringsAsFactors = F) %>% 
-  filter(Group != "NA") %>% #Exclude the mock community
-  mutate(Group = case_when(Group == 'O+' ~ 'PPI', #make group names readable by humans
-                           Group == 'C+' ~ 'Clindamycin',
-                           Group == 'CO+' ~ 'Clindamycin + PPI'))
+  filter(Group != "NA") #Exclude the mock community
 
 # Import taxonomy into data frame and clean up taxonomy names
 taxonomy <- read_tsv(file="data/mothur/ppi.taxonomy") %>%
@@ -108,6 +105,14 @@ agg_genus_data <- agg_taxa_data %>%
   inner_join(., metadata, by = "shared_names") %>% 
   ungroup() 
 
+# Summarize relative abundance data for family level
+agg_family_data <- agg_taxa_data %>% 
+  group_by(shared_names, family) %>% 
+  summarize(agg_rel_abund=sum(rel_abund)) %>% 
+  # Merge relative abundance data to phyla data
+  inner_join(., metadata, by = "shared_names") %>% 
+  ungroup() 
+
 # Use top_n function to determine top 10 genera in our samples
 top_genus <- agg_genus_data %>% 
   group_by(genus) %>% 
@@ -184,6 +189,34 @@ agg_phylum_data %>%
   scale_y_log10(breaks=c(1e-4, 1e-3, 1e-2, 1e-1, 1), labels=c(1e-2, 1e-1, 1, 10, 100))+
   theme_classic()
 
+#Kruskal_wallis test for family differences across treatment groups with Benjamini-Hochburg correction 
+family_tests <- agg_family_data %>% 
+  group_by(family) %>% 
+  do(tidy(kruskal.test(agg_rel_abund~factor(Group), data=.))) %>% ungroup() %>% 
+  mutate(p.value.adj=p.adjust(p.value, method="BH")) %>% 
+  arrange(p.value.adj)
+
+#List the significant family based on treatment groups after Benjamini-Hochburg correction
[…]

### 2019-06-20 code/ppi_cfu_over_time.R
Save plots to make preliminary figures

@@ -1,53 +1,50 @@
 #Load packages
 library(tidyverse)
-library(dtw)
 library(vegan)
+library(cowplot)
 
-#Define input files
-cfu <- 'data/08_20_18_Omep.#2/08_20_18_Omep._#2_exp.xlsx'
+# Import metadata into data frame
+metadata <- read.table('data/process/ppi_metadata.txt', header = T, sep = '\t', stringsAsFactors = F) %>% 
+  filter(Group != "NA") #Exclude the mock community'
 
-# Define output files
-plot_file <- 'results/figures/ppi2_cfu.pdf'
-
-#Load in data
-ppi2_data <- readxl::read_excel("data/08_20_18_Omep.#2/08_20_18_Omep._#2_exp.xlsx", sheet = "cfu_final")
+#Function to have y-axis in scientific notation
+fancy_scientific <- function(l) {
+  # turn in to character string in scientific notation
+  l <- format(l, scientific = TRUE)
+  # quote the part before the exponent to keep all the digits
+  l <- gsub("^(.*)e", "'\\1'e", l)
+  # turn the 'e+' into plotmath format
+  l <- gsub("e", "%*%10^", l)
+  # return this as an expression
+  parse(text=l)
+}
 
 #Create data frame of CFU data by selecting group column & columns ending with CFU/g
-ppi2_cfu <- select(ppi2_data, Group, ends_with("CFU/g")) 
-#make sure Group column is a factor vector and all CFU/g are treated as numeric vectors:
-ppi2_cfu$Group <- as.factor(ppi2_cfu$Group)
-str(ppi2_cfu)
+ppi_cfu <- select(metadata, Group, ends_with("CFU.g")) %>% 
+  gather(Day, CFU, -Group) %>% 
+  mutate(Group=factor(Group, levels=c("Clindamycin", "Clindamycin + PPI", "PPI")),
+         Day = as.numeric(str_replace(Day, "^[^.](\\d\\d?)[\\w\\.\\s\\/]+", "\\1")),
+         Day = Day - 7, #Modify day notation so that C. difficile challenge day is represented as day 0.
[…]

### 2019-06-20 code/ppi_pcoa.R

@@ -13,6 +13,7 @@
 #####################
 
 library(tidyverse)
+library(cowplot)
 
 metadata <- read.table('data/process/ppi_metadata.txt', header = T, sep = '\t', stringsAsFactors = F)
 
@@ -65,12 +66,14 @@ pcoa %>% filter(Group == "Clindamycin + PPI") %>%
   theme(plot.title = element_text(hjust = 0.5))
 
 # plot for before abx
-pcoa %>% filter(day < -1) %>% 
+before_abx <- pcoa %>% filter(day < -1) %>% 
   ggplot(aes(x=axis1, y=axis2, color=Group, alpha = day)) +
   geom_point() +
   theme_classic() +
   labs(title="Before antibiotic treatment") +
-  theme(plot.title = element_text(hjust = 0.5))
+  theme(plot.title = element_text(hjust = 0.5))+
+  ggsave("results/figures/before_abx.png")
+  
   
 
 #plot before abx with D0 removed
@@ -79,15 +82,16 @@ pcoa %>% filter(day < -1) %>% filter(day > -7) %>%
   geom_point() +
   theme_classic()+
   labs(title="Before antibiotic treatment without initial baseline day") +
-  theme(plot.title = element_text(hjust = 0.5))
+  theme(plot.title = element_text(hjust = 0.5)))
 
 # plot day after abx
-pcoa %>%	filter(day == 0) %>% 
+before_plus_day_after_abx <- pcoa %>%	filter(day < 1) %>% 
   ggplot(aes(x=axis1, y=axis2, color=Group, alpha = day)) +
   geom_point() +
   theme_classic()+
-  labs(title="Day after antibiotic treatment") +
-  theme(plot.title = element_text(hjust = 0.5))
+  labs(title="Before antibiotic treatment + 1 day after") +
+  theme(plot.title = element_text(hjust = 0.5))+
+  ggsave("results/figures/before_plus_day_after_abx.png")
 
 # plot 1st 7 days: Includes abx treatment day
 pcoa %>% filter(day < 1) %>%	
@@ -103,5 +107,6 @@ pcoa %>% filter(day > 1) %>%
   geom_point() +
   theme_classic() +
   labs(title="After antibiotic treatment & C. difficile challenge") +
-  theme(plot.title = element_text(hjust = 0.5))
+  theme(plot.title = element_text(hjust = 0.5))+
+  ggsave("results/figures/after_abx_C.diff.png")
 

### 2019-06-20 code/ppi_taxa.R

@@ -17,6 +17,7 @@
 #Setup----
 library(tidyverse)
 library(broom)
+library(cowplot)
 
 # Import metadata into data frame
 metadata <- read.table('data/process/ppi_metadata.txt', header = T, sep = '\t', stringsAsFactors = F) %>% 
@@ -82,7 +83,7 @@ top_phyla <- agg_phylum_data %>%
   top_n(5, median) %>% 
   pull(phylum)
 
-# Plot of phylum data with log-scaled y-axis: 
+# Plot of top phylum data with log-scaled y-axis: 
 agg_phylum_data %>% 
   filter(phylum %in% top_phyla) %>% 
   mutate(agg_rel_abund = agg_rel_abund + 1/6000) %>% 
@@ -121,7 +122,7 @@ top_genus <- agg_genus_data %>%
   top_n(10, median) %>% 
   pull(genus)
 
-# Plot of genus data with log-scaled y-axis: 
+# Plot of top genus data with log-scaled y-axis: 
 agg_genus_data %>% 
   filter(genus %in% top_genus) %>% 
   mutate(agg_rel_abund = agg_rel_abund + 1/6000) %>% 
@@ -136,7 +137,7 @@ agg_genus_data %>%
   coord_flip()+
   theme_classic()
 
-# Plot of genus over time
+# Plot of top genus over time
 agg_genus_data %>% 
   filter(genus %in% top_genus) %>% 
   mutate(agg_rel_abund = agg_rel_abund + 1/6000) %>%
@@ -243,7 +244,8 @@ agg_genus_data %>%
        y="Relative abundance (%)")+
   scale_y_log10(breaks=c(1e-4, 1e-3, 1e-2, 1e-1, 1), labels=c(1e-2, 1e-1, 1, 10, 100))+
   coord_flip()+
-  theme_classic()
+  theme_classic()+
+  ggsave("results/figures/genera_assoc_w_treatment.png")
    
 #Kruskal_wallis test with Benjamini-Hochburg correction for phylum differences across time in the PPI treatment group 
 ppi_phylum_tests <- agg_phylum_data %>% 
@@ -277,7 +279,7 @@ agg_phylum_data %>%
 #Kruskal_wallis test for genus differences across time in the PPI group with Benjamini-Hochburg correction 
 ppi_genus_tests <- agg_genus_data %>% 
   group_by(genus) %>% 
[…]

### 2019-06-20 submission/manuscript.Rmd
Preliminary draft

@@
 difficile* colonization resistance.
-Multiple epidemiological studies have suggested an association between PPI use and incidence of Clostridium difficile infections (CDIs) [@tariq2017association; @nehra2018proton; @Naito2018].
+Multiple epidemiological studies have suggested an association between PPI use and incidence of Clostridium difficile infections (CDIs) [@tariq2017association; @nehra2018proton; @Naito2018; @Elias2019].
 However, most of the studies suggesting a link between PPIs and *C.
-difficile* are retrospective and did not evaluate the microbiome [@tariq2017association; @nehra2018proton].
+difficile* are retrospective and did not evaluate the microbiome [@tariq2017association; @nehra2018proton; @Elias2019].
 Thus, it is unclear whether the gastrointestinal microbiome changes associated with PPI use play a role in the association between PPIs and CDI incidence, severity or recurrence.

### 2019-06-20 code/figure_1.R
Preliminary versions of figures

@@ -1,0 +1,12 @@
+library(tidyverse)
+library(magick)
+library(cowplot)
+
+a <- ggdraw() + draw_image("results/figures/PPI_Exp_Scheme.png")
+b <- ggdraw() + draw_image("results/figures/before_abx.png")
+c <- ggdraw() + draw_image("results/figures/before_plus_day_after_abx.png")
+
+plot_grid(a, b, c, labels = "AUTO")+
+  ggsave("results/figures/figure_1.pdf", width=6.5, height=6)
+
+

### 2019-06-20 code/figure_2.R

@@ -1,0 +1,10 @@
+library(tidyverse)
+library(magick)
+library(cowplot)
+
+a <- ggdraw() + draw_image("results/figures/ppi_cfu.png")
+b <- ggdraw() + draw_image("results/figures/after_abx_C.diff.png")
+c <- ggdraw() + draw_image("results/figures/genera_assoc_w_treatment.png")
+
+plot_grid(a, b, c, labels = "AUTO")+
+  ggsave("results/figures/figure_2.pdf", width=6.5, height=6)

### 2019-06-24 code/figure_1.R
Add components to initial figures

@@ -4,9 +4,10 @@ library(cowplot)
 
 a <- ggdraw() + draw_image("results/figures/PPI_Exp_Scheme.png")
 b <- ggdraw() + draw_image("results/figures/before_abx.png")
-c <- ggdraw() + draw_image("results/figures/before_plus_day_after_abx.png")
+c <- ggdraw() + draw_image("results/figures/families_prev_assoc_w_PPIs.png")
+d <- ggdraw() + draw_image("results/figures/before_plus_day_after_abx.png")
 
-plot_grid(a, b, c, labels = "AUTO")+
+plot_grid(a, b, c, d, labels = "AUTO")+
   ggsave("results/figures/figure_1.pdf", width=6.5, height=6)
 
 

### 2019-06-24 code/figure_2.R

@@ -5,6 +5,7 @@ library(cowplot)
 a <- ggdraw() + draw_image("results/figures/ppi_cfu.png")
 b <- ggdraw() + draw_image("results/figures/after_abx_C.diff.png")
 c <- ggdraw() + draw_image("results/figures/genera_assoc_w_treatment.png")
+d <- ggdraw() + draw_image("results/figures/family_assoc_w_treatment.png")
 
-plot_grid(a, b, c, labels = "AUTO")+
+plot_grid(a, b, c, d, labels = "AUTO")+
   ggsave("results/figures/figure_2.pdf", width=6.5, height=6)

### 2019-06-24 code/ppi_taxa.R

@@ -216,7 +216,26 @@ agg_family_data %>%
        y="Relative abundance (%)")+
   scale_y_log10(breaks=c(1e-4, 1e-3, 1e-2, 1e-1, 1), labels=c(1e-2, 1e-1, 1, 10, 100))+
   coord_flip()+
-  theme_classic()
+  theme_classic()+
+  ggsave("results/figures/family_assoc_w_treatment.png")
+
+#Graph the families associated with human PPI use according to the literature [@Imhann2017]
+ppi_family <- c("Enterococcaceae", "Lactobacillaceae", "Micrococcaceae", "Staphylococcaceae", "Streptococcaceae", "Ruminococcaceae")
+agg_family_data %>% 
+  filter(family %in% ppi_family) %>% 
+  mutate(family=factor(family, levels=ppi_family)) %>% 
+  mutate(agg_rel_abund = agg_rel_abund + 1/6000) %>% 
+  ggplot(aes(x= reorder(family, agg_rel_abund), y=agg_rel_abund, color=Group, alpha = day))+
+  geom_hline(yintercept=1/3000, color="gray")+
+  geom_boxplot(outlier.shape = NA)+
+  geom_jitter(shape=19, size=1, alpha=0.7, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.2)) +
+  labs(title="Families previously associated with human PPI use", 
+       x=NULL,
+       y="Relative abundance (%)")+
+  scale_y_log10(breaks=c(1e-4, 1e-3, 1e-2, 1e-1, 1), labels=c(1e-2, 1e-1, 1, 10, 100))+
+  coord_flip()+
+  theme_classic()+
+  ggsave("results/figures/families_prev_assoc_w_PPIs.png")
 
 #Kruskal_wallis test for genus differences across treatment groups with Benjamini-Hochburg correction 
 genus_tests <- agg_genus_data %>% 
@@ -351,11 +370,11 @@ agg_family_data %>%
   geom_hline(yintercept=1/3000, color="gray")+
   theme_classic()
 
+#Lactobacillaceae family relative abundance over time
 agg_family_data %>% 
   filter(family == "Lactobacillaceae") %>% 
   mutate(agg_rel_abund = agg_rel_abund + 1/6000) %>%
   select(Group, day, agg_rel_abund, family) %>% 
-  filter(Group == "PPI") %>%
[…]

### 2019-06-24 code/ppi_cfu_stats.R
Starting statisical analysis

@@ -1,0 +1,48 @@
+library(tidyverse)
+library(broom)
+
+#Load in metadata file and tidy CFU/g data from each timepoint
+ppi_cfu <- select(metadata, Group, ends_with("CFU.g")) %>% 
+  gather(Day, CFU, -Group) %>% 
+  mutate(Group=factor(Group, levels=c("PPI", "Clindamycin", "Clindamycin + PPI")),
+         Day = as.numeric(str_replace(Day, "^[^.](\\d\\d?)[\\w\\.\\s\\/]+", "\\1")),
+         Day = Day - 7, #Modify day notation so that C. difficile challenge day is represented as day 0.
+         CFU=as.numeric(CFU) + 1) %>% 
+  ungroup
+
+#Analysis
+get_mean_cfu <- function(x){
+      x %>% 
+        group_by(Group) %>% 
+        summarize(mean=mean(CFU)) %>% 
+        spread(key=Group, value=mean)
+}
+
+select(ppi_cfu, Group, Day, CFU) %>% 
+  group_by(Day) %>% 
+  nest() %>% 
+  mutate(model=map(data, ~kruskal.test(x=.x$CFU, g=as.factor(.x$Group)) %>% tidy())) %>% 
+  mutate(mean= map(data, get_mean_cfu)) %>% 
+  unnest(model, mean) %>% 
+  mutate(model=map(data,
+                   ~wilcox.test(x=.x$CFU, g=as.factor(.x$Group), p.adjust.method="BH") %>% 
+                     tidy() %>% 
+                     mutate(compare=paste(group1, group2, sep="-")) %>% 
+##                     select(-group1, -group2) %>% 
+##                     spread(key=compare, value=p.value)
+                   )
+         ) %>% 
+##  unnest(model) %>% 
+##  select(-data, -parameter, -statistic) %>% 
+  write_tsv("data/process/ppi_cfu_stats.tsv")
+
+cfu_tests <- ppi_cfu %>% 
+  group_by(Day) %>% 
+  do(tidy(kruskal.test(CFU~Group, data=.))) %>% 
+  ungroup() %>% 
+  mutate(p.value.adj=p.adjust(p.value, method="BH")) %>% 
+  arrange(p.value.adj)
+
+for (i in ppi_cfu$Day){
+print(pairwise.wilcox.test(ppi_cfu$CFU, ppi_cfu$Group, p.adjust.method = "BH"))
+}

### 2019-06-24 submission/manuscript.Rmd
Generate initial manuscript draft

@@
 Antibiotics are a major risk factor for *Clostridium difficile* infections (CDIs), but it is less clear what other factors contribute to CDI incidence, severity, and recurrence.
-However, other medications besides antibiotics have been shown to alter the microbiota.
-Proton pump inhibitors (PPIs) have been shown to alter the human intestinal microbiota in observational and interventional stduies and PPI use has also been associated with CDIs.
+Interestingly, other medications besides antibiotics have been shown to alter the microbiota.
+Proton pump inhibitors (PPIs) were associated with alterations in the human intestinal microbiota through observational and interventional stduies and PPI use has also been associated with CDIs.
 We evaluated the ability of the PPI omeprazole to alter the murine intestinal microbiota and disrupt resistance to *C.
 difficile*.
-We found PPI treatment had minimal impact on the murine intestinal microbiota and was not enough of a disruption to promote C.
-difficile colonization after challenge.
-Further studies are needed to determine whether other factors such as the composition of the staring bacterial community, comorbidities, and use of additional medications contribute to the association between PPIs and CDIs.
+We found PPI treatment had minimal impact on the murine intestinal microbiota and was not enough to promote *C.
+difficile* colonization after challenge.
+Further studies are needed to determine whether other factors such as the composition of the starting bacterial community, comorbidities, and use of additional medications contribute to the association between PPIs and CDIs.
 
@@
 ## Introduction
-Proton pump inhibitors (PPIs) are one of the most widely used drugs and are available over the counter.
-PPIs are approved to treat a variety of conditions including gastroesophageal reflux disease, ulcers, and prophylaxis for nonsteroidal anti-inflammatory drug (NSAID)-associated gastrointestinal bleeding [@nehra2018proton].
+Proton pump inhibitors (PPIs) are one of the most widely used drugs and are available over the counter [@Singh2018].
+PPIs are approved to treat a variety of conditions including gastroesophageal reflux disease, ulcers, or as prophylaxis for nonsteroidal anti-inflammatory drug (NSAID)-associated gastrointestinal bleeding [@nehra2018proton].
 Unfortunately, there have been a number of adverse events associated with their use including small intestinal bacterial overgrowth, decreased calcium absorption, and increased risk of enteric infections [@nehra2018proton].
-There’s also been a number of large cohort studies as well as interventional clinical trials that have demonstrated specific alterations in the intestinal microbiome associated with PPI use [@Imhann2017; @Naito2018].
-PPI-associated microbiota changes have been attributed to the PPIs ability to increase stomach pH which may promote the survival of oral and pathogenic bacteria.
+There’s also been a number of large cohort studies as well as interventional clinical trials that have demonstrated specific alterations in the intestinal microbiome were associated with PPI use [@Imhann2017; @Naito2018].
+PPI-associated microbiota changes have been attributed to the ability of PPIs to increase stomach acid pH which may promote the survival of oral and pathogenic bacteria [@Imhann2017; @Naito2018].
 Human fecal microbiota changes with PPI use include increases in Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae with decreases in Ruminococcaceae [@Imhann2017].
-An in vitro screen, demonstrated that 3 PPIs inhibit the growth of 7-8 strains of Lachnospiraceae and Ruminococcaceae, which represented ~16% of the 40 intestinal bacteria strains that were screened [@maier2018extensive].
-Antibiotics have a large impact on the intestinal microbiome and are a primary risk factor for developing CDIs [@Schubert2015].
+An *in vitro screen*, demonstrated that 3 PPIs inhibit the growth of 7-8 strains of Lachnospiraceae and Ruminococcaceae, which represented ~16% of the 40 intestinal bacteria strains that were screened [@maier2018extensive].
+Antibiotics have a large impact on the intestinal microbiome and are a primary risk factor for developing *Clostridium difficile* infections (CDIs) [@Schubert2015].
 It is less clear whether other human medications that impact the microbiota also influence *C.
 difficile* colonization resistance.
-Multiple epidemiological studies have suggested an association between PPI use and incidence of Clostridium difficile infections (CDIs) [@tariq2017association; @nehra2018proton; @Naito2018; @Elias2019].
+Multiple epidemiological studies have suggested an association between PPI use and incidence or recurrence of CDIs [@tariq2017association; @nehra2018proton; @Naito2018; @Elias2019].
 However, most of the studies suggesting a link between PPIs and *C.
 difficile* are retrospective and did not evaluate the microbiome [@tariq2017association; @nehra2018proton; @Elias2019].
-Thus, it is unclear whether the gastrointestinal microbiome changes associated with PPI use play a role in the association between PPIs and CDI incidence, severity or recurrence.
-Additionally, epidemiological studies have a limited capacity to address potential confounders and comorbidities in patients that were on PPIs and developed CDI infections.
+Thus, it is unclear whether the gastrointestinal microbiome changes associated with PPI use play a role in the association between PPIs and CDI incidence or recurrence.
+Additionally, epidemiological studies have a limited capacity to address potential confounders and comorbidities in patients that were on PPIs and developed CDI infections or recurrent CDIs [@Elias2019; @tariq2017association].
 Although there have been a few *C.
-difficile* mouse model studies that have demonstrated PPIs have some effect on *C.
-difficile* colonization with or without additional antibiotic treatment, none of these studies looked at the interplay with the intestinal microbiome [@hung2015proton; @Kaur2007; @kaur2011effect].
-Here, we evaluated the impact of a PPI on the murine microbiome and susceptibility to C.
-difficile colonization in relation to clindamycin, an antibiotic that perturbs the microbiome enough to allow C.
-difficile to colonize but is mild enough that C.
-difficile is cleared within 10 days [@Jenior2018].
+difficile* mouse model studies that have demonstrated PPIs have some effect on CDIs with or without additional antibiotic treatment, none of these studies looked at the interplay with the intestinal microbiome [@hung2015proton; @Kaur2007; @kaur2011effect].
+Here, we evaluated the impact of a PPI on the murine microbiome and susceptibility to *C.
+difficile* colonization in relation to clindamycin, an antibiotic that perturbs the microbiome enough to allow *C.
+difficile* to colonize but is mild enough that *C.
+difficile* is cleared within 10 days [@Jenior2018].
 
@@
 **Murine fecal microbiomes were minimally affected by PPI treatment **
-To test if PPI treatment alters the microbiome and promotes susceptibility to CDIs, we gavaged mice daily with a high dose of omeprazole for 7 days before C.
-difficile challenge (Figure 1A) and compared the PPI-treated mice to a group that received the antibiotic, clindamycin, and a group that received both PPI and clindamycin treatment.
+To test if PPI treatment alters the microbiome and promotes susceptibility to CDIs, we gavaged mice daily with a high dose of omeprazole for 7 days before *C.
+difficile* challenge (Figure 1A) and compared the PPI-treated mice to a group that received the antibiotic, clindamycin, and a group that received both PPI and clindamycin treatment.
 A principle coordinates analysis (PCoA) of the Bray-Curtis distances over the initial 7 days of treatment revealed the bacterial communities of PPI-treated mice remained relatively unchanged (Figure 1B).
-In contrast, the microbiomes from the 2 groups of mice that received clindamycin started to shift away from the untreated and PPI-treated mice the day after treatment (Figure 1B).
+Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes that have previously been impacted by PPIs in human studies, but we see no significant fluctuations in the stool of our PPI-treated mice throughout the course of the 16-day experiment (Figure 1C).In contrast, the microbiomes from the 2 groups of mice that received clindamycin started to shift away from the untreated and PPI-treated mice the day after treatment (Figure 1D).
 
-**PPI-treatment did not promote susceptibility to Clostridium difficile infection in mice**
-After 7 days of PPI treatment or 1 day after clindamycin treatment, mice were challenged with C.
-difficile 630 spores.
-While all 4 of the clindamycin-treated mice were colonized with C.
-difficile, all of the PPI-treated mice were resistant to C.
-difficile colonization (Figure 2A).
+**PPI-treatment did not promote susceptibility to *C. difficile infection* in mice**
+After 7 days of PPI treatment or 1 day after clindamycin treatment, mice were challenged with *C.
+difficile* 630 spores.
+While all 4 of the clindamycin-treated mice were colonized with *C.
+difficile*, all of the PPI-treated mice were resistant to *C.
+difficile* colonization (Figure 2A).
 Interestingly, only 1 cage of the PPI and clindamycin-treated mice were colonized, while the other cage was not (Figure 2A).
-PCoAs of the bacterial communities from the 3 groups of mice after C.
-difficile challenge, showed that the greatest shifts in bacterial communities occurred in the clindamycin-treated mice (Figure 2B).
-Within 5 days, all of the mice had cleared C.
-difficile, suggesting there was no difference in rate of clearance for the mice that were initially colonized with C.
-difficile.
-Streptococcus, ___, ____ are all taxa that have previously been impacted by PPIs in human studies, but we see no significant fluctuations in the stool of our PPI-treated mice throughout the course of the 16-day experiment (Figure 2C).
-Together these results suggest that PPI treatment alone had minor effects on microbial community resistance to C.
-difficile colonization in mice.
-Instead most of the differences between our 3 groups appear to be driven by clindamycin treatment and included decreased Alistipes, Barnesiella, Porhyromonadaceae, Ruminococcaceae (Schubert), taxa previously found to be altered in clindamycin-treated mice that were challenged with C.
-difficile.
+PCoAs of the bacterial communities from the 3 groups of mice after *C.
+difficile* challenge, showed that the greatest shifts in bacterial communities occurred in the clindamycin-treated mice (Figure 2B).
+Within 5 days, all of the mice had cleared *C.
+difficile*, suggesting there was no difference in rate of clearance for the mice that were initially colonized with *C.
+difficile*.
+Our results suggest that PPI treatment alone had minor effects on microbial community resistance to *C.
+difficile* colonization in mice.
+Instead most of the differences between our 3 treatment groups appear to be driven by clindamycin administration and included decreased *Alistipes*, *Barnesiella*, Porhyromonadaceae, Ruminococcaceae [@Schubert2015], taxa previously found to be altered in clindamycin-treated mice that were challenged with *C.
+difficile*.
 
@@
 
-Our results demonstrate the PPI omeprazole had minimal impact on the stool microbiomes of mice and did not promote susceptibility to C.
-difficile colonization.
+Our results demonstrate the PPI omeprazole had minimal impact on the stool microbiomes of mice and did not promote susceptibility to *C.
+difficile* colonization.
 In contrast, clindamycin-treated mice were initially colonized, but cleared the infection within 10 days.
-Combining PPI and clindamycin treatment had a mixed effect, one cage was colonized with C.
-difficile while the other cage remained resistant.
-Our findings that PPI treatment had minimal impact on the fecal microbiome are comparable to several other mouse studies that indicate PPIs have more of an effect on the small intestinal microbiota compared to the fecal microbiota (Yasutomi).
-Our findings differ from the taxa observed to be significantly impacted by PPI use in human studies, but a closer examinations suggest that the significance of these associations is driven by some individuals more than others and the effect size can be relatively small with overall differences on a PCoA plot difficult to distinguish (Imhann, Freedberg).
-Also, most of the studies to date compare different individuals (PPI users versus non-users) or the same individuals before and after treatment (Naito 2018, Imhann), unfortunately these limited microbiota snapshots may be skewed by day-to-day microbiota variations (Llorens-Rico & Raes Nature 2019), which become apparent in the longitudinally sampled mice.
-Enterococcus and Streptococcus are frequently mentioned in the context of human PPI use and human CDIs (Imhann 2017), however we did not find much of either taxa in our mice.
-Previous work that administered the PPI lansoprazole daily for 2 weeks to mice and then challenged with C.
-difficile demonstrated that PPI treatment alone resulted in detectable C.
-difficile CFUs in the stool 1 week after challenge, but also showed there was detectable C.
-difficile in mice not treated with antibiotics.
-The presence of C.
-difficile in mice that were not treated with either antibiotics or PPIs could be attributed to the higher dose of vegetative cells (108 CFU) used to challenge the mice and may partly explain their observation of C.
-difficile in PPI-treated mice.
-We have previously shown mice from are colony that are not given antibiotics are resistant to C.
-difficile 630 when challenged with 103 spores (Jenior 2017).
-Another group also used 106 spores or 107 vegetative cells to challenge antibiotic-treated mice or mice treated with both the PPI esomeprazole for 2 days and antibiotics and demonstrated the antibiotic/PPI-treated mice experience more severe CDI.
-Our study extended previous work examining PPIs and C.
-difficile in mice to examine the contribution of the intestinal microbiota.
-Unfortunately, we found the PPI omeprazole had minimal impact on the murine intestinal microbiota and did not promote C.
-difficile colonization.
-Our 16S rRNA sequencing suggests that Streptococcus and Enterococcus are rare taxa in our C57BL/6 mouse colony.
-If these taxa are important contributors, to the associations between PPIs and C.
-difficile infection in humans, this may be part of the reason we did not see an effect in our CDI mouse model.
-There could also be differences at the species level for taxa that are impacted with PPIs in humans compared to mice, which may also contribute to the association.
-Other murine starting communities or the addition of human-associated strains may be needed to determine whether PPIs can increase Enterococcus and Streptococcus and subsequently promote or worsen CDIs.
-Other factors such as age, body mass index, comorbidities, and use of other medications in human studies may also be contributing to the association between PPIs and CDIs.
-NSAID treatment can increase CDI severity in mice (Maseda 2019), given that PPIs are sometimes used in combination with NSAIDs, it may be worth examining  This study only looked at the impact of PPIs with or without antibiotics, but future work may want to incorporate age, other comorbidities and bacterial strains that are less common in mice to evaluate whether PPIs in conjunction with other factors can increase the risk of C.
-difficile.
+Combining PPI and clindamycin treatment had a mixed effect, one cage was colonized with *C.
+difficile* while the other cage remained resistant.
+Our findings that PPI treatment had minimal impact on the fecal microbiome are comparable to several other mouse studies that indicate PPIs have more of an effect on the small intestinal microbiota compared to the fecal microbiota [@Yasutomi2018].
+We did not find significant changes for the taxa observed to be significantly impacted by PPI use in human studies, but a closer examination suggests that the significance of these associations is driven by some individuals more than others and the effect size can be relatively small with overall differences on a PCoA plot difficult to distinguish [@Freedberg2015, @Imhann2015].
+Also, most of the studies to date compare different individuals (PPI users versus non-users) or the same individuals before and after treatment [@Naito2018; @Imhann2017], unfortunately these limited microbiota snapshots may be skewed by day-to-day microbiota variations [@LlornsRico2019], which become apparent in the longitudinally sampled mice.
+Previous work that administered the PPI lansoprazole daily for 2 weeks to mice and then challenged with *C.
+difficile* demonstrated that PPI treatment alone resulted in detectable *C.
+difficile* in the stool 1 week after challenge, but also showed there was detectable *C.
+difficile* in mice not treated with antibiotics [@Kaur2007; @kaur2011effect].
+The presence of *C.
+difficile* in mice that were not treated with either antibiotics or PPIs could be attributed to the higher dose of vegetative cells (10^8^ CFU) and the different *C.
+difficile* strain used to challenge the mice and may partly explain their observation of *C.
+difficile* in PPI-treated mice [@Kaur2007; @kaur2011effect].
+We have previously shown mice from are colony that are not given antibiotics are resistant to *C.
+difficile* 630 when challenged with 10^3^ spores [@Jenior2017].
+The other group also used a higher dose of 10^6^ spores or 10^7^ vegetative cells to challenge antibiotic-treated mice or mice treated with both the PPI esomeprazole for 2 days and antibiotics and demonstrated the antibiotic/PPI-treated mice developed more severe CDIs [@hung2015proton].
+Our study extended previous work examining PPIs and *C.
+difficile* in mice to examine the contribution of the intestinal microbiota.
+We found the PPI omeprazole had minimal impact on the murine intestinal microbiota and in contrast to previous work, PPIs did not promote *C.
+difficile* colonization.
+Our 16S rRNA sequencing suggested that *Streptococcus* and *Enterococcus* are rare taxa in our C57BL/6 mouse colony.
+If these taxa are important contributors to the associations between PPIs and CDIs in humans, this may be part of the reason we did not see an effect in our CDI mouse model.
+There could also be differences at the species level for taxa that are altered with PPIs in humans compared to mice, which may also contribute to the association.
+Other murine starting communities or the addition of human PPI treatment-associated strains may be needed to determine whether PPIs can alter taxa and subsequently promote CDIs.
+Factors such as age, body mass index, comorbidities, and use of other medications in human studies may also be contributing to the association between PPIs and CDIs.
+NSAID treatment can increase CDI severity in mice [@Maseda2019].
+Given that PPIs are sometimes used in combination with NSAIDs [@nehra2018proton], it may be worth examining both medications together using a mouse model.
+This study only looked at the impact of PPIs with or without antibiotics, but future work may want to incorporate age, other comorbidities and bacterial strains that are less common in mice to evaluate whether PPIs in conjunction with other factors can increase the risk of CDIs.
 
@@
 Omeprazole was prepared from 20 mg/mL frozen aliquots and diluted to an 8mg/mL prior to gavage.
-All mice received ~40mg/kg omeprazole or vehicle solution once per day through the duration of the experiment starting 7 days before C.
-difficile challenge.
-One day prior to C.
-difficile challenge, mice received an intraperitoneal injection of 10 mg/kg clindamycin or sterile saline vehicle.
+All mice received ~40 mg/kg omeprazole or vehicle solution once per day through the duration of the experiment with treatment starting 7 days before *C.
+difficile* challenge (Figure 1A).
+One day prior to *C.
+difficile* challenge, mice received an intraperitoneal injection of 10 mg/kg clindamycin or sterile saline vehicle.
 All drugs were filter sterilized through a 0.22 micron syringe filter before administration to animals.
@@
 **C. difficile infection model**
-Mice were challenged with C.
-difficile 630 seven days after the start of omeprazole treatment and one day after clindamycin treatment.
-Mice were challenged with 103 spores in ultrapure distilled water.
-Samples were collected for 16S rRNA sequencing or quantifying C.
-difficile CFU throughout the duration of the experiments at the indicated timepoints (Figure 1A).
-Samples for 16S rRNA sequencing were flash frozen in liquid nitrogen and stored at -80oC until DNA extraction, while samples for CFU quantification were transferred into an anaerobic chamber and serially diluted in PBS.
-Diluted samples were plated on TCCFA (taurocholate, cycloserine, cefoxitin, fructose agar) plates and incubated at 37oC for 24 hours under anaerobic conditions to quantify C.
-difficile CFU.
+Mice were challenged with *C.
+difficile* 630 seven days after the start of omeprazole treatment and one day after clindamycin treatment.
+Mice were challenged with 10^3^ spores in ultrapure distilled water.
+Samples were collected for 16S rRNA sequencing or *C.
+difficile* CFU quantification throughout the duration of the experiments at the indicated timepoints (Figure 1A).
+Samples for 16S rRNA sequencing were flash frozen in liquid nitrogen and stored at -80&deg;C until DNA extraction, while samples for CFU quantification were transferred into an anaerobic chamber and serially diluted in PBS.
+Diluted samples were plated on TCCFA (taurocholate, cycloserine, cefoxitin, fructose agar) plates and incubated at 37&deg;C for 24 hours under anaerobic conditions to quantify *C.
+difficile* CFU.
 
-**16S rRNA gene sequencing** DNA for 16S rRNA gene sequencing was extracted from 10-50 mg fecal pellets from each mouse using the DNeasy Powersoil HTP 96 Kit (Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf).
-The V4 hypervariable region of the 16S rRNA gene was amplified with Accuprime Pfx DNA polymerase (Thermo Fisher Scientific)) using previously described custom barcoded primers (Kozich et al 2013).
+**16S rRNA gene sequencing** DNA for 16S rRNA gene sequencing was extracted from 10-50 mg fecal pellet from each mouse using the DNeasy Powersoil HTP 96 Kit (Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf).
+The V4 hypervariable region of the 16S rRNA gene was amplified with Accuprime Pfx DNA polymerase (Thermo Fisher Scientific) using previously described custom barcoded primers [@Kozich2013].
 The 16S rRNA amplicon library was sequenced with the MiSeq (Illumina).
@@
 
-**16S rRNA gene sequence analysis**Mothur (v1.40.5) was used for all sequence processing steps (Schloss2009) using the previously published protocol (Kozich2013).
+**16S rRNA gene sequence analysis**Mothur (v1.40.5) was used for all sequence processing steps [@Schloss2009] using the previously published protocol [@Kozich2013].
 R (v.3.5.1) was used to generate figures and perform statistical analysis.
@@
 
-**Figure 1. PPI treatment had minimal impact on the murine fecal microbiota** A. Mouse experiment timeline and logistics.
+**Figure 1.
+PPI treatment had minimal impact on the murine fecal microbiota** A.
+Mouse experiment timeline and logistics.
+B.
+Principal Coordinates Analaysis (PCoA) of Bray-curtis distances from stool samples for all treatment groups during the intial 7 days of treatment with the PPI omeprazole or vehicle cluster together and do not chage much over time.
+C.
+Relative abundances of families previously associated with PPI use in humans do not change much over the 16-day course of treatment with PPIs (7 days before *C.
+difficile* challenge through 9 days post *C.
+difficile* challenge).
+D.
+PCoA of stool samples collected before and 1 day after antibiotic treatment, demonstrating the fecal microbiota of mice treated with clindamycin starts to separate from the rest of the samples.
+
+**Figure 2.
+PPI treatment alone does not promote CDIs in mice** A.
+C.
+difficile CFUs/g stool measured each day post C.
+difficile challenge for clindamycin, clindamycin/PPI, and PPI-treated mice.
+B.
+PCoA of stool samples collected after antibiotic treatments.
+C.
+Genera significantly associated with treatment groups.
+D.
+Families significantly associated with treatment groups.
 

### 2019-06-24 submission/manuscript.tex
Generate initial manuscript draft

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{}
    \pretitle{\vspace{\droptitle}}
  \posttitle{}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}

\begin{document}

\vspace{35mm}

\section{\texorpdfstring{Proton pump inhibitor administration does not
promote \emph{Clostridium difficile} colonization in a murine
model}{Proton pump inhibitor administration does not promote Clostridium difficile colonization in a murine model}}\label{proton-pump-inhibitor-administration-does-not-promote-clostridium-difficile-colonization-in-a-murine-model}

\vspace{35mm}

Sarah Tomkovich\({^1}\), Nicholas A. Lesniak\({^1}\), Yuan Li\({^1}\),
Lucas Bishop\({^1}\), Madison J. Fitzgerald\({^1}\), Patrick D.
Schloss\textsuperscript{1\(\dagger\)}

\vspace{40mm}

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}

\(1\) Department of Microbiology and Immunology, University of Michigan,
Ann Arbor, MI 48109

\newpage

\linenumbers

\subsection{Abstract}\label{abstract}

Proton pump inhibitor (PPI) use has been associated with microbiota
alterations and susceptibility to \emph{Clostridium difficile}
infections (CDIs) in humans. We assessed how PPI treatment alters the
fecal microbiota and whether PPIs promote CDIs using a CDI mouse model.
Mice were treated with the PPI omeprazole for 7 days prior to \emph{C.
difficile} challenge and were compared to mice that were treated with
the antibiotic clindamycin or both omeprazole and clindamycin. We found
that omeprazole was not sufficient to promote \emph{C. difficile}
colonization. When omeprazole treatment was combined with the
antibiotic, there was a mixed level of \emph{C. difficile} colonization,
where one cage of mice remained resistant and the other cage was
colonized. 16S rRNA sequencing analysis revealed that omeprazole had
minimal impact on the murine microbiota throughout the 16 days of PPI
exposure. \emph{C. difficile} colonization of clindamycin-treated mice
was associated with changes in \emph{Alistipes}, \emph{Barnesiella},
Porphyromonadaceae, and Ruminococcaceae. These results suggest PPI
treatment alone is not sufficient to disrupt microbiota resistance to C.
difficle infection in mice that are normally resistant in the absence of
antibiotic treatment.

\subsection{Importance}\label{importance}

Antibiotics are a major risk factor for \emph{Clostridium difficile}
infections (CDIs), but it is less clear what other factors contribute to
CDI incidence, severity, and recurrence. Interestingly, other
medications besides antibiotics have been shown to alter the microbiota.
Proton pump inhibitors (PPIs) were associated with alterations in the
human intestinal microbiota through observational and interventional
stduies and PPI use has also been associated with CDIs. We evaluated the
ability of the PPI omeprazole to alter the murine intestinal microbiota
and disrupt resistance to \emph{C. difficile}. We found PPI treatment
had minimal impact on the murine intestinal microbiota and was not
enough to promote \emph{C. difficile} colonization after challenge.
Further studies are needed to determine whether other factors such as
the composition of the starting bacterial community, comorbidities, and
use of additional medications contribute to the association between PPIs
and CDIs.

\newpage

\subsection{Introduction}\label{introduction}

Proton pump inhibitors (PPIs) are one of the most widely used drugs and
are available over the counter (1). PPIs are approved to treat a variety
of conditions including gastroesophageal reflux disease, ulcers, or as
prophylaxis for nonsteroidal anti-inflammatory drug (NSAID)-associated
gastrointestinal bleeding (2). Unfortunately, there have been a number
of adverse events associated with their use including small intestinal
bacterial overgrowth, decreased calcium absorption, and increased risk
of enteric infections (2). There's also been a number of large cohort
studies as well as interventional clinical trials that have demonstrated
specific alterations in the intestinal microbiome were associated with
PPI use (3, 4). PPI-associated microbiota changes have been attributed
to the ability of PPIs to increase stomach acid pH which may promote the
survival of oral and pathogenic bacteria (3, 4). Human fecal microbiota
changes with PPI use include increases in Enterococcaceae,
Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae
with decreases in Ruminococcaceae (3). An \emph{in vitro screen},
demonstrated that 3 PPIs inhibit the growth of 7-8 strains of
Lachnospiraceae and Ruminococcaceae, which represented
\textasciitilde{}16\% of the 40 intestinal bacteria strains that were
screened (5). Antibiotics have a large impact on the intestinal
microbiome and are a primary risk factor for developing
\emph{Clostridium difficile} infections (CDIs) (6). It is less clear
whether other human medications that impact the microbiota also
influence \emph{C. difficile} colonization resistance. Multiple
epidemiological studies have suggested an association between PPI use
and incidence or recurrence of CDIs (2, 4, 7, 8). However, most of the
studies suggesting a link between PPIs and \emph{C. difficile} are
retrospective and did not evaluate the microbiome (2, 7, 8). Thus, it is
unclear whether the gastrointestinal microbiome changes associated with
PPI use play a role in the association between PPIs and CDI incidence or
recurrence. Additionally, epidemiological studies have a limited
capacity to address potential confounders and comorbidities in patients
that were on PPIs and developed CDI infections or recurrent CDIs (7, 8).
Although there have been a few \emph{C. difficile} mouse model studies
that have demonstrated PPIs have some effect on CDIs with or without
additional antibiotic treatment, none of these studies looked at the
interplay with the intestinal microbiome (9--11). Here, we evaluated the
impact of a PPI on the murine microbiome and susceptibility to \emph{C.
difficile} colonization in relation to clindamycin, an antibiotic that
perturbs the microbiome enough to allow \emph{C. difficile} to colonize
but is mild enough that \emph{C. difficile} is cleared within 10 days
(12).

\subsection{Results}\label{results}

\textbf{Murine fecal microbiomes were minimally affected by PPI
treatment } To test if PPI treatment alters the microbiome and promotes
susceptibility to CDIs, we gavaged mice daily with a high dose of
omeprazole for 7 days before \emph{C. difficile} challenge (Figure 1A)
and compared the PPI-treated mice to a group that received the
antibiotic, clindamycin, and a group that received both PPI and
clindamycin treatment. A principle coordinates analysis (PCoA) of the
Bray-Curtis distances over the initial 7 days of treatment revealed the
bacterial communities of PPI-treated mice remained relatively unchanged
(Figure 1B). Enterococcaceae, Lactobacillaceae, Micrococcaceae,
Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes
that have previously been impacted by PPIs in human studies, but we see
no significant fluctuations in the stool of our PPI-treated mice
throughout the course of the 16-day experiment (Figure 1C).In contrast,
the microbiomes from the 2 groups of mice that received clindamycin
started to shift away from the untreated and PPI-treated mice the day
after treatment (Figure 1D).

\textbf{PPI-treatment did not promote susceptibility to \emph{C.
difficile infection} in mice} After 7 days of PPI treatment or 1 day
after clindamycin treatment, mice were challenged with \emph{C.
difficile} 630 spores. While all 4 of the clindamycin-treated mice were
colonized with \emph{C. difficile}, all of the PPI-treated mice were
resistant to \emph{C. difficile} colonization (Figure 2A).
Interestingly, only 1 cage of the PPI and clindamycin-treated mice were
colonized, while the other cage was not (Figure 2A). PCoAs of the
bacterial communities from the 3 groups of mice after \emph{C.
difficile} challenge, showed that the greatest shifts in bacterial
communities occurred in the clindamycin-treated mice (Figure 2B). Within
5 days, all of the mice had cleared \emph{C. difficile}, suggesting
there was no difference in rate of clearance for the mice that were
initially colonized with \emph{C. difficile}. Our results suggest that
PPI treatment alone had minor effects on microbial community resistance
to \emph{C. difficile} colonization in mice. Instead most of the
differences between our 3 treatment groups appear to be driven by
clindamycin administration and included decreased \emph{Alistipes},
\emph{Barnesiella}, Porhyromonadaceae, Ruminococcaceae (6), taxa
previously found to be altered in clindamycin-treated mice that were
challenged with \emph{C. difficile}.

\subsection{Discussion and
Conclusions}\label{discussion-and-conclusions}

Our results demonstrate the PPI omeprazole had minimal impact on the
stool microbiomes of mice and did not promote susceptibility to \emph{C.
difficile} colonization. In contrast, clindamycin-treated mice were
initially colonized, but cleared the infection within 10 days. Combining
PPI and clindamycin treatment had a mixed effect, one cage was colonized
with \emph{C. difficile} while the other cage remained resistant. Our
findings that PPI treatment had minimal impact on the fecal microbiome
are comparable to several other mouse studies that indicate PPIs have
more of an effect on the small intestinal microbiota compared to the
fecal microbiota (13). We did not find significant changes for the taxa
observed to be significantly impacted by PPI use in human studies, but a
closer examination suggests that the significance of these associations
is driven by some individuals more than others and the effect size can
be relatively small with overall differences on a PCoA plot difficult to
distinguish (14). Also, most of the studies to date compare different
individuals (PPI users versus non-users) or the same individuals before
and after treatment (3, 4), unfortunately these limited microbiota
snapshots may be skewed by day-to-day microbiota variations (16), which
become apparent in the longitudinally sampled mice. Previous work that
administered the PPI lansoprazole daily for 2 weeks to mice and then
challenged with \emph{C. difficile} demonstrated that PPI treatment
alone resulted in detectable \emph{C. difficile} in the stool 1 week
after challenge, but also showed there was detectable \emph{C.
difficile} in mice not treated with antibiotics (10, 11). The presence
of \emph{C. difficile} in mice that were not treated with either
antibiotics or PPIs could be attributed to the higher dose of vegetative
cells (10\textsuperscript{8} CFU) and the different \emph{C. difficile}
strain used to challenge the mice and may partly explain their
observation of \emph{C. difficile} in PPI-treated mice (10, 11). We have
previously shown mice from are colony that are not given antibiotics are
resistant to \emph{C. difficile} 630 when challenged with
10\textsuperscript{3} spores (17). The other group also used a higher
dose of 10\textsuperscript{6} spores or 10\textsuperscript{7} vegetative
cells to challenge antibiotic-treated mice or mice treated with both the
PPI esomeprazole for 2 days and antibiotics and demonstrated the
antibiotic/PPI-treated mice developed more severe CDIs (9). Our study
extended previous work examining PPIs and \emph{C. difficile} in mice to
examine the contribution of the intestinal microbiota. We found the PPI
omeprazole had minimal impact on the murine intestinal microbiota and in
contrast to previous work, PPIs did not promote \emph{C. difficile}
colonization. Our 16S rRNA sequencing suggested that
\emph{Streptococcus} and \emph{Enterococcus} are rare taxa in our
C57BL/6 mouse colony. If these taxa are important contributors to the
associations between PPIs and CDIs in humans, this may be part of the
reason we did not see an effect in our CDI mouse model. There could also
be differences at the species level for taxa that are altered with PPIs
in humans compared to mice, which may also contribute to the
association. Other murine starting communities or the addition of human
PPI treatment-associated strains may be needed to determine whether PPIs
can alter taxa and subsequently promote CDIs.\\
Factors such as age, body mass index, comorbidities, and use of other
medications in human studies may also be contributing to the association
between PPIs and CDIs. NSAID treatment can increase CDI severity in mice
(18). Given that PPIs are sometimes used in combination with NSAIDs (2),
it may be worth examining both medications together using a mouse model.
This study only looked at the impact of PPIs with or without
antibiotics, but future work may want to incorporate age, other
comorbidities and bacterial strains that are less common in mice to
evaluate whether PPIs in conjunction with other factors can increase the
risk of CDIs.

\subsection{Materials and Methods}\label{materials-and-methods}

\textbf{Animals} All mouse experiments were performed with 7- to
12-week-old C57BL/6 male and female mice. 2-3 mice were housed per cage
and male and female mice were housed separately. All animal experiments
were approved by the University of Michigan Animal Care and Use
Committee (IACUC) under protocol number PRO00006983.

\textbf{Drug treatments} Omeprazole (Sigma Aldrich) was prepared in a
vehicle solution of 40\% polyethylene glycol 400 (Sigma-Aldrich) in
phosphate buffered saline. Omeprazole was prepared from 20 mg/mL frozen
aliquots and diluted to an 8mg/mL prior to gavage. All mice received
\textasciitilde{}40 mg/kg omeprazole or vehicle solution once per day
through the duration of the experiment with treatment starting 7 days
before \emph{C. difficile} challenge (Figure 1A). One day prior to
\emph{C. difficile} challenge, mice received an intraperitoneal
injection of 10 mg/kg clindamycin or sterile saline vehicle. All drugs
were filter sterilized through a 0.22 micron syringe filter before
administration to animals.

\textbf{C. difficile infection model} Mice were challenged with \emph{C.
difficile} 630 seven days after the start of omeprazole treatment and
one day after clindamycin treatment. Mice were challenged with
10\textsuperscript{3} spores in ultrapure distilled water. Samples were
collected for 16S rRNA sequencing or \emph{C. difficile} CFU
quantification throughout the duration of the experiments at the
indicated timepoints (Figure 1A). Samples for 16S rRNA sequencing were
flash frozen in liquid nitrogen and stored at -80°C until DNA
extraction, while samples for CFU quantification were transferred into
an anaerobic chamber and serially diluted in PBS. Diluted samples were
plated on TCCFA (taurocholate, cycloserine, cefoxitin, fructose agar)
plates and incubated at 37°C for 24 hours under anaerobic conditions to
quantify \emph{C. difficile} CFU.

\textbf{16S rRNA gene sequencing} DNA for 16S rRNA gene sequencing was
extracted from 10-50 mg fecal pellet from each mouse using the DNeasy
Powersoil HTP 96 Kit (Qiagen) and an EpMotion 5075 automated pipetting
system (Eppendorf). The V4 hypervariable region of the 16S rRNA gene was
amplified with Accuprime Pfx DNA polymerase (Thermo Fisher Scientific)
using previously described custom barcoded primers (19). The 16S rRNA
amplicon library was sequenced with the MiSeq (Illumina). Amplicons were
cleaned up and normalized with the SequalPrep Normalization Plate Kit
(ThermoFisher Scientific) and pooled amplicons were quantified with the
KAPA library quantification kit (KAPA Biosystems).

\textbf{16S rRNA gene sequence analysis}Mothur (v1.40.5) was used for
all sequence processing steps (20) using the previously published
protocol (19). R (v.3.5.1) was used to generate figures and perform
statistical analysis.

\textbf{Statistical Analysis}

\textbf{Code availability} The code for all sequence processing and
analysis step as well as an Rmarkdown version of this manuscript is
available at
\url{https://github.com/SchlossLab/Tomkovich_PPI_XXXX_2019}.

\textbf{Data availability} The 16S rRNA sequencing data have been
deposited in the NCBI Sequence Read Archive (Accession no.
\_\_\_\_\_\_\_\_\_)

\newpage

\subsection{Figures}\label{figures}

\textbf{Figure 1. PPI treatment had minimal impact on the murine fecal
microbiota} A. Mouse experiment timeline and logistics. B. Principal
Coordinates Analaysis (PCoA) of Bray-curtis distances from stool samples
for all treatment groups during the intial 7 days of treatment with the
PPI omeprazole or vehicle cluster together and do not chage much over
time. C. Relative abundances of families previously associated with PPI
use in humans do not change much over the 16-day course of treatment
with PPIs (7 days before \emph{C. difficile} challenge through 9 days
post \emph{C. difficile} challenge). D. PCoA of stool samples collected
before and 1 day after antibiotic treatment, demonstrating the fecal
microbiota of mice treated with clindamycin starts to separate from the
rest of the samples.

\textbf{Figure 2. PPI treatment alone does not promote CDIs in mice} A.
C. difficile CFUs/g stool measured each day post C. difficile challenge
for clindamycin, clindamycin/PPI, and PPI-treated mice. B. PCoA of stool
samples collected after antibiotic treatments. C. Genera significantly
associated with treatment groups. D. Families significantly associated
with treatment groups.

\newpage

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\hypertarget{ref-Singh2018}{}
1. \textbf{Singh A}, \textbf{Cresci GA}, \textbf{Kirby DF}. 2018. Proton
pump inhibitors: Risks and rewards and emerging consequences to the gut
microbiome. Nutrition in Clinical Practice \textbf{33}:614--624.
doi:\href{https://doi.org/10.1002/ncp.10181}{10.1002/ncp.10181}.

\hypertarget{ref-nehra2018proton}{}
2. \textbf{Nehra AK}, \textbf{Alexander JA}, \textbf{Loftus CG},
\textbf{Nehra V}. 2018. Proton pump inhibitors: Review of emerging
concerns, pp. 240--246. \emph{In} Mayo clinic proceedings. Elsevier.

\hypertarget{ref-Imhann2017}{}
3. \textbf{Imhann F}, \textbf{Vila AV}, \textbf{Bonder MJ},
\textbf{Manosalva AGL}, \textbf{Koonen DP}, \textbf{Fu J},
\textbf{Wijmenga C}, \textbf{Zhernakova A}, \textbf{Weersma RK}. 2017.
The influence of proton pump inhibitors and other commonly used
medication on the gut microbiota. Gut Microbes \textbf{8}:351--358.
doi:\href{https://doi.org/10.1080/19490976.2017.1284732}{10.1080/19490976.2017.1284732}.

\hypertarget{ref-Naito2018}{}
4. \textbf{Naito Y}, \textbf{Kashiwagi K}, \textbf{Takagi T},
\textbf{Andoh A}, \textbf{Inoue R}. 2018. Intestinal dysbiosis secondary
to proton-pump inhibitor use. Digestion \textbf{97}:195--204.
doi:\href{https://doi.org/10.1159/000481813}{10.1159/000481813}.

\hypertarget{ref-maier2018extensive}{}
5. \textbf{Maier L}, \textbf{Pruteanu M}, \textbf{Kuhn M},
\textbf{Zeller G}, \textbf{Telzerow A}, \textbf{Anderson EE},
\textbf{Brochado AR}, \textbf{Fernandez KC}, \textbf{Dose H},
\textbf{Mori H}, \textbf{others}. 2018. Extensive impact of
non-antibiotic drugs on human gut bacteria. Nature \textbf{555}:623.

\hypertarget{ref-Schubert2015}{}
6. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
Antibiotic-induced alterations of the murine gut microbiota and
subsequent effects on colonization resistance against clostridium
difficile. mBio \textbf{6}.
doi:\href{https://doi.org/10.1128/mbio.00974-15}{10.1128/mbio.00974-15}.

\hypertarget{ref-tariq2017association}{}
7. \textbf{Tariq R}, \textbf{Singh S}, \textbf{Gupta A}, \textbf{Pardi
DS}, \textbf{Khanna S}. 2017. Association of gastric acid suppression
with recurrent clostridium difficile infection: A systematic review and
meta-analysis. JAMA internal medicine \textbf{177}:784--791.

\hypertarget{ref-Elias2019}{}
8. \textbf{Elias E}, \textbf{Targownik LE}. 2019. The clinician's guide
to proton pump inhibitor related adverse events. Drugs
\textbf{79}:715--731.
doi:\href{https://doi.org/10.1007/s40265-019-01110-3}{10.1007/s40265-019-01110-3}.

\hypertarget{ref-hung2015proton}{}
9. \textbf{Hung Y-P}, \textbf{Ko W-C}, \textbf{Chou P-H}, \textbf{Chen
Y-H}, \textbf{Lin H-J}, \textbf{Liu Y-H}, \textbf{Tsai H-W}, \textbf{Lee
J-C}, \textbf{Tsai P-J}. 2015. Proton-pump inhibitor exposure aggravates
clostridium difficile--associated colitis: Evidence from a mouse model.
The Journal of infectious diseases \textbf{212}:654--663.

\hypertarget{ref-Kaur2007}{}
10. \textbf{Kaur S}, \textbf{Vaishnavi C}, \textbf{Prasad KK},
\textbf{Ray P}, \textbf{Kochhar R}. 2007. Comparative role of antibiotic
and proton pump inhibitor in ExperimentalClostridium difficileInfection
in mice. Microbiology and Immunology \textbf{51}:1209--1214.
doi:\href{https://doi.org/10.1111/j.1348-0421.2007.tb04016.x}{10.1111/j.1348-0421.2007.tb04016.x}.

\hypertarget{ref-kaur2011effect}{}
11. \textbf{Kaur S}, \textbf{Vaishnavi C}, \textbf{Prasad KK},
\textbf{Ray P}, \textbf{Kochhar R}. 2011. Effect of lactobacillus
acidophilus \& epidermal growth factor on experimentally induced
clostridium difficile infection. The Indian journal of medical research
\textbf{133}:434.

\hypertarget{ref-Jenior2018}{}
12. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
\textbf{Schloss PD}. 2018. Clostridium difficile alters the structure
and metabolism of distinct cecal microbiomes during initial infection to
promote sustained colonization. mSphere \textbf{3}.
doi:\href{https://doi.org/10.1128/msphere.00261-18}{10.1128/msphere.00261-18}.

\hypertarget{ref-Yasutomi2018}{}
13. \textbf{Yasutomi E}, \textbf{Hoshi N}, \textbf{Adachi S},
\textbf{Otsuka T}, \textbf{Kong L}, \textbf{Ku Y}, \textbf{Yamairi H},
\textbf{Inoue J}, \textbf{Ishida T}, \textbf{Watanabe D}, \textbf{Ooi
M}, \textbf{Yoshida M}, \textbf{Tsukimi T}, \textbf{Fukuda S},
\textbf{Azuma T}. 2018. Proton pump inhibitors increase the
susceptibility of mice to oral infection with enteropathogenic bacteria.
Digestive Diseases and Sciences \textbf{63}:881--889.
doi:\href{https://doi.org/10.1007/s10620-017-4905-3}{10.1007/s10620-017-4905-3}.

\hypertarget{ref-Freedberg2015}{}
14. \textbf{Freedberg DE}, \textbf{Toussaint NC}, \textbf{Chen SP},
\textbf{Ratner AJ}, \textbf{Whittier S}, \textbf{Wang TC}, \textbf{Wang
HH}, \textbf{Abrams JA}. 2015. Proton pump inhibitors alter specific
taxa in the human gastrointestinal microbiome: A crossover trial.
Gastroenterology \textbf{149}:883--885.e9.
doi:\href{https://doi.org/10.1053/j.gastro.2015.06.043}{10.1053/j.gastro.2015.06.043}.

\hypertarget{ref-Imhann2015}{}
15. \textbf{Imhann F}, \textbf{Bonder MJ}, \textbf{Vila AV}, \textbf{Fu
J}, \textbf{Mujagic Z}, \textbf{Vork L}, \textbf{Tigchelaar EF},
\textbf{Jankipersadsing SA}, \textbf{Cenit MC}, \textbf{Harmsen HJM},
\textbf{Dijkstra G}, \textbf{Franke L}, \textbf{Xavier RJ},
\textbf{Jonkers D}, \textbf{Wijmenga C}, \textbf{Weersma RK},
\textbf{Zhernakova A}. 2015. Proton pump inhibitors affect the gut
microbiome. Gut \textbf{65}:740--748.
doi:\href{https://doi.org/10.1136/gutjnl-2015-310376}{10.1136/gutjnl-2015-310376}.

\hypertarget{ref-LlornsRico2019}{}
16. \textbf{Lloréns-Rico V}, \textbf{Raes J}. 2019. Tracking humans and
microbes. Nature \textbf{569}:632--633.
doi:\href{https://doi.org/10.1038/d41586-019-01591-y}{10.1038/d41586-019-01591-y}.

\hypertarget{ref-Jenior2017}{}
17. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
\textbf{Schloss PD}. 2017. Clostridium difficile colonizes alternative
nutrient niches during infection across distinct murine gut microbiomes.
mSystems \textbf{2}.
doi:\href{https://doi.org/10.1128/msystems.00063-17}{10.1128/msystems.00063-17}.

\hypertarget{ref-Maseda2019}{}
18. \textbf{Maseda D}, \textbf{Zackular JP}, \textbf{Trindade B},
\textbf{Kirk L}, \textbf{Roxas JL}, \textbf{Rogers LM},
\textbf{Washington MK}, \textbf{Du L}, \textbf{Koyama T},
\textbf{Viswanathan VK}, \textbf{Vedantam G}, \textbf{Schloss PD},
\textbf{Crofford LJ}, \textbf{Skaar EP}, \textbf{Aronoff DM}. 2019.
Nonsteroidal anti-inflammatory drugs alter the microbiota and exacerbate
clostridium difficile colitis while dysregulating the inflammatory
response. mBio \textbf{10}.
doi:\href{https://doi.org/10.1128/mbio.02282-18}{10.1128/mbio.02282-18}.

\hypertarget{ref-Kozich2013}{}
19. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.

\hypertarget{ref-Schloss2009}{}
20. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology
\textbf{75}:7537--7541.
doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.

\end{document}

### 2019-06-24 submission/manuscript.Rmd
Fix paragraph spacing and make some edits

@@
 ## Importance
-Antibiotics are a major risk factor for *Clostridium difficile* infections (CDIs), but it is less clear what other factors contribute to CDI incidence, severity, and recurrence.
+Antibiotics are a major risk factor for *Clostridium difficile* infections (CDIs), but it is less clear what other factors contribute to CDI incidence and recurrence.
 Interestingly, other medications besides antibiotics have been shown to alter the microbiota.
-Proton pump inhibitors (PPIs) were associated with alterations in the human intestinal microbiota through observational and interventional stduies and PPI use has also been associated with CDIs.
-We evaluated the ability of the PPI omeprazole to alter the murine intestinal microbiota and disrupt resistance to *C.
+Proton pump inhibitors (PPIs) are associated with alterations in the human intestinal microbiota through observational and interventional stduies and PPI use has also been associated with CDIs.
+We evaluated the ability of the PPI omeprazole to alter the murine intestinal microbiota and disrupt colonization resistance to *C.
 difficile*.
@@
 Thus, it is unclear whether the gastrointestinal microbiome changes associated with PPI use play a role in the association between PPIs and CDI incidence or recurrence.
-Additionally, epidemiological studies have a limited capacity to address potential confounders and comorbidities in patients that were on PPIs and developed CDI infections or recurrent CDIs [@Elias2019; @tariq2017association].
+Additionally, epidemiological studies have a limited capacity to address potential confounders and comorbidities in patients that were on PPIs and developed CDIs or recurrent CDIs [@Elias2019; @tariq2017association].
 Although there have been a few *C.
@@
 A principle coordinates analysis (PCoA) of the Bray-Curtis distances over the initial 7 days of treatment revealed the bacterial communities of PPI-treated mice remained relatively unchanged (Figure 1B).
-Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes that have previously been impacted by PPIs in human studies, but we see no significant fluctuations in the stool of our PPI-treated mice throughout the course of the 16-day experiment (Figure 1C).In contrast, the microbiomes from the 2 groups of mice that received clindamycin started to shift away from the untreated and PPI-treated mice the day after treatment (Figure 1D).
+Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes that have previously been impacted by PPIs in human studies, but we see no significant fluctuations in the stool of our PPI-treated mice throughout the course of the 16-day experiment (Figure 1C).
+In contrast, the microbiomes from the 2 groups of mice that received clindamycin started to shift away from the untreated and PPI-treated mice the day after treatment (Figure 1D).
 
@@
 difficile*, suggesting there was no difference in rate of clearance for the mice that were initially colonized with *C.
-difficile*.
+difficile* despite continuing to receive PPI treatment throughout the course of the experiment.
 Our results suggest that PPI treatment alone had minor effects on microbial community resistance to *C.
 difficile* colonization in mice.
-Instead most of the differences between our 3 treatment groups appear to be driven by clindamycin administration and included decreased *Alistipes*, *Barnesiella*, Porhyromonadaceae, Ruminococcaceae [@Schubert2015], taxa previously found to be altered in clindamycin-treated mice that were challenged with *C.
-difficile*.
+Instead most of the differences between our 3 treatment groups appear to be driven by clindamycin administration and included decreased *Alistipes*, *Barnesiella*, Porhyromonadaceae, Ruminococcaceae, taxa previously found to be altered in clindamycin-treated mice that were challenged with *C.
+difficile* [@Schubert2015].
 
@@
 difficile* while the other cage remained resistant.
-Our findings that PPI treatment had minimal impact on the fecal microbiome are comparable to several other mouse studies that indicate PPIs have more of an effect on the small intestinal microbiota compared to the fecal microbiota [@Yasutomi2018].
-We did not find significant changes for the taxa observed to be significantly impacted by PPI use in human studies, but a closer examination suggests that the significance of these associations is driven by some individuals more than others and the effect size can be relatively small with overall differences on a PCoA plot difficult to distinguish [@Freedberg2015, @Imhann2015].
-Also, most of the studies to date compare different individuals (PPI users versus non-users) or the same individuals before and after treatment [@Naito2018; @Imhann2017], unfortunately these limited microbiota snapshots may be skewed by day-to-day microbiota variations [@LlornsRico2019], which become apparent in the longitudinally sampled mice.
+Our findings that PPI treatment had minimal impact on the fecal microbiome are comparable to another PPI mouse study that indicated PPIs have more of an effect on the small intestinal microbiota compared to the fecal microbiota [@Yasutomi2018].
+We did not find significant changes for the taxa observed to be significantly impacted by PPI use in human studies, but a closer examination suggests that the significance of these associations in humans is driven by some individuals more than others and the effect size can be relatively small with overall differences on a PCoA plot difficult to distinguish [@Freedberg2015, @Imhann2015].
+Also, most of the studies to date compare different individuals (PPI users versus non-users) or the same individuals before and after treatment [@Naito2018; @Imhann2017], unfortunately these limited microbiota snapshots may be skewed by day-to-day microbiota variations [@LlornsRico2019] that would be revealed with more frequent longitudinal sampling.
 Previous work that administered the PPI lansoprazole daily for 2 weeks to mice and then challenged with *C.
@@
 difficile* 630 when challenged with 10^3^ spores [@Jenior2017].
-The other group also used a higher dose of 10^6^ spores or 10^7^ vegetative cells to challenge antibiotic-treated mice or mice treated with both the PPI esomeprazole for 2 days and antibiotics and demonstrated the antibiotic/PPI-treated mice developed more severe CDIs [@hung2015proton].
+Another group also used a higher dose of 10^6^ spores or 10^7^ vegetative cells and a different *C.
+difficile* strain to challenge antibiotic-treated mice or mice treated with both the PPI esomeprazole for 2 days and antibiotics and demonstrated the antibiotic/PPI-treated mice developed more severe CDIs [@hung2015proton].
 Our study extended previous work examining PPIs and *C.

### 2019-06-24 submission/manuscript.tex
Fix paragraph spacing and make some edits

@@
 infections (CDIs), but it is less clear what other factors contribute to
-CDI incidence, severity, and recurrence. Interestingly, other
-medications besides antibiotics have been shown to alter the microbiota.
-Proton pump inhibitors (PPIs) were associated with alterations in the
-human intestinal microbiota through observational and interventional
-stduies and PPI use has also been associated with CDIs. We evaluated the
-ability of the PPI omeprazole to alter the murine intestinal microbiota
-and disrupt resistance to \emph{C. difficile}. We found PPI treatment
-had minimal impact on the murine intestinal microbiota and was not
-enough to promote \emph{C. difficile} colonization after challenge.
+CDI incidence and recurrence. Interestingly, other medications besides
+antibiotics have been shown to alter the microbiota. Proton pump
+inhibitors (PPIs) are associated with alterations in the human
+intestinal microbiota through observational and interventional stduies
+and PPI use has also been associated with CDIs. We evaluated the ability
+of the PPI omeprazole to alter the murine intestinal microbiota and
+disrupt colonization resistance to \emph{C. difficile}. We found PPI
+treatment had minimal impact on the murine intestinal microbiota and was
+not enough to promote \emph{C. difficile} colonization after challenge.
 Further studies are needed to determine whether other factors such as
@@
 \textasciitilde{}16\% of the 40 intestinal bacteria strains that were
-screened (5). Antibiotics have a large impact on the intestinal
-microbiome and are a primary risk factor for developing
-\emph{Clostridium difficile} infections (CDIs) (6). It is less clear
-whether other human medications that impact the microbiota also
-influence \emph{C. difficile} colonization resistance. Multiple
-epidemiological studies have suggested an association between PPI use
-and incidence or recurrence of CDIs (2, 4, 7, 8). However, most of the
-studies suggesting a link between PPIs and \emph{C. difficile} are
-retrospective and did not evaluate the microbiome (2, 7, 8). Thus, it is
-unclear whether the gastrointestinal microbiome changes associated with
-PPI use play a role in the association between PPIs and CDI incidence or
-recurrence. Additionally, epidemiological studies have a limited
-capacity to address potential confounders and comorbidities in patients
-that were on PPIs and developed CDI infections or recurrent CDIs (7, 8).
+screened (5).\\
+Antibiotics have a large impact on the intestinal microbiome and are a
+primary risk factor for developing \emph{Clostridium difficile}
+infections (CDIs) (6). It is less clear whether other human medications
+that impact the microbiota also influence \emph{C. difficile}
+colonization resistance. Multiple epidemiological studies have suggested
+an association between PPI use and incidence or recurrence of CDIs (2,
+4, 7, 8). However, most of the studies suggesting a link between PPIs
+and \emph{C. difficile} are retrospective and did not evaluate the
+microbiome (2, 7, 8). Thus, it is unclear whether the gastrointestinal
+microbiome changes associated with PPI use play a role in the
+association between PPIs and CDI incidence or recurrence. Additionally,
+epidemiological studies have a limited capacity to address potential
+confounders and comorbidities in patients that were on PPIs and
+developed CDIs or recurrent CDIs (7, 8).\\
 Although there have been a few \emph{C. difficile} mouse model studies
@@
 no significant fluctuations in the stool of our PPI-treated mice
-throughout the course of the 16-day experiment (Figure 1C).In contrast,
+throughout the course of the 16-day experiment (Figure 1C). In contrast,
 the microbiomes from the 2 groups of mice that received clindamycin
@@
 there was no difference in rate of clearance for the mice that were
-initially colonized with \emph{C. difficile}. Our results suggest that
-PPI treatment alone had minor effects on microbial community resistance
-to \emph{C. difficile} colonization in mice. Instead most of the
-differences between our 3 treatment groups appear to be driven by
-clindamycin administration and included decreased \emph{Alistipes},
-\emph{Barnesiella}, Porhyromonadaceae, Ruminococcaceae (6), taxa
-previously found to be altered in clindamycin-treated mice that were
-challenged with \emph{C. difficile}.
+initially colonized with \emph{C. difficile} despite continuing to
+receive PPI treatment throughout the course of the experiment. Our
+results suggest that PPI treatment alone had minor effects on microbial
+community resistance to \emph{C. difficile} colonization in mice.
+Instead most of the differences between our 3 treatment groups appear to
+be driven by clindamycin administration and included decreased
+\emph{Alistipes}, \emph{Barnesiella}, Porhyromonadaceae,
+Ruminococcaceae, taxa previously found to be altered in
+clindamycin-treated mice that were challenged with \emph{C. difficile}
+(6).
 
@@
 PPI and clindamycin treatment had a mixed effect, one cage was colonized
-with \emph{C. difficile} while the other cage remained resistant. Our
-findings that PPI treatment had minimal impact on the fecal microbiome
-are comparable to several other mouse studies that indicate PPIs have
-more of an effect on the small intestinal microbiota compared to the
-fecal microbiota (13). We did not find significant changes for the taxa
-observed to be significantly impacted by PPI use in human studies, but a
-closer examination suggests that the significance of these associations
-is driven by some individuals more than others and the effect size can
-be relatively small with overall differences on a PCoA plot difficult to
-distinguish (14). Also, most of the studies to date compare different
-individuals (PPI users versus non-users) or the same individuals before
-and after treatment (3, 4), unfortunately these limited microbiota
-snapshots may be skewed by day-to-day microbiota variations (16), which
-become apparent in the longitudinally sampled mice. Previous work that
-administered the PPI lansoprazole daily for 2 weeks to mice and then
-challenged with \emph{C. difficile} demonstrated that PPI treatment
-alone resulted in detectable \emph{C. difficile} in the stool 1 week
-after challenge, but also showed there was detectable \emph{C.
-difficile} in mice not treated with antibiotics (10, 11). The presence
-of \emph{C. difficile} in mice that were not treated with either
-antibiotics or PPIs could be attributed to the higher dose of vegetative
-cells (10\textsuperscript{8} CFU) and the different \emph{C. difficile}
-strain used to challenge the mice and may partly explain their
-observation of \emph{C. difficile} in PPI-treated mice (10, 11). We have
-previously shown mice from are colony that are not given antibiotics are
-resistant to \emph{C. difficile} 630 when challenged with
-10\textsuperscript{3} spores (17). The other group also used a higher
+with \emph{C. difficile} while the other cage remained resistant.\\
+Our findings that PPI treatment had minimal impact on the fecal
+microbiome are comparable to another PPI mouse study that indicated PPIs
+have more of an effect on the small intestinal microbiota compared to
+the fecal microbiota (13). We did not find significant changes for the
+taxa observed to be significantly impacted by PPI use in human studies,
+but a closer examination suggests that the significance of these
+associations in humans is driven by some individuals more than others
+and the effect size can be relatively small with overall differences on
+a PCoA plot difficult to distinguish (14). Also, most of the studies to
+date compare different individuals (PPI users versus non-users) or the
+same individuals before and after treatment (3, 4), unfortunately these
+limited microbiota snapshots may be skewed by day-to-day microbiota
+variations (16) that would be revealed with more frequent longitudinal
+sampling.\\
+Previous work that administered the PPI lansoprazole daily for 2 weeks
+to mice and then challenged with \emph{C. difficile} demonstrated that
+PPI treatment alone resulted in detectable \emph{C. difficile} in the
+stool 1 week after challenge, but also showed there was detectable
+\emph{C. difficile} in mice not treated with antibiotics (10, 11). The
+presence of \emph{C. difficile} in mice that were not treated with
+either antibiotics or PPIs could be attributed to the higher dose of
+vegetative cells (10\textsuperscript{8} CFU) and the different \emph{C.
+difficile} strain used to challenge the mice and may partly explain
+their observation of \emph{C. difficile} in PPI-treated mice (10, 11).
+We have previously shown mice from are colony that are not given
+antibiotics are resistant to \emph{C. difficile} 630 when challenged
+with 10\textsuperscript{3} spores (17). Another group also used a higher
 dose of 10\textsuperscript{6} spores or 10\textsuperscript{7} vegetative
-cells to challenge antibiotic-treated mice or mice treated with both the
-PPI esomeprazole for 2 days and antibiotics and demonstrated the
-antibiotic/PPI-treated mice developed more severe CDIs (9). Our study
-extended previous work examining PPIs and \emph{C. difficile} in mice to
-examine the contribution of the intestinal microbiota. We found the PPI
-omeprazole had minimal impact on the murine intestinal microbiota and in
-contrast to previous work, PPIs did not promote \emph{C. difficile}
-colonization. Our 16S rRNA sequencing suggested that
+cells and a different \emph{C. difficile} strain to challenge
+antibiotic-treated mice or mice treated with both the PPI esomeprazole
+for 2 days and antibiotics and demonstrated the antibiotic/PPI-treated
+mice developed more severe CDIs (9).\\
+Our study extended previous work examining PPIs and \emph{C. difficile}
+in mice to examine the contribution of the intestinal microbiota. We
+found the PPI omeprazole had minimal impact on the murine intestinal
+microbiota and in contrast to previous work, PPIs did not promote
+\emph{C. difficile} colonization. Our 16S rRNA sequencing suggested that
 \emph{Streptococcus} and \emph{Enterococcus} are rare taxa in our

### 2019-06-24 code/ppi_taxa.R
Fix wrong figure component

@@ -211,7 +211,7 @@ agg_family_data %>%
   geom_hline(yintercept=1/3000, color="gray")+
   geom_boxplot(outlier.shape = NA)+
   geom_jitter(shape=19, size=1, alpha=0.7, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.2)) +
-  labs(title="Family significantly associated with treatment group", 
+  labs(title="Families significantly associated with treatment group", 
        x=NULL,
        y="Relative abundance (%)")+
   scale_y_log10(breaks=c(1e-4, 1e-3, 1e-2, 1e-1, 1), labels=c(1e-2, 1e-1, 1, 10, 100))+

### 2019-07-01 code/figure_s1.R
Adding supplemental figure 1

@@ -1,0 +1,11 @@
+library(tidyverse)
+library(magick)
+library(cowplot)
+
+a <- ggdraw() + draw_image("results/figures/lactobacillaceae_time.png")
+b <- ggdraw() + draw_image("results/figures/ruminococcaceae_time.png")
+
+plot_grid(a, b, labels = "AUTO")+
+  ggsave("results/figures/figure_s1.pdf", width=6.5, height=3)
+
+

### 2019-07-01 code/ppi_taxa.R

@@ -379,7 +379,9 @@ agg_family_data %>%
   geom_point()+
   geom_line()+
   geom_hline(yintercept=1/3000, color="gray")+
-  theme_classic()
+  labs(title="Lactobacillaceae") +
+  theme_classic()+
+  ggsave("results/figures/lactobacillaceae_time.png")
 
 agg_family_data %>% 
   filter(family == "Micrococcaceae") %>% 
@@ -423,7 +425,9 @@ agg_family_data %>%
   geom_point()+
   geom_line()+
   geom_hline(yintercept=1/3000, color="gray")+
-  theme_classic()
+  labs(title="Ruminococcaceae") +
+  theme_classic()+
+  ggsave("results/figures/ruminococcaceae_time.png")
 
 agg_family_data %>% 
   filter(family == "Lachnospiraceae") %>% 

### 2019-07-01 submission/manuscript.Rmd
Revisions and addressing Nick's comments

@@
 We assessed how PPI treatment alters the fecal microbiota and whether PPIs promote CDIs using a CDI mouse model.
-Mice were treated with the PPI omeprazole for 7 days prior to *C.
+Mice were treated with a high daily dose of the PPI omeprazole for 7 days prior to *C.
 difficile* challenge and were compared to mice that were treated with the antibiotic clindamycin or both omeprazole and clindamycin.
+High-dose PPI treatment was maintained throughout the experiment, which included *C.
+difficile* challenge and the following 9 days.
 We found that omeprazole was not sufficient to promote *C.
@@
 Interestingly, other medications besides antibiotics have been shown to alter the microbiota.
-Proton pump inhibitors (PPIs) are associated with alterations in the human intestinal microbiota through observational and interventional stduies and PPI use has also been associated with CDIs.
-We evaluated the ability of the PPI omeprazole to alter the murine intestinal microbiota and disrupt colonization resistance to *C.
+Proton pump inhibitors (PPIs) are associated with alterations in the human intestinal microbiota through observational and interventional studies and PPI use has also been associated with CDI incidence and recurrence.
+We evaluated the ability of a high dose of the PPI omeprazole to alter the murine intestinal microbiota and disrupt colonization resistance to *C.
 difficile*.
-We found PPI treatment had minimal impact on the murine intestinal microbiota and was not enough to promote *C.
+We found daily PPI treatment had minimal impact on the murine intestinal microbiota and was not enough to promote *C.
 difficile* colonization after challenge.
@@
 difficile* mouse model studies that have demonstrated PPIs have some effect on CDIs with or without additional antibiotic treatment, none of these studies looked at the interplay with the intestinal microbiome [@hung2015proton; @Kaur2007; @kaur2011effect].
-Here, we evaluated the impact of a PPI on the murine microbiome and susceptibility to *C.
+Here, we evaluated the impact of a daily high dose PPI treatment on the murine microbiome and susceptibility to *C.
 difficile* colonization in relation to clindamycin, an antibiotic that perturbs the microbiome enough to allow *C.
@@
 A principle coordinates analysis (PCoA) of the Bray-Curtis distances over the initial 7 days of treatment revealed the bacterial communities of PPI-treated mice remained relatively unchanged (Figure 1B).
-Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes that have previously been impacted by PPIs in human studies, but we see no significant fluctuations in the stool of our PPI-treated mice throughout the course of the 16-day experiment (Figure 1C).
-In contrast, the microbiomes from the 2 groups of mice that received clindamycin started to shift away from the untreated and PPI-treated mice the day after treatment (Figure 1D).
+Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes that have previously been impacted by PPIs in human studies, but we see no significant fluctuations in the stool of our PPI-treated mice throughout the course of the 16-day experiment (Figure 1C, Figure S1).
+In contrast, the microbiomes from the 2 groups of mice that received clindamycin started to shift away from the PPI-treated mice and the samples from previous days one day after treatment (Figure 1D).
 
@@
 
-Our results demonstrate the PPI omeprazole had minimal impact on the stool microbiomes of mice and did not promote susceptibility to *C.
+Our results demonstrate that daily treatment with a high dose of the PPI omeprazole had minimal impact on the stool microbiomes of mice and did not promote susceptibility to *C.
 difficile* colonization.
@@
 Combining PPI and clindamycin treatment had a mixed effect, one cage was colonized with *C.
-difficile* while the other cage remained resistant.
+difficile* and cleared within 10 days, while the other cage remained resistant.
 Our findings that PPI treatment had minimal impact on the fecal microbiome are comparable to another PPI mouse study that indicated PPIs have more of an effect on the small intestinal microbiota compared to the fecal microbiota [@Yasutomi2018].
-We did not find significant changes for the taxa observed to be significantly impacted by PPI use in human studies, but a closer examination suggests that the significance of these associations in humans is driven by some individuals more than others and the effect size can be relatively small with overall differences on a PCoA plot difficult to distinguish [@Freedberg2015, @Imhann2015].
+We did not find significant changes for the taxa observed to be significantly impacted by PPI use in human studies.
+However, 3 of the families that typically increase were absent or at low abundance in the mice from the beginning and may be one contributing factor to why we did not see any changes.
+Additionally, some of the significance of PPI associations in human interventional trials appears to be driven by a handful of specific taxa with overall differences on a PCoA plot difficult to distinguish [@Freedberg2015, @Imhann2015].
 Also, most of the studies to date compare different individuals (PPI users versus non-users) or the same individuals before and after treatment [@Naito2018; @Imhann2017], unfortunately these limited microbiota snapshots may be skewed by day-to-day microbiota variations [@LlornsRico2019] that would be revealed with more frequent longitudinal sampling.
-Previous work that administered the PPI lansoprazole daily for 2 weeks to mice and then challenged with *C.
+Previous work that administered 0.5 mg/kg of the PPI lansoprazole daily for 2 weeks to mice and then challenged with *C.
 difficile* demonstrated that PPI treatment alone resulted in detectable *C.
@@
 The presence of *C.
-difficile* in mice that were not treated with either antibiotics or PPIs could be attributed to the higher dose of vegetative cells (10^8^ CFU) and the different *C.
-difficile* strain used to challenge the mice and may partly explain their observation of *C.
+difficile* in mice that were not treated with either antibiotics or PPIs could be attributed to the higher dose of vegetative cells (10^8^ CFU) used to challenge the mice and may partly explain their observation of *C.
 difficile* in PPI-treated mice [@Kaur2007; @kaur2011effect].
-We have previously shown mice from are colony that are not given antibiotics are resistant to *C.
+We have previously shown mice from our colony that are not given antibiotics are resistant to *C.
 difficile* 630 when challenged with 10^3^ spores [@Jenior2017].
-Another group also used a higher dose of 10^6^ spores or 10^7^ vegetative cells and a different *C.
-difficile* strain to challenge antibiotic-treated mice or mice treated with both the PPI esomeprazole for 2 days and antibiotics and demonstrated the antibiotic/PPI-treated mice developed more severe CDIs [@hung2015proton].
+Another group also used a higher dose of 10^6^ spores or 10^7^ vegetative cells to challenge antibiotic-treated mice or mice treated with both the PPI esomeprazole (40mg/kg dose) for 2 days and antibiotics and demonstrated the antibiotic/PPI-treated mice developed more severe CDIs [@hung2015proton].
 Our study extended previous work examining PPIs and *C.
@@
 difficile* colonization.
-Our 16S rRNA sequencing suggested that *Streptococcus* and *Enterococcus* are rare taxa in our C57BL/6 mouse colony.
-If these taxa are important contributors to the associations between PPIs and CDIs in humans, this may be part of the reason we did not see an effect in our CDI mouse model.
+Our 16S rRNA sequencing suggested that *Streptococcus* and *Enterococcus* are rare genera in our C57BL/6 mouse colony.
+If these genera are important contributors to the associations between PPIs and CDIs in humans, this may be part of the reason we did not see an effect in our CDI mouse model.
 There could also be differences at the species level for taxa that are altered with PPIs in humans compared to mice, which may also contribute to the association.
 Other murine starting communities or the addition of human PPI treatment-associated strains may be needed to determine whether PPIs can alter taxa and subsequently promote CDIs.
-Factors such as age, body mass index, comorbidities, and use of other medications in human studies may also be contributing to the association between PPIs and CDIs.
+Factors such as age, body mass index, comorbidities, and use of other medications in human studies may also be contributing to the association between PPIs and CDI incidence or recurrence.
 NSAID treatment can increase CDI severity in mice [@Maseda2019].
 Given that PPIs are sometimes used in combination with NSAIDs [@nehra2018proton], it may be worth examining both medications together using a mouse model.
-This study only looked at the impact of PPIs with or without antibiotics, but future work may want to incorporate age, other comorbidities and bacterial strains that are less common in mice to evaluate whether PPIs in conjunction with other factors can increase the risk of CDIs.
+This study only looked at the impact of PPIs with or without antibiotics, but future work may want to incorporate age, other comorbidities and bacterial strains that are less common in mice to evaluate whether PPIs in conjunction with other factors can increase the risk of CDIs or recurrent CDIs.
+
+## Acknowledgements
+
+This research was supported by NIH grant U01AI12455.
+We would also like to thank the Unit for Laboratory Animal Medicine at the University of Michigan for maintaining our mouse colony and providing the infrastructure and support for performing our mouse expeirments.
+The authors are also thankful to members of the Schloss lab for helpful discussions throughout the process of designing the experiment, analyzing the results and drafting of the manuscript.
+
+\newpage
 
@@
 Omeprazole was prepared from 20 mg/mL frozen aliquots and diluted to an 8mg/mL prior to gavage.
-All mice received ~40 mg/kg omeprazole or vehicle solution once per day through the duration of the experiment with treatment starting 7 days before *C.
+All mice received 40 mg/kg omeprazole or vehicle solution once per day through the duration of the experiment with treatment starting 7 days before *C.
 difficile* challenge (Figure 1A).
 One day prior to *C.
-difficile* challenge, mice received an intraperitoneal injection of 10 mg/kg clindamycin or sterile saline vehicle.
+difficile* challenge, 2 groups of mice received an intraperitoneal injection of 10 mg/kg clindamycin or sterile saline vehicle.
 All drugs were filter sterilized through a 0.22 micron syringe filter before administration to animals.
 
-**C. difficile infection model**
+***C. difficile* infection model**
 Mice were challenged with *C.
@@
 Mice were challenged with 10^3^ spores in ultrapure distilled water.
-Samples were collected for 16S rRNA sequencing or *C.
+Stool samples were collected for 16S rRNA sequencing or *C.
 difficile* CFU quantification throughout the duration of the experiments at the indicated timepoints (Figure 1A).
@@
 
-**Statistical Analysis**
+**Statistical Analysis** To assess for differences in relative abundance in families and genera across our 3 different treatment groups (Clindamycin, Clindamycin + PPI, and PPI), we used a Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
 
@@
 **Data availability** The 16S rRNA sequencing data have been deposited in the NCBI Sequence Read Archive (Accession no. _________)
+
+\newpage
+
+## References
+
+<div id="refs"></div>
 
@@
 Mouse experiment timeline and logistics.
+Stools for 16S rRNA sequencing analysis were collected on the days that are labeled (Day -7, -5, -3, -1, 0, 1, 2, 3, 4, 5, 7, 9).
 B.
@@
 difficile challenge for clindamycin, clindamycin/PPI, and PPI-treated mice.
+Lines represent the mean CFU/g for each treatment group while points represent CFU/g for individual mice within each group.
 B.
@@
 C.
-Genera significantly associated with treatment groups.
+Genera significantly associated with treatment groups for samples across all sequenced timepoints.
 D.
-Families significantly associated with treatment groups.
+Families significantly associated with treatment groups for samples across all sequenced timepoints.
+C,D data were analyzed by Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
 
-\newpage
-
-## References
+**Figure S1.
+Previous human PPI-associated families fluctuate over time with no overall trend in either direction** Relative abundance over time for Lactobacillaceae (A) and Ruminococcaceae (B), 2 of the PPI-associated familes from human PPI studies across all 3 treatment groups.
+Each point represents the relative abundance for an individual mouse

### 2019-07-01 submission/manuscript.tex
Revisions and addressing Nick's comments

@@
 fecal microbiota and whether PPIs promote CDIs using a CDI mouse model.
-Mice were treated with the PPI omeprazole for 7 days prior to \emph{C.
-difficile} challenge and were compared to mice that were treated with
-the antibiotic clindamycin or both omeprazole and clindamycin. We found
-that omeprazole was not sufficient to promote \emph{C. difficile}
-colonization. When omeprazole treatment was combined with the
-antibiotic, there was a mixed level of \emph{C. difficile} colonization,
-where one cage of mice remained resistant and the other cage was
-colonized. 16S rRNA sequencing analysis revealed that omeprazole had
-minimal impact on the murine microbiota throughout the 16 days of PPI
-exposure. \emph{C. difficile} colonization of clindamycin-treated mice
-was associated with changes in \emph{Alistipes}, \emph{Barnesiella},
-Porphyromonadaceae, and Ruminococcaceae. These results suggest PPI
-treatment alone is not sufficient to disrupt microbiota resistance to C.
-difficle infection in mice that are normally resistant in the absence of
-antibiotic treatment.
+Mice were treated with a high daily dose of the PPI omeprazole for 7
+days prior to \emph{C. difficile} challenge and were compared to mice
+that were treated with the antibiotic clindamycin or both omeprazole and
+clindamycin. High-dose PPI treatment was maintained throughout the
+experiment, which included \emph{C. difficile} challenge and the
+following 9 days. We found that omeprazole was not sufficient to promote
+\emph{C. difficile} colonization. When omeprazole treatment was combined
+with the antibiotic, there was a mixed level of \emph{C. difficile}
+colonization, where one cage of mice remained resistant and the other
+cage was colonized. 16S rRNA sequencing analysis revealed that
+omeprazole had minimal impact on the murine microbiota throughout the 16
+days of PPI exposure. \emph{C. difficile} colonization of
+clindamycin-treated mice was associated with changes in
+\emph{Alistipes}, \emph{Barnesiella}, Porphyromonadaceae, and
+Ruminococcaceae. These results suggest PPI treatment alone is not
+sufficient to disrupt microbiota resistance to C. difficle infection in
+mice that are normally resistant in the absence of antibiotic treatment.
 
@@
 inhibitors (PPIs) are associated with alterations in the human
-intestinal microbiota through observational and interventional stduies
-and PPI use has also been associated with CDIs. We evaluated the ability
-of the PPI omeprazole to alter the murine intestinal microbiota and
-disrupt colonization resistance to \emph{C. difficile}. We found PPI
-treatment had minimal impact on the murine intestinal microbiota and was
-not enough to promote \emph{C. difficile} colonization after challenge.
-Further studies are needed to determine whether other factors such as
-the composition of the starting bacterial community, comorbidities, and
-use of additional medications contribute to the association between PPIs
-and CDIs.
+intestinal microbiota through observational and interventional studies
+and PPI use has also been associated with CDI incidence and recurrence.
+We evaluated the ability of a high dose of the PPI omeprazole to alter
+the murine intestinal microbiota and disrupt colonization resistance to
+\emph{C. difficile}. We found daily PPI treatment had minimal impact on
+the murine intestinal microbiota and was not enough to promote \emph{C.
+difficile} colonization after challenge. Further studies are needed to
+determine whether other factors such as the composition of the starting
+bacterial community, comorbidities, and use of additional medications
+contribute to the association between PPIs and CDIs.
 
@@
 interplay with the intestinal microbiome (9--11). Here, we evaluated the
-impact of a PPI on the murine microbiome and susceptibility to \emph{C.
-difficile} colonization in relation to clindamycin, an antibiotic that
-perturbs the microbiome enough to allow \emph{C. difficile} to colonize
-but is mild enough that \emph{C. difficile} is cleared within 10 days
-(12).
+impact of a daily high dose PPI treatment on the murine microbiome and
+susceptibility to \emph{C. difficile} colonization in relation to
+clindamycin, an antibiotic that perturbs the microbiome enough to allow
+\emph{C. difficile} to colonize but is mild enough that \emph{C.
+difficile} is cleared within 10 days (12).
 
@@
 no significant fluctuations in the stool of our PPI-treated mice
-throughout the course of the 16-day experiment (Figure 1C). In contrast,
-the microbiomes from the 2 groups of mice that received clindamycin
-started to shift away from the untreated and PPI-treated mice the day
-after treatment (Figure 1D).
+throughout the course of the 16-day experiment (Figure 1C, Figure S1).
+In contrast, the microbiomes from the 2 groups of mice that received
+clindamycin started to shift away from the PPI-treated mice and the
+samples from previous days one day after treatment (Figure 1D).
 
@@
 
-Our results demonstrate the PPI omeprazole had minimal impact on the
-stool microbiomes of mice and did not promote susceptibility to \emph{C.
-difficile} colonization. In contrast, clindamycin-treated mice were
-initially colonized, but cleared the infection within 10 days. Combining
-PPI and clindamycin treatment had a mixed effect, one cage was colonized
-with \emph{C. difficile} while the other cage remained resistant.\\
+Our results demonstrate that daily treatment with a high dose of the PPI
+omeprazole had minimal impact on the stool microbiomes of mice and did
+not promote susceptibility to \emph{C. difficile} colonization. In
+contrast, clindamycin-treated mice were initially colonized, but cleared
+the infection within 10 days. Combining PPI and clindamycin treatment
+had a mixed effect, one cage was colonized with \emph{C. difficile} and
+cleared within 10 days, while the other cage remained resistant.\\
 Our findings that PPI treatment had minimal impact on the fecal
@@
 the fecal microbiota (13). We did not find significant changes for the
-taxa observed to be significantly impacted by PPI use in human studies,
-but a closer examination suggests that the significance of these
-associations in humans is driven by some individuals more than others
-and the effect size can be relatively small with overall differences on
-a PCoA plot difficult to distinguish (14). Also, most of the studies to
+taxa observed to be significantly impacted by PPI use in human studies.
+However, 3 of the families that typically increase were absent or at low
+abundance in the mice from the beginning and may be one contributing
+factor to why we did not see any changes. Additionally, some of the
+significance of PPI associations in human interventional trials appears
+to be driven by a handful of specific taxa with overall differences on a
+PCoA plot difficult to distinguish (14). Also, most of the studies to
 date compare different individuals (PPI users versus non-users) or the
@@
 sampling.\\
-Previous work that administered the PPI lansoprazole daily for 2 weeks
-to mice and then challenged with \emph{C. difficile} demonstrated that
-PPI treatment alone resulted in detectable \emph{C. difficile} in the
-stool 1 week after challenge, but also showed there was detectable
-\emph{C. difficile} in mice not treated with antibiotics (10, 11). The
-presence of \emph{C. difficile} in mice that were not treated with
-either antibiotics or PPIs could be attributed to the higher dose of
-vegetative cells (10\textsuperscript{8} CFU) and the different \emph{C.
-difficile} strain used to challenge the mice and may partly explain
-their observation of \emph{C. difficile} in PPI-treated mice (10, 11).
-We have previously shown mice from are colony that are not given
-antibiotics are resistant to \emph{C. difficile} 630 when challenged
-with 10\textsuperscript{3} spores (17). Another group also used a higher
-dose of 10\textsuperscript{6} spores or 10\textsuperscript{7} vegetative
-cells and a different \emph{C. difficile} strain to challenge
-antibiotic-treated mice or mice treated with both the PPI esomeprazole
-for 2 days and antibiotics and demonstrated the antibiotic/PPI-treated
-mice developed more severe CDIs (9).\\
+Previous work that administered 0.5 mg/kg of the PPI lansoprazole daily
+for 2 weeks to mice and then challenged with \emph{C. difficile}
+demonstrated that PPI treatment alone resulted in detectable \emph{C.
+difficile} in the stool 1 week after challenge, but also showed there
+was detectable \emph{C. difficile} in mice not treated with antibiotics
+(10, 11). The presence of \emph{C. difficile} in mice that were not
+treated with either antibiotics or PPIs could be attributed to the
+higher dose of vegetative cells (10\textsuperscript{8} CFU) used to
+challenge the mice and may partly explain their observation of \emph{C.
+difficile} in PPI-treated mice (10, 11). We have previously shown mice
+from our colony that are not given antibiotics are resistant to \emph{C.
+difficile} 630 when challenged with 10\textsuperscript{3} spores (17).
+Another group also used a higher dose of 10\textsuperscript{6} spores or
+10\textsuperscript{7} vegetative cells to challenge antibiotic-treated
+mice or mice treated with both the PPI esomeprazole (40mg/kg dose) for 2
+days and antibiotics and demonstrated the antibiotic/PPI-treated mice
+developed more severe CDIs (9).\\
 Our study extended previous work examining PPIs and \emph{C. difficile}
@@
 \emph{C. difficile} colonization. Our 16S rRNA sequencing suggested that
-\emph{Streptococcus} and \emph{Enterococcus} are rare taxa in our
-C57BL/6 mouse colony. If these taxa are important contributors to the
+\emph{Streptococcus} and \emph{Enterococcus} are rare genera in our
+C57BL/6 mouse colony. If these genera are important contributors to the
 associations between PPIs and CDIs in humans, this may be part of the
@@
 medications in human studies may also be contributing to the association
-between PPIs and CDIs. NSAID treatment can increase CDI severity in mice
-(18). Given that PPIs are sometimes used in combination with NSAIDs (2),
-it may be worth examining both medications together using a mouse model.
-This study only looked at the impact of PPIs with or without
-antibiotics, but future work may want to incorporate age, other
-comorbidities and bacterial strains that are less common in mice to
-evaluate whether PPIs in conjunction with other factors can increase the
-risk of CDIs.
+between PPIs and CDI incidence or recurrence. NSAID treatment can
+increase CDI severity in mice (18). Given that PPIs are sometimes used
+in combination with NSAIDs (2), it may be worth examining both
+medications together using a mouse model. This study only looked at the
+impact of PPIs with or without antibiotics, but future work may want to
+incorporate age, other comorbidities and bacterial strains that are less
+common in mice to evaluate whether PPIs in conjunction with other
+factors can increase the risk of CDIs or recurrent CDIs.
+
+\subsection{Acknowledgements}\label{acknowledgements}
+
+This research was supported by NIH grant U01AI12455. We would also like
+to thank the Unit for Laboratory Animal Medicine at the University of
+Michigan for maintaining our mouse colony and providing the
+infrastructure and support for performing our mouse expeirments. The
+authors are also thankful to members of the Schloss lab for helpful
+discussions throughout the process of designing the experiment,
+analyzing the results and drafting of the manuscript.
+
+\newpage
 
@@
 phosphate buffered saline. Omeprazole was prepared from 20 mg/mL frozen
-aliquots and diluted to an 8mg/mL prior to gavage. All mice received
-\textasciitilde{}40 mg/kg omeprazole or vehicle solution once per day
-through the duration of the experiment with treatment starting 7 days
-before \emph{C. difficile} challenge (Figure 1A). One day prior to
-\emph{C. difficile} challenge, mice received an intraperitoneal
-injection of 10 mg/kg clindamycin or sterile saline vehicle. All drugs
-were filter sterilized through a 0.22 micron syringe filter before
-administration to animals.
+aliquots and diluted to an 8mg/mL prior to gavage. All mice received 40
+mg/kg omeprazole or vehicle solution once per day through the duration
+of the experiment with treatment starting 7 days before \emph{C.
+difficile} challenge (Figure 1A). One day prior to \emph{C. difficile}
+challenge, 2 groups of mice received an intraperitoneal injection of 10
+mg/kg clindamycin or sterile saline vehicle. All drugs were filter
+sterilized through a 0.22 micron syringe filter before administration to
+animals.
 
-\textbf{C. difficile infection model} Mice were challenged with \emph{C.
-difficile} 630 seven days after the start of omeprazole treatment and
-one day after clindamycin treatment. Mice were challenged with
-10\textsuperscript{3} spores in ultrapure distilled water. Samples were
-collected for 16S rRNA sequencing or \emph{C. difficile} CFU
-quantification throughout the duration of the experiments at the
+\textbf{\emph{C. difficile} infection model} Mice were challenged with
+\emph{C. difficile} 630 seven days after the start of omeprazole
+treatment and one day after clindamycin treatment. Mice were challenged
+with 10\textsuperscript{3} spores in ultrapure distilled water. Stool
+samples were collected for 16S rRNA sequencing or \emph{C. difficile}
+CFU quantification throughout the duration of the experiments at the
 indicated timepoints (Figure 1A). Samples for 16S rRNA sequencing were
@@
 
-\textbf{Statistical Analysis}
+\textbf{Statistical Analysis} To assess for differences in relative
+abundance in families and genera across our 3 different treatment groups
+(Clindamycin, Clindamycin + PPI, and PPI), we used a Kruskal-Wallis test
+with a Benjamini-Hochberg correction for multiple comparisons.
 
@@
 
-\subsection{Figures}\label{figures}
-
-\textbf{Figure 1. PPI treatment had minimal impact on the murine fecal
-microbiota} A. Mouse experiment timeline and logistics. B. Principal
-Coordinates Analaysis (PCoA) of Bray-curtis distances from stool samples
-for all treatment groups during the intial 7 days of treatment with the
-PPI omeprazole or vehicle cluster together and do not chage much over
-time. C. Relative abundances of families previously associated with PPI
-use in humans do not change much over the 16-day course of treatment
-with PPIs (7 days before \emph{C. difficile} challenge through 9 days
-post \emph{C. difficile} challenge). D. PCoA of stool samples collected
-before and 1 day after antibiotic treatment, demonstrating the fecal
-microbiota of mice treated with clindamycin starts to separate from the
-rest of the samples.
-
-\textbf{Figure 2. PPI treatment alone does not promote CDIs in mice} A.
-C. difficile CFUs/g stool measured each day post C. difficile challenge
-for clindamycin, clindamycin/PPI, and PPI-treated mice. B. PCoA of stool
-samples collected after antibiotic treatments. C. Genera significantly
-associated with treatment groups. D. Families significantly associated
-with treatment groups.
-
-\newpage
-
-\subsection*{References}\label{references}
-\addcontentsline{toc}{subsection}{References}
+\subsection{References}\label{references}
 
@@
 
+\newpage
+
+\subsection{Figures}\label{figures}
+
+\textbf{Figure 1. PPI treatment had minimal impact on the murine fecal
+microbiota} A. Mouse experiment timeline and logistics. Stools for 16S
+rRNA sequencing analysis were collected on the days that are labeled
+(Day -7, -5, -3, -1, 0, 1, 2, 3, 4, 5, 7, 9). B. Principal Coordinates
+Analaysis (PCoA) of Bray-curtis distances from stool samples for all
+treatment groups during the intial 7 days of treatment with the PPI
+omeprazole or vehicle cluster together and do not chage much over time.
+C. Relative abundances of families previously associated with PPI use in
+humans do not change much over the 16-day course of treatment with PPIs
+(7 days before \emph{C. difficile} challenge through 9 days post
+\emph{C. difficile} challenge). D. PCoA of stool samples collected
+before and 1 day after antibiotic treatment, demonstrating the fecal
+microbiota of mice treated with clindamycin starts to separate from the
+rest of the samples.
+
+\textbf{Figure 2. PPI treatment alone does not promote CDIs in mice} A.
+C. difficile CFUs/g stool measured each day post C. difficile challenge
+for clindamycin, clindamycin/PPI, and PPI-treated mice. Lines represent
+the mean CFU/g for each treatment group while points represent CFU/g for
+individual mice within each group. B. PCoA of stool samples collected
+after antibiotic treatments. C. Genera significantly associated with
+treatment groups for samples across all sequenced timepoints. D.
+Families significantly associated with treatment groups for samples
+across all sequenced timepoints. C,D data were analyzed by
+Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple
+comparisons.
+
+\textbf{Figure S1. Previous human PPI-associated families fluctuate over
+time with no overall trend in either direction} Relative abundance over
+time for Lactobacillaceae (A) and Ruminococcaceae (B), 2 of the
+PPI-associated familes from human PPI studies across all 3 treatment
+groups. Each point represents the relative abundance for an individual
+mouse
+
 \end{document}

### 2019-07-01 code/ppi_pcoa.R

@@ -74,8 +74,6 @@ before_abx <- pcoa %>% filter(day < -1) %>%
   theme(plot.title = element_text(hjust = 0.5))+
   ggsave("results/figures/before_abx.png")
   
-  
-
 #plot before abx with D0 removed
 pcoa %>% filter(day < -1) %>% filter(day > -7) %>% 
   ggplot(aes(x=axis1, y=axis2, color=Group, alpha = day)) +

### 2019-07-11 code/figure_1.R
Saved plots to submission folder so figures can go into markdown

@@ -8,6 +8,7 @@ c <- ggdraw() + draw_image("results/figures/families_prev_assoc_w_PPIs.png")
 d <- ggdraw() + draw_image("results/figures/before_plus_day_after_abx.png")
 
 plot_grid(a, b, c, d, labels = "AUTO")+
-  ggsave("results/figures/figure_1.pdf", width=6.5, height=6)
+  ggsave("results/figures/figure_1.pdf", width=6.5, height=6)+
+  ggsave("submission/figure_1.pdf", width=6.5, height=6)
 
 

### 2019-07-11 code/figure_2.R

@@ -8,4 +8,6 @@ c <- ggdraw() + draw_image("results/figures/genera_assoc_w_treatment.png")
 d <- ggdraw() + draw_image("results/figures/family_assoc_w_treatment.png")
 
 plot_grid(a, b, c, d, labels = "AUTO")+
-  ggsave("results/figures/figure_2.pdf", width=6.5, height=6)
+  ggsave("results/figures/figure_2.pdf", width=6.5, height=6)+
+  ggsave("submission/figure_2.pdf", width=6.5, height=6)
+

### 2019-07-11 code/figure_s1.R

@@ -6,6 +6,8 @@ a <- ggdraw() + draw_image("results/figures/lactobacillaceae_time.png")
 b <- ggdraw() + draw_image("results/figures/ruminococcaceae_time.png")
 
 plot_grid(a, b, labels = "AUTO")+
-  ggsave("results/figures/figure_s1.pdf", width=6.5, height=3)
+  ggsave("results/figures/figure_s1.pdf", width=6.5, height=3)+
+  ggsave("submission/figure_s1.pdf", width=6.5, height=3)
+
 
 

### 2019-07-11 code/mimarks_ncbi_info.R
Grabbing some metadata columns for the mimarks file to submit to NCBI

@@ -1,0 +1,21 @@
+#Load packages
+library(tidyverse)
+library(lubridate)
+library(janitor)
+
+# Import metadata into data frame
+metadata <- read.table('data/process/ppi_metadata.txt', header = T, sep = '\t', stringsAsFactors = F) %>% 
+  #Unite 2 tube label columns into a new column "description". This conveys group name, mouse ID, and collection date.
+  unite(sample_title, Tube.label, Tube.label__1) %>% 
+  mutate(description = Group) %>% 
+  mutate(collection_date = excel_numeric_to_date(Date.Collected)) %>% #Convert excel numeric values to date values with excel_numeric_to_date function from the janitor package
+  select(sample_title, description, collection_date, shared_names) #Select columns that are missing from mimarks tsv file
+
+
+# Import MIMARKS ppi.tsv file into data frame
+mimarks <- read_tsv('data/raw/ppi.tsv')
+# Add metadata columns to mimarks file to finish filling out the information for all samples
+mimarks <- left_join(mimarks, metadata, by = c("#This is a tab-delimited file. Additional Documentation can be found at http://www.mothur.org/wiki/MIMarks_Data_Packages." = "shared_names"))
+write_tsv(mimarks, 'data/raw/ppi.tsv', na = " ")
+          
+                     

### 2019-07-11 submission/manuscript.Rmd
Responding to Pat's comments and making additional edits

@@
 difficile* challenge and were compared to mice that were treated with the antibiotic clindamycin or both omeprazole and clindamycin.
-High-dose PPI treatment was maintained throughout the experiment, which included *C.
-difficile* challenge and the following 9 days.
+High-dose PPI treatment was maintained throughout the experiment, which included pre-treatment, the day of *C.
+difficile* challenge, and the following 9 days.
 We found that omeprazole was not sufficient to promote *C.
 difficile* colonization.
-When omeprazole treatment was combined with the antibiotic, there was a mixed level of *C.
-difficile* colonization, where one cage of mice remained resistant and the other cage was colonized.
+When omeprazole treatment was combined with the antibiotic,one cage of mice remained resistant to *C.
+difficile* colonization, while the other cage was colonized.
 16S rRNA sequencing analysis revealed that omeprazole had minimal impact on the murine microbiota throughout the 16 days of PPI exposure.
-*C.
-difficile* colonization of clindamycin-treated mice was associated with changes in *Alistipes*, *Barnesiella*, Porphyromonadaceae, and Ruminococcaceae.
-These results suggest PPI treatment alone is not sufficient to disrupt microbiota resistance to C.
-difficle infection in mice that are normally resistant in the absence of antibiotic treatment.
+These results suggest PPI treatment alone is not sufficient to disrupt microbiota resistance to *C.
+difficle* infection in mice that are normally resistant in the absence of antibiotic treatment.
 
 ## Importance
-Antibiotics are a major risk factor for *Clostridium difficile* infections (CDIs), but it is less clear what other factors contribute to CDI incidence and recurrence.
-Interestingly, other medications besides antibiotics have been shown to alter the microbiota.
-Proton pump inhibitors (PPIs) are associated with alterations in the human intestinal microbiota through observational and interventional studies and PPI use has also been associated with CDI incidence and recurrence.
+Antibiotics are a major risk factor for *Clostridium difficile* infections (CDIs), but other factors may also contribute to CDI incidence and recurrence.
+Interestingly, other medications besides antibiotics can impact the microbiota.
+Proton pump inhibitors (PPIs) are associated with alterations in the human intestinal microbiota through observational and interventional studies and PPI use has also been associated with CDI incidence and recurrence in epidemiological studies.
 We evaluated the ability of a high dose of the PPI omeprazole to alter the murine intestinal microbiota and disrupt colonization resistance to *C.
 difficile*.
-We found daily PPI treatment had minimal impact on the murine intestinal microbiota and was not enough to promote *C.
+We found daily PPI treatment had minimal impact on the murine fecal microbiota and was not enough to promote *C.
 difficile* colonization after challenge.
-Further studies are needed to determine whether other factors such as the composition of the starting bacterial community, comorbidities, and use of additional medications contribute to the association between PPIs and CDIs.
+Further studies are needed to determine whether other factors such as the composition of the starting bacterial community, comorbidities, and use of additional medications contribute to the association between PPIs and CDIs seen in humans.
 
@@
 ## Introduction
-Proton pump inhibitors (PPIs) are one of the most widely used drugs and are available over the counter [@Singh2018].
-PPIs are approved to treat a variety of conditions including gastroesophageal reflux disease, ulcers, or as prophylaxis for nonsteroidal anti-inflammatory drug (NSAID)-associated gastrointestinal bleeding [@nehra2018proton].
-Unfortunately, there have been a number of adverse events associated with their use including small intestinal bacterial overgrowth, decreased calcium absorption, and increased risk of enteric infections [@nehra2018proton].
-There’s also been a number of large cohort studies as well as interventional clinical trials that have demonstrated specific alterations in the intestinal microbiome were associated with PPI use [@Imhann2017; @Naito2018].
-PPI-associated microbiota changes have been attributed to the ability of PPIs to increase stomach acid pH which may promote the survival of oral and pathogenic bacteria [@Imhann2017; @Naito2018].
-Human fecal microbiota changes with PPI use include increases in Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae with decreases in Ruminococcaceae [@Imhann2017].
-An *in vitro screen*, demonstrated that 3 PPIs inhibit the growth of 7-8 strains of Lachnospiraceae and Ruminococcaceae, which represented ~16% of the 40 intestinal bacteria strains that were screened [@maier2018extensive].
 Antibiotics have a large impact on the intestinal microbiome and are a primary risk factor for developing *Clostridium difficile* infections (CDIs) [@Schubert2015].
@@
 difficile* colonization resistance.
-Multiple epidemiological studies have suggested an association between PPI use and incidence or recurrence of CDIs [@tariq2017association; @nehra2018proton; @Naito2018; @Elias2019].
-However, most of the studies suggesting a link between PPIs and *C.
-difficile* are retrospective and did not evaluate the microbiome [@tariq2017association; @nehra2018proton; @Elias2019].
+Multiple epidemiological studies have suggested an association between proton pump inhibitor (PPI) use and incidence or recurrence of CDIs [@tariq2017association; @nehra2018proton; @Naito2018; @Elias2019].
+There have also been a number of large cohort studies as well as interventional clinical trials that have demonstrated specific alterations in the intestinal microbiome were associated with PPI use [@Imhann2017; @Naito2018].
+PPI-associated microbiota changes have been attributed to the ability of PPIs to increase stomach acid pH which may promote the survival of oral and pathogenic bacteria [@Imhann2017; @Naito2018].
+Human fecal microbiota changes with PPI use include increases in Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae and decreases in Ruminococcaceae [@Imhann2017].
+Unfortunately, most of the studies suggesting a link between PPIs and *C.
+difficile* were retrospective and did not evaluate the microbiome [@tariq2017association; @nehra2018proton; @Elias2019].
 Thus, it is unclear whether the gastrointestinal microbiome changes associated with PPI use play a role in the association between PPIs and CDI incidence or recurrence.
 Additionally, epidemiological studies have a limited capacity to address potential confounders and comorbidities in patients that were on PPIs and developed CDIs or recurrent CDIs [@Elias2019; @tariq2017association].
-Although there have been a few *C.
-difficile* mouse model studies that have demonstrated PPIs have some effect on CDIs with or without additional antibiotic treatment, none of these studies looked at the interplay with the intestinal microbiome [@hung2015proton; @Kaur2007; @kaur2011effect].
 Here, we evaluated the impact of a daily high dose PPI treatment on the murine microbiome and susceptibility to *C.
@@
 
-## Results
-
 **Murine fecal microbiomes were minimally affected by PPI treatment **
 To test if PPI treatment alters the microbiome and promotes susceptibility to CDIs, we gavaged mice daily with a high dose of omeprazole for 7 days before *C.
-difficile* challenge (Figure 1A) and compared the PPI-treated mice to a group that received the antibiotic, clindamycin, and a group that received both PPI and clindamycin treatment.
+difficile* challenge (Figure 1A).
+The fecal bacterial communities of PPI-treated mice were compared to a group of mice that received the antibiotic, clindamycin, and a group that received both PPI and clindamycin treatment using 16S rRNA gene sequencing.
 A principle coordinates analysis (PCoA) of the Bray-Curtis distances over the initial 7 days of treatment revealed the bacterial communities of PPI-treated mice remained relatively unchanged (Figure 1B).
-Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes that have previously been impacted by PPIs in human studies, but we see no significant fluctuations in the stool of our PPI-treated mice throughout the course of the 16-day experiment (Figure 1C, Figure S1).
+Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes that have previously been impacted by PPIs in human studies, but we observed no significant fluctuations in the stool of our PPI-treated mice throughout the course of the 16-day experiment (Figure 1C, Figure S1).
 In contrast, the microbiomes from the 2 groups of mice that received clindamycin started to shift away from the PPI-treated mice and the samples from previous days one day after treatment (Figure 1D).
+These results demonstrated that high dose PPI treatment alone had no significant impact on the murine fecal bacterial community.
 
 **PPI-treatment did not promote susceptibility to *C. difficile infection* in mice**
+Next, we examined whether high dose PPI treatment altered susceptibility to *C.
+difficile* infection in mice.
 After 7 days of PPI treatment or 1 day after clindamycin treatment, mice were challenged with *C.
@@
 difficile* despite continuing to receive PPI treatment throughout the course of the experiment.
-Our results suggest that PPI treatment alone had minor effects on microbial community resistance to *C.
+Our results suggest that PPI treatment alone had no effects on bacterial community resistance to *C.
 difficile* colonization in mice.
@@
 difficile* [@Schubert2015].
+These findings demonstrated that high dose PPI treatment did not promote susceptibility to *C.
+difficile* colonization.
 
 ## Discussion and Conclusions
-
-Our results demonstrate that daily treatment with a high dose of the PPI omeprazole had minimal impact on the stool microbiomes of mice and did not promote susceptibility to *C.
-difficile* colonization.
-In contrast, clindamycin-treated mice were initially colonized, but cleared the infection within 10 days.
-Combining PPI and clindamycin treatment had a mixed effect, one cage was colonized with *C.
-difficile* and cleared within 10 days, while the other cage remained resistant.
-Our findings that PPI treatment had minimal impact on the fecal microbiome are comparable to another PPI mouse study that indicated PPIs have more of an effect on the small intestinal microbiota compared to the fecal microbiota [@Yasutomi2018].
+Our findings that PPI treatment had minimal impact on the fecal microbiome were comparable to another PPI mouse study that indicated PPIs have more of an effect on the small intestinal microbiota compared to the fecal microbiota [@Yasutomi2018].
+The same group demonstrated PPI treatment increased the stomach pH in mice [@Yasutomi2018], which could improve survival of bacteria passing through the stomach.
 We did not find significant changes for the taxa observed to be significantly impacted by PPI use in human studies.
-However, 3 of the families that typically increase were absent or at low abundance in the mice from the beginning and may be one contributing factor to why we did not see any changes.
+However, 3 of the families that typically increase were absent or at low abundance in the mice from the beginning and may be one contributing factor to our observation that PPIs had no significant effects on bacterial taxa.
 Additionally, some of the significance of PPI associations in human interventional trials appears to be driven by a handful of specific taxa with overall differences on a PCoA plot difficult to distinguish [@Freedberg2015, @Imhann2015].
-Also, most of the studies to date compare different individuals (PPI users versus non-users) or the same individuals before and after treatment [@Naito2018; @Imhann2017], unfortunately these limited microbiota snapshots may be skewed by day-to-day microbiota variations [@LlornsRico2019] that would be revealed with more frequent longitudinal sampling.
-Previous work that administered 0.5 mg/kg of the PPI lansoprazole daily for 2 weeks to mice and then challenged with *C.
+Also, most of the studies to date compared different individuals (PPI users versus non-users) or the same individuals before and after treatment [@Naito2018; @Imhann2017], unfortunately these limited microbiota snapshots may be skewed by day-to-day microbiota variations [@LlornsRico2019] that would be revealed with more frequent longitudinal sampling.
+ALthough there have been a few *C.
+difficile* mouse model studies that have demonstrated PPIs have some effect on CDIs with or without additional antibiotic treatment[@Kaur2007; @kaur2011effect; @hung2015proton], there were some key methodological differences between the previous studies and our own.
+One group administered 0.5 mg/kg of the PPI lansoprazole daily for 2 weeks to mice and then challenged with *C.
 difficile* demonstrated that PPI treatment alone resulted in detectable *C.
@@
 The presence of *C.
-difficile* in mice that were not treated with either antibiotics or PPIs could be attributed to the higher dose of vegetative cells (10^8^ CFU) used to challenge the mice and may partly explain their observation of *C.
+difficile* in mice that were not treated with either antibiotics or PPIs could be attributed to the higher dose of vegetative cells (10^8^ CFU) used to challenge the mice and may partly explain the observation of *C.
 difficile* in PPI-treated mice [@Kaur2007; @kaur2011effect].
-We have previously shown mice from our colony that are not given antibiotics are resistant to *C.
+We have previously shown mice from our colony that were not given antibiotics were resistant to *C.
 difficile* 630 when challenged with 10^3^ spores [@Jenior2017].
-Another group also used a higher dose of 10^6^ spores or 10^7^ vegetative cells to challenge antibiotic-treated mice or mice treated with both the PPI esomeprazole (40mg/kg dose) for 2 days and antibiotics and demonstrated the antibiotic/PPI-treated mice developed more severe CDIs [@hung2015proton].
+The other group also used a higher dose of 10^6^ spores or 10^7^ vegetative cells to challenge antibiotic-treated mice or mice treated with both the PPI esomeprazole (40mg/kg dose) for 2 days and antibiotics and demonstrated the antibiotic/PPI-treated mice developed more severe CDIs [@hung2015proton].
 Our study extended previous work examining PPIs and *C.
@@
 difficile* colonization.
-Our 16S rRNA sequencing suggested that *Streptococcus* and *Enterococcus* are rare genera in our C57BL/6 mouse colony.
-If these genera are important contributors to the associations between PPIs and CDIs in humans, this may be part of the reason we did not see an effect in our CDI mouse model.
-There could also be differences at the species level for taxa that are altered with PPIs in humans compared to mice, which may also contribute to the association.
+16S rRNA sequencing suggested that *Streptococcus* and *Enterococcus* are rare genera in our C57BL/6 mouse colony.
+These genera could be important contributors to the associations between PPIs and CDIs in humans, and could be a contributing factor to why we observed that PPI treatment had no effect on *C.
+difficile* colonization in our CDI mouse model.
+There could also be differences at the species level for taxa that are altered with PPIs in humans compared to mice.
 Other murine starting communities or the addition of human PPI treatment-associated strains may be needed to determine whether PPIs can alter taxa and subsequently promote CDIs.
 Factors such as age, body mass index, comorbidities, and use of other medications in human studies may also be contributing to the association between PPIs and CDI incidence or recurrence.
-NSAID treatment can increase CDI severity in mice [@Maseda2019].
-Given that PPIs are sometimes used in combination with NSAIDs [@nehra2018proton], it may be worth examining both medications together using a mouse model.
-This study only looked at the impact of PPIs with or without antibiotics, but future work may want to incorporate age, other comorbidities and bacterial strains that are less common in mice to evaluate whether PPIs in conjunction with other factors can increase the risk of CDIs or recurrent CDIs.
+For example, NSAID treatment can increase CDI severity in mice [@Maseda2019] and PPIs are sometimes used in combination with NSAIDs [@nehra2018proton].
+This study addressed the impact of PPIs with or without antibiotics on a murine model of CDI, and found PPIs did not promote *C.
+difficile* colonization.
+Future studies are needed to determine whether age, other comorbidities and bacterial strains that are less common in mice can increase the risk of CDIs or recurrent CDIs when combined with PPI treatment.
 
@@
 **Animals** All mouse experiments were performed with 7- to 12-week-old C57BL/6 male and female mice.
-2-3 mice were housed per cage and male and female mice were housed separately.
+Each experimental group of mice was spit between 2 cages with 2-3 mice housed per cage and male and female mice were housed separately.
 All animal experiments were approved by the University of Michigan Animal Care and Use Committee (IACUC) under protocol number PRO00006983.
@@
 ## Figures
-
+\includegraphics{figure_1.pdf}
 **Figure 1.
@@
 
+\newpage
+\includegraphics{figure_2.pdf}
 **Figure 2.
@@
 
+\newpage
+\includegraphics{figure_s1.pdf}
 **Figure S1.

### 2019-07-11 submission/manuscript.tex
Responding to Pat's comments and making additional edits

@@
 clindamycin. High-dose PPI treatment was maintained throughout the
-experiment, which included \emph{C. difficile} challenge and the
-following 9 days. We found that omeprazole was not sufficient to promote
-\emph{C. difficile} colonization. When omeprazole treatment was combined
-with the antibiotic, there was a mixed level of \emph{C. difficile}
-colonization, where one cage of mice remained resistant and the other
-cage was colonized. 16S rRNA sequencing analysis revealed that
-omeprazole had minimal impact on the murine microbiota throughout the 16
-days of PPI exposure. \emph{C. difficile} colonization of
-clindamycin-treated mice was associated with changes in
-\emph{Alistipes}, \emph{Barnesiella}, Porphyromonadaceae, and
-Ruminococcaceae. These results suggest PPI treatment alone is not
-sufficient to disrupt microbiota resistance to C. difficle infection in
-mice that are normally resistant in the absence of antibiotic treatment.
+experiment, which included pre-treatment, the day of \emph{C. difficile}
+challenge, and the following 9 days. We found that omeprazole was not
+sufficient to promote \emph{C. difficile} colonization. When omeprazole
+treatment was combined with the antibiotic,one cage of mice remained
+resistant to \emph{C. difficile} colonization, while the other cage was
+colonized. 16S rRNA sequencing analysis revealed that omeprazole had
+minimal impact on the murine microbiota throughout the 16 days of PPI
+exposure. These results suggest PPI treatment alone is not sufficient to
+disrupt microbiota resistance to \emph{C. difficle} infection in mice
+that are normally resistant in the absence of antibiotic treatment.
 
@@
 intestinal microbiota through observational and interventional studies
-and PPI use has also been associated with CDI incidence and recurrence.
-We evaluated the ability of a high dose of the PPI omeprazole to alter
-the murine intestinal microbiota and disrupt colonization resistance to
-\emph{C. difficile}. We found daily PPI treatment had minimal impact on
-the murine intestinal microbiota and was not enough to promote \emph{C.
-difficile} colonization after challenge. Further studies are needed to
-determine whether other factors such as the composition of the starting
-bacterial community, comorbidities, and use of additional medications
-contribute to the association between PPIs and CDIs.
+and PPI use has also been associated with CDI incidence and recurrence
+in hospital and epidemiological studies. We evaluated the ability of a
+high dose of the PPI omeprazole to alter the murine intestinal
+microbiota and disrupt colonization resistance to \emph{C. difficile}.
+We found daily PPI treatment had minimal impact on the murine intestinal
+microbiota and was not enough to promote \emph{C. difficile}
+colonization after challenge. Further studies are needed to determine
+whether other factors such as the composition of the starting bacterial
+community, comorbidities, and use of additional medications contribute
+to the association between PPIs and CDIs that is seen in humans.
 
@@
 
-Proton pump inhibitors (PPIs) are one of the most widely used drugs and
-are available over the counter (1). PPIs are approved to treat a variety
-of conditions including gastroesophageal reflux disease, ulcers, or as
-prophylaxis for nonsteroidal anti-inflammatory drug (NSAID)-associated
-gastrointestinal bleeding (2). Unfortunately, there have been a number
-of adverse events associated with their use including small intestinal
-bacterial overgrowth, decreased calcium absorption, and increased risk
-of enteric infections (2). There's also been a number of large cohort
-studies as well as interventional clinical trials that have demonstrated
-specific alterations in the intestinal microbiome were associated with
-PPI use (3, 4). PPI-associated microbiota changes have been attributed
-to the ability of PPIs to increase stomach acid pH which may promote the
-survival of oral and pathogenic bacteria (3, 4). Human fecal microbiota
-changes with PPI use include increases in Enterococcaceae,
-Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae
-with decreases in Ruminococcaceae (3). An \emph{in vitro screen},
-demonstrated that 3 PPIs inhibit the growth of 7-8 strains of
+Antibiotics have a large impact on the intestinal microbiome and are a
+primary risk factor for developing \emph{Clostridium difficile}
+infections (CDIs) (1). It is less clear whether other human medications
+that impact the microbiota also influence \emph{C. difficile}
+colonization resistance. Multiple epidemiological studies have suggested
+an association between proton pump inhibitor (PPI) use and incidence or
+recurrence of CDIs (2--5). PPIs are one of the most widely used drugs in
+the U.S. and are available over the counter (6). PPIs are approved to
+treat a variety of conditions including gastroesophageal reflux disease,
+ulcers, or as prophylaxis for nonsteroidal anti-inflammatory drug
+(NSAID)-associated gastrointestinal bleeding (3). Unfortunately, there
+have been a number of adverse events associated with their use including
+small intestinal bacterial overgrowth, decreased calcium absorption, and
+increased risk of enteric infections (3). There have also been a number
+of large cohort studies as well as interventional clinical trials that
+have demonstrated specific alterations in the intestinal microbiome were
+associated with PPI use (4, 7). PPI-associated microbiota changes have
+been attributed to the ability of PPIs to increase stomach acid pH which
+may promote the survival of oral and pathogenic bacteria (4, 7). Human
+fecal microbiota changes with PPI use include increases in
+Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae and
+Streptococcaceae and decreases in Ruminococcaceae (7). An \emph{in vitro
+screen} demonstrated that 3 PPIs inhibit the growth of 7-8 strains of
 Lachnospiraceae and Ruminococcaceae, which represented
 \textasciitilde{}16\% of the 40 intestinal bacteria strains that were
-screened (5).\\
-Antibiotics have a large impact on the intestinal microbiome and are a
-primary risk factor for developing \emph{Clostridium difficile}
-infections (CDIs) (6). It is less clear whether other human medications
-that impact the microbiota also influence \emph{C. difficile}
-colonization resistance. Multiple epidemiological studies have suggested
-an association between PPI use and incidence or recurrence of CDIs (2,
-4, 7, 8). However, most of the studies suggesting a link between PPIs
-and \emph{C. difficile} are retrospective and did not evaluate the
-microbiome (2, 7, 8). Thus, it is unclear whether the gastrointestinal
-microbiome changes associated with PPI use play a role in the
-association between PPIs and CDI incidence or recurrence. Additionally,
-epidemiological studies have a limited capacity to address potential
-confounders and comorbidities in patients that were on PPIs and
-developed CDIs or recurrent CDIs (7, 8).\\
-Although there have been a few \emph{C. difficile} mouse model studies
-that have demonstrated PPIs have some effect on CDIs with or without
-additional antibiotic treatment, none of these studies looked at the
-interplay with the intestinal microbiome (9--11). Here, we evaluated the
-impact of a daily high dose PPI treatment on the murine microbiome and
-susceptibility to \emph{C. difficile} colonization in relation to
-clindamycin, an antibiotic that perturbs the microbiome enough to allow
-\emph{C. difficile} to colonize but is mild enough that \emph{C.
-difficile} is cleared within 10 days (12).
-
-\subsection{Results}\label{results}
+screened (8), suggesting direct action on bacteria might also play a
+role.\\
+However, most of the studies suggesting a link between PPIs and \emph{C.
+difficile} were retrospective and did not evaluate the microbiome (2, 3,
+5). Thus, it is unclear whether the gastrointestinal microbiome changes
+associated with PPI use play a role in the association between PPIs and
+CDI incidence or recurrence. Additionally, epidemiological studies have
+a limited capacity to address potential confounders and comorbidities in
+patients that were on PPIs and developed CDIs or recurrent CDIs (2, 5).
+Here, we evaluated the impact of a daily high dose PPI treatment on the
+murine microbiome and susceptibility to \emph{C. difficile} colonization
+in relation to clindamycin, an antibiotic that perturbs the microbiome
+enough to allow \emph{C. difficile} to colonize but is mild enough that
+\emph{C. difficile} is cleared within 10 days (9).
 
@@
 susceptibility to CDIs, we gavaged mice daily with a high dose of
-omeprazole for 7 days before \emph{C. difficile} challenge (Figure 1A)
-and compared the PPI-treated mice to a group that received the
+omeprazole for 7 days before \emph{C. difficile} challenge (Figure 1A).
+PPI-treated mice were compared to a group of mice that received the
 antibiotic, clindamycin, and a group that received both PPI and
@@
 Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes
-that have previously been impacted by PPIs in human studies, but we see
-no significant fluctuations in the stool of our PPI-treated mice
-throughout the course of the 16-day experiment (Figure 1C, Figure S1).
-In contrast, the microbiomes from the 2 groups of mice that received
-clindamycin started to shift away from the PPI-treated mice and the
-samples from previous days one day after treatment (Figure 1D).
+that have previously been impacted by PPIs in human studies, but we
+observed no significant fluctuations in the stool of our PPI-treated
+mice throughout the course of the 16-day experiment (Figure 1C, Figure
+S1). In contrast, the microbiomes from the 2 groups of mice that
+received clindamycin started to shift away from the PPI-treated mice and
+the samples from previous days one day after treatment (Figure 1D).
+These results demonstrated that high dose PPI treatment alone had no
+significant impact on the murine fecal bacterial community.
 
 \textbf{PPI-treatment did not promote susceptibility to \emph{C.
-difficile infection} in mice} After 7 days of PPI treatment or 1 day
-after clindamycin treatment, mice were challenged with \emph{C.
-difficile} 630 spores. While all 4 of the clindamycin-treated mice were
-colonized with \emph{C. difficile}, all of the PPI-treated mice were
-resistant to \emph{C. difficile} colonization (Figure 2A).
-Interestingly, only 1 cage of the PPI and clindamycin-treated mice were
-colonized, while the other cage was not (Figure 2A). PCoAs of the
-bacterial communities from the 3 groups of mice after \emph{C.
-difficile} challenge, showed that the greatest shifts in bacterial
-communities occurred in the clindamycin-treated mice (Figure 2B). Within
-5 days, all of the mice had cleared \emph{C. difficile}, suggesting
-there was no difference in rate of clearance for the mice that were
-initially colonized with \emph{C. difficile} despite continuing to
-receive PPI treatment throughout the course of the experiment. Our
-results suggest that PPI treatment alone had minor effects on microbial
-community resistance to \emph{C. difficile} colonization in mice.
-Instead most of the differences between our 3 treatment groups appear to
-be driven by clindamycin administration and included decreased
-\emph{Alistipes}, \emph{Barnesiella}, Porhyromonadaceae,
-Ruminococcaceae, taxa previously found to be altered in
-clindamycin-treated mice that were challenged with \emph{C. difficile}
-(6).
+difficile infection} in mice} Next, we examined whether high dose PPI
+treatment altered susceptibility to \emph{C. difficile} infection in
+mice. After 7 days of PPI treatment or 1 day after clindamycin
+treatment, mice were challenged with \emph{C. difficile} 630 spores.
+While all 4 of the clindamycin-treated mice were colonized with \emph{C.
+difficile}, all of the PPI-treated mice were resistant to \emph{C.
+difficile} colonization (Figure 2A). Interestingly, only 1 cage of the
+PPI and clindamycin-treated mice were colonized, while the other cage
+was not (Figure 2A). PCoAs of the bacterial communities from the 3
+groups of mice after \emph{C. difficile} challenge, showed that the
+greatest shifts in bacterial communities occurred in the
+clindamycin-treated mice (Figure 2B). Within 5 days, all of the mice had
+cleared \emph{C. difficile}, suggesting there was no difference in rate
+of clearance for the mice that were initially colonized with \emph{C.
+difficile} despite continuing to receive PPI treatment throughout the
+course of the experiment. Our results suggest that PPI treatment alone
+had minor effects on microbial community resistance to \emph{C.
+difficile} colonization in mice. Instead most of the differences between
+our 3 treatment groups appear to be driven by clindamycin administration
+and included decreased \emph{Alistipes}, \emph{Barnesiella},
+Porhyromonadaceae, Ruminococcaceae, taxa previously found to be altered
+in clindamycin-treated mice that were challenged with \emph{C.
+difficile} (1). These findings demonstrated that high dose PPI treatment
+did not promote susceptibility to \emph{C. difficile} colonization.
 
@@
 
-Our results demonstrate that daily treatment with a high dose of the PPI
-omeprazole had minimal impact on the stool microbiomes of mice and did
-not promote susceptibility to \emph{C. difficile} colonization. In
-contrast, clindamycin-treated mice were initially colonized, but cleared
-the infection within 10 days. Combining PPI and clindamycin treatment
-had a mixed effect, one cage was colonized with \emph{C. difficile} and
-cleared within 10 days, while the other cage remained resistant.\\
 Our findings that PPI treatment had minimal impact on the fecal
-microbiome are comparable to another PPI mouse study that indicated PPIs
-have more of an effect on the small intestinal microbiota compared to
-the fecal microbiota (13). We did not find significant changes for the
-taxa observed to be significantly impacted by PPI use in human studies.
-However, 3 of the families that typically increase were absent or at low
-abundance in the mice from the beginning and may be one contributing
-factor to why we did not see any changes. Additionally, some of the
-significance of PPI associations in human interventional trials appears
-to be driven by a handful of specific taxa with overall differences on a
-PCoA plot difficult to distinguish (14). Also, most of the studies to
-date compare different individuals (PPI users versus non-users) or the
-same individuals before and after treatment (3, 4), unfortunately these
-limited microbiota snapshots may be skewed by day-to-day microbiota
-variations (16) that would be revealed with more frequent longitudinal
-sampling.\\
-Previous work that administered 0.5 mg/kg of the PPI lansoprazole daily
-for 2 weeks to mice and then challenged with \emph{C. difficile}
-demonstrated that PPI treatment alone resulted in detectable \emph{C.
-difficile} in the stool 1 week after challenge, but also showed there
-was detectable \emph{C. difficile} in mice not treated with antibiotics
-(10, 11). The presence of \emph{C. difficile} in mice that were not
-treated with either antibiotics or PPIs could be attributed to the
-higher dose of vegetative cells (10\textsuperscript{8} CFU) used to
-challenge the mice and may partly explain their observation of \emph{C.
-difficile} in PPI-treated mice (10, 11). We have previously shown mice
-from our colony that are not given antibiotics are resistant to \emph{C.
-difficile} 630 when challenged with 10\textsuperscript{3} spores (17).
-Another group also used a higher dose of 10\textsuperscript{6} spores or
-10\textsuperscript{7} vegetative cells to challenge antibiotic-treated
-mice or mice treated with both the PPI esomeprazole (40mg/kg dose) for 2
-days and antibiotics and demonstrated the antibiotic/PPI-treated mice
-developed more severe CDIs (9).\\
+microbiome were comparable to another PPI mouse study that indicated
+PPIs have more of an effect on the small intestinal microbiota compared
+to the fecal microbiota (10). We did not find significant changes for
+the taxa observed to be significantly impacted by PPI use in human
+studies. However, 3 of the families that typically increase were absent
+or at low abundance in the mice from the beginning and may be one
+contributing factor to why we did not see any changes. Additionally,
+some of the significance of PPI associations in human interventional
+trials appears to be driven by a handful of specific taxa with overall
+differences on a PCoA plot difficult to distinguish (11). Also, most of
+the studies to date compareed different individuals (PPI users versus
+non-users) or the same individuals before and after treatment (4, 7),
+unfortunately these limited microbiota snapshots may be skewed by
+day-to-day microbiota variations (13) that would be revealed with more
+frequent longitudinal sampling.\\
+ALthough there have been a few \emph{C. difficile} mouse model studies
+that have demonstrated PPIs have some effect on CDIs with or without
+additional antibiotic treatment, none of these studies looked at the
+interplay with the intestinal microbiome (14--16). Additionally, there
+were some key methodical differences between the previous studies and
+our own. Previous work that administered 0.5 mg/kg of the PPI
+lansoprazole daily for 2 weeks to mice and then challenged with \emph{C.
+difficile} demonstrated that PPI treatment alone resulted in detectable
+\emph{C. difficile} in the stool 1 week after challenge, but also showed
+there was detectable \emph{C. difficile} in mice not treated with
+antibiotics (14, 15). The presence of \emph{C. difficile} in mice that
+were not treated with either antibiotics or PPIs could be attributed to
+the higher dose of vegetative cells (10\textsuperscript{8} CFU) used to
+challenge the mice and may partly explain the observation of \emph{C.
+difficile} in PPI-treated mice (14, 15). We have previously shown mice
+from our colony that were not given antibiotics were resistant to
+\emph{C. difficile} 630 when challenged with 10\textsuperscript{3}
+spores (17). Another group also used a higher dose of
+10\textsuperscript{6} spores or 10\textsuperscript{7} vegetative cells
+to challenge antibiotic-treated mice or mice treated with both the PPI
+esomeprazole (40mg/kg dose) for 2 days and antibiotics and demonstrated
+the antibiotic/PPI-treated mice developed more severe CDIs (16).\\
 Our study extended previous work examining PPIs and \emph{C. difficile}
@@
 medications in human studies may also be contributing to the association
-between PPIs and CDI incidence or recurrence. NSAID treatment can
-increase CDI severity in mice (18). Given that PPIs are sometimes used
-in combination with NSAIDs (2), it may be worth examining both
-medications together using a mouse model. This study only looked at the
-impact of PPIs with or without antibiotics, but future work may want to
-incorporate age, other comorbidities and bacterial strains that are less
-common in mice to evaluate whether PPIs in conjunction with other
-factors can increase the risk of CDIs or recurrent CDIs.
+between PPIs and CDI incidence or recurrence. For example, NSAID
+treatment can increase CDI severity in mice (18) and PPIs are sometimes
+used in combination with NSAIDs (3). This study addressed the impact of
+PPIs with or without antibiotics on a murine model of CDI, and found
+PPIs did not promote \emph{C. difficile} colonization. Future studies
+are needed to determine whether age, other comorbidities and bacterial
+strains that are less common in mice can increase the risk of CDIs or
+recurrent CDIs when combined with PPI treatment.
 
@@
 \textbf{Animals} All mouse experiments were performed with 7- to
-12-week-old C57BL/6 male and female mice. 2-3 mice were housed per cage
-and male and female mice were housed separately. All animal experiments
-were approved by the University of Michigan Animal Care and Use
-Committee (IACUC) under protocol number PRO00006983.
+12-week-old C57BL/6 male and female mice. Each experimental group of
+mice was spit between 2 cages with 2-3 mice housed per cage and male and
+female mice were housed separately. All animal experiments were approved
+by the University of Michigan Animal Care and Use Committee (IACUC)
+under protocol number PRO00006983.
 
@@
 \hypertarget{refs}{}
-\hypertarget{ref-Singh2018}{}
-1. \textbf{Singh A}, \textbf{Cresci GA}, \textbf{Kirby DF}. 2018. Proton
-pump inhibitors: Risks and rewards and emerging consequences to the gut
-microbiome. Nutrition in Clinical Practice \textbf{33}:614--624.
-doi:\href{https://doi.org/10.1002/ncp.10181}{10.1002/ncp.10181}.
+\hypertarget{ref-Schubert2015}{}
+1. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
+Antibiotic-induced alterations of the murine gut microbiota and
+subsequent effects on colonization resistance against clostridium
+difficile. mBio \textbf{6}.
+doi:\href{https://doi.org/10.1128/mbio.00974-15}{10.1128/mbio.00974-15}.
+
+\hypertarget{ref-tariq2017association}{}
+2. \textbf{Tariq R}, \textbf{Singh S}, \textbf{Gupta A}, \textbf{Pardi
+DS}, \textbf{Khanna S}. 2017. Association of gastric acid suppression
+with recurrent clostridium difficile infection: A systematic review and
+meta-analysis. JAMA internal medicine \textbf{177}:784--791.
 
 \hypertarget{ref-nehra2018proton}{}
-2. \textbf{Nehra AK}, \textbf{Alexander JA}, \textbf{Loftus CG},
+3. \textbf{Nehra AK}, \textbf{Alexander JA}, \textbf{Loftus CG},
 \textbf{Nehra V}. 2018. Proton pump inhibitors: Review of emerging
 concerns, pp. 240--246. \emph{In} Mayo clinic proceedings. Elsevier.
-
-\hypertarget{ref-Imhann2017}{}
-3. \textbf{Imhann F}, \textbf{Vila AV}, \textbf{Bonder MJ},
-\textbf{Manosalva AGL}, \textbf{Koonen DP}, \textbf{Fu J},
-\textbf{Wijmenga C}, \textbf{Zhernakova A}, \textbf{Weersma RK}. 2017.
-The influence of proton pump inhibitors and other commonly used
-medication on the gut microbiota. Gut Microbes \textbf{8}:351--358.
-doi:\href{https://doi.org/10.1080/19490976.2017.1284732}{10.1080/19490976.2017.1284732}.
 
@@
 
+\hypertarget{ref-Elias2019}{}
+5. \textbf{Elias E}, \textbf{Targownik LE}. 2019. The clinician's guide
+to proton pump inhibitor related adverse events. Drugs
+\textbf{79}:715--731.
+doi:\href{https://doi.org/10.1007/s40265-019-01110-3}{10.1007/s40265-019-01110-3}.
+
+\hypertarget{ref-Singh2018}{}
+6. \textbf{Singh A}, \textbf{Cresci GA}, \textbf{Kirby DF}. 2018. Proton
+pump inhibitors: Risks and rewards and emerging consequences to the gut
+microbiome. Nutrition in Clinical Practice \textbf{33}:614--624.
+doi:\href{https://doi.org/10.1002/ncp.10181}{10.1002/ncp.10181}.
+
+\hypertarget{ref-Imhann2017}{}
+7. \textbf{Imhann F}, \textbf{Vila AV}, \textbf{Bonder MJ},
+\textbf{Manosalva AGL}, \textbf{Koonen DP}, \textbf{Fu J},
+\textbf{Wijmenga C}, \textbf{Zhernakova A}, \textbf{Weersma RK}. 2017.
+The influence of proton pump inhibitors and other commonly used
+medication on the gut microbiota. Gut Microbes \textbf{8}:351--358.
+doi:\href{https://doi.org/10.1080/19490976.2017.1284732}{10.1080/19490976.2017.1284732}.
+
 \hypertarget{ref-maier2018extensive}{}
-5. \textbf{Maier L}, \textbf{Pruteanu M}, \textbf{Kuhn M},
+8. \textbf{Maier L}, \textbf{Pruteanu M}, \textbf{Kuhn M},
 \textbf{Zeller G}, \textbf{Telzerow A}, \textbf{Anderson EE},
@@
 
-\hypertarget{ref-Schubert2015}{}
-6. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
-Antibiotic-induced alterations of the murine gut microbiota and
-subsequent effects on colonization resistance against clostridium
-difficile. mBio \textbf{6}.
-doi:\href{https://doi.org/10.1128/mbio.00974-15}{10.1128/mbio.00974-15}.
-
-\hypertarget{ref-tariq2017association}{}
-7. \textbf{Tariq R}, \textbf{Singh S}, \textbf{Gupta A}, \textbf{Pardi
-DS}, \textbf{Khanna S}. 2017. Association of gastric acid suppression
-with recurrent clostridium difficile infection: A systematic review and
-meta-analysis. JAMA internal medicine \textbf{177}:784--791.
-
-\hypertarget{ref-Elias2019}{}
-8. \textbf{Elias E}, \textbf{Targownik LE}. 2019. The clinician's guide
-to proton pump inhibitor related adverse events. Drugs
-\textbf{79}:715--731.
-doi:\href{https://doi.org/10.1007/s40265-019-01110-3}{10.1007/s40265-019-01110-3}.
-
-\hypertarget{ref-hung2015proton}{}
-9. \textbf{Hung Y-P}, \textbf{Ko W-C}, \textbf{Chou P-H}, \textbf{Chen
-Y-H}, \textbf{Lin H-J}, \textbf{Liu Y-H}, \textbf{Tsai H-W}, \textbf{Lee
-J-C}, \textbf{Tsai P-J}. 2015. Proton-pump inhibitor exposure aggravates
-clostridium difficile--associated colitis: Evidence from a mouse model.
-The Journal of infectious diseases \textbf{212}:654--663.
-
-\hypertarget{ref-Kaur2007}{}
-10. \textbf{Kaur S}, \textbf{Vaishnavi C}, \textbf{Prasad KK},
-\textbf{Ray P}, \textbf{Kochhar R}. 2007. Comparative role of antibiotic
-and proton pump inhibitor in ExperimentalClostridium difficileInfection
-in mice. Microbiology and Immunology \textbf{51}:1209--1214.
-doi:\href{https://doi.org/10.1111/j.1348-0421.2007.tb04016.x}{10.1111/j.1348-0421.2007.tb04016.x}.
-
-\hypertarget{ref-kaur2011effect}{}
-11. \textbf{Kaur S}, \textbf{Vaishnavi C}, \textbf{Prasad KK},
-\textbf{Ray P}, \textbf{Kochhar R}. 2011. Effect of lactobacillus
-acidophilus \& epidermal growth factor on experimentally induced
-clostridium difficile infection. The Indian journal of medical research
-\textbf{133}:434.
-
 \hypertarget{ref-Jenior2018}{}
-12. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+9. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
 \textbf{Schloss PD}. 2018. Clostridium difficile alters the structure
@@
 \hypertarget{ref-Yasutomi2018}{}
-13. \textbf{Yasutomi E}, \textbf{Hoshi N}, \textbf{Adachi S},
+10. \textbf{Yasutomi E}, \textbf{Hoshi N}, \textbf{Adachi S},
 \textbf{Otsuka T}, \textbf{Kong L}, \textbf{Ku Y}, \textbf{Yamairi H},
@@
 \hypertarget{ref-Freedberg2015}{}
-14. \textbf{Freedberg DE}, \textbf{Toussaint NC}, \textbf{Chen SP},
+11. \textbf{Freedberg DE}, \textbf{Toussaint NC}, \textbf{Chen SP},
 \textbf{Ratner AJ}, \textbf{Whittier S}, \textbf{Wang TC}, \textbf{Wang
@@
 \hypertarget{ref-Imhann2015}{}
-15. \textbf{Imhann F}, \textbf{Bonder MJ}, \textbf{Vila AV}, \textbf{Fu
+12. \textbf{Imhann F}, \textbf{Bonder MJ}, \textbf{Vila AV}, \textbf{Fu
 J}, \textbf{Mujagic Z}, \textbf{Vork L}, \textbf{Tigchelaar EF},
@@
 \hypertarget{ref-LlornsRico2019}{}
-16. \textbf{Lloréns-Rico V}, \textbf{Raes J}. 2019. Tracking humans and
+13. \textbf{Lloréns-Rico V}, \textbf{Raes J}. 2019. Tracking humans and
 microbes. Nature \textbf{569}:632--633.
 doi:\href{https://doi.org/10.1038/d41586-019-01591-y}{10.1038/d41586-019-01591-y}.
+
+\hypertarget{ref-Kaur2007}{}
+14. \textbf{Kaur S}, \textbf{Vaishnavi C}, \textbf{Prasad KK},
+\textbf{Ray P}, \textbf{Kochhar R}. 2007. Comparative role of antibiotic
+and proton pump inhibitor in ExperimentalClostridium difficileInfection
+in mice. Microbiology and Immunology \textbf{51}:1209--1214.
+doi:\href{https://doi.org/10.1111/j.1348-0421.2007.tb04016.x}{10.1111/j.1348-0421.2007.tb04016.x}.
+
+\hypertarget{ref-kaur2011effect}{}
+15. \textbf{Kaur S}, \textbf{Vaishnavi C}, \textbf{Prasad KK},
+\textbf{Ray P}, \textbf{Kochhar R}. 2011. Effect of lactobacillus
+acidophilus \& epidermal growth factor on experimentally induced
+clostridium difficile infection. The Indian journal of medical research
+\textbf{133}:434.
+
+\hypertarget{ref-hung2015proton}{}
+16. \textbf{Hung Y-P}, \textbf{Ko W-C}, \textbf{Chou P-H}, \textbf{Chen
+Y-H}, \textbf{Lin H-J}, \textbf{Liu Y-H}, \textbf{Tsai H-W}, \textbf{Lee
+J-C}, \textbf{Tsai P-J}. 2015. Proton-pump inhibitor exposure aggravates
+clostridium difficile--associated colitis: Evidence from a mouse model.
+The Journal of infectious diseases \textbf{212}:654--663.
 
@@
 
-\textbf{Figure 1. PPI treatment had minimal impact on the murine fecal
-microbiota} A. Mouse experiment timeline and logistics. Stools for 16S
-rRNA sequencing analysis were collected on the days that are labeled
-(Day -7, -5, -3, -1, 0, 1, 2, 3, 4, 5, 7, 9). B. Principal Coordinates
-Analaysis (PCoA) of Bray-curtis distances from stool samples for all
-treatment groups during the intial 7 days of treatment with the PPI
-omeprazole or vehicle cluster together and do not chage much over time.
-C. Relative abundances of families previously associated with PPI use in
-humans do not change much over the 16-day course of treatment with PPIs
-(7 days before \emph{C. difficile} challenge through 9 days post
-\emph{C. difficile} challenge). D. PCoA of stool samples collected
-before and 1 day after antibiotic treatment, demonstrating the fecal
-microbiota of mice treated with clindamycin starts to separate from the
-rest of the samples.
+\includegraphics{figure_1.pdf} \textbf{Figure 1. PPI treatment had
+minimal impact on the murine fecal microbiota} A. Mouse experiment
+timeline and logistics. Stools for 16S rRNA sequencing analysis were
+collected on the days that are labeled (Day -7, -5, -3, -1, 0, 1, 2, 3,
+4, 5, 7, 9). B. Principal Coordinates Analaysis (PCoA) of Bray-curtis
+distances from stool samples for all treatment groups during the intial
+7 days of treatment with the PPI omeprazole or vehicle cluster together
+and do not chage much over time. C. Relative abundances of families
+previously associated with PPI use in humans do not change much over the
+16-day course of treatment with PPIs (7 days before \emph{C. difficile}
+challenge through 9 days post \emph{C. difficile} challenge). D. PCoA of
+stool samples collected before and 1 day after antibiotic treatment,
+demonstrating the fecal microbiota of mice treated with clindamycin
+starts to separate from the rest of the samples.
 
-\textbf{Figure 2. PPI treatment alone does not promote CDIs in mice} A.
-C. difficile CFUs/g stool measured each day post C. difficile challenge
-for clindamycin, clindamycin/PPI, and PPI-treated mice. Lines represent
-the mean CFU/g for each treatment group while points represent CFU/g for
-individual mice within each group. B. PCoA of stool samples collected
-after antibiotic treatments. C. Genera significantly associated with
-treatment groups for samples across all sequenced timepoints. D.
-Families significantly associated with treatment groups for samples
-across all sequenced timepoints. C,D data were analyzed by
-Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple
-comparisons.
+\newpage
 
-\textbf{Figure S1. Previous human PPI-associated families fluctuate over
-time with no overall trend in either direction} Relative abundance over
-time for Lactobacillaceae (A) and Ruminococcaceae (B), 2 of the
-PPI-associated familes from human PPI studies across all 3 treatment
-groups. Each point represents the relative abundance for an individual
-mouse
+\includegraphics{results/figures/figure_2.pdf} \textbf{Figure 2. PPI
+treatment alone does not promote CDIs in mice} A. C. difficile CFUs/g
+stool measured each day post C. difficile challenge for clindamycin,
+clindamycin/PPI, and PPI-treated mice. Lines represent the mean CFU/g
+for each treatment group while points represent CFU/g for individual
+mice within each group. B. PCoA of stool samples collected after
+antibiotic treatments. C. Genera significantly associated with treatment
+groups for samples across all sequenced timepoints. D. Families
+significantly associated with treatment groups for samples across all
+sequenced timepoints. C,D data were analyzed by Kruskal-Wallis test with
+a Benjamini-Hochberg correction for multiple comparisons.
+
+\newpage
+
+\includegraphics{figure_s1.pdf} \textbf{Figure S1. Previous human
+PPI-associated families fluctuate over time with no overall trend in
+either direction} Relative abundance over time for Lactobacillaceae (A)
+and Ruminococcaceae (B), 2 of the PPI-associated familes from human PPI
+studies across all 3 treatment groups. Each point represents the
+relative abundance for an individual mouse
 

### 2019-07-11 submission/manuscript.Rmd
Additional edits before sharing with Pat

@@
 difficile* colonization (Figure 2A).
-Interestingly, only 1 cage of the PPI and clindamycin-treated mice were colonized, while the other cage was not (Figure 2A).
+Interestingly, only 1 cage of the PPI and clindamycin-treated mice were colonized, while the other cage was resistant (Figure 2A).
 PCoAs of the bacterial communities from the 3 groups of mice after *C.
@@
 ## Discussion and Conclusions
-Our findings that PPI treatment had minimal impact on the fecal microbiome were comparable to another PPI mouse study that indicated PPIs have more of an effect on the small intestinal microbiota compared to the fecal microbiota [@Yasutomi2018].
-The same group demonstrated PPI treatment increased the stomach pH in mice [@Yasutomi2018], which could improve survival of bacteria passing through the stomach.
+Our findings that PPI treatment had minimal impact on the fecal microbiome were comparable to another PPI mouse study that indicated PPIs had more of an effect on the small intestinal microbiota compared to the fecal microbiota [@Yasutomi2018].
+The same group demonstrated PPI treatment increased the stomach pH in mice [@Yasutomi2018], which may improve survival of bacteria passing through the stomach.
 We did not find significant changes for the taxa observed to be significantly impacted by PPI use in human studies.
@@
 ALthough there have been a few *C.
-difficile* mouse model studies that have demonstrated PPIs have some effect on CDIs with or without additional antibiotic treatment[@Kaur2007; @kaur2011effect; @hung2015proton], there were some key methodological differences between the previous studies and our own.
+difficile* mouse model studies that have demonstrated PPIs have some effect on CDIs with or without additional antibiotic treatment[@Kaur2007; @kaur2011effect; @hung2015proton], there were some key methodological differences between these studies and our own.
 One group administered 0.5 mg/kg of the PPI lansoprazole daily for 2 weeks to mice and then challenged with *C.
@@
 16S rRNA sequencing suggested that *Streptococcus* and *Enterococcus* are rare genera in our C57BL/6 mouse colony.
-These genera could be important contributors to the associations between PPIs and CDIs in humans, and could be a contributing factor to why we observed that PPI treatment had no effect on *C.
+These genera could be important contributors to the associations between PPIs and CDIs in humans, and could be a contributing factor to our observation that PPI treatment had no effect on *C.
 difficile* colonization in our CDI mouse model.

### 2019-07-11 submission/manuscript.tex
Additional edits before sharing with Pat

@@
 Antibiotics are a major risk factor for \emph{Clostridium difficile}
-infections (CDIs), but it is less clear what other factors contribute to
-CDI incidence and recurrence. Interestingly, other medications besides
-antibiotics have been shown to alter the microbiota. Proton pump
-inhibitors (PPIs) are associated with alterations in the human
-intestinal microbiota through observational and interventional studies
-and PPI use has also been associated with CDI incidence and recurrence
-in hospital and epidemiological studies. We evaluated the ability of a
-high dose of the PPI omeprazole to alter the murine intestinal
-microbiota and disrupt colonization resistance to \emph{C. difficile}.
-We found daily PPI treatment had minimal impact on the murine intestinal
-microbiota and was not enough to promote \emph{C. difficile}
-colonization after challenge. Further studies are needed to determine
-whether other factors such as the composition of the starting bacterial
-community, comorbidities, and use of additional medications contribute
-to the association between PPIs and CDIs that is seen in humans.
+infections (CDIs), but other factors may also contribute to CDI
+incidence and recurrence. Interestingly, other medications besides
+antibiotics can impact the microbiota. Proton pump inhibitors (PPIs) are
+associated with alterations in the human intestinal microbiota through
+observational and interventional studies and PPI use has also been
+associated with CDI incidence and recurrence in epidemiological studies.
+We evaluated the ability of a high dose of the PPI omeprazole to alter
+the murine intestinal microbiota and disrupt colonization resistance to
+\emph{C. difficile}. We found daily PPI treatment had minimal impact on
+the murine fecal microbiota and was not enough to promote \emph{C.
+difficile} colonization after challenge. Further studies are needed to
+determine whether other factors such as the composition of the starting
+bacterial community, comorbidities, and use of additional medications
+contribute to the association between PPIs and CDIs seen in humans.
 
@@
 an association between proton pump inhibitor (PPI) use and incidence or
-recurrence of CDIs (2--5). PPIs are one of the most widely used drugs in
-the U.S. and are available over the counter (6). PPIs are approved to
-treat a variety of conditions including gastroesophageal reflux disease,
-ulcers, or as prophylaxis for nonsteroidal anti-inflammatory drug
-(NSAID)-associated gastrointestinal bleeding (3). Unfortunately, there
-have been a number of adverse events associated with their use including
-small intestinal bacterial overgrowth, decreased calcium absorption, and
-increased risk of enteric infections (3). There have also been a number
-of large cohort studies as well as interventional clinical trials that
-have demonstrated specific alterations in the intestinal microbiome were
-associated with PPI use (4, 7). PPI-associated microbiota changes have
-been attributed to the ability of PPIs to increase stomach acid pH which
-may promote the survival of oral and pathogenic bacteria (4, 7). Human
-fecal microbiota changes with PPI use include increases in
-Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae and
-Streptococcaceae and decreases in Ruminococcaceae (7). An \emph{in vitro
-screen} demonstrated that 3 PPIs inhibit the growth of 7-8 strains of
-Lachnospiraceae and Ruminococcaceae, which represented
-\textasciitilde{}16\% of the 40 intestinal bacteria strains that were
-screened (8), suggesting direct action on bacteria might also play a
-role.\\
-However, most of the studies suggesting a link between PPIs and \emph{C.
-difficile} were retrospective and did not evaluate the microbiome (2, 3,
-5). Thus, it is unclear whether the gastrointestinal microbiome changes
-associated with PPI use play a role in the association between PPIs and
-CDI incidence or recurrence. Additionally, epidemiological studies have
-a limited capacity to address potential confounders and comorbidities in
-patients that were on PPIs and developed CDIs or recurrent CDIs (2, 5).
-Here, we evaluated the impact of a daily high dose PPI treatment on the
-murine microbiome and susceptibility to \emph{C. difficile} colonization
-in relation to clindamycin, an antibiotic that perturbs the microbiome
-enough to allow \emph{C. difficile} to colonize but is mild enough that
-\emph{C. difficile} is cleared within 10 days (9).
+recurrence of CDIs (2--5). There have also been a number of large cohort
+studies as well as interventional clinical trials that have demonstrated
+specific alterations in the intestinal microbiome were associated with
+PPI use (4, 6). PPI-associated microbiota changes have been attributed
+to the ability of PPIs to increase stomach acid pH which may promote the
+survival of oral and pathogenic bacteria (4, 6). Human fecal microbiota
+changes with PPI use include increases in Enterococcaceae,
+Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae
+and decreases in Ruminococcaceae (6).\\
+Unfortunately, most of the studies suggesting a link between PPIs and
+\emph{C. difficile} were retrospective and did not evaluate the
+microbiome (2, 3, 5). Thus, it is unclear whether the gastrointestinal
+microbiome changes associated with PPI use play a role in the
+association between PPIs and CDI incidence or recurrence. Additionally,
+epidemiological studies have a limited capacity to address potential
+confounders and comorbidities in patients that were on PPIs and
+developed CDIs or recurrent CDIs (2, 5). Here, we evaluated the impact
+of a daily high dose PPI treatment on the murine microbiome and
+susceptibility to \emph{C. difficile} colonization in relation to
+clindamycin, an antibiotic that perturbs the microbiome enough to allow
+\emph{C. difficile} to colonize but is mild enough that \emph{C.
+difficile} is cleared within 10 days (7).
 
@@
 omeprazole for 7 days before \emph{C. difficile} challenge (Figure 1A).
-PPI-treated mice were compared to a group of mice that received the
-antibiotic, clindamycin, and a group that received both PPI and
-clindamycin treatment. A principle coordinates analysis (PCoA) of the
-Bray-Curtis distances over the initial 7 days of treatment revealed the
-bacterial communities of PPI-treated mice remained relatively unchanged
-(Figure 1B). Enterococcaceae, Lactobacillaceae, Micrococcaceae,
+The fecal bacterial communities of PPI-treated mice were compared to a
+group of mice that received the antibiotic, clindamycin, and a group
+that received both PPI and clindamycin treatment using 16S rRNA gene
+sequencing. A principle coordinates analysis (PCoA) of the Bray-Curtis
+distances over the initial 7 days of treatment revealed the bacterial
+communities of PPI-treated mice remained relatively unchanged (Figure
+1B). Enterococcaceae, Lactobacillaceae, Micrococcaceae,
 Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes
@@
 PPI and clindamycin-treated mice were colonized, while the other cage
-was not (Figure 2A). PCoAs of the bacterial communities from the 3
+was resistant (Figure 2A). PCoAs of the bacterial communities from the 3
 groups of mice after \emph{C. difficile} challenge, showed that the
@@
 course of the experiment. Our results suggest that PPI treatment alone
-had minor effects on microbial community resistance to \emph{C.
-difficile} colonization in mice. Instead most of the differences between
-our 3 treatment groups appear to be driven by clindamycin administration
-and included decreased \emph{Alistipes}, \emph{Barnesiella},
+had no effects on bacterial community resistance to \emph{C. difficile}
+colonization in mice. Instead most of the differences between our 3
+treatment groups appear to be driven by clindamycin administration and
+included decreased \emph{Alistipes}, \emph{Barnesiella},
 Porhyromonadaceae, Ruminococcaceae, taxa previously found to be altered
@@
 microbiome were comparable to another PPI mouse study that indicated
-PPIs have more of an effect on the small intestinal microbiota compared
-to the fecal microbiota (10). We did not find significant changes for
-the taxa observed to be significantly impacted by PPI use in human
-studies. However, 3 of the families that typically increase were absent
-or at low abundance in the mice from the beginning and may be one
-contributing factor to why we did not see any changes. Additionally,
-some of the significance of PPI associations in human interventional
-trials appears to be driven by a handful of specific taxa with overall
-differences on a PCoA plot difficult to distinguish (11). Also, most of
-the studies to date compareed different individuals (PPI users versus
-non-users) or the same individuals before and after treatment (4, 7),
-unfortunately these limited microbiota snapshots may be skewed by
-day-to-day microbiota variations (13) that would be revealed with more
-frequent longitudinal sampling.\\
+PPIs had more of an effect on the small intestinal microbiota compared
+to the fecal microbiota (8). The same group demonstrated PPI treatment
+increased the stomach pH in mice (8), which may improve survival of
+bacteria passing through the stomach. We did not find significant
+changes for the taxa observed to be significantly impacted by PPI use in
+human studies. However, 3 of the families that typically increase were
+absent or at low abundance in the mice from the beginning and may be one
+contributing factor to our observation that PPIs had no significant
+effects on bacterial taxa. Additionally, some of the significance of PPI
+associations in human interventional trials appears to be driven by a
+handful of specific taxa with overall differences on a PCoA plot
+difficult to distinguish (9). Also, most of the studies to date compared
+different individuals (PPI users versus non-users) or the same
+individuals before and after treatment (4, 6), unfortunately these
+limited microbiota snapshots may be skewed by day-to-day microbiota
+variations (11) that would be revealed with more frequent longitudinal
+sampling.\\
 ALthough there have been a few \emph{C. difficile} mouse model studies
 that have demonstrated PPIs have some effect on CDIs with or without
-additional antibiotic treatment, none of these studies looked at the
-interplay with the intestinal microbiome (14--16). Additionally, there
-were some key methodical differences between the previous studies and
-our own. Previous work that administered 0.5 mg/kg of the PPI
-lansoprazole daily for 2 weeks to mice and then challenged with \emph{C.
-difficile} demonstrated that PPI treatment alone resulted in detectable
-\emph{C. difficile} in the stool 1 week after challenge, but also showed
-there was detectable \emph{C. difficile} in mice not treated with
-antibiotics (14, 15). The presence of \emph{C. difficile} in mice that
-were not treated with either antibiotics or PPIs could be attributed to
-the higher dose of vegetative cells (10\textsuperscript{8} CFU) used to
-challenge the mice and may partly explain the observation of \emph{C.
-difficile} in PPI-treated mice (14, 15). We have previously shown mice
-from our colony that were not given antibiotics were resistant to
-\emph{C. difficile} 630 when challenged with 10\textsuperscript{3}
-spores (17). Another group also used a higher dose of
-10\textsuperscript{6} spores or 10\textsuperscript{7} vegetative cells
-to challenge antibiotic-treated mice or mice treated with both the PPI
-esomeprazole (40mg/kg dose) for 2 days and antibiotics and demonstrated
-the antibiotic/PPI-treated mice developed more severe CDIs (16).\\
+additional antibiotic treatment(12--14), there were some key
+methodological differences between these studies and our own. One group
+administered 0.5 mg/kg of the PPI lansoprazole daily for 2 weeks to mice
+and then challenged with \emph{C. difficile} demonstrated that PPI
+treatment alone resulted in detectable \emph{C. difficile} in the stool
+1 week after challenge, but also showed there was detectable \emph{C.
+difficile} in mice not treated with antibiotics (12, 13). The presence
+of \emph{C. difficile} in mice that were not treated with either
+antibiotics or PPIs could be attributed to the higher dose of vegetative
+cells (10\textsuperscript{8} CFU) used to challenge the mice and may
+partly explain the observation of \emph{C. difficile} in PPI-treated
+mice (12, 13). We have previously shown mice from our colony that were
+not given antibiotics were resistant to \emph{C. difficile} 630 when
+challenged with 10\textsuperscript{3} spores (15). The other group also
+used a higher dose of 10\textsuperscript{6} spores or
+10\textsuperscript{7} vegetative cells to challenge antibiotic-treated
+mice or mice treated with both the PPI esomeprazole (40mg/kg dose) for 2
+days and antibiotics and demonstrated the antibiotic/PPI-treated mice
+developed more severe CDIs (14).\\
 Our study extended previous work examining PPIs and \emph{C. difficile}
@@
 microbiota and in contrast to previous work, PPIs did not promote
-\emph{C. difficile} colonization. Our 16S rRNA sequencing suggested that
+\emph{C. difficile} colonization. 16S rRNA sequencing suggested that
 \emph{Streptococcus} and \emph{Enterococcus} are rare genera in our
-C57BL/6 mouse colony. If these genera are important contributors to the
-associations between PPIs and CDIs in humans, this may be part of the
-reason we did not see an effect in our CDI mouse model. There could also
-be differences at the species level for taxa that are altered with PPIs
-in humans compared to mice, which may also contribute to the
-association. Other murine starting communities or the addition of human
-PPI treatment-associated strains may be needed to determine whether PPIs
-can alter taxa and subsequently promote CDIs.\\
+C57BL/6 mouse colony. These genera could be important contributors to
+the associations between PPIs and CDIs in humans, and could be a
+contributing factor to our observation that PPI treatment had no effect
+on \emph{C. difficile} colonization in our CDI mouse model. There could
+also be differences at the species level for taxa that are altered with
+PPIs in humans compared to mice. Other murine starting communities or
+the addition of human PPI treatment-associated strains may be needed to
+determine whether PPIs can alter taxa and subsequently promote CDIs.\\
 Factors such as age, body mass index, comorbidities, and use of other
@@
 between PPIs and CDI incidence or recurrence. For example, NSAID
-treatment can increase CDI severity in mice (18) and PPIs are sometimes
+treatment can increase CDI severity in mice (16) and PPIs are sometimes
 used in combination with NSAIDs (3). This study addressed the impact of
@@
 amplified with Accuprime Pfx DNA polymerase (Thermo Fisher Scientific)
-using previously described custom barcoded primers (19). The 16S rRNA
+using previously described custom barcoded primers (17). The 16S rRNA
 amplicon library was sequenced with the MiSeq (Illumina). Amplicons were
@@
 \textbf{16S rRNA gene sequence analysis}Mothur (v1.40.5) was used for
-all sequence processing steps (20) using the previously published
-protocol (19). R (v.3.5.1) was used to generate figures and perform
+all sequence processing steps (18) using the previously published
+protocol (17). R (v.3.5.1) was used to generate figures and perform
 statistical analysis.
@@
 
-\hypertarget{ref-Singh2018}{}
-6. \textbf{Singh A}, \textbf{Cresci GA}, \textbf{Kirby DF}. 2018. Proton
-pump inhibitors: Risks and rewards and emerging consequences to the gut
-microbiome. Nutrition in Clinical Practice \textbf{33}:614--624.
-doi:\href{https://doi.org/10.1002/ncp.10181}{10.1002/ncp.10181}.
-
 \hypertarget{ref-Imhann2017}{}
-7. \textbf{Imhann F}, \textbf{Vila AV}, \textbf{Bonder MJ},
+6. \textbf{Imhann F}, \textbf{Vila AV}, \textbf{Bonder MJ},
 \textbf{Manosalva AGL}, \textbf{Koonen DP}, \textbf{Fu J},
@@
 
-\hypertarget{ref-maier2018extensive}{}
-8. \textbf{Maier L}, \textbf{Pruteanu M}, \textbf{Kuhn M},
-\textbf{Zeller G}, \textbf{Telzerow A}, \textbf{Anderson EE},
-\textbf{Brochado AR}, \textbf{Fernandez KC}, \textbf{Dose H},
-\textbf{Mori H}, \textbf{others}. 2018. Extensive impact of
-non-antibiotic drugs on human gut bacteria. Nature \textbf{555}:623.
-
 \hypertarget{ref-Jenior2018}{}
-9. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+7. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
 \textbf{Schloss PD}. 2018. Clostridium difficile alters the structure
@@
 \hypertarget{ref-Yasutomi2018}{}
-10. \textbf{Yasutomi E}, \textbf{Hoshi N}, \textbf{Adachi S},
+8. \textbf{Yasutomi E}, \textbf{Hoshi N}, \textbf{Adachi S},
 \textbf{Otsuka T}, \textbf{Kong L}, \textbf{Ku Y}, \textbf{Yamairi H},
@@
 \hypertarget{ref-Freedberg2015}{}
-11. \textbf{Freedberg DE}, \textbf{Toussaint NC}, \textbf{Chen SP},
+9. \textbf{Freedberg DE}, \textbf{Toussaint NC}, \textbf{Chen SP},
 \textbf{Ratner AJ}, \textbf{Whittier S}, \textbf{Wang TC}, \textbf{Wang
@@
 \hypertarget{ref-Imhann2015}{}
-12. \textbf{Imhann F}, \textbf{Bonder MJ}, \textbf{Vila AV}, \textbf{Fu
+10. \textbf{Imhann F}, \textbf{Bonder MJ}, \textbf{Vila AV}, \textbf{Fu
 J}, \textbf{Mujagic Z}, \textbf{Vork L}, \textbf{Tigchelaar EF},
@@
 \hypertarget{ref-LlornsRico2019}{}
-13. \textbf{Lloréns-Rico V}, \textbf{Raes J}. 2019. Tracking humans and
+11. \textbf{Lloréns-Rico V}, \textbf{Raes J}. 2019. Tracking humans and
 microbes. Nature \textbf{569}:632--633.
@@
 \hypertarget{ref-Kaur2007}{}
-14. \textbf{Kaur S}, \textbf{Vaishnavi C}, \textbf{Prasad KK},
+12. \textbf{Kaur S}, \textbf{Vaishnavi C}, \textbf{Prasad KK},
 \textbf{Ray P}, \textbf{Kochhar R}. 2007. Comparative role of antibiotic
@@
 \hypertarget{ref-kaur2011effect}{}
-15. \textbf{Kaur S}, \textbf{Vaishnavi C}, \textbf{Prasad KK},
+13. \textbf{Kaur S}, \textbf{Vaishnavi C}, \textbf{Prasad KK},
 \textbf{Ray P}, \textbf{Kochhar R}. 2011. Effect of lactobacillus
@@
 \hypertarget{ref-hung2015proton}{}
-16. \textbf{Hung Y-P}, \textbf{Ko W-C}, \textbf{Chou P-H}, \textbf{Chen
+14. \textbf{Hung Y-P}, \textbf{Ko W-C}, \textbf{Chou P-H}, \textbf{Chen
 Y-H}, \textbf{Lin H-J}, \textbf{Liu Y-H}, \textbf{Tsai H-W}, \textbf{Lee
@@
 \hypertarget{ref-Jenior2017}{}
-17. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+15. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
 \textbf{Schloss PD}. 2017. Clostridium difficile colonizes alternative
@@
 \hypertarget{ref-Maseda2019}{}
-18. \textbf{Maseda D}, \textbf{Zackular JP}, \textbf{Trindade B},
+16. \textbf{Maseda D}, \textbf{Zackular JP}, \textbf{Trindade B},
 \textbf{Kirk L}, \textbf{Roxas JL}, \textbf{Rogers LM},
@@
 \hypertarget{ref-Kozich2013}{}
-19. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+17. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \hypertarget{ref-Schloss2009}{}
-20. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+18. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 
-\includegraphics{results/figures/figure_2.pdf} \textbf{Figure 2. PPI
-treatment alone does not promote CDIs in mice} A. C. difficile CFUs/g
-stool measured each day post C. difficile challenge for clindamycin,
-clindamycin/PPI, and PPI-treated mice. Lines represent the mean CFU/g
-for each treatment group while points represent CFU/g for individual
-mice within each group. B. PCoA of stool samples collected after
-antibiotic treatments. C. Genera significantly associated with treatment
-groups for samples across all sequenced timepoints. D. Families
-significantly associated with treatment groups for samples across all
-sequenced timepoints. C,D data were analyzed by Kruskal-Wallis test with
-a Benjamini-Hochberg correction for multiple comparisons.
+\includegraphics{figure_2.pdf} \textbf{Figure 2. PPI treatment alone
+does not promote CDIs in mice} A. C. difficile CFUs/g stool measured
+each day post C. difficile challenge for clindamycin, clindamycin/PPI,
+and PPI-treated mice. Lines represent the mean CFU/g for each treatment
+group while points represent CFU/g for individual mice within each
+group. B. PCoA of stool samples collected after antibiotic treatments.
+C. Genera significantly associated with treatment groups for samples
+across all sequenced timepoints. D. Families significantly associated
+with treatment groups for samples across all sequenced timepoints. C,D
+data were analyzed by Kruskal-Wallis test with a Benjamini-Hochberg
+correction for multiple comparisons.
 

### 2019-07-29 submission/manuscript.Rmd
Responding to edits & comments from Pat

@@
 bibliography: references.bib
+csl: mbio.csl
+fontsize: 11pt
+geometry: margin=1.0in
 output:
   pdf_document:
-    keep_tex: true
     includes:
       in_header: header.tex
-csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
-fontsize: 11pt
-geometry: margin=1.0in
+    keep_tex: yes
+  word_document: default
 ---
@@
 Proton pump inhibitor (PPI) use has been associated with microbiota alterations and susceptibility to *Clostridium difficile* infections (CDIs) in humans.
-We assessed how PPI treatment alters the fecal microbiota and whether PPIs promote CDIs using a CDI mouse model.
-Mice were treated with a high daily dose of the PPI omeprazole for 7 days prior to *C.
-difficile* challenge and were compared to mice that were treated with the antibiotic clindamycin or both omeprazole and clindamycin.
-High-dose PPI treatment was maintained throughout the experiment, which included pre-treatment, the day of *C.
+We assessed how PPI treatment alters the fecal microbiota and whether PPIs promote CDIs in a mouse model.
+Mice receiving the PPI treatment were gavaged with 40 mg/kg of omeprazole during a 7-day pretreatment phase, the day of *C.
 difficile* challenge, and the following 9 days.
-We found that omeprazole was not sufficient to promote *C.
+We found that mice treated with omeprazole were not colonized by *C.
+difficile*.
+When omeprazole treatment was combined with a single clindamycin treatment, one cage of mice remained resistant to *C.
+difficile* colonization, while the other cage was colonized.
+Treatming mice with clindamycin and no omeprazole followed by challenge resulted in  *C.
 difficile* colonization.
-When omeprazole treatment was combined with the antibiotic,one cage of mice remained resistant to *C.
-difficile* colonization, while the other cage was colonized.
-16S rRNA sequencing analysis revealed that omeprazole had minimal impact on the murine microbiota throughout the 16 days of PPI exposure.
+16S rRNA gene sequencing analysis revealed that omeprazole had minimal impact on the structure of the murine microbiota throughout the 16 days of PPI exposure.
 These results suggest PPI treatment alone is not sufficient to disrupt microbiota resistance to *C.
@@
 ## Importance
-Antibiotics are a major risk factor for *Clostridium difficile* infections (CDIs), but other factors may also contribute to CDI incidence and recurrence.
-Interestingly, other medications besides antibiotics can impact the microbiota.
-Proton pump inhibitors (PPIs) are associated with alterations in the human intestinal microbiota through observational and interventional studies and PPI use has also been associated with CDI incidence and recurrence in epidemiological studies.
-We evaluated the ability of a high dose of the PPI omeprazole to alter the murine intestinal microbiota and disrupt colonization resistance to *C.
+Antibiotics are a major risk factor for *Clostridium difficile* infections (CDIs), but other factors may also contribute.
+In epidemiological studies, PPI use has been associated with CDI incidence and recurrence.
+Proton pump inhibitors (PPIs) have also been associated with alterations in the human intestinal microbiota in observational and interventional studies.
+We evaluated the effects of the PPI omeprazole on the structure of the murine intestinal microbiota and its ability to disrupt colonization resistance to *C.
 difficile*.
-We found daily PPI treatment had minimal impact on the murine fecal microbiota and was not enough to promote *C.
-difficile* colonization after challenge.
-Further studies are needed to determine whether other factors such as the composition of the starting bacterial community, comorbidities, and use of additional medications contribute to the association between PPIs and CDIs seen in humans.
+We found the PPI treatment had minimal impact on the murine fecal microbiota and did not promote *C.
+difficile* colonization.
+Further studies are needed to determine whether other factors such as the composition of the starting bacterial community, comorbidities, and use of additional medications contribute to the association between PPIs and CDIs seen in humans or whether aspects of murine physiology may limit its utility to test these types of hypotheses.
 
@@
 
-## Introduction
 Antibiotics have a large impact on the intestinal microbiome and are a primary risk factor for developing *Clostridium difficile* infections (CDIs) [@Schubert2015].
@@
 Multiple epidemiological studies have suggested an association between proton pump inhibitor (PPI) use and incidence or recurrence of CDIs [@tariq2017association; @nehra2018proton; @Naito2018; @Elias2019].
-There have also been a number of large cohort studies as well as interventional clinical trials that have demonstrated specific alterations in the intestinal microbiome were associated with PPI use [@Imhann2017; @Naito2018].
+There have also been a number of large cohort studies as well as interventional clinical trials that have demonstrated that specific alterations in the intestinal microbiome were associated with PPI use [@Imhann2017; @Naito2018].
 PPI-associated microbiota changes have been attributed to the ability of PPIs to increase stomach acid pH which may promote the survival of oral and pathogenic bacteria [@Imhann2017; @Naito2018].
-Human fecal microbiota changes with PPI use include increases in Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae and decreases in Ruminococcaceae [@Imhann2017].
+Human fecal microbiota changes with PPI use include increases in Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae and decreases in Ruminococcaceae [@Imhann2017; @maier2018extensive].
+Several of these taxa have been associated with C.
+difficile colonization in humans [@Schuberte01021-14]
 Unfortunately, most of the studies suggesting a link between PPIs and *C.
 difficile* were retrospective and did not evaluate the microbiome [@tariq2017association; @nehra2018proton; @Elias2019].
-Thus, it is unclear whether the gastrointestinal microbiome changes associated with PPI use play a role in the association between PPIs and CDI incidence or recurrence.
+Thus, it is unclear whether the gastrointestinal microbiome changes associated with PPI use explain the association between PPIs and CDIs.
 Additionally, epidemiological studies have a limited capacity to address potential confounders and comorbidities in patients that were on PPIs and developed CDIs or recurrent CDIs [@Elias2019; @tariq2017association].
@@
 
-**Murine fecal microbiomes were minimally affected by PPI treatment **
-To test if PPI treatment alters the microbiome and promotes susceptibility to CDIs, we gavaged mice daily with a high dose of omeprazole for 7 days before *C.
+**Murine fecal microbiomes were minimally affected by PPI treatment ** To test whether PPI treatment alters the microbiome and promotes susceptibility to CDIs, we gavaged mice with 40 mg/kg of omeprazole for 7 days before *C.
 difficile* challenge (Figure 1A).
-The fecal bacterial communities of PPI-treated mice were compared to a group of mice that received the antibiotic, clindamycin, and a group that received both PPI and clindamycin treatment using 16S rRNA gene sequencing.
-A principle coordinates analysis (PCoA) of the Bray-Curtis distances over the initial 7 days of treatment revealed the bacterial communities of PPI-treated mice remained relatively unchanged (Figure 1B).
-Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes that have previously been impacted by PPIs in human studies, but we observed no significant fluctuations in the stool of our PPI-treated mice throughout the course of the 16-day experiment (Figure 1C, Figure S1).
-In contrast, the microbiomes from the 2 groups of mice that received clindamycin started to shift away from the PPI-treated mice and the samples from previous days one day after treatment (Figure 1D).
-These results demonstrated that high dose PPI treatment alone had no significant impact on the murine fecal bacterial community.
+A principle coordinates analysis (PCoA) of the Bray-Curtis distances over the initial 7 days of treatment revealed the bacterial communities of PPI-treated mice remained relatively unchanged relative to animals that did not recieve the PPI (Figure 1B).
+We observed no significant fluctuations in the relative abundance of those taxa previously shown to respond to PPI treatment throughout the course of the 16-day experiment (Figure 1C, Figure S1A-B).
+We also observed no significant fluctuations in relative abundances at the family and genus level over the beginning, day of *C.
+difficile* challenge, and end of the experiment for the PPI-treated mice (Figure 1D, Figure S1C).
+Next, the fecal bacterial communities of PPI-treated mice were compared to a group of mice that received a single 10 mg/kg dose of clindamycin and a group that received both PPI and clindamycin.
+In contrast to the dynamics of mice only receiving omeprazole, the microbiomes from the 2 groups of mice that received clindamycin started to shift away from the PPI-treated mice and the samples from previous days one day after treatment (Figure 1D).
+These results demonstrated that the PPI treatment alone had a minimal impact on the murine fecal bacterial community relative to the impact of clindamycin treatment.
 
-**PPI-treatment did not promote susceptibility to *C. difficile infection* in mice**
-Next, we examined whether high dose PPI treatment altered susceptibility to *C.
+**PPI-treatment did not promote susceptibility to *C.
+difficile infection* in mice** Next, we examined whether PPI treatment altered susceptibility to *C.
 difficile* infection in mice.
-After 7 days of PPI treatment or 1 day after clindamycin treatment, mice were challenged with *C.
+After PPI treatment or clindamycin treatment, mice were challenged with *C.
 difficile* 630 spores.
-While all 4 of the clindamycin-treated mice were colonized with *C.
-difficile*, all of the PPI-treated mice were resistant to *C.
-difficile* colonization (Figure 2A).
-Interestingly, only 1 cage of the PPI and clindamycin-treated mice were colonized, while the other cage was resistant (Figure 2A).
-PCoAs of the bacterial communities from the 3 groups of mice after *C.
-difficile* challenge, showed that the greatest shifts in bacterial communities occurred in the clindamycin-treated mice (Figure 2B).
-Within 5 days, all of the mice had cleared *C.
-difficile*, suggesting there was no difference in rate of clearance for the mice that were initially colonized with *C.
-difficile* despite continuing to receive PPI treatment throughout the course of the experiment.
-Our results suggest that PPI treatment alone had no effects on bacterial community resistance to *C.
+While *C.
+difficile* colonized the clindamycin treated mice, it was unable to colonize the PPI-treated mice (Figure 2A).
+Interestingly, only 1 cage of mice that received both the PPI and clindamycin were colonized, while the mice in other cage wwere resistant (Figure 2A).
+The greatest shifts in bacterial communities occurred in the clindamycin-treated mice (Figure 2B, Figure S2).
+Regardless of whether the mice became colonized, all of the mice had cleared *C.
+difficile* within 5 days (Figure 2A), suggesting that omeprazole did not affect the rate of clearance.
+Our results suggest that PPI treatment had no effects on bacterial community resistance to *C.
 difficile* colonization in mice.
-Instead most of the differences between our 3 treatment groups appear to be driven by clindamycin administration and included decreased *Alistipes*, *Barnesiella*, Porhyromonadaceae, Ruminococcaceae, taxa previously found to be altered in clindamycin-treated mice that were challenged with *C.
-difficile* [@Schubert2015].
+Instead most of the differences between our 3 treatment groups appear to be driven by clindamycin administration (Figure 2C-D) and included previously described decreases in *Alistipes*, *Barnesiella*, Porhyromonadaceae, Ruminococcaceae [@Schubert2015].
 These findings demonstrated that high dose PPI treatment did not promote susceptibility to *C.
@@
 
-## Discussion and Conclusions
+## Conclusions
+We found the PPI omeprazole did not impact the gut microbiota and did not promote *C.
+difficile* infection in mice.
 Our findings that PPI treatment had minimal impact on the fecal microbiome were comparable to another PPI mouse study that indicated PPIs had more of an effect on the small intestinal microbiota compared to the fecal microbiota [@Yasutomi2018].
@@
 We did not find significant changes for the taxa observed to be significantly impacted by PPI use in human studies.
-However, 3 of the families that typically increase were absent or at low abundance in the mice from the beginning and may be one contributing factor to our observation that PPIs had no significant effects on bacterial taxa.
+However, 3 of the human-associated taxa were absent or at low abundance in our mice.
 Additionally, some of the significance of PPI associations in human interventional trials appears to be driven by a handful of specific taxa with overall differences on a PCoA plot difficult to distinguish [@Freedberg2015, @Imhann2015].
-Also, most of the studies to date compared different individuals (PPI users versus non-users) or the same individuals before and after treatment [@Naito2018; @Imhann2017], unfortunately these limited microbiota snapshots may be skewed by day-to-day microbiota variations [@LlornsRico2019] that would be revealed with more frequent longitudinal sampling.
-ALthough there have been a few *C.
-difficile* mouse model studies that have demonstrated PPIs have some effect on CDIs with or without additional antibiotic treatment[@Kaur2007; @kaur2011effect; @hung2015proton], there were some key methodological differences between these studies and our own.
+The PPI-treated mice fecal communities did not change overall and no differences in the relative abundances of individual families or genera were observed over time.
+While our fecal microbiota findings are comparable to what's been shown in another mouse study, whether PPI-induced changes in specific bacterial abundances observed in humans play a role in CDIs remains to be determined.
+Although there have been a few *C.
+difficile* mouse model studies that have demonstrated PPIs have some effect on CDIs with or without additional antibiotic treatment [@Kaur2007; @kaur2011effect; @hung2015proton], there were some key methodological differences between these studies and our own.
 One group administered 0.5 mg/kg of the PPI lansoprazole daily for 2 weeks to mice and then challenged with *C.
@@
 The presence of *C.
-difficile* in mice that were not treated with either antibiotics or PPIs could be attributed to the higher dose of vegetative cells (10^8^ CFU) used to challenge the mice and may partly explain the observation of *C.
-difficile* in PPI-treated mice [@Kaur2007; @kaur2011effect].
-We have previously shown mice from our colony that were not given antibiotics were resistant to *C.
-difficile* 630 when challenged with 10^3^ spores [@Jenior2017].
-The other group also used a higher dose of 10^6^ spores or 10^7^ vegetative cells to challenge antibiotic-treated mice or mice treated with both the PPI esomeprazole (40mg/kg dose) for 2 days and antibiotics and demonstrated the antibiotic/PPI-treated mice developed more severe CDIs [@hung2015proton].
-Our study extended previous work examining PPIs and *C.
-difficile* in mice to examine the contribution of the intestinal microbiota.
-We found the PPI omeprazole had minimal impact on the murine intestinal microbiota and in contrast to previous work, PPIs did not promote *C.
-difficile* colonization.
-16S rRNA sequencing suggested that *Streptococcus* and *Enterococcus* are rare genera in our C57BL/6 mouse colony.
+difficile* in mice not treated with antibiotics may be partly attributed to the higher dose of vegetative cells (10^8^ CFU) used, but nevertheless makes it difficult to interpret the extent to which PPI treatment contributed to *C.
+difficile* colonization in their mouse model [@Kaur2007; @kaur2011effect].
+In contrast, we have previously shown mice from our colony that were not given antibiotics were resistant to *C.
+difficile* 630 when challenged with 10^3^ spores [@Jenior2017] and previous work by our collaborators demonstrated mice not given antibiotics were resistant to 10^5^ vegetative cells of *C.
+difficile * VPI  10463 [@Reeves2011].
+The other PPI CDI mouse study also used a higher dose of 10^6^ spores or 10^7^ vegetative cells to challenge antibiotic-treated mice or mice treated with both antibiotics and the PPI esomeprazole (40mg/kg dose) for 2 days and demonstrated the antibiotic/PPI-treated mice developed more severe CDIs [@hung2015proton].
+We tested the same high 40 mg/kg PPI dose and expanded pre-treatment to 7 days before challenge to test the impact of PPI treatment alone on our CDI mouse model.
+Our study extended the previous work examining PPIs and *C.
+difficile* in mice by evaluating the contribution of the intestinal microbiota.
+We found the PPI omeprazole had no significant impact on bacterial taxa within the murine intestinal microbiota over the 16-day experiment and in contrast to previous work, PPIs did not alter *C.
+difficile* colonization resistance.
+16S rRNA sequencing suggested that *Streptococcus* and *Enterococcus* are rare genera in our C57BL/6 mouse colony and variation in these 2 genera have been observed across other facilities and vendors [@Rausch2016; @Robertson2019].
 These genera could be important contributors to the associations between PPIs and CDIs in humans, and could be a contributing factor to our observation that PPI treatment had no effect on *C.
 difficile* colonization in our CDI mouse model.
-There could also be differences at the species level for taxa that are altered with PPIs in humans compared to mice.
-Other murine starting communities or the addition of human PPI treatment-associated strains may be needed to determine whether PPIs can alter taxa and subsequently promote CDIs.
-Factors such as age, body mass index, comorbidities, and use of other medications in human studies may also be contributing to the association between PPIs and CDI incidence or recurrence.
-For example, NSAID treatment can increase CDI severity in mice [@Maseda2019] and PPIs are sometimes used in combination with NSAIDs [@nehra2018proton].
+Beyond microbiome differences, factors such as age, body mass index, comorbidities, and use of other medications in human studies may also be contributing to the association between PPIs and CDI incidence or recurrence.
 This study addressed the impact of PPIs with or without antibiotics on a murine model of CDI, and found PPIs did not promote *C.
 difficile* colonization.
+The epidemiological evidence linking PPIs to CDIs is primarily from observational studies, which makes determining causality and whether other risk factors play a role challenging [@Eze2017].
 Future studies are needed to determine whether age, other comorbidities and bacterial strains that are less common in mice can increase the risk of CDIs or recurrent CDIs when combined with PPI treatment.
@@
 We would also like to thank the Unit for Laboratory Animal Medicine at the University of Michigan for maintaining our mouse colony and providing the infrastructure and support for performing our mouse expeirments.
-The authors are also thankful to members of the Schloss lab for helpful discussions throughout the process of designing the experiment, analyzing the results and drafting of the manuscript.
+The authors are also thankful to members of the Schloss lab for helpful discussions throughout the process of designing the experiment, analyzing the results, crafting the figures, and drafting of the manuscript.
 
@@
 **Animals** All mouse experiments were performed with 7- to 12-week-old C57BL/6 male and female mice.
-Each experimental group of mice was spit between 2 cages with 2-3 mice housed per cage and male and female mice were housed separately.
+Each experimental group of mice was spit between 2 cages with 2-3 mice housed per cage and male and female mice housed separately.
 All animal experiments were approved by the University of Michigan Animal Care and Use Committee (IACUC) under protocol number PRO00006983.
@@
 **16S rRNA gene sequencing** DNA for 16S rRNA gene sequencing was extracted from 10-50 mg fecal pellet from each mouse using the DNeasy Powersoil HTP 96 Kit (Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf).
+The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used as a mock community [@Sze2019] and water was used as a negative control.
 The V4 hypervariable region of the 16S rRNA gene was amplified with Accuprime Pfx DNA polymerase (Thermo Fisher Scientific) using previously described custom barcoded primers [@Kozich2013].
@@
 **16S rRNA gene sequence analysis**Mothur (v1.40.5) was used for all sequence processing steps [@Schloss2009] using the previously published protocol [@Kozich2013].
+In brief, forward and reverse reads for each sample were combined and low-quality sequences and chimeras were removed.
+Duplicate sequences were merged, before taxonomy assignment using a modified version (v16) of the Ribosomal Database Project reference database (v11.5) with an 80% cutoff.
+Operational taxonomic units (OTUs) were assigned with the opticlust clustering algorithm using a 97% similarity threshold.
+To adjust for uneven sequencing across samples, all samples were rarefied to 3000 sequences, 1,000 times.
+PCoAs were generated based on Bray-Curtis distance.
 R (v.3.5.1) was used to generate figures and perform statistical analysis.
 
-**Statistical Analysis** To assess for differences in relative abundance in families and genera across our 3 different treatment groups (Clindamycin, Clindamycin + PPI, and PPI), we used a Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
+**Statistical Analysis** To test for differences in relative abundances in families and genera across our 3 different treatment groups (Clindamycin, Clindamycin + PPI, and PPI) or within the PPI treatment group across 3 timepoints (Day -7, 0, and 9), we used a Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
 
@@
 
-**Data availability** The 16S rRNA sequencing data have been deposited in the NCBI Sequence Read Archive (Accession no. _________)
+**Data availability** The 16S rRNA sequencing data have been deposited in the NCBI Sequence Read Archive (Accession no. PRJNA554866).
 
@@
 **Figure 1.
-PPI treatment had minimal impact on the murine fecal microbiota** A.
+PPI treatment had minimal impact on the murine fecal microbiota.** A.
 Mouse experiment timeline and logistics.
+The PPI omeprazole was administered throughout the duration of the experiment.
+Clindamycin was administered 1 day before *C.
+difficile* challenge on Day 0.
 Stools for 16S rRNA sequencing analysis were collected on the days that are labeled (Day -7, -5, -3, -1, 0, 1, 2, 3, 4, 5, 7, 9).
+*C.
+difficile* CFU in the stool was quantified daily through 6 days post-infection by anaerobic culture.
 B.
-Principal Coordinates Analaysis (PCoA) of Bray-curtis distances from stool samples for all treatment groups during the intial 7 days of treatment with the PPI omeprazole or vehicle cluster together and do not chage much over time.
+Principal Coordinates Analaysis (PCoA) of Bray-curtis distances from stool samples for all treatment groups during the intial 7 days of the experiment.
+Each symbol represents a stool sample from an individual mouse on a given day.
+Transparency of the symbol corresponds to treatment day.
+Circles represent samples before antibiotic treatment (administered on Day -1), while open diamonds represent samples from the day after antibiotic treatment (Day 0).
+Mice treated with the PPI omeprazole or vehicle cluster together and do not change much over time while mice that receive clindamycin have samples that start to separate from the rest of the samples the day after antibiotic treatment.
 C.
@@
 D.
-PCoA of stool samples collected before and 1 day after antibiotic treatment, demonstrating the fecal microbiota of mice treated with clindamycin starts to separate from the rest of the samples.
+Boxplots showing the relative abundances of 7 familes within PPI-treated mice that had the lowest adjusted P values after comparing across 3 timepoints.
+There were no significant differences acoss time for any of the identified families (All P > 0.05) after analysis by Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
+See Fig.
+S1C for genera with lowest adjusted P values (All P > 0.05).
+For C-D, the grey line indicates the limit of detection.
 
@@
 **Figure 2.
-PPI treatment alone does not promote CDIs in mice** A.
+PPI treatment alone does not promote CDIs in mice.** A.
+*C.
+difficile* CFUs/g stool measured each day post *C.
+difficile* challenge for clindamycin, clindamycin/PPI, and PPI-treated mice.
+Lines represent the mean CFU/g for each treatment group while points represent CFU/g for individual mice within each group.
+The black dashed line indicates the limit of detection.
+B.
+PCoA of of Bray-curtis distances from stool samples collected after antibiotic treatment (last 9 days of the experiment).
+Transparency of the symbol corresponds to treatment day.
+Symbols represent the *C.
+difficile* colonization status of the mice measured 2 days post-infection.
+Circles represent resistant mice (*C.
+difficile* was undetectable in stool samples), while X-shapes represent mice that were colonized with *C.
+difficile*, although all mice cleared *C.
+difficile* within 5 days of infection.
+PPI-treated fecal samples primarily cluster together throughout the experiment.
 C.
-difficile CFUs/g stool measured each day post C.
-difficile challenge for clindamycin, clindamycin/PPI, and PPI-treated mice.
-Lines represent the mean CFU/g for each treatment group while points represent CFU/g for individual mice within each group.
-B.
-PCoA of stool samples collected after antibiotic treatments.
-C.
-Genera significantly associated with treatment groups for samples across all sequenced timepoints.
+Genera significantly associated with treatment groups of stool samples across all sequenced timepoints.
 D.
-Families significantly associated with treatment groups for samples across all sequenced timepoints.
+Families significantly associated with treatment groups of stool samples across all sequenced timepoints.
 C,D data were analyzed by Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
+For C-D, the grey line indicates the limit of detection.
 
@@
 **Figure S1.
-Previous human PPI-associated families fluctuate over time with no overall trend in either direction** Relative abundance over time for Lactobacillaceae (A) and Ruminococcaceae (B), 2 of the PPI-associated familes from human PPI studies across all 3 treatment groups.
-Each point represents the relative abundance for an individual mouse
+Taxa within PPI treated mice fluctuate over time with no overall trend in either direction.** Relative abundance over time for Lactobacillaceae (A) and Ruminococcaceae (B), 2 of the PPI-associated familes from human PPI studies across all 3 treatment groups.
+Each point represents the relative abundance for an individual mouse, while the lines represent the mean relative abundances for each treatment group of mice.
+C.
+Boxplots showing the relative abundances of 13 genera within PPI-treated mice that had the lowest adjusted P values after comparing across 3 timepoints.
+There were no significant differences acoss time for any of the identified families (All P = 0.568) after analysis by Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
+For A-C, the grey lines indicate the limit of detection.
+
+\newpage
+\includegraphics{figure_s2.pdf}
+**Figure S2.
+Microbiota diversity and richness decrease with antibiotic treatment but remain relatively constant with PPI treatment.** A.
+Boxplots of the Shannon Index (A) and richness (B) values for each group of mice over 3 timepoints (Day -7, 0, and 9).
+Each circle represents the value for an individual mouse.

### 2019-07-29 submission/manuscript.tex
Responding to edits & comments from Pat

@@
 infections (CDIs) in humans. We assessed how PPI treatment alters the
-fecal microbiota and whether PPIs promote CDIs using a CDI mouse model.
-Mice were treated with a high daily dose of the PPI omeprazole for 7
-days prior to \emph{C. difficile} challenge and were compared to mice
-that were treated with the antibiotic clindamycin or both omeprazole and
-clindamycin. High-dose PPI treatment was maintained throughout the
-experiment, which included pre-treatment, the day of \emph{C. difficile}
-challenge, and the following 9 days. We found that omeprazole was not
-sufficient to promote \emph{C. difficile} colonization. When omeprazole
-treatment was combined with the antibiotic,one cage of mice remained
-resistant to \emph{C. difficile} colonization, while the other cage was
-colonized. 16S rRNA sequencing analysis revealed that omeprazole had
-minimal impact on the murine microbiota throughout the 16 days of PPI
-exposure. These results suggest PPI treatment alone is not sufficient to
-disrupt microbiota resistance to \emph{C. difficle} infection in mice
-that are normally resistant in the absence of antibiotic treatment.
+fecal microbiota and whether PPIs promote CDIs in a mouse model. Mice
+receiving the PPI treatment were gavaged with 40 mg/kg of omeprazole
+during a 7-day pretreatment phase, the day of \emph{C. difficile}
+challenge, and the following 9 days. We found that mice treated with
+omeprazole were not colonized by \emph{C. difficile}. When omeprazole
+treatment was combined with a single clindamycin treatment, one cage of
+mice remained resistant to \emph{C. difficile} colonization, while the
+other cage was colonized. Treatming mice with clindamycin and no
+omeprazole followed by challenge resulted in \emph{C. difficile}
+colonization. 16S rRNA gene sequencing analysis revealed that omeprazole
+had minimal impact on the structure of the murine microbiota throughout
+the 16 days of PPI exposure. These results suggest PPI treatment alone
+is not sufficient to disrupt microbiota resistance to \emph{C. difficle}
+infection in mice that are normally resistant in the absence of
+antibiotic treatment.
 
@@
 Antibiotics are a major risk factor for \emph{Clostridium difficile}
-infections (CDIs), but other factors may also contribute to CDI
-incidence and recurrence. Interestingly, other medications besides
-antibiotics can impact the microbiota. Proton pump inhibitors (PPIs) are
-associated with alterations in the human intestinal microbiota through
-observational and interventional studies and PPI use has also been
-associated with CDI incidence and recurrence in epidemiological studies.
-We evaluated the ability of a high dose of the PPI omeprazole to alter
-the murine intestinal microbiota and disrupt colonization resistance to
-\emph{C. difficile}. We found daily PPI treatment had minimal impact on
-the murine fecal microbiota and was not enough to promote \emph{C.
-difficile} colonization after challenge. Further studies are needed to
+infections (CDIs), but other factors may also contribute. In
+epidemiological studies, PPI use has been associated with CDI incidence
+and recurrence. Proton pump inhibitors (PPIs) have also been associated
+with alterations in the human intestinal microbiota in observational and
+interventional studies. We evaluated the effects of the PPI omeprazole
+on the structure of the murine intestinal microbiota and its ability to
+disrupt colonization resistance to \emph{C. difficile}. We found the PPI
+treatment had minimal impact on the murine fecal microbiota and did not
+promote \emph{C. difficile} colonization. Further studies are needed to
 determine whether other factors such as the composition of the starting
 bacterial community, comorbidities, and use of additional medications
-contribute to the association between PPIs and CDIs seen in humans.
+contribute to the association between PPIs and CDIs seen in humans or
+whether aspects of murine physiology may limit its utility to test these
+types of hypotheses.
 
 \newpage
-
-\subsection{Introduction}\label{introduction}
 
@@
 studies as well as interventional clinical trials that have demonstrated
-specific alterations in the intestinal microbiome were associated with
-PPI use (4, 6). PPI-associated microbiota changes have been attributed
-to the ability of PPIs to increase stomach acid pH which may promote the
-survival of oral and pathogenic bacteria (4, 6). Human fecal microbiota
-changes with PPI use include increases in Enterococcaceae,
+that specific alterations in the intestinal microbiome were associated
+with PPI use (4, 6). PPI-associated microbiota changes have been
+attributed to the ability of PPIs to increase stomach acid pH which may
+promote the survival of oral and pathogenic bacteria (4, 6). Human fecal
+microbiota changes with PPI use include increases in Enterococcaceae,
 Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae
-and decreases in Ruminococcaceae (6).\\
+and decreases in Ruminococcaceae (6, 7). Several of these taxa have been
+associated with C. difficile colonization in humans (8)\\
 Unfortunately, most of the studies suggesting a link between PPIs and
@@
 microbiome (2, 3, 5). Thus, it is unclear whether the gastrointestinal
-microbiome changes associated with PPI use play a role in the
-association between PPIs and CDI incidence or recurrence. Additionally,
-epidemiological studies have a limited capacity to address potential
-confounders and comorbidities in patients that were on PPIs and
-developed CDIs or recurrent CDIs (2, 5). Here, we evaluated the impact
-of a daily high dose PPI treatment on the murine microbiome and
-susceptibility to \emph{C. difficile} colonization in relation to
-clindamycin, an antibiotic that perturbs the microbiome enough to allow
-\emph{C. difficile} to colonize but is mild enough that \emph{C.
-difficile} is cleared within 10 days (7).
+microbiome changes associated with PPI use explain the association
+between PPIs and CDIs. Additionally, epidemiological studies have a
+limited capacity to address potential confounders and comorbidities in
+patients that were on PPIs and developed CDIs or recurrent CDIs (2, 5).
+Here, we evaluated the impact of a daily high dose PPI treatment on the
+murine microbiome and susceptibility to \emph{C. difficile} colonization
+in relation to clindamycin, an antibiotic that perturbs the microbiome
+enough to allow \emph{C. difficile} to colonize but is mild enough that
+\emph{C. difficile} is cleared within 10 days (9).
 
 \textbf{Murine fecal microbiomes were minimally affected by PPI
-treatment } To test if PPI treatment alters the microbiome and promotes
-susceptibility to CDIs, we gavaged mice daily with a high dose of
+treatment } To test whether PPI treatment alters the microbiome and
+promotes susceptibility to CDIs, we gavaged mice with 40 mg/kg of
 omeprazole for 7 days before \emph{C. difficile} challenge (Figure 1A).
-The fecal bacterial communities of PPI-treated mice were compared to a
-group of mice that received the antibiotic, clindamycin, and a group
-that received both PPI and clindamycin treatment using 16S rRNA gene
-sequencing. A principle coordinates analysis (PCoA) of the Bray-Curtis
-distances over the initial 7 days of treatment revealed the bacterial
-communities of PPI-treated mice remained relatively unchanged (Figure
-1B). Enterococcaceae, Lactobacillaceae, Micrococcaceae,
-Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes
-that have previously been impacted by PPIs in human studies, but we
-observed no significant fluctuations in the stool of our PPI-treated
-mice throughout the course of the 16-day experiment (Figure 1C, Figure
-S1). In contrast, the microbiomes from the 2 groups of mice that
+A principle coordinates analysis (PCoA) of the Bray-Curtis distances
+over the initial 7 days of treatment revealed the bacterial communities
+of PPI-treated mice remained relatively unchanged relative to animals
+that did not recieve the PPI (Figure 1B). We observed no significant
+fluctuations in the relative abundance of those taxa previously shown to
+respond to PPI treatment throughout the course of the 16-day experiment
+(Figure 1C, Figure S1A-B). We also observed no significant fluctuations
+in relative abundances at the family and genus level over the beginning,
+day of \emph{C. difficile} challenge, and end of the experiment for the
+PPI-treated mice (Figure 1D, Figure S1C). Next, the fecal bacterial
+communities of PPI-treated mice were compared to a group of mice that
+received a single 10 mg/kg dose of clindamycin and a group that received
+both PPI and clindamycin. In contrast to the dynamics of mice only
+receiving omeprazole, the microbiomes from the 2 groups of mice that
 received clindamycin started to shift away from the PPI-treated mice and
 the samples from previous days one day after treatment (Figure 1D).
-These results demonstrated that high dose PPI treatment alone had no
-significant impact on the murine fecal bacterial community.
+These results demonstrated that the PPI treatment alone had a minimal
+impact on the murine fecal bacterial community relative to the impact of
+clindamycin treatment.
 
 \textbf{PPI-treatment did not promote susceptibility to \emph{C.
-difficile infection} in mice} Next, we examined whether high dose PPI
-treatment altered susceptibility to \emph{C. difficile} infection in
-mice. After 7 days of PPI treatment or 1 day after clindamycin
-treatment, mice were challenged with \emph{C. difficile} 630 spores.
-While all 4 of the clindamycin-treated mice were colonized with \emph{C.
-difficile}, all of the PPI-treated mice were resistant to \emph{C.
-difficile} colonization (Figure 2A). Interestingly, only 1 cage of the
-PPI and clindamycin-treated mice were colonized, while the other cage
-was resistant (Figure 2A). PCoAs of the bacterial communities from the 3
-groups of mice after \emph{C. difficile} challenge, showed that the
-greatest shifts in bacterial communities occurred in the
-clindamycin-treated mice (Figure 2B). Within 5 days, all of the mice had
-cleared \emph{C. difficile}, suggesting there was no difference in rate
-of clearance for the mice that were initially colonized with \emph{C.
-difficile} despite continuing to receive PPI treatment throughout the
-course of the experiment. Our results suggest that PPI treatment alone
-had no effects on bacterial community resistance to \emph{C. difficile}
-colonization in mice. Instead most of the differences between our 3
-treatment groups appear to be driven by clindamycin administration and
-included decreased \emph{Alistipes}, \emph{Barnesiella},
-Porhyromonadaceae, Ruminococcaceae, taxa previously found to be altered
-in clindamycin-treated mice that were challenged with \emph{C.
-difficile} (1). These findings demonstrated that high dose PPI treatment
-did not promote susceptibility to \emph{C. difficile} colonization.
+difficile infection} in mice} Next, we examined whether PPI treatment
+altered susceptibility to \emph{C. difficile} infection in mice. After
+PPI treatment or clindamycin treatment, mice were challenged with
+\emph{C. difficile} 630 spores. While \emph{C. difficile} colonized the
+clindamycin treated mice, it was unable to colonize the PPI-treated mice
+(Figure 2A). Interestingly, only 1 cage of mice that received both the
+PPI and clindamycin were colonized, while the mice in other cage wwere
+resistant (Figure 2A). The greatest shifts in bacterial communities
+occurred in the clindamycin-treated mice (Figure 2B, Figure S2).
+Regardless of whether the mice became colonized, all of the mice had
+cleared \emph{C. difficile} within 5 days (Figure 2A), suggesting that
+omeprazole did not affect the rate of clearance. Our results suggest
+that PPI treatment had no effects on bacterial community resistance to
+\emph{C. difficile} colonization in mice. Instead most of the
+differences between our 3 treatment groups appear to be driven by
+clindamycin administration (Figure 2C-D) and included previously
+described decreases in \emph{Alistipes}, \emph{Barnesiella},
+Porhyromonadaceae, Ruminococcaceae (1). These findings demonstrated that
+high dose PPI treatment did not promote susceptibility to \emph{C.
+difficile} colonization.
 
-\subsection{Discussion and
-Conclusions}\label{discussion-and-conclusions}
+\subsection{Conclusions}\label{conclusions}
 
-Our findings that PPI treatment had minimal impact on the fecal
-microbiome were comparable to another PPI mouse study that indicated
-PPIs had more of an effect on the small intestinal microbiota compared
-to the fecal microbiota (8). The same group demonstrated PPI treatment
-increased the stomach pH in mice (8), which may improve survival of
-bacteria passing through the stomach. We did not find significant
-changes for the taxa observed to be significantly impacted by PPI use in
-human studies. However, 3 of the families that typically increase were
-absent or at low abundance in the mice from the beginning and may be one
-contributing factor to our observation that PPIs had no significant
-effects on bacterial taxa. Additionally, some of the significance of PPI
-associations in human interventional trials appears to be driven by a
-handful of specific taxa with overall differences on a PCoA plot
-difficult to distinguish (9). Also, most of the studies to date compared
-different individuals (PPI users versus non-users) or the same
-individuals before and after treatment (4, 6), unfortunately these
-limited microbiota snapshots may be skewed by day-to-day microbiota
-variations (11) that would be revealed with more frequent longitudinal
-sampling.\\
-ALthough there have been a few \emph{C. difficile} mouse model studies
+We found the PPI omeprazole did not impact the gut microbiota and did
+not promote \emph{C. difficile} infection in mice. Our findings that PPI
+treatment had minimal impact on the fecal microbiome were comparable to
+another PPI mouse study that indicated PPIs had more of an effect on the
+small intestinal microbiota compared to the fecal microbiota (10). The
+same group demonstrated PPI treatment increased the stomach pH in mice
+(10), which may improve survival of bacteria passing through the
+stomach. We did not find significant changes for the taxa observed to be
+significantly impacted by PPI use in human studies. However, 3 of the
+human-associated taxa were absent or at low abundance in our mice.
+Additionally, some of the significance of PPI associations in human
+interventional trials appears to be driven by a handful of specific taxa
+with overall differences on a PCoA plot difficult to distinguish (11).
+The PPI-treated mice fecal communities did not change overall and no
+differences in the relative abundances of individual families or genera
+were observed over time. While our fecal microbiota findings are
+comparable to what's been shown in another mouse study, whether
+PPI-induced changes in specific bacterial abundances observed in humans
+play a role in CDIs remains to be determined.\\
+Although there have been a few \emph{C. difficile} mouse model studies
 that have demonstrated PPIs have some effect on CDIs with or without
-additional antibiotic treatment(12--14), there were some key
+additional antibiotic treatment (13--15), there were some key
 methodological differences between these studies and our own. One group
@@
 1 week after challenge, but also showed there was detectable \emph{C.
-difficile} in mice not treated with antibiotics (12, 13). The presence
-of \emph{C. difficile} in mice that were not treated with either
-antibiotics or PPIs could be attributed to the higher dose of vegetative
-cells (10\textsuperscript{8} CFU) used to challenge the mice and may
-partly explain the observation of \emph{C. difficile} in PPI-treated
-mice (12, 13). We have previously shown mice from our colony that were
-not given antibiotics were resistant to \emph{C. difficile} 630 when
-challenged with 10\textsuperscript{3} spores (15). The other group also
-used a higher dose of 10\textsuperscript{6} spores or
-10\textsuperscript{7} vegetative cells to challenge antibiotic-treated
-mice or mice treated with both the PPI esomeprazole (40mg/kg dose) for 2
-days and antibiotics and demonstrated the antibiotic/PPI-treated mice
-developed more severe CDIs (14).\\
-Our study extended previous work examining PPIs and \emph{C. difficile}
-in mice to examine the contribution of the intestinal microbiota. We
-found the PPI omeprazole had minimal impact on the murine intestinal
-microbiota and in contrast to previous work, PPIs did not promote
-\emph{C. difficile} colonization. 16S rRNA sequencing suggested that
+difficile} in mice not treated with antibiotics (13, 14). The presence
+of \emph{C. difficile} in mice not treated with antibiotics may be
+partly attributed to the higher dose of vegetative cells
+(10\textsuperscript{8} CFU) used, but nevertheless makes it difficult to
+interpret the extent to which PPI treatment contributed to \emph{C.
+difficile} colonization in their mouse model (13, 14). In contrast, we
+have previously shown mice from our colony that were not given
+antibiotics were resistant to \emph{C. difficile} 630 when challenged
+with 10\textsuperscript{3} spores (16) and previous work by our
+collaborators demonstrated mice not given antibiotics were resistant to
+10\textsuperscript{5} vegetative cells of \emph{C. difficile } VPI 10463
+(17). The other PPI CDI mouse study also used a higher dose of
+10\textsuperscript{6} spores or 10\textsuperscript{7} vegetative cells
+to challenge antibiotic-treated mice or mice treated with both
+antibiotics and the PPI esomeprazole (40mg/kg dose) for 2 days and
+demonstrated the antibiotic/PPI-treated mice developed more severe CDIs
+(15). We tested the same high 40 mg/kg PPI dose and expanded
+pre-treatment to 7 days before challenge to test the impact of PPI
+treatment alone on our CDI mouse model.\\
+Our study extended the previous work examining PPIs and \emph{C.
+difficile} in mice by evaluating the contribution of the intestinal
+microbiota. We found the PPI omeprazole had no significant impact on
+bacterial taxa within the murine intestinal microbiota over the 16-day
+experiment and in contrast to previous work, PPIs did not alter \emph{C.
+difficile} colonization resistance. 16S rRNA sequencing suggested that
 \emph{Streptococcus} and \emph{Enterococcus} are rare genera in our
-C57BL/6 mouse colony. These genera could be important contributors to
-the associations between PPIs and CDIs in humans, and could be a
-contributing factor to our observation that PPI treatment had no effect
-on \emph{C. difficile} colonization in our CDI mouse model. There could
-also be differences at the species level for taxa that are altered with
-PPIs in humans compared to mice. Other murine starting communities or
-the addition of human PPI treatment-associated strains may be needed to
-determine whether PPIs can alter taxa and subsequently promote CDIs.\\
-Factors such as age, body mass index, comorbidities, and use of other
-medications in human studies may also be contributing to the association
-between PPIs and CDI incidence or recurrence. For example, NSAID
-treatment can increase CDI severity in mice (16) and PPIs are sometimes
-used in combination with NSAIDs (3). This study addressed the impact of
-PPIs with or without antibiotics on a murine model of CDI, and found
-PPIs did not promote \emph{C. difficile} colonization. Future studies
-are needed to determine whether age, other comorbidities and bacterial
-strains that are less common in mice can increase the risk of CDIs or
-recurrent CDIs when combined with PPI treatment.
+C57BL/6 mouse colony and variation in these 2 genera have been observed
+across other facilities and vendors (18, 19). These genera could be
+important contributors to the associations between PPIs and CDIs in
+humans, and could be a contributing factor to our observation that PPI
+treatment had no effect on \emph{C. difficile} colonization in our CDI
+mouse model.\\
+Beyond microbiome differences, factors such as age, body mass index,
+comorbidities, and use of other medications in human studies may also be
+contributing to the association between PPIs and CDI incidence or
+recurrence. This study addressed the impact of PPIs with or without
+antibiotics on a murine model of CDI, and found PPIs did not promote
+\emph{C. difficile} colonization. The epidemiological evidence linking
+PPIs to CDIs is primarily from observational studies, which makes
+determining causality and whether other risk factors play a role
+challenging (20). Future studies are needed to determine whether age,
+other comorbidities and bacterial strains that are less common in mice
+can increase the risk of CDIs or recurrent CDIs when combined with PPI
+treatment.
 
@@
 discussions throughout the process of designing the experiment,
-analyzing the results and drafting of the manuscript.
+analyzing the results, crafting the figures, and drafting of the
+manuscript.
 
@@
 mice was spit between 2 cages with 2-3 mice housed per cage and male and
-female mice were housed separately. All animal experiments were approved
-by the University of Michigan Animal Care and Use Committee (IACUC)
-under protocol number PRO00006983.
+female mice housed separately. All animal experiments were approved by
+the University of Michigan Animal Care and Use Committee (IACUC) under
+protocol number PRO00006983.
 
@@
 Powersoil HTP 96 Kit (Qiagen) and an EpMotion 5075 automated pipetting
-system (Eppendorf). The V4 hypervariable region of the 16S rRNA gene was
-amplified with Accuprime Pfx DNA polymerase (Thermo Fisher Scientific)
-using previously described custom barcoded primers (17). The 16S rRNA
-amplicon library was sequenced with the MiSeq (Illumina). Amplicons were
-cleaned up and normalized with the SequalPrep Normalization Plate Kit
-(ThermoFisher Scientific) and pooled amplicons were quantified with the
-KAPA library quantification kit (KAPA Biosystems).
+system (Eppendorf). The ZymoBIOMICS\textsuperscript{TM} Microbial
+Community DNA Standard (Zymo, CA, USA) was used as a mock community (21)
+and water was used as a negative control. The V4 hypervariable region of
+the 16S rRNA gene was amplified with Accuprime Pfx DNA polymerase
+(Thermo Fisher Scientific) using previously described custom barcoded
+primers (22). The 16S rRNA amplicon library was sequenced with the MiSeq
+(Illumina). Amplicons were cleaned up and normalized with the SequalPrep
+Normalization Plate Kit (ThermoFisher Scientific) and pooled amplicons
+were quantified with the KAPA library quantification kit (KAPA
+Biosystems).
 
 \textbf{16S rRNA gene sequence analysis}Mothur (v1.40.5) was used for
-all sequence processing steps (18) using the previously published
-protocol (17). R (v.3.5.1) was used to generate figures and perform
-statistical analysis.
+all sequence processing steps (23) using the previously published
+protocol (22). In brief, forward and reverse reads for each sample were
+combined and low-quality sequences and chimeras were removed. Duplicate
+sequences were merged, before taxonomy assignment using a modified
+version (v16) of the Ribosomal Database Project reference database
+(v11.5) with an 80\% cutoff. Operational taxonomic units (OTUs) were
+assigned with the opticlust clustering algorithm using a 97\% similarity
+threshold. To adjust for uneven sequencing across samples, all samples
+were rarefied to 3000 sequences, 1,000 times. PCoAs were generated based
+on Bray-Curtis distance. R (v.3.5.1) was used to generate figures and
+perform statistical analysis.
 
-\textbf{Statistical Analysis} To assess for differences in relative
-abundance in families and genera across our 3 different treatment groups
-(Clindamycin, Clindamycin + PPI, and PPI), we used a Kruskal-Wallis test
-with a Benjamini-Hochberg correction for multiple comparisons.
+\textbf{Statistical Analysis} To test for differences in relative
+abundances in families and genera across our 3 different treatment
+groups (Clindamycin, Clindamycin + PPI, and PPI) or within the PPI
+treatment group across 3 timepoints (Day -7, 0, and 9), we used a
+Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple
+comparisons.
 
@@
 \textbf{Data availability} The 16S rRNA sequencing data have been
-deposited in the NCBI Sequence Read Archive (Accession no.
-\_\_\_\_\_\_\_\_\_)
+deposited in the NCBI Sequence Read Archive (Accession no. PRJNA554866).
 
@@
 
+\hypertarget{ref-maier2018extensive}{}
+7. \textbf{Maier L}, \textbf{Pruteanu M}, \textbf{Kuhn M},
+\textbf{Zeller G}, \textbf{Telzerow A}, \textbf{Anderson EE},
+\textbf{Brochado AR}, \textbf{Fernandez KC}, \textbf{Dose H},
+\textbf{Mori H}, \textbf{others}. 2018. Extensive impact of
+non-antibiotic drugs on human gut bacteria. Nature \textbf{555}:623.
+
+\hypertarget{ref-Schuberte01021-14}{}
+8. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
+\textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
+\textbf{Aronoff DM}, \textbf{Schloss PD}. 2014. Microbiome data
+distinguish patients with clostridium difficile infection and non-c.
+difficile-associated diarrhea from healthy controls. mBio \textbf{5}.
+doi:\href{https://doi.org/10.1128/mBio.01021-14}{10.1128/mBio.01021-14}.
+
 \hypertarget{ref-Jenior2018}{}
-7. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+9. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
 \textbf{Schloss PD}. 2018. Clostridium difficile alters the structure
@@
 \hypertarget{ref-Yasutomi2018}{}
-8. \textbf{Yasutomi E}, \textbf{Hoshi N}, \textbf{Adachi S},
+10. \textbf{Yasutomi E}, \textbf{Hoshi N}, \textbf{Adachi S},
 \textbf{Otsuka T}, \textbf{Kong L}, \textbf{Ku Y}, \textbf{Yamairi H},
@@
 \hypertarget{ref-Freedberg2015}{}
-9. \textbf{Freedberg DE}, \textbf{Toussaint NC}, \textbf{Chen SP},
+11. \textbf{Freedberg DE}, \textbf{Toussaint NC}, \textbf{Chen SP},
 \textbf{Ratner AJ}, \textbf{Whittier S}, \textbf{Wang TC}, \textbf{Wang
@@
 \hypertarget{ref-Imhann2015}{}
-10. \textbf{Imhann F}, \textbf{Bonder MJ}, \textbf{Vila AV}, \textbf{Fu
+12. \textbf{Imhann F}, \textbf{Bonder MJ}, \textbf{Vila AV}, \textbf{Fu
 J}, \textbf{Mujagic Z}, \textbf{Vork L}, \textbf{Tigchelaar EF},
@@
 
-\hypertarget{ref-LlornsRico2019}{}
-11. \textbf{Lloréns-Rico V}, \textbf{Raes J}. 2019. Tracking humans and
-microbes. Nature \textbf{569}:632--633.
-doi:\href{https://doi.org/10.1038/d41586-019-01591-y}{10.1038/d41586-019-01591-y}.
-
 \hypertarget{ref-Kaur2007}{}
-12. \textbf{Kaur S}, \textbf{Vaishnavi C}, \textbf{Prasad KK},
+13. \textbf{Kaur S}, \textbf{Vaishnavi C}, \textbf{Prasad KK},
 \textbf{Ray P}, \textbf{Kochhar R}. 2007. Comparative role of antibiotic
@@
 \hypertarget{ref-kaur2011effect}{}
-13. \textbf{Kaur S}, \textbf{Vaishnavi C}, \textbf{Prasad KK},
+14. \textbf{Kaur S}, \textbf{Vaishnavi C}, \textbf{Prasad KK},
 \textbf{Ray P}, \textbf{Kochhar R}. 2011. Effect of lactobacillus
@@
 \hypertarget{ref-hung2015proton}{}
-14. \textbf{Hung Y-P}, \textbf{Ko W-C}, \textbf{Chou P-H}, \textbf{Chen
+15. \textbf{Hung Y-P}, \textbf{Ko W-C}, \textbf{Chou P-H}, \textbf{Chen
 Y-H}, \textbf{Lin H-J}, \textbf{Liu Y-H}, \textbf{Tsai H-W}, \textbf{Lee
@@
 \hypertarget{ref-Jenior2017}{}
-15. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+16. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
 \textbf{Schloss PD}. 2017. Clostridium difficile colonizes alternative
@@
 
-\hypertarget{ref-Maseda2019}{}
-16. \textbf{Maseda D}, \textbf{Zackular JP}, \textbf{Trindade B},
-\textbf{Kirk L}, \textbf{Roxas JL}, \textbf{Rogers LM},
-\textbf{Washington MK}, \textbf{Du L}, \textbf{Koyama T},
-\textbf{Viswanathan VK}, \textbf{Vedantam G}, \textbf{Schloss PD},
-\textbf{Crofford LJ}, \textbf{Skaar EP}, \textbf{Aronoff DM}. 2019.
-Nonsteroidal anti-inflammatory drugs alter the microbiota and exacerbate
-clostridium difficile colitis while dysregulating the inflammatory
-response. mBio \textbf{10}.
-doi:\href{https://doi.org/10.1128/mbio.02282-18}{10.1128/mbio.02282-18}.
+\hypertarget{ref-Reeves2011}{}
+17. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
+\textbf{Huffnagle GB}, \textbf{Schloss PD}, \textbf{Young VB}. 2011. The
+interplay between microbiome dynamics and pathogen dynamics in a murine
+model ofClostridium difficileInfection. Gut Microbes
+\textbf{2}:145--158.
+doi:\href{https://doi.org/10.4161/gmic.2.3.16333}{10.4161/gmic.2.3.16333}.
+
+\hypertarget{ref-Rausch2016}{}
+18. \textbf{Rausch P}, \textbf{Basic M}, \textbf{Batra A},
+\textbf{Bischoff SC}, \textbf{Blaut M}, \textbf{Clavel T},
+\textbf{Gläsner J}, \textbf{Gopalakrishnan S}, \textbf{Grassl GA},
+\textbf{Günther C}, \textbf{Haller D}, \textbf{Hirose M},
+\textbf{Ibrahim S}, \textbf{Loh G}, \textbf{Mattner J}, \textbf{Nagel
+S}, \textbf{Pabst O}, \textbf{Schmidt F}, \textbf{Siegmund B},
+\textbf{Strowig T}, \textbf{Volynets V}, \textbf{Wirtz S},
+\textbf{Zeissig S}, \textbf{Zeissig Y}, \textbf{Bleich A},
+\textbf{Baines JF}. 2016. Analysis of factors contributing to variation
+in the c57bl/6J fecal microbiota across german animal facilities.
+International Journal of Medical Microbiology \textbf{306}:343--355.
+doi:\href{https://doi.org/10.1016/j.ijmm.2016.03.004}{10.1016/j.ijmm.2016.03.004}.
+
+\hypertarget{ref-Robertson2019}{}
+19. \textbf{Robertson SJ}, \textbf{Lemire P}, \textbf{Maughan H},
+\textbf{Goethel A}, \textbf{Turpin W}, \textbf{Bedrani L},
+\textbf{Guttman DS}, \textbf{Croitoru K}, \textbf{Girardin SE},
+\textbf{Philpott DJ}. 2019. Comparison of co-housing and littermate
+methods for microbiota standardization in mouse models. Cell Reports
+\textbf{27}:1910--1919.e2.
+doi:\href{https://doi.org/10.1016/j.celrep.2019.04.023}{10.1016/j.celrep.2019.04.023}.
+
+\hypertarget{ref-Eze2017}{}
+20. \textbf{Eze P}, \textbf{Balsells E}, \textbf{Kyaw MH}, \textbf{Nair
+H}. 2017. Risk factors for clostridium difficile infections an overview
+of the evidence base and challenges in data synthesis. Journal of Global
+Health \textbf{7}.
+doi:\href{https://doi.org/10.7189/jogh.07.010417}{10.7189/jogh.07.010417}.
+
+\hypertarget{ref-Sze2019}{}
+21. \textbf{Sze MA}, \textbf{Schloss PD}. 2019. The impact of DNA
+polymerase and number of rounds of amplification in PCR on 16S rRNA gene
+sequence data. mSphere \textbf{4}.
+doi:\href{https://doi.org/10.1128/msphere.00163-19}{10.1128/msphere.00163-19}.
 
 \hypertarget{ref-Kozich2013}{}
-17. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+22. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \hypertarget{ref-Schloss2009}{}
-18. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+23. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \includegraphics{figure_1.pdf} \textbf{Figure 1. PPI treatment had
-minimal impact on the murine fecal microbiota} A. Mouse experiment
-timeline and logistics. Stools for 16S rRNA sequencing analysis were
-collected on the days that are labeled (Day -7, -5, -3, -1, 0, 1, 2, 3,
-4, 5, 7, 9). B. Principal Coordinates Analaysis (PCoA) of Bray-curtis
+minimal impact on the murine fecal microbiota.} A. Mouse experiment
+timeline and logistics. The PPI omeprazole was administered throughout
+the duration of the experiment. Clindamycin was administered 1 day
+before \emph{C. difficile} challenge on Day 0. Stools for 16S rRNA
+sequencing analysis were collected on the days that are labeled (Day -7,
+-5, -3, -1, 0, 1, 2, 3, 4, 5, 7, 9). \emph{C. difficile} CFU in the
+stool was quantified daily through 6 days post-infection by anaerobic
+culture. B. Principal Coordinates Analaysis (PCoA) of Bray-curtis
 distances from stool samples for all treatment groups during the intial
-7 days of treatment with the PPI omeprazole or vehicle cluster together
-and do not chage much over time. C. Relative abundances of families
-previously associated with PPI use in humans do not change much over the
-16-day course of treatment with PPIs (7 days before \emph{C. difficile}
-challenge through 9 days post \emph{C. difficile} challenge). D. PCoA of
-stool samples collected before and 1 day after antibiotic treatment,
-demonstrating the fecal microbiota of mice treated with clindamycin
-starts to separate from the rest of the samples.
+7 days of the experiment. Each symbol represents a stool sample from an
+individual mouse on a given day. Transparency of the symbol corresponds
+to treatment day. Circles represent samples before antibiotic treatment
+(administered on Day -1), while open diamonds represent samples from the
+day after antibiotic treatment (Day 0). Mice treated with the PPI
+omeprazole or vehicle cluster together and do not change much over time
+while mice that receive clindamycin have samples that start to separate
+from the rest of the samples the day after antibiotic treatment. C.
+Relative abundances of families previously associated with PPI use in
+humans do not change much over the 16-day course of treatment with PPIs
+(7 days before \emph{C. difficile} challenge through 9 days post
+\emph{C. difficile} challenge). D. Boxplots showing the relative
+abundances of 7 familes within PPI-treated mice that had the lowest
+adjusted P values after comparing across 3 timepoints. There were no
+significant differences acoss time for any of the identified families
+(All P \textgreater{} 0.05) after analysis by Kruskal-Wallis test with a
+Benjamini-Hochberg correction for multiple comparisons. See Fig. S1C for
+genera with lowest adjusted P values (All P \textgreater{} 0.05). For
+C-D, the grey line indicates the limit of detection.
 
@@
 \includegraphics{figure_2.pdf} \textbf{Figure 2. PPI treatment alone
-does not promote CDIs in mice} A. C. difficile CFUs/g stool measured
-each day post C. difficile challenge for clindamycin, clindamycin/PPI,
-and PPI-treated mice. Lines represent the mean CFU/g for each treatment
-group while points represent CFU/g for individual mice within each
-group. B. PCoA of stool samples collected after antibiotic treatments.
-C. Genera significantly associated with treatment groups for samples
-across all sequenced timepoints. D. Families significantly associated
-with treatment groups for samples across all sequenced timepoints. C,D
-data were analyzed by Kruskal-Wallis test with a Benjamini-Hochberg
-correction for multiple comparisons.
+does not promote CDIs in mice.} A. \emph{C. difficile} CFUs/g stool
+measured each day post \emph{C. difficile} challenge for clindamycin,
+clindamycin/PPI, and PPI-treated mice. Lines represent the mean CFU/g
+for each treatment group while points represent CFU/g for individual
+mice within each group. The black dashed line indicates the limit of
+detection. B. PCoA of of Bray-curtis distances from stool samples
+collected after antibiotic treatment (last 9 days of the experiment).
+Transparency of the symbol corresponds to treatment day. Symbols
+represent the \emph{C. difficile} colonization status of the mice
+measured 2 days post-infection. Circles represent resistant mice
+(\emph{C. difficile} was undetectable in stool samples), while X-shapes
+represent mice that were colonized with \emph{C. difficile}, although
+all mice cleared \emph{C. difficile} within 5 days of infection.
+PPI-treated fecal samples primarily cluster together throughout the
+experiment. C. Genera significantly associated with treatment groups of
+stool samples across all sequenced timepoints. D. Families significantly
+associated with treatment groups of stool samples across all sequenced
+timepoints. C,D data were analyzed by Kruskal-Wallis test with a
+Benjamini-Hochberg correction for multiple comparisons. For C-D, the
+grey line indicates the limit of detection.
 
@@
 
-\includegraphics{figure_s1.pdf} \textbf{Figure S1. Previous human
-PPI-associated families fluctuate over time with no overall trend in
-either direction} Relative abundance over time for Lactobacillaceae (A)
-and Ruminococcaceae (B), 2 of the PPI-associated familes from human PPI
+\includegraphics{figure_s1.pdf} \textbf{Figure S1. Taxa within PPI
+treated mice fluctuate over time with no overall trend in either
+direction.} Relative abundance over time for Lactobacillaceae (A) and
+Ruminococcaceae (B), 2 of the PPI-associated familes from human PPI
 studies across all 3 treatment groups. Each point represents the
-relative abundance for an individual mouse
+relative abundance for an individual mouse, while the lines represent
+the mean relative abundances for each treatment group of mice. C.
+Boxplots showing the relative abundances of 13 genera within PPI-treated
+mice that had the lowest adjusted P values after comparing across 3
+timepoints. There were no significant differences acoss time for any of
+the identified families (All P = 0.568) after analysis by Kruskal-Wallis
+test with a Benjamini-Hochberg correction for multiple comparisons. For
+A-C, the grey lines indicate the limit of detection.
+
+\newpage
+
+\includegraphics{figure_s2.pdf} \textbf{Figure S2. Microbiota diversity
+and richness decrease with antibiotic treatment but remain relatively
+constant with PPI treatment.} A. Boxplots of the Shannon Index (A) and
+richness (B) values for each group of mice over 3 timepoints (Day -7, 0,
+and 9). Each circle represents the value for an individual mouse.
 

### 2019-08-07 submission/manuscript.Rmd
Changes in response to Pat's comments & edits

@@
 output:
+  word_document: default
   pdf_document:
@@
     keep_tex: yes
-  word_document: default
 ---
@@
 
-# Proton pump inhibitor administration does not promote *Clostridium difficile* colonization in a murine model
+# The proton pump inhibitor omeprazole does not promote *Clostridium difficile* colonization in a murine model
 
@@
 Proton pump inhibitor (PPI) use has been associated with microbiota alterations and susceptibility to *Clostridium difficile* infections (CDIs) in humans.
-We assessed how PPI treatment alters the fecal microbiota and whether PPIs promote CDIs in a mouse model.
-Mice receiving the PPI treatment were gavaged with 40 mg/kg of omeprazole during a 7-day pretreatment phase, the day of *C.
+We assessed how PPI treatment alters the fecal microbiota and whether treatment promotes CDIs in a mouse model.
+Mice receiving a PPI treatment were gavaged with 40 mg/kg of omeprazole during a 7-day pretreatment phase, the day of *C.
 difficile* challenge, and the following 9 days.
@@
 difficile* colonization, while the other cage was colonized.
-Treatming mice with clindamycin and no omeprazole followed by challenge resulted in  *C.
+Treating mice with only clindamycin followed by challenge resulted in  *C.
 difficile* colonization.
-16S rRNA gene sequencing analysis revealed that omeprazole had minimal impact on the structure of the murine microbiota throughout the 16 days of PPI exposure.
-These results suggest PPI treatment alone is not sufficient to disrupt microbiota resistance to *C.
+16S rRNA gene sequencing analysis revealed that omeprazole had minimal impact on the structure of the murine microbiota throughout the 16 days of omeprazole exposure.
+These results suggest omeprazole treatment alone is not sufficient to disrupt microbiota resistance to *C.
 difficle* infection in mice that are normally resistant in the absence of antibiotic treatment.
@@
 ## Importance
-Antibiotics are a major risk factor for *Clostridium difficile* infections (CDIs), but other factors may also contribute.
+Antibiotics are the primary risk factor for *Clostridium difficile* infections (CDIs), but other factors may also increase a person's risk.
 In epidemiological studies, PPI use has been associated with CDI incidence and recurrence.
@@
 difficile*.
-We found the PPI treatment had minimal impact on the murine fecal microbiota and did not promote *C.
+We found omeprazole treatment had minimal impact on the murine fecal microbiota and did not promote *C.
 difficile* colonization.
@@
 Multiple epidemiological studies have suggested an association between proton pump inhibitor (PPI) use and incidence or recurrence of CDIs [@tariq2017association; @nehra2018proton; @Naito2018; @Elias2019].
-There have also been a number of large cohort studies as well as interventional clinical trials that have demonstrated that specific alterations in the intestinal microbiome were associated with PPI use [@Imhann2017; @Naito2018].
+There have also been a number of large cohort studies and interventional clinical trials that demonstrated specific alterations in the intestinal microbiome were associated with PPI use [@Imhann2017; @Naito2018].
 PPI-associated microbiota changes have been attributed to the ability of PPIs to increase stomach acid pH which may promote the survival of oral and pathogenic bacteria [@Imhann2017; @Naito2018].
-Human fecal microbiota changes with PPI use include increases in Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae and decreases in Ruminococcaceae [@Imhann2017; @maier2018extensive].
-Several of these taxa have been associated with C.
-difficile colonization in humans [@Schuberte01021-14]
+In human fecal samples, PPI use results in increases in Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae and decreases in Ruminococcaceae [@Imhann2015; @Imhann2017; @Freedberg2015; @maier2018extensive].
+Several of these taxa have also been associated with *C.
+difficile* colonization in humans [@Schuberte01021-14].
 Unfortunately, most of the studies suggesting a link between PPIs and *C.
-difficile* were retrospective and did not evaluate the microbiome [@tariq2017association; @nehra2018proton; @Elias2019].
+difficile* were retrospective and did not evaluate changes in the microbiome [@tariq2017association; @nehra2018proton; @Elias2019].
 Thus, it is unclear whether the gastrointestinal microbiome changes associated with PPI use explain the association between PPIs and CDIs.
 Additionally, epidemiological studies have a limited capacity to address potential confounders and comorbidities in patients that were on PPIs and developed CDIs or recurrent CDIs [@Elias2019; @tariq2017association].
-Here, we evaluated the impact of a daily high dose PPI treatment on the murine microbiome and susceptibility to *C.
+Here, we evaluated the impact of daily PPI treatment with omeprazole on the murine microbiome and susceptibility to *C.
 difficile* colonization in relation to clindamycin, an antibiotic that perturbs the microbiome enough to allow *C.
@@
 
-**Murine fecal microbiomes were minimally affected by PPI treatment ** To test whether PPI treatment alters the microbiome and promotes susceptibility to CDIs, we gavaged mice with 40 mg/kg of omeprazole for 7 days before *C.
+**Murine fecal microbiomes were minimally affected by omeprazole treatment ** To test whether omeprazole treatment alters the microbiome and promotes susceptibility to CDIs, we gavaged mice with 40 mg/kg of omeprazole for 7 days before *C.
 difficile* challenge (Figure 1A).
-A principle coordinates analysis (PCoA) of the Bray-Curtis distances over the initial 7 days of treatment revealed the bacterial communities of PPI-treated mice remained relatively unchanged relative to animals that did not recieve the PPI (Figure 1B).
-We observed no significant fluctuations in the relative abundance of those taxa previously shown to respond to PPI treatment throughout the course of the 16-day experiment (Figure 1C, Figure S1A-B).
-We also observed no significant fluctuations in relative abundances at the family and genus level over the beginning, day of *C.
-difficile* challenge, and end of the experiment for the PPI-treated mice (Figure 1D, Figure S1C).
-Next, the fecal bacterial communities of PPI-treated mice were compared to a group of mice that received a single 10 mg/kg dose of clindamycin and a group that received both PPI and clindamycin.
-In contrast to the dynamics of mice only receiving omeprazole, the microbiomes from the 2 groups of mice that received clindamycin started to shift away from the PPI-treated mice and the samples from previous days one day after treatment (Figure 1D).
-These results demonstrated that the PPI treatment alone had a minimal impact on the murine fecal bacterial community relative to the impact of clindamycin treatment.
+A principle coordinates analysis (PCoA) of the Bray-Curtis distances over the initial 7 days of treatment revealed the bacterial communities of omeprazole-treated mice remained relatively unchanged (Figure 1B).
+We observed no significant fluctuations in the relative abundance of those taxa previously shown to respond to PPI treatment throughout the course of the 16-day experiment (Figure 1C-D, S1).
+We also observed no significant changes in relative abundances at the family and genus level over the course of the experiment for the omeprazole-treated mice (data not shown).
+These results demonstrated that the omeprazole treatment alone had a minimal impact on the murine fecal bacterial community after 7 days of pretreatment.
 
-**PPI-treatment did not promote susceptibility to *C.
-difficile infection* in mice** Next, we examined whether PPI treatment altered susceptibility to *C.
+**Omeprazole treatment did not promote susceptibility to *C.
+difficile infection* in mice** Next, we examined whether omeprazole treatment altered susceptibility to *C.
 difficile* infection in mice.
-After PPI treatment or clindamycin treatment, mice were challenged with *C.
+After omeprazole treatment or clindamycin treatment, mice were challenged with 10^3^ *C.
 difficile* 630 spores.
 While *C.
-difficile* colonized the clindamycin treated mice, it was unable to colonize the PPI-treated mice (Figure 2A).
-Interestingly, only 1 cage of mice that received both the PPI and clindamycin were colonized, while the mice in other cage wwere resistant (Figure 2A).
-The greatest shifts in bacterial communities occurred in the clindamycin-treated mice (Figure 2B, Figure S2).
+difficile* colonized the clindamycin-treated mice, it did not colonize the omeprazole-treated mice (Figure 2A).
+Interestingly, only 1 cage of mice that received both omeprazole and clindamycin were colonized, while the other cage of mice were resistant (Figure 2A).
+The greatest shifts in bacterial communities occurred in the clindamycin-treated mice (Figure 2B, S2).
 Regardless of whether the mice became colonized, all of the mice had cleared *C.
 difficile* within 5 days (Figure 2A), suggesting that omeprazole did not affect the rate of clearance.
-Our results suggest that PPI treatment had no effects on bacterial community resistance to *C.
+Our results suggest that omeprazole treatment had no effect on bacterial community resistance to *C.
 difficile* colonization in mice.
-Instead most of the differences between our 3 treatment groups appear to be driven by clindamycin administration (Figure 2C-D) and included previously described decreases in *Alistipes*, *Barnesiella*, Porhyromonadaceae, Ruminococcaceae [@Schubert2015].
-These findings demonstrated that high dose PPI treatment did not promote susceptibility to *C.
+Instead most of the differences between the 3 treatment groups appeared to be driven by clindamycin administration (Figure 2C, S2).
+These findings demonstrated that high dose omeprazole treatment did not promote susceptibility to *C.
 difficile* colonization.
@@
 difficile* infection in mice.
-Our findings that PPI treatment had minimal impact on the fecal microbiome were comparable to another PPI mouse study that indicated PPIs had more of an effect on the small intestinal microbiota compared to the fecal microbiota [@Yasutomi2018].
-The same group demonstrated PPI treatment increased the stomach pH in mice [@Yasutomi2018], which may improve survival of bacteria passing through the stomach.
+Our findings that omeprazole treatment had minimal impact on the fecal microbiome were comparable to another PPI mouse study that indicated the PPI lansoprazole had more of an effect on the small intestinal microbiota compared to the fecal microbiota [@Yasutomi2018].
+The same group demonstrated lansoprazole treatment increased the stomach pH in mice [@Yasutomi2018], which may improve survival of bacteria passing through the stomach.
 We did not find significant changes for the taxa observed to be significantly impacted by PPI use in human studies.
 However, 3 of the human-associated taxa were absent or at low abundance in our mice.
-Additionally, some of the significance of PPI associations in human interventional trials appears to be driven by a handful of specific taxa with overall differences on a PCoA plot difficult to distinguish [@Freedberg2015, @Imhann2015].
-The PPI-treated mice fecal communities did not change overall and no differences in the relative abundances of individual families or genera were observed over time.
-While our fecal microbiota findings are comparable to what's been shown in another mouse study, whether PPI-induced changes in specific bacterial abundances observed in humans play a role in CDIs remains to be determined.
-Although there have been a few *C.
-difficile* mouse model studies that have demonstrated PPIs have some effect on CDIs with or without additional antibiotic treatment [@Kaur2007; @kaur2011effect; @hung2015proton], there were some key methodological differences between these studies and our own.
+Although our fecal microbiota findings are comparable to what has been shown in another mouse study [@Yasutomi2018], whether PPI-induced changes in specific bacterial abundances observed in humans play a role in CDIs remains to be determined.
+While several *C.
+difficile* mouse model studies have shown that PPIs have an effect on CDIs with or without additional antibiotic treatment [@Kaur2007; @kaur2011effect; @hung2015proton], there were insufficient controls to attribute the effect solely to PPI treatment.
 One group administered 0.5 mg/kg of the PPI lansoprazole daily for 2 weeks to mice and then challenged with *C.
 difficile* demonstrated that PPI treatment alone resulted in detectable *C.
-difficile* in the stool 1 week after challenge, but also showed there was detectable *C.
+difficile* in the stool 1 week after challenge, however there was detectable *C.
 difficile* in mice not treated with antibiotics [@Kaur2007; @kaur2011effect].
-The presence of *C.
-difficile* in mice not treated with antibiotics may be partly attributed to the higher dose of vegetative cells (10^8^ CFU) used, but nevertheless makes it difficult to interpret the extent to which PPI treatment contributed to *C.
-difficile* colonization in their mouse model [@Kaur2007; @kaur2011effect].
-In contrast, we have previously shown mice from our colony that were not given antibiotics were resistant to *C.
-difficile* 630 when challenged with 10^3^ spores [@Jenior2017] and previous work by our collaborators demonstrated mice not given antibiotics were resistant to 10^5^ vegetative cells of *C.
-difficile * VPI  10463 [@Reeves2011].
-The other PPI CDI mouse study also used a higher dose of 10^6^ spores or 10^7^ vegetative cells to challenge antibiotic-treated mice or mice treated with both antibiotics and the PPI esomeprazole (40mg/kg dose) for 2 days and demonstrated the antibiotic/PPI-treated mice developed more severe CDIs [@hung2015proton].
-We tested the same high 40 mg/kg PPI dose and expanded pre-treatment to 7 days before challenge to test the impact of PPI treatment alone on our CDI mouse model.
-Our study extended the previous work examining PPIs and *C.
-difficile* in mice by evaluating the contribution of the intestinal microbiota.
-We found the PPI omeprazole had no significant impact on bacterial taxa within the murine intestinal microbiota over the 16-day experiment and in contrast to previous work, PPIs did not alter *C.
+The other mouse study demonstrated the antibiotic/PPI (esomeprazole)-treated mice developed more severe CDIs compared to antibiotic-treated mice, but did not have a group treated with just esomeprazole for comparison [@hung2015proton].
+We tested the same high 40 mg/kg PPI dose and expanded pre-treatment to 7 days before challenge to test the impact of omeprazole treatment alone on our CDI mouse model.
+Additionally, we've previously demonstrated that mice from our colony were resistant to *C.
+difficile* 630 colonization without antibiotic treatment [@Jenior2017], ensuring there was not already partial susceptiblity to *C.
+difficile* before treatment.
+The additional controls in our study allowed us to assess the contibution of omeprazole alone to *C.
+difficile* susceptibility in mice.
+Our study also extended previous work examining PPIs and *C.
+difficile* in mice by incorporating the contribution of the intestinal microbiota.
+We found the PPI omeprazole had no significant impact on bacterial taxa within the murine intestinal microbiota over the 16-day experiment.
+In contrast to previous work, omeprazole did not alter *C.
 difficile* colonization resistance.
@@
 difficile* colonization in our CDI mouse model.
+Gastrointestinal physiological differences, particularly the higher stomach pH in mice (pH 3-4) compared to humans (ph 1) [@Hugenholtz2017] could also explain why omeprazole had a limited impact on the murine microbiome.
+The microbiota and physiological differences between humans and mice may limit the usefulness of employing mouse models to study the impact of PPIs on the microbiota and CDIs.
 Beyond microbiome differences, factors such as age, body mass index, comorbidities, and use of other medications in human studies may also be contributing to the association between PPIs and CDI incidence or recurrence.
@@
 
-**Animals** All mouse experiments were performed with 7- to 12-week-old C57BL/6 male and female mice.
+**Animals.** All mouse experiments were performed with 7- to 12-week-old C57BL/6 male and female mice.
 Each experimental group of mice was spit between 2 cages with 2-3 mice housed per cage and male and female mice housed separately.
@@
 
-**Drug treatments**
+**Drug treatments.**
 Omeprazole (Sigma Aldrich) was prepared in a vehicle solution of 40% polyethylene glycol 400 (Sigma-Aldrich) in phosphate buffered saline.
-Omeprazole was prepared from 20 mg/mL frozen aliquots and diluted to an 8mg/mL prior to gavage.
-All mice received 40 mg/kg omeprazole or vehicle solution once per day through the duration of the experiment with treatment starting 7 days before *C.
+Omeprazole was prepared from 20 mg/mL frozen aliquots and diluted to an 8 mg/mL prior to gavage.
+All mice received 40 mg/kg omeprazole (a dose previously used in mouse experiments [@hung2015proton]) or vehicle solution once per day through the duration of the experiment with treatment starting 7 days before *C.
 difficile* challenge (Figure 1A).
 One day prior to *C.
-difficile* challenge, 2 groups of mice received an intraperitoneal injection of 10 mg/kg clindamycin or sterile saline vehicle.
+difficile* challenge, 2 groups of mice received an intraperitoneal injection of 10 mg/kg clindamycin or sterile saline vehicle [@Jenior2018].
 All drugs were filter sterilized through a 0.22 micron syringe filter before administration to animals.
 
-***C. difficile* infection model**
+***C. difficile* infection model.**
 Mice were challenged with *C.
 difficile* 630 seven days after the start of omeprazole treatment and one day after clindamycin treatment.
-Mice were challenged with 10^3^ spores in ultrapure distilled water.
+Mice were challenged with 10^3^ spores in ultrapure distilled water as described previously [@Jenior2018].
 Stool samples were collected for 16S rRNA sequencing or *C.
@@
 
-**16S rRNA gene sequencing** DNA for 16S rRNA gene sequencing was extracted from 10-50 mg fecal pellet from each mouse using the DNeasy Powersoil HTP 96 Kit (Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf).
-The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used as a mock community [@Sze2019] and water was used as a negative control.
+**16S rRNA gene sequencing.** DNA for 16S rRNA gene sequencing was extracted from 10-50 mg fecal pellet from each mouse using the DNeasy Powersoil HTP 96 Kit (Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf).
+The 16S rRNA sequencing libary was prepared as described previously [@Kozich2013].
+In brief, the ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used as a mock community [@Sze2019] and water was used as a negative control.
 The V4 hypervariable region of the 16S rRNA gene was amplified with Accuprime Pfx DNA polymerase (Thermo Fisher Scientific) using previously described custom barcoded primers [@Kozich2013].
@@
 
-**16S rRNA gene sequence analysis**Mothur (v1.40.5) was used for all sequence processing steps [@Schloss2009] using the previously published protocol [@Kozich2013].
+**16S rRNA gene sequence analysis.** mothur (v1.40.5) was used for all sequence processing steps [@Schloss2009] using a previously published protocol [@Kozich2013].
 In brief, forward and reverse reads for each sample were combined and low-quality sequences and chimeras were removed.
@@
 Operational taxonomic units (OTUs) were assigned with the opticlust clustering algorithm using a 97% similarity threshold.
-To adjust for uneven sequencing across samples, all samples were rarefied to 3000 sequences, 1,000 times.
+To adjust for uneven sequencing across samples, all samples were rarefied to 3,000 sequences, 1,000 times.
 PCoAs were generated based on Bray-Curtis distance.
@@
 
-**Statistical Analysis** To test for differences in relative abundances in families and genera across our 3 different treatment groups (Clindamycin, Clindamycin + PPI, and PPI) or within the PPI treatment group across 3 timepoints (Day -7, 0, and 9), we used a Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
+**Statistical Analysis.** To test for differences in relative abundances in families and genera across our 3 different treatment groups (Clindamycin, Clindamycin + PPI, and PPI) or within the PPI treatment group across 3 timepoints (Day -7, 0, and 9), we used a Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
 
-**Code availability** The code for all sequence processing and analysis step as well as an Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Tomkovich_PPI_XXXX_2019.
+**Code availability.** The code for all sequence processing and analysis step as well as an Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Tomkovich_PPI_mSphere_2019.
 
-**Data availability** The 16S rRNA sequencing data have been deposited in the NCBI Sequence Read Archive (Accession no. PRJNA554866).
+**Data availability.** The 16S rRNA sequencing data have been deposited in the NCBI Sequence Read Archive (Accession no. PRJNA554866).
 
@@
 **Figure 1.
-PPI treatment had minimal impact on the murine fecal microbiota.** A.
+Omeprazole treatment had minimal impact on the murine fecal microbiota.** A.
 Mouse experiment timeline and logistics.
@@
 B.
-Principal Coordinates Analaysis (PCoA) of Bray-curtis distances from stool samples for all treatment groups during the intial 7 days of the experiment.
-Each symbol represents a stool sample from an individual mouse on a given day.
-Transparency of the symbol corresponds to treatment day.
-Circles represent samples before antibiotic treatment (administered on Day -1), while open diamonds represent samples from the day after antibiotic treatment (Day 0).
-Mice treated with the PPI omeprazole or vehicle cluster together and do not change much over time while mice that receive clindamycin have samples that start to separate from the rest of the samples the day after antibiotic treatment.
-C.
-Relative abundances of families previously associated with PPI use in humans do not change much over the 16-day course of treatment with PPIs (7 days before *C.
-difficile* challenge through 9 days post *C.
-difficile* challenge).
-D.
-Boxplots showing the relative abundances of 7 familes within PPI-treated mice that had the lowest adjusted P values after comparing across 3 timepoints.
-There were no significant differences acoss time for any of the identified families (All P > 0.05) after analysis by Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
-See Fig.
-S1C for genera with lowest adjusted P values (All P > 0.05).
-For C-D, the grey line indicates the limit of detection.
+Principal Coordinates Analaysis (PCoA) of Bray-curtis distances from stool samples of mice in the omeprazole treatment group during the intial 7 days of the experiment.
+Each color represents stool samples from the same mouse and lines connect sequentially collected samples.
+C-D.
+Relative abundances of families previously associated with PPI use in humans at the start of the experiment (C) and after 7 days of omeprazole treatment (D).
+Each circle represents an individual mouse.
+There were no significant differences acoss treatment groups for any of the identified families in the seqeunce data at day -7 and day 0, analyzed by Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
+For C-D, the grey vertical line indicates the limit of detection.
 
@@
 **Figure 2.
-PPI treatment alone does not promote CDIs in mice.** A.
+Omeprazole treatment alone does not promote CDIs in mice.** A.
 *C.
@@
 difficile* within 5 days of infection.
-PPI-treated fecal samples primarily cluster together throughout the experiment.
+Omeprazole treated fecal samples primarily cluster together throughout the experiment.
 C.
-Genera significantly associated with treatment groups of stool samples across all sequenced timepoints.
-D.
-Families significantly associated with treatment groups of stool samples across all sequenced timepoints.
-C,D data were analyzed by Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
-For C-D, the grey line indicates the limit of detection.
+Genera that vary the most across treatment groups for stool samples collected from mice 2 days post-infection.
+Data were analyzed by Kruskal-Wallis test, and no P values were significant after Benjamini-Hochberg correction for multiple comparisons.
+The grey vertical line indicates the limit of detection.
 
@@
 **Figure S1.
-Taxa within PPI treated mice fluctuate over time with no overall trend in either direction.** Relative abundance over time for Lactobacillaceae (A) and Ruminococcaceae (B), 2 of the PPI-associated familes from human PPI studies across all 3 treatment groups.
-Each point represents the relative abundance for an individual mouse, while the lines represent the mean relative abundances for each treatment group of mice.
-C.
-Boxplots showing the relative abundances of 13 genera within PPI-treated mice that had the lowest adjusted P values after comparing across 3 timepoints.
-There were no significant differences acoss time for any of the identified families (All P = 0.568) after analysis by Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
-For A-C, the grey lines indicate the limit of detection.
+Families within omeprazole treated mice fluctuate over time with no overall trend in either direction.** Relative abundance over time for Lactobacillaceae (A) and Ruminococcaceae (B), 2 of the PPI-associated familes from human PPI studies across all 3 treatment groups.
+Each point represents the relative abundance for an individual mouse stool sample, while the lines represent the mean relative abundances for each treatment group of mice.
+The grey horizontal lines indicate the limit of detection.
 
@@
 **Figure S2.
-Microbiota diversity and richness decrease with antibiotic treatment but remain relatively constant with PPI treatment.** A.
-Boxplots of the Shannon Index (A) and richness (B) values for each group of mice over 3 timepoints (Day -7, 0, and 9).
-Each circle represents the value for an individual mouse.
+Microbiota diversity and richness decrease with antibiotic treatment but remain relatively constant with omeprazole treatment.** A.
+Boxplots of the Shannon Diversity Index values (A) and number of observed OTUs (B) for each group of mice over 3 timepoints (Day -7, 0, and 9).
+Each circle represents the value for a stool sample from an individual mouse.

### 2019-08-07 submission/manuscript.tex
Changes in response to Pat's comments & edits

@@
 
-\section{\texorpdfstring{Proton pump inhibitor administration does not
+\section{\texorpdfstring{The proton pump inhibitor omeprazole does not
 promote \emph{Clostridium difficile} colonization in a murine
-model}{Proton pump inhibitor administration does not promote Clostridium difficile colonization in a murine model}}\label{proton-pump-inhibitor-administration-does-not-promote-clostridium-difficile-colonization-in-a-murine-model}
+model}{The proton pump inhibitor omeprazole does not promote Clostridium difficile colonization in a murine model}}\label{the-proton-pump-inhibitor-omeprazole-does-not-promote-clostridium-difficile-colonization-in-a-murine-model}
 
@@
 infections (CDIs) in humans. We assessed how PPI treatment alters the
-fecal microbiota and whether PPIs promote CDIs in a mouse model. Mice
-receiving the PPI treatment were gavaged with 40 mg/kg of omeprazole
+fecal microbiota and whether treatment promotes CDIs in a mouse model.
+Mice receiving a PPI treatment were gavaged with 40 mg/kg of omeprazole
 during a 7-day pretreatment phase, the day of \emph{C. difficile}
@@
 mice remained resistant to \emph{C. difficile} colonization, while the
-other cage was colonized. Treatming mice with clindamycin and no
-omeprazole followed by challenge resulted in \emph{C. difficile}
-colonization. 16S rRNA gene sequencing analysis revealed that omeprazole
-had minimal impact on the structure of the murine microbiota throughout
-the 16 days of PPI exposure. These results suggest PPI treatment alone
-is not sufficient to disrupt microbiota resistance to \emph{C. difficle}
+other cage was colonized. Treating mice with only clindamycin followed
+by challenge resulted in \emph{C. difficile} colonization. 16S rRNA gene
+sequencing analysis revealed that omeprazole had minimal impact on the
+structure of the murine microbiota throughout the 16 days of omeprazole
+exposure. These results suggest omeprazole treatment alone is not
+sufficient to disrupt microbiota resistance to \emph{C. difficle}
 infection in mice that are normally resistant in the absence of
@@
 
-Antibiotics are a major risk factor for \emph{Clostridium difficile}
-infections (CDIs), but other factors may also contribute. In
-epidemiological studies, PPI use has been associated with CDI incidence
-and recurrence. Proton pump inhibitors (PPIs) have also been associated
-with alterations in the human intestinal microbiota in observational and
-interventional studies. We evaluated the effects of the PPI omeprazole
-on the structure of the murine intestinal microbiota and its ability to
-disrupt colonization resistance to \emph{C. difficile}. We found the PPI
-treatment had minimal impact on the murine fecal microbiota and did not
-promote \emph{C. difficile} colonization. Further studies are needed to
-determine whether other factors such as the composition of the starting
-bacterial community, comorbidities, and use of additional medications
-contribute to the association between PPIs and CDIs seen in humans or
-whether aspects of murine physiology may limit its utility to test these
-types of hypotheses.
+Antibiotics are the primary risk factor for \emph{Clostridium difficile}
+infections (CDIs), but other factors may also increase a person's risk.
+In epidemiological studies, PPI use has been associated with CDI
+incidence and recurrence. Proton pump inhibitors (PPIs) have also been
+associated with alterations in the human intestinal microbiota in
+observational and interventional studies. We evaluated the effects of
+the PPI omeprazole on the structure of the murine intestinal microbiota
+and its ability to disrupt colonization resistance to \emph{C.
+difficile}. We found omeprazole treatment had minimal impact on the
+murine fecal microbiota and did not promote \emph{C. difficile}
+colonization. Further studies are needed to determine whether other
+factors such as the composition of the starting bacterial community,
+comorbidities, and use of additional medications contribute to the
+association between PPIs and CDIs seen in humans or whether aspects of
+murine physiology may limit its utility to test these types of
+hypotheses.
 
@@
 recurrence of CDIs (2--5). There have also been a number of large cohort
-studies as well as interventional clinical trials that have demonstrated
-that specific alterations in the intestinal microbiome were associated
-with PPI use (4, 6). PPI-associated microbiota changes have been
-attributed to the ability of PPIs to increase stomach acid pH which may
-promote the survival of oral and pathogenic bacteria (4, 6). Human fecal
-microbiota changes with PPI use include increases in Enterococcaceae,
-Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae
-and decreases in Ruminococcaceae (6, 7). Several of these taxa have been
-associated with C. difficile colonization in humans (8)\\
+studies and interventional clinical trials that demonstrated specific
+alterations in the intestinal microbiome were associated with PPI use
+(4, 6). PPI-associated microbiota changes have been attributed to the
+ability of PPIs to increase stomach acid pH which may promote the
+survival of oral and pathogenic bacteria (4, 6). In human fecal samples,
+PPI use results in increases in Enterococcaceae, Lactobacillaceae,
+Micrococcaceae, Staphylococcaceae and Streptococcaceae and decreases in
+Ruminococcaceae (6--9). Several of these taxa have also been associated
+with \emph{C. difficile} colonization in humans (10).\\
 Unfortunately, most of the studies suggesting a link between PPIs and
-\emph{C. difficile} were retrospective and did not evaluate the
-microbiome (2, 3, 5). Thus, it is unclear whether the gastrointestinal
-microbiome changes associated with PPI use explain the association
-between PPIs and CDIs. Additionally, epidemiological studies have a
-limited capacity to address potential confounders and comorbidities in
-patients that were on PPIs and developed CDIs or recurrent CDIs (2, 5).
-Here, we evaluated the impact of a daily high dose PPI treatment on the
-murine microbiome and susceptibility to \emph{C. difficile} colonization
-in relation to clindamycin, an antibiotic that perturbs the microbiome
-enough to allow \emph{C. difficile} to colonize but is mild enough that
-\emph{C. difficile} is cleared within 10 days (9).
+\emph{C. difficile} were retrospective and did not evaluate changes in
+the microbiome (2, 3, 5). Thus, it is unclear whether the
+gastrointestinal microbiome changes associated with PPI use explain the
+association between PPIs and CDIs. Additionally, epidemiological studies
+have a limited capacity to address potential confounders and
+comorbidities in patients that were on PPIs and developed CDIs or
+recurrent CDIs (2, 5). Here, we evaluated the impact of daily PPI
+treatment with omeprazole on the murine microbiome and susceptibility to
+\emph{C. difficile} colonization in relation to clindamycin, an
+antibiotic that perturbs the microbiome enough to allow \emph{C.
+difficile} to colonize but is mild enough that \emph{C. difficile} is
+cleared within 10 days (11).
 
-\textbf{Murine fecal microbiomes were minimally affected by PPI
-treatment } To test whether PPI treatment alters the microbiome and
-promotes susceptibility to CDIs, we gavaged mice with 40 mg/kg of
+\textbf{Murine fecal microbiomes were minimally affected by omeprazole
+treatment } To test whether omeprazole treatment alters the microbiome
+and promotes susceptibility to CDIs, we gavaged mice with 40 mg/kg of
 omeprazole for 7 days before \emph{C. difficile} challenge (Figure 1A).
@@
 over the initial 7 days of treatment revealed the bacterial communities
-of PPI-treated mice remained relatively unchanged relative to animals
-that did not recieve the PPI (Figure 1B). We observed no significant
-fluctuations in the relative abundance of those taxa previously shown to
-respond to PPI treatment throughout the course of the 16-day experiment
-(Figure 1C, Figure S1A-B). We also observed no significant fluctuations
-in relative abundances at the family and genus level over the beginning,
-day of \emph{C. difficile} challenge, and end of the experiment for the
-PPI-treated mice (Figure 1D, Figure S1C). Next, the fecal bacterial
-communities of PPI-treated mice were compared to a group of mice that
-received a single 10 mg/kg dose of clindamycin and a group that received
-both PPI and clindamycin. In contrast to the dynamics of mice only
-receiving omeprazole, the microbiomes from the 2 groups of mice that
-received clindamycin started to shift away from the PPI-treated mice and
-the samples from previous days one day after treatment (Figure 1D).
-These results demonstrated that the PPI treatment alone had a minimal
-impact on the murine fecal bacterial community relative to the impact of
-clindamycin treatment.
+of omeprazole-treated mice remained relatively unchanged (Figure 1B). We
+observed no significant fluctuations in the relative abundance of those
+taxa previously shown to respond to PPI treatment throughout the course
+of the 16-day experiment (Figure 1C-D, S1). We also observed no
+significant changes in relative abundances at the family and genus level
+over the course of the experiment for the omeprazole-treated mice (data
+not shown). These results demonstrated that the omeprazole treatment
+alone had a minimal impact on the murine fecal bacterial community after
+7 days of pretreatment.
 
-\textbf{PPI-treatment did not promote susceptibility to \emph{C.
-difficile infection} in mice} Next, we examined whether PPI treatment
-altered susceptibility to \emph{C. difficile} infection in mice. After
-PPI treatment or clindamycin treatment, mice were challenged with
-\emph{C. difficile} 630 spores. While \emph{C. difficile} colonized the
-clindamycin treated mice, it was unable to colonize the PPI-treated mice
-(Figure 2A). Interestingly, only 1 cage of mice that received both the
-PPI and clindamycin were colonized, while the mice in other cage wwere
-resistant (Figure 2A). The greatest shifts in bacterial communities
-occurred in the clindamycin-treated mice (Figure 2B, Figure S2).
-Regardless of whether the mice became colonized, all of the mice had
-cleared \emph{C. difficile} within 5 days (Figure 2A), suggesting that
-omeprazole did not affect the rate of clearance. Our results suggest
-that PPI treatment had no effects on bacterial community resistance to
-\emph{C. difficile} colonization in mice. Instead most of the
-differences between our 3 treatment groups appear to be driven by
-clindamycin administration (Figure 2C-D) and included previously
-described decreases in \emph{Alistipes}, \emph{Barnesiella},
-Porhyromonadaceae, Ruminococcaceae (1). These findings demonstrated that
-high dose PPI treatment did not promote susceptibility to \emph{C.
-difficile} colonization.
+\textbf{Omeprazole treatment did not promote susceptibility to \emph{C.
+difficile infection} in mice} Next, we examined whether omeprazole
+treatment altered susceptibility to \emph{C. difficile} infection in
+mice. After omeprazole treatment or clindamycin treatment, mice were
+challenged with 10\textsuperscript{3} \emph{C. difficile} 630 spores.
+While \emph{C. difficile} colonized the clindamycin-treated mice, it did
+not colonize the omeprazole-treated mice (Figure 2A). Interestingly,
+only 1 cage of mice that received both omeprazole and clindamycin were
+colonized, while the other cage of mice were resistant (Figure 2A). The
+greatest shifts in bacterial communities occurred in the
+clindamycin-treated mice (Figure 2B, S2). Regardless of whether the mice
+became colonized, all of the mice had cleared \emph{C. difficile} within
+5 days (Figure 2A), suggesting that omeprazole did not affect the rate
+of clearance. Our results suggest that omeprazole treatment had no
+effect on bacterial community resistance to \emph{C. difficile}
+colonization in mice. Instead most of the differences between the 3
+treatment groups appeared to be driven by clindamycin administration
+(Figure 2C, S2). These findings demonstrated that high dose omeprazole
+treatment did not promote susceptibility to \emph{C. difficile}
+colonization.
 
@@
 We found the PPI omeprazole did not impact the gut microbiota and did
-not promote \emph{C. difficile} infection in mice. Our findings that PPI
-treatment had minimal impact on the fecal microbiome were comparable to
-another PPI mouse study that indicated PPIs had more of an effect on the
-small intestinal microbiota compared to the fecal microbiota (10). The
-same group demonstrated PPI treatment increased the stomach pH in mice
-(10), which may improve survival of bacteria passing through the
-stomach. We did not find significant changes for the taxa observed to be
-significantly impacted by PPI use in human studies. However, 3 of the
-human-associated taxa were absent or at low abundance in our mice.
-Additionally, some of the significance of PPI associations in human
-interventional trials appears to be driven by a handful of specific taxa
-with overall differences on a PCoA plot difficult to distinguish (11).
-The PPI-treated mice fecal communities did not change overall and no
-differences in the relative abundances of individual families or genera
-were observed over time. While our fecal microbiota findings are
-comparable to what's been shown in another mouse study, whether
-PPI-induced changes in specific bacterial abundances observed in humans
-play a role in CDIs remains to be determined.\\
-Although there have been a few \emph{C. difficile} mouse model studies
-that have demonstrated PPIs have some effect on CDIs with or without
-additional antibiotic treatment (13--15), there were some key
-methodological differences between these studies and our own. One group
-administered 0.5 mg/kg of the PPI lansoprazole daily for 2 weeks to mice
-and then challenged with \emph{C. difficile} demonstrated that PPI
-treatment alone resulted in detectable \emph{C. difficile} in the stool
-1 week after challenge, but also showed there was detectable \emph{C.
-difficile} in mice not treated with antibiotics (13, 14). The presence
-of \emph{C. difficile} in mice not treated with antibiotics may be
-partly attributed to the higher dose of vegetative cells
-(10\textsuperscript{8} CFU) used, but nevertheless makes it difficult to
-interpret the extent to which PPI treatment contributed to \emph{C.
-difficile} colonization in their mouse model (13, 14). In contrast, we
-have previously shown mice from our colony that were not given
-antibiotics were resistant to \emph{C. difficile} 630 when challenged
-with 10\textsuperscript{3} spores (16) and previous work by our
-collaborators demonstrated mice not given antibiotics were resistant to
-10\textsuperscript{5} vegetative cells of \emph{C. difficile } VPI 10463
-(17). The other PPI CDI mouse study also used a higher dose of
-10\textsuperscript{6} spores or 10\textsuperscript{7} vegetative cells
-to challenge antibiotic-treated mice or mice treated with both
-antibiotics and the PPI esomeprazole (40mg/kg dose) for 2 days and
-demonstrated the antibiotic/PPI-treated mice developed more severe CDIs
-(15). We tested the same high 40 mg/kg PPI dose and expanded
-pre-treatment to 7 days before challenge to test the impact of PPI
-treatment alone on our CDI mouse model.\\
-Our study extended the previous work examining PPIs and \emph{C.
-difficile} in mice by evaluating the contribution of the intestinal
+not promote \emph{C. difficile} infection in mice. Our findings that
+omeprazole treatment had minimal impact on the fecal microbiome were
+comparable to another PPI mouse study that indicated the PPI
+lansoprazole had more of an effect on the small intestinal microbiota
+compared to the fecal microbiota (12). The same group demonstrated
+lansoprazole treatment increased the stomach pH in mice (12), which may
+improve survival of bacteria passing through the stomach. We did not
+find significant changes for the taxa observed to be significantly
+impacted by PPI use in human studies. However, 3 of the human-associated
+taxa were absent or at low abundance in our mice. Although our fecal
+microbiota findings are comparable to what has been shown in another
+mouse study (12), whether PPI-induced changes in specific bacterial
+abundances observed in humans play a role in CDIs remains to be
+determined.\\
+While several \emph{C. difficile} mouse model studies have shown that
+PPIs have an effect on CDIs with or without additional antibiotic
+treatment (13--15), there were insufficient controls to attribute the
+effect solely to PPI treatment. One group administered 0.5 mg/kg of the
+PPI lansoprazole daily for 2 weeks to mice and then challenged with
+\emph{C. difficile} demonstrated that PPI treatment alone resulted in
+detectable \emph{C. difficile} in the stool 1 week after challenge,
+however there was detectable \emph{C. difficile} in mice not treated
+with antibiotics (13, 14). The other mouse study demonstrated the
+antibiotic/PPI (esomeprazole)-treated mice developed more severe CDIs
+compared to antibiotic-treated mice, but did not have a group treated
+with just esomeprazole for comparison (15). We tested the same high 40
+mg/kg PPI dose and expanded pre-treatment to 7 days before challenge to
+test the impact of omeprazole treatment alone on our CDI mouse model.
+Additionally, we've previously demonstrated that mice from our colony
+were resistant to \emph{C. difficile} 630 colonization without
+antibiotic treatment (16), ensuring there was not already partial
+susceptiblity to \emph{C. difficile} before treatment. The additional
+controls in our study allowed us to assess the contibution of omeprazole
+alone to \emph{C. difficile} susceptibility in mice.\\
+Our study also extended previous work examining PPIs and \emph{C.
+difficile} in mice by incorporating the contribution of the intestinal
 microbiota. We found the PPI omeprazole had no significant impact on
 bacterial taxa within the murine intestinal microbiota over the 16-day
-experiment and in contrast to previous work, PPIs did not alter \emph{C.
-difficile} colonization resistance. 16S rRNA sequencing suggested that
-\emph{Streptococcus} and \emph{Enterococcus} are rare genera in our
-C57BL/6 mouse colony and variation in these 2 genera have been observed
-across other facilities and vendors (18, 19). These genera could be
-important contributors to the associations between PPIs and CDIs in
-humans, and could be a contributing factor to our observation that PPI
-treatment had no effect on \emph{C. difficile} colonization in our CDI
-mouse model.\\
+experiment. In contrast to previous work, omeprazole did not alter
+\emph{C. difficile} colonization resistance. 16S rRNA sequencing
+suggested that \emph{Streptococcus} and \emph{Enterococcus} are rare
+genera in our C57BL/6 mouse colony and variation in these 2 genera have
+been observed across other facilities and vendors (17, 18). These genera
+could be important contributors to the associations between PPIs and
+CDIs in humans, and could be a contributing factor to our observation
+that PPI treatment had no effect on \emph{C. difficile} colonization in
+our CDI mouse model. Gastrointestinal physiological differences,
+particularly the higher stomach pH in mice (pH 3-4) compared to humans
+(ph 1) (19) could also explain why omeprazole had a limited impact on
+the murine microbiome. The microbiota and physiological differences
+between humans and mice may limit the usefulness of employing mouse
+models to study the impact of PPIs on the microbiota and CDIs.\\
 Beyond microbiome differences, factors such as age, body mass index,
@@
 
-\textbf{Animals} All mouse experiments were performed with 7- to
+\textbf{Animals.} All mouse experiments were performed with 7- to
 12-week-old C57BL/6 male and female mice. Each experimental group of
@@
 
-\textbf{Drug treatments} Omeprazole (Sigma Aldrich) was prepared in a
+\textbf{Drug treatments.} Omeprazole (Sigma Aldrich) was prepared in a
 vehicle solution of 40\% polyethylene glycol 400 (Sigma-Aldrich) in
 phosphate buffered saline. Omeprazole was prepared from 20 mg/mL frozen
-aliquots and diluted to an 8mg/mL prior to gavage. All mice received 40
-mg/kg omeprazole or vehicle solution once per day through the duration
-of the experiment with treatment starting 7 days before \emph{C.
-difficile} challenge (Figure 1A). One day prior to \emph{C. difficile}
-challenge, 2 groups of mice received an intraperitoneal injection of 10
-mg/kg clindamycin or sterile saline vehicle. All drugs were filter
-sterilized through a 0.22 micron syringe filter before administration to
-animals.
+aliquots and diluted to an 8 mg/mL prior to gavage. All mice received 40
+mg/kg omeprazole (a dose previously used in mouse experiments (15)) or
+vehicle solution once per day through the duration of the experiment
+with treatment starting 7 days before \emph{C. difficile} challenge
+(Figure 1A). One day prior to \emph{C. difficile} challenge, 2 groups of
+mice received an intraperitoneal injection of 10 mg/kg clindamycin or
+sterile saline vehicle (11). All drugs were filter sterilized through a
+0.22 micron syringe filter before administration to animals.
 
-\textbf{\emph{C. difficile} infection model} Mice were challenged with
+\textbf{\emph{C. difficile} infection model.} Mice were challenged with
 \emph{C. difficile} 630 seven days after the start of omeprazole
 treatment and one day after clindamycin treatment. Mice were challenged
-with 10\textsuperscript{3} spores in ultrapure distilled water. Stool
-samples were collected for 16S rRNA sequencing or \emph{C. difficile}
-CFU quantification throughout the duration of the experiments at the
-indicated timepoints (Figure 1A). Samples for 16S rRNA sequencing were
-flash frozen in liquid nitrogen and stored at -80°C until DNA
-extraction, while samples for CFU quantification were transferred into
-an anaerobic chamber and serially diluted in PBS. Diluted samples were
-plated on TCCFA (taurocholate, cycloserine, cefoxitin, fructose agar)
-plates and incubated at 37°C for 24 hours under anaerobic conditions to
-quantify \emph{C. difficile} CFU.
+with 10\textsuperscript{3} spores in ultrapure distilled water as
+described previously (11). Stool samples were collected for 16S rRNA
+sequencing or \emph{C. difficile} CFU quantification throughout the
+duration of the experiments at the indicated timepoints (Figure 1A).
+Samples for 16S rRNA sequencing were flash frozen in liquid nitrogen and
+stored at -80°C until DNA extraction, while samples for CFU
+quantification were transferred into an anaerobic chamber and serially
+diluted in PBS. Diluted samples were plated on TCCFA (taurocholate,
+cycloserine, cefoxitin, fructose agar) plates and incubated at 37°C for
+24 hours under anaerobic conditions to quantify \emph{C. difficile} CFU.
 
-\textbf{16S rRNA gene sequencing} DNA for 16S rRNA gene sequencing was
+\textbf{16S rRNA gene sequencing.} DNA for 16S rRNA gene sequencing was
 extracted from 10-50 mg fecal pellet from each mouse using the DNeasy
 Powersoil HTP 96 Kit (Qiagen) and an EpMotion 5075 automated pipetting
-system (Eppendorf). The ZymoBIOMICS\textsuperscript{TM} Microbial
-Community DNA Standard (Zymo, CA, USA) was used as a mock community (21)
-and water was used as a negative control. The V4 hypervariable region of
-the 16S rRNA gene was amplified with Accuprime Pfx DNA polymerase
-(Thermo Fisher Scientific) using previously described custom barcoded
-primers (22). The 16S rRNA amplicon library was sequenced with the MiSeq
-(Illumina). Amplicons were cleaned up and normalized with the SequalPrep
-Normalization Plate Kit (ThermoFisher Scientific) and pooled amplicons
-were quantified with the KAPA library quantification kit (KAPA
-Biosystems).
+system (Eppendorf). The 16S rRNA sequencing libary was prepared as
+described previously (21). In brief, the ZymoBIOMICS\textsuperscript{TM}
+Microbial Community DNA Standard (Zymo, CA, USA) was used as a mock
+community (22) and water was used as a negative control. The V4
+hypervariable region of the 16S rRNA gene was amplified with Accuprime
+Pfx DNA polymerase (Thermo Fisher Scientific) using previously described
+custom barcoded primers (21). The 16S rRNA amplicon library was
+sequenced with the MiSeq (Illumina). Amplicons were cleaned up and
+normalized with the SequalPrep Normalization Plate Kit (ThermoFisher
+Scientific) and pooled amplicons were quantified with the KAPA library
+quantification kit (KAPA Biosystems).
 
-\textbf{16S rRNA gene sequence analysis}Mothur (v1.40.5) was used for
-all sequence processing steps (23) using the previously published
-protocol (22). In brief, forward and reverse reads for each sample were
-combined and low-quality sequences and chimeras were removed. Duplicate
-sequences were merged, before taxonomy assignment using a modified
-version (v16) of the Ribosomal Database Project reference database
-(v11.5) with an 80\% cutoff. Operational taxonomic units (OTUs) were
-assigned with the opticlust clustering algorithm using a 97\% similarity
-threshold. To adjust for uneven sequencing across samples, all samples
-were rarefied to 3000 sequences, 1,000 times. PCoAs were generated based
-on Bray-Curtis distance. R (v.3.5.1) was used to generate figures and
+\textbf{16S rRNA gene sequence analysis.} mothur (v1.40.5) was used for
+all sequence processing steps (23) using a previously published protocol
+(21). In brief, forward and reverse reads for each sample were combined
+and low-quality sequences and chimeras were removed. Duplicate sequences
+were merged, before taxonomy assignment using a modified version (v16)
+of the Ribosomal Database Project reference database (v11.5) with an
+80\% cutoff. Operational taxonomic units (OTUs) were assigned with the
+opticlust clustering algorithm using a 97\% similarity threshold. To
+adjust for uneven sequencing across samples, all samples were rarefied
+to 3,000 sequences, 1,000 times. PCoAs were generated based on
+Bray-Curtis distance. R (v.3.5.1) was used to generate figures and
 perform statistical analysis.
 
-\textbf{Statistical Analysis} To test for differences in relative
+\textbf{Statistical Analysis.} To test for differences in relative
 abundances in families and genera across our 3 different treatment
@@
 
-\textbf{Code availability} The code for all sequence processing and
+\textbf{Code availability.} The code for all sequence processing and
 analysis step as well as an Rmarkdown version of this manuscript is
 available at
-\url{https://github.com/SchlossLab/Tomkovich_PPI_XXXX_2019}.
+\url{https://github.com/SchlossLab/Tomkovich_PPI_mSphere_2019}.
 
-\textbf{Data availability} The 16S rRNA sequencing data have been
+\textbf{Data availability.} The 16S rRNA sequencing data have been
 deposited in the NCBI Sequence Read Archive (Accession no. PRJNA554866).
@@
 
+\hypertarget{ref-Imhann2015}{}
+7. \textbf{Imhann F}, \textbf{Bonder MJ}, \textbf{Vila AV}, \textbf{Fu
+J}, \textbf{Mujagic Z}, \textbf{Vork L}, \textbf{Tigchelaar EF},
+\textbf{Jankipersadsing SA}, \textbf{Cenit MC}, \textbf{Harmsen HJM},
+\textbf{Dijkstra G}, \textbf{Franke L}, \textbf{Xavier RJ},
+\textbf{Jonkers D}, \textbf{Wijmenga C}, \textbf{Weersma RK},
+\textbf{Zhernakova A}. 2015. Proton pump inhibitors affect the gut
+microbiome. Gut \textbf{65}:740--748.
+doi:\href{https://doi.org/10.1136/gutjnl-2015-310376}{10.1136/gutjnl-2015-310376}.
+
+\hypertarget{ref-Freedberg2015}{}
+8. \textbf{Freedberg DE}, \textbf{Toussaint NC}, \textbf{Chen SP},
+\textbf{Ratner AJ}, \textbf{Whittier S}, \textbf{Wang TC}, \textbf{Wang
+HH}, \textbf{Abrams JA}. 2015. Proton pump inhibitors alter specific
+taxa in the human gastrointestinal microbiome: A crossover trial.
+Gastroenterology \textbf{149}:883--885.e9.
+doi:\href{https://doi.org/10.1053/j.gastro.2015.06.043}{10.1053/j.gastro.2015.06.043}.
+
 \hypertarget{ref-maier2018extensive}{}
-7. \textbf{Maier L}, \textbf{Pruteanu M}, \textbf{Kuhn M},
+9. \textbf{Maier L}, \textbf{Pruteanu M}, \textbf{Kuhn M},
 \textbf{Zeller G}, \textbf{Telzerow A}, \textbf{Anderson EE},
@@
 \hypertarget{ref-Schuberte01021-14}{}
-8. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
+10. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
 \textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
@@
 \hypertarget{ref-Jenior2018}{}
-9. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+11. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
 \textbf{Schloss PD}. 2018. Clostridium difficile alters the structure
@@
 \hypertarget{ref-Yasutomi2018}{}
-10. \textbf{Yasutomi E}, \textbf{Hoshi N}, \textbf{Adachi S},
+12. \textbf{Yasutomi E}, \textbf{Hoshi N}, \textbf{Adachi S},
 \textbf{Otsuka T}, \textbf{Kong L}, \textbf{Ku Y}, \textbf{Yamairi H},
@@
 doi:\href{https://doi.org/10.1007/s10620-017-4905-3}{10.1007/s10620-017-4905-3}.
-
-\hypertarget{ref-Freedberg2015}{}
-11. \textbf{Freedberg DE}, \textbf{Toussaint NC}, \textbf{Chen SP},
-\textbf{Ratner AJ}, \textbf{Whittier S}, \textbf{Wang TC}, \textbf{Wang
-HH}, \textbf{Abrams JA}. 2015. Proton pump inhibitors alter specific
-taxa in the human gastrointestinal microbiome: A crossover trial.
-Gastroenterology \textbf{149}:883--885.e9.
-doi:\href{https://doi.org/10.1053/j.gastro.2015.06.043}{10.1053/j.gastro.2015.06.043}.
-
-\hypertarget{ref-Imhann2015}{}
-12. \textbf{Imhann F}, \textbf{Bonder MJ}, \textbf{Vila AV}, \textbf{Fu
-J}, \textbf{Mujagic Z}, \textbf{Vork L}, \textbf{Tigchelaar EF},
-\textbf{Jankipersadsing SA}, \textbf{Cenit MC}, \textbf{Harmsen HJM},
-\textbf{Dijkstra G}, \textbf{Franke L}, \textbf{Xavier RJ},
-\textbf{Jonkers D}, \textbf{Wijmenga C}, \textbf{Weersma RK},
-\textbf{Zhernakova A}. 2015. Proton pump inhibitors affect the gut
-microbiome. Gut \textbf{65}:740--748.
-doi:\href{https://doi.org/10.1136/gutjnl-2015-310376}{10.1136/gutjnl-2015-310376}.
 
@@
 
-\hypertarget{ref-Reeves2011}{}
-17. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
-\textbf{Huffnagle GB}, \textbf{Schloss PD}, \textbf{Young VB}. 2011. The
-interplay between microbiome dynamics and pathogen dynamics in a murine
-model ofClostridium difficileInfection. Gut Microbes
-\textbf{2}:145--158.
-doi:\href{https://doi.org/10.4161/gmic.2.3.16333}{10.4161/gmic.2.3.16333}.
-
 \hypertarget{ref-Rausch2016}{}
-18. \textbf{Rausch P}, \textbf{Basic M}, \textbf{Batra A},
+17. \textbf{Rausch P}, \textbf{Basic M}, \textbf{Batra A},
 \textbf{Bischoff SC}, \textbf{Blaut M}, \textbf{Clavel T},
@@
 \hypertarget{ref-Robertson2019}{}
-19. \textbf{Robertson SJ}, \textbf{Lemire P}, \textbf{Maughan H},
+18. \textbf{Robertson SJ}, \textbf{Lemire P}, \textbf{Maughan H},
 \textbf{Goethel A}, \textbf{Turpin W}, \textbf{Bedrani L},
@@
 doi:\href{https://doi.org/10.1016/j.celrep.2019.04.023}{10.1016/j.celrep.2019.04.023}.
+
+\hypertarget{ref-Hugenholtz2017}{}
+19. \textbf{Hugenholtz F}, \textbf{Vos WM de}. 2017. Mouse models for
+human intestinal microbiota research: A critical evaluation. Cellular
+and Molecular Life Sciences \textbf{75}:149--160.
+doi:\href{https://doi.org/10.1007/s00018-017-2693-8}{10.1007/s00018-017-2693-8}.
 
@@
 
-\hypertarget{ref-Sze2019}{}
-21. \textbf{Sze MA}, \textbf{Schloss PD}. 2019. The impact of DNA
-polymerase and number of rounds of amplification in PCR on 16S rRNA gene
-sequence data. mSphere \textbf{4}.
-doi:\href{https://doi.org/10.1128/msphere.00163-19}{10.1128/msphere.00163-19}.
-
 \hypertarget{ref-Kozich2013}{}
-22. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+21. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+
+\hypertarget{ref-Sze2019}{}
+22. \textbf{Sze MA}, \textbf{Schloss PD}. 2019. The impact of DNA
+polymerase and number of rounds of amplification in PCR on 16S rRNA gene
+sequence data. mSphere \textbf{4}.
+doi:\href{https://doi.org/10.1128/msphere.00163-19}{10.1128/msphere.00163-19}.
 
@@
 
-\includegraphics{figure_1.pdf} \textbf{Figure 1. PPI treatment had
-minimal impact on the murine fecal microbiota.} A. Mouse experiment
+\includegraphics{figure_1.pdf} \textbf{Figure 1. Omeprazole treatment
+had minimal impact on the murine fecal microbiota.} A. Mouse experiment
 timeline and logistics. The PPI omeprazole was administered throughout
@@
 culture. B. Principal Coordinates Analaysis (PCoA) of Bray-curtis
-distances from stool samples for all treatment groups during the intial
-7 days of the experiment. Each symbol represents a stool sample from an
-individual mouse on a given day. Transparency of the symbol corresponds
-to treatment day. Circles represent samples before antibiotic treatment
-(administered on Day -1), while open diamonds represent samples from the
-day after antibiotic treatment (Day 0). Mice treated with the PPI
-omeprazole or vehicle cluster together and do not change much over time
-while mice that receive clindamycin have samples that start to separate
-from the rest of the samples the day after antibiotic treatment. C.
-Relative abundances of families previously associated with PPI use in
-humans do not change much over the 16-day course of treatment with PPIs
-(7 days before \emph{C. difficile} challenge through 9 days post
-\emph{C. difficile} challenge). D. Boxplots showing the relative
-abundances of 7 familes within PPI-treated mice that had the lowest
-adjusted P values after comparing across 3 timepoints. There were no
-significant differences acoss time for any of the identified families
-(All P \textgreater{} 0.05) after analysis by Kruskal-Wallis test with a
-Benjamini-Hochberg correction for multiple comparisons. See Fig. S1C for
-genera with lowest adjusted P values (All P \textgreater{} 0.05). For
-C-D, the grey line indicates the limit of detection.
+distances from stool samples of mice in the omeprazole treatment group
+during the intial 7 days of the experiment. Each color represents stool
+samples from the same mouse and lines connect sequentially collected
+samples. C-D. Relative abundances of families previously associated with
+PPI use in humans at the start of the experiment (C) and after 7 days of
+omeprazole treatment (D). Each circle represents an individual mouse.
+There were no significant differences acoss treatment groups for any of
+the identified families in the seqeunce data at day -7 and day 0,
+analyzed by Kruskal-Wallis test with a Benjamini-Hochberg correction for
+multiple comparisons. For C-D, the grey vertical line indicates the
+limit of detection.
 
@@
 
-\includegraphics{figure_2.pdf} \textbf{Figure 2. PPI treatment alone
-does not promote CDIs in mice.} A. \emph{C. difficile} CFUs/g stool
-measured each day post \emph{C. difficile} challenge for clindamycin,
-clindamycin/PPI, and PPI-treated mice. Lines represent the mean CFU/g
-for each treatment group while points represent CFU/g for individual
-mice within each group. The black dashed line indicates the limit of
-detection. B. PCoA of of Bray-curtis distances from stool samples
-collected after antibiotic treatment (last 9 days of the experiment).
-Transparency of the symbol corresponds to treatment day. Symbols
-represent the \emph{C. difficile} colonization status of the mice
-measured 2 days post-infection. Circles represent resistant mice
+\includegraphics{figure_2.pdf} \textbf{Figure 2. Omeprazole treatment
+alone does not promote CDIs in mice.} A. \emph{C. difficile} CFUs/g
+stool measured each day post \emph{C. difficile} challenge for
+clindamycin, clindamycin/PPI, and PPI-treated mice. Lines represent the
+mean CFU/g for each treatment group while points represent CFU/g for
+individual mice within each group. The black dashed line indicates the
+limit of detection. B. PCoA of of Bray-curtis distances from stool
+samples collected after antibiotic treatment (last 9 days of the
+experiment). Transparency of the symbol corresponds to treatment day.
+Symbols represent the \emph{C. difficile} colonization status of the
+mice measured 2 days post-infection. Circles represent resistant mice
 (\emph{C. difficile} was undetectable in stool samples), while X-shapes
@@
 all mice cleared \emph{C. difficile} within 5 days of infection.
-PPI-treated fecal samples primarily cluster together throughout the
-experiment. C. Genera significantly associated with treatment groups of
-stool samples across all sequenced timepoints. D. Families significantly
-associated with treatment groups of stool samples across all sequenced
-timepoints. C,D data were analyzed by Kruskal-Wallis test with a
-Benjamini-Hochberg correction for multiple comparisons. For C-D, the
-grey line indicates the limit of detection.
+Omeprazole treated fecal samples primarily cluster together throughout
+the experiment. C. Genera that vary the most across treatment groups for
+stool samples collected from mice 2 days post-infection. Data were
+analyzed by Kruskal-Wallis test, and no P values were significant after
+Benjamini-Hochberg correction for multiple comparisons. The grey
+vertical line indicates the limit of detection.
 
@@
 
-\includegraphics{figure_s1.pdf} \textbf{Figure S1. Taxa within PPI
-treated mice fluctuate over time with no overall trend in either
-direction.} Relative abundance over time for Lactobacillaceae (A) and
-Ruminococcaceae (B), 2 of the PPI-associated familes from human PPI
+\includegraphics{figure_s1.pdf} \textbf{Figure S1. Families within
+omeprazole treated mice fluctuate over time with no overall trend in
+either direction.} Relative abundance over time for Lactobacillaceae (A)
+and Ruminococcaceae (B), 2 of the PPI-associated familes from human PPI
 studies across all 3 treatment groups. Each point represents the
-relative abundance for an individual mouse, while the lines represent
-the mean relative abundances for each treatment group of mice. C.
-Boxplots showing the relative abundances of 13 genera within PPI-treated
-mice that had the lowest adjusted P values after comparing across 3
-timepoints. There were no significant differences acoss time for any of
-the identified families (All P = 0.568) after analysis by Kruskal-Wallis
-test with a Benjamini-Hochberg correction for multiple comparisons. For
-A-C, the grey lines indicate the limit of detection.
+relative abundance for an individual mouse stool sample, while the lines
+represent the mean relative abundances for each treatment group of mice.
+The grey horizontal lines indicate the limit of detection.
 
@@
 and richness decrease with antibiotic treatment but remain relatively
-constant with PPI treatment.} A. Boxplots of the Shannon Index (A) and
-richness (B) values for each group of mice over 3 timepoints (Day -7, 0,
-and 9). Each circle represents the value for an individual mouse.
+constant with omeprazole treatment.} A. Boxplots of the Shannon
+Diversity Index values (A) and number of observed OTUs (B) for each
+group of mice over 3 timepoints (Day -7, 0, and 9). Each circle
+represents the value for a stool sample from an individual mouse.
 

### 2019-08-07 code/figure_1.R

@@ -4,8 +4,8 @@ library(cowplot)
 
 a <- ggdraw() + draw_image("results/figures/PPI_Exp_Scheme.png")
 b <- ggdraw() + draw_image("results/figures/before_C._diff_challenge.png")
-c <- ggdraw() + draw_image("results/figures/families_prev_assoc_w_PPIs.png")
-d <- ggdraw() + draw_image("results/figures/ppi_family_time.png")
+c <- ggdraw() + draw_image("results/figures/families_prev_assoc_w_PPIs_-7.png")
+d <- ggdraw() + draw_image("results/figures/families_prev_assoc_w_PPIs_0.png")
 
 plot_grid(a, b, c, d, labels = "AUTO", label_size = 12)+
   ggsave("results/figures/figure_1.pdf", width=6.5, height=3.8)+

### 2019-08-07 code/figure_2.R

@@ -5,9 +5,8 @@ library(cowplot)
 a <- ggdraw() + draw_image("results/figures/ppi_cfu.png")
 b <- ggdraw() + draw_image("results/figures/after_abx_C.diff.png")
 c <- ggdraw() + draw_image("results/figures/genera_assoc_w_treatment.png")
-d <- ggdraw() + draw_image("results/figures/family_assoc_w_treatment.png")
 
-plot_grid(a, b, c, d, labels = "AUTO", label_size = 12)+
+plot_grid(a, b, c, labels = "AUTO", label_size = 12)+
   ggsave("results/figures/figure_2.pdf", width=6.5, height=3.8)+
   ggsave("submission/figure_2.pdf", width=6.5, height=3.8)
 

### 2019-08-07 code/figure_s1.R

@@ -4,11 +4,10 @@ library(cowplot)
 
 a <- ggdraw() + draw_image("results/figures/lactobacillaceae_time.png")
 b <- ggdraw() + draw_image("results/figures/ruminococcaceae_time.png")
-c <- ggdraw() + draw_image("results/figures/ppi_genera_time.png")
 
-plot_grid(a, b, c, labels = "AUTO", label_size = 12)+
-  ggsave("results/figures/figure_s1.pdf", width=6.5, height=3)+
-  ggsave("submission/figure_s1.pdf", width=6.5, height=3)
+plot_grid(a, b, labels = "AUTO", label_size = 12)+
+  ggsave("results/figures/figure_s1.pdf", width=6.5, height=1.5)+
+  ggsave("submission/figure_s1.pdf", width=6.5, height=1.5)
 
 
 

### 2019-08-07 code/ppi_alpha.R

@@ -103,7 +103,7 @@ shapiro.test(ppi_day_alpha$shannon) # Test for Shannon diversity
 shapiro.test(ppi_day_alpha$sobs)
 # P > 0.05, means the data are normally distributed and we should do parametric test (aov/ANOVA)
 
-#Hypothesis testing between groups, all timepoints analysis of variance (ANOVA)
+#Hypothesis testing across 3 timepoints in the PPI group by analysis of variance (ANOVA)
 #shannon
 ppi_day_shannon_aov <- aov(shannon~day, data = ppi_day_alpha) 
 summary(ppi_day_shannon_aov)
@@ -115,6 +115,52 @@ TukeyHSD(ppi_day_shannon_aov)
 ppi_day_sobs_aov <- aov(sobs~day, data = ppi_day_alpha) 
 summary(ppi_day_sobs_aov)
 #Since P = 0.265, do not do Tukeys honest significance difference test to determine which comparisons between group are significant.
+
+#Test if Shannon diversity and richness is significantly different over D-7, 0, and 9 of the experiment within the clindamycin group----
+clind_day_alpha <- meta_alpha %>% 
+  filter( Group == "Clindamycin") %>% # select just clindamycin group 
+  filter(day == -7 | day == 0 | day ==9) %>% # select -7, 0, and -9 timepoints (beginning, middle and end of the experiment)
+  mutate(day = as.factor(day))# turn day into factor so we can do post hoc comparisons
+
+# Test if PPI Shannon data are normally distributed using Shapiro-Wilk normality test
+shapiro.test(clind_day_alpha$shannon) # Test for Shannon diversity
+shapiro.test(clind_day_alpha$sobs)
+# P < 0.05, means the data are not normally distributed and we should do nonparametric test (Kruskal-Wallis)
+
+#Hypothesis testing across 3 timepoints in the Clindamycin group
+#shannon
+clind_shannon_kruskal <- kruskal.test(shannon~day, data=clind_day_alpha)
+pairwise.wilcox.test(g=clind_day_alpha[["day"]], x=clind_day_alpha[["shannon"]], p.adjust.method="BH")
+# No pairwise comparisons between timepoints were significant
+
+#richness
+clind_sobs_kruskal <- kruskal.test(sobs~day, data=clind_day_alpha)
[…]

### 2019-08-07 code/ppi_cfu_over_time.R

@@ -1,6 +1,5 @@
 #Load packages
 library(tidyverse)
-library(vegan)
 library(cowplot)
 
 # Define color palette:
@@ -53,6 +52,5 @@ cfu_time <- ggplot(NULL) +
   scale_y_log10(labels=fancy_scientific, breaks = c(10, 100, 10^3, 10^4, 10^5, 10^6, 10^7, 10^8, 10^9))+
   theme_classic()+
   theme(legend.position = c(.9, .8)) +
-  labs(title=title) +
-  theme(plot.title=element_text(hjust=0.5))
+  theme(legend.title=element_blank(), plot.title=element_text(hjust=0.5))
 save_plot("results/figures/ppi_cfu.png", cfu_time, base_aspect_ratio = 2)

### 2019-08-07 code/ppi_pcoa.R

@@ -16,8 +16,9 @@ library(tidyverse)
 library(cowplot)
 
 metadata <- read.table('data/process/ppi_metadata.txt', header = T, sep = '\t', stringsAsFactors = F) %>% 
-  mutate(abx_status = if_else(day > -8 & day < 0, "pre", "post")) %>% # Create a column to differentiate between timepoints that are from before or after exposure to the antibiotic clindamycin
-  mutate(c.diff_colonized = if_else(D9.C..difficile.CFU.g > 0, "colonized", "resistant")) # Create a column to differentiate mice that were colonized with C. difficile from mice that were resistant. Based off of D9 (2 day post challenge) CFU counts.
+  mutate(c.diff_colonized = if_else(D9.C..difficile.CFU.g > 0, "colonized", "resistant")) %>% # Create a column to differentiate mice that were colonized with C. difficile from mice that were resistant. Based off of D9 (2 day post challenge) CFU counts.
+  mutate(Mouse.ID = as.factor(Mouse.ID)) # Make sure mouse.ID is treated as a factor to use a discrete color scale.
+
 
 # read in pcoa data
 pcoa <- read_tsv('data/mothur/ppi.opti_mcc.braycurtis.0.03.lt.ave.pcoa.axes') %>%
@@ -31,6 +32,7 @@ color_scheme <- c("#d95f02", "#1b9e77", "#7570b3")
 color_ppi <-  c("#7570b3") # Use for graphs looking at just the PPI group over time
 color_cppi <- c("#1b9e77") # Use for graphs looking at just the Clindamycin + PPI group over time
 color_c <- c("#d95f02") # Use for graphs looking at just the Clindamycin group over time
+color_mouse <- c("#dadaeb", "#bcbddc", "#9e9ac8", "#756bb1", "#54278f") # To differentiate between 5 individual mice
 
 #plot all samples
 pcoa %>%  
@@ -120,22 +122,18 @@ before_plus_day_after_abx <- pcoa %>%	filter(day < 1) %>%
 # plot 1st 7 days: Includes abx treatment day. Pre-clindamycin treatment represented by circles.
 # Post-clindamycin treatment represented by open diamonds.
 pcoa_before_challenge <- pcoa %>% 
-  filter(day < 1) %>%	
-  ggplot(aes(x=axis1, y=axis2, color=Group, alpha = day, shape = abx_status)) +
[…]

### 2019-08-07 code/ppi_taxa.R

@@ -179,139 +179,188 @@ agg_genus_data %>%
   theme_classic()
 
 #Hypothesis testing----
-#Kruskal_wallis test for phylum differences across treatment groups with Benjamini-Hochburg correction 
-phylum_tests <- agg_phylum_data %>% 
+#Kruskal_wallis test for phylum differences across treatment groups with Benjamini-Hochburg correction on Day -7 & Day 0 of the experiment
+#Day -7 (Start day of the experiment)
+phylum_tests_start_day <- agg_phylum_data %>% 
+  filter(day == -7) %>% 
   group_by(phylum) %>% 
   do(tidy(kruskal.test(agg_rel_abund~factor(Group), data=.))) %>% ungroup() %>% 
   mutate(p.value.adj=p.adjust(p.value, method="BH")) %>% 
   arrange(p.value.adj)
 
-#List the significant phyla based on treatment groups after Benjamini-Hochburg correction
-sig_phyla <- phylum_tests %>% 
+phylum_tests_day0 <- agg_phylum_data %>% 
+  filter(day == 0) %>% 
+  group_by(phylum) %>% 
+  do(tidy(kruskal.test(agg_rel_abund~factor(Group), data=.))) %>% ungroup() %>% 
+  mutate(p.value.adj=p.adjust(p.value, method="BH")) %>% 
+  arrange(p.value.adj)
+
+#List the significant phyla based on treatment groups after Benjamini-Hochburg correction on Day -7 & Day 0 of the experiment
+sig_phyla_start_day <- phylum_tests_start_day %>% 
   filter(p.value.adj <= 0.05) %>% 
   pull(phylum)
 
-#Pairwise wilcox test to determine which comparisons are significantly different
-wilcox_phylum_tests <- agg_phylum_data %>% 
-  group_by(phylum) %>% 
-  filter(phylum %in% sig_phyla) %>% 
-  select(phylum, Group, agg_rel_abund) 
-wilcox_phylum_tests <- wilcox_phylum_tests %>% 
-  do(tidy(pairwise.wilcox.test(g=wilcox_phylum_tests[["Group"]], x=wilcox_phylum_tests[["agg_rel_abund"]], p.adjust.method = "BH")))
-# Warning messages: cannot compute exact p value with ties
+sig_phyla_day_0 <- phylum_tests_day0 %>% 
+  filter(p.value.adj <= 0.05) %>% 
+  pull(phylum)
[…]

### 2019-08-21 code/ppi_pcoa.R
Started to subset distance matrix in R, but ended up just using mothur, and making a mothur/subset directory

@@ -4,6 +4,7 @@
 # 
 # input:
 #	data/mothur/ppi.opti_mcc.braycurtis.0.03.lt.ave.pcoa.axes
+# data/mothur/ppi.opti_mcc.braycurtis.0.03.lt.ave.dist
 #	data/process/ppi_metadata.txt
 #
 # output:
@@ -17,8 +18,8 @@ library(cowplot)
 
 metadata <- read.table('data/process/ppi_metadata.txt', header = T, sep = '\t', stringsAsFactors = F) %>% 
   mutate(c.diff_colonized = if_else(D9.C..difficile.CFU.g > 0, "colonized", "resistant")) %>% # Create a column to differentiate mice that were colonized with C. difficile from mice that were resistant. Based off of D9 (2 day post challenge) CFU counts.
-  mutate(Mouse.ID = as.factor(Mouse.ID)) # Make sure mouse.ID is treated as a factor to use a discrete color scale.
-
+  mutate(Mouse.ID = as.factor(Mouse.ID)) %>% # Make sure mouse.ID is treated as a factor to use a discrete color scale.
+  mutate(Group=factor(Group, levels=c("Clindamycin", "Clind. + Omep.", "Omeprazole"))) #Make sure Group is treated as a factor
 
 # read in pcoa data
 pcoa <- read_tsv('data/mothur/ppi.opti_mcc.braycurtis.0.03.lt.ave.pcoa.axes') %>%
@@ -28,16 +29,20 @@ pcoa <- read_tsv('data/mothur/ppi.opti_mcc.braycurtis.0.03.lt.ave.pcoa.axes') %>
 pcoa <- pcoa %>% filter(Group != "NA")
 
 # Define color palette:
-color_scheme <- c("#d95f02", "#1b9e77", "#7570b3")
+color_scheme <- c("#d95f02", "#1b9e77",  "#7570b3")
 color_ppi <-  c("#7570b3") # Use for graphs looking at just the PPI group over time
 color_cppi <- c("#1b9e77") # Use for graphs looking at just the Clindamycin + PPI group over time
 color_c <- c("#d95f02") # Use for graphs looking at just the Clindamycin group over time
-color_mouse <- c("#dadaeb", "#bcbddc", "#9e9ac8", "#756bb1", "#54278f") # To differentiate between 5 individual mice
+color_mouse <- c("#7fc97f", "#beaed4", "#fdc086", "#386cb0", "#f0027f") # To differentiate between 5 individual mice
[…]

### 2019-08-21 code/ppi_pcoa.R
Remove code about how to import distance matrix into R

@@ -124,42 +124,22 @@ before_plus_day_after_abx <- pcoa %>%	filter(day < 1) %>%
   ggsave("results/figures/before_plus_day_after_abx.png")
 
 # Figure 1B----
-# plot 1st 7 days: Includes abx treatment day. Pull just these samples from the distance matrix: data/mothur/ppi.opti_mcc.braycurtis.0.03.lt.ave.dist and then analyze that file with
-# mothur
-# Function for reading distance matrix into R. Source: https://forum.mothur.org/t/importing-dist-matrix-into-r/1266/6
-parseDistanceDF = function(phylip_file) {
-  
-  # Read the first line of the phylip file to find out how many sequences/samples it contains
-  temp_connection = file(phylip_file, 'r')
-  len = readLines(temp_connection, n=1)
-  len = as.numeric(len)
-  len = len +1
-  close(temp_connection)
-  
-  
-  phylip_data = read.table(phylip_file, fill=T, row.names=1, skip=1, col.names=1:len)
-  colnames(phylip_data) <- row.names(phylip_data)
-  return(phylip_data)
-}
-#Read in PPI distance matrix using parseDistanceMatrix function
-distance_matrix <- parseDistanceDF('data/mothur/ppi.opti_mcc.braycurtis.0.03.lt.ave.dist')
-#Get a list of the shared_names from metadata file that 
+# plot 1st 7 days: Includes abx treatment day. 
+#Get a list of the shared_names from metadata file that are just the omeprazole treated mice before antibiotic treatment
 ppi_before_challenge_subset <- metadata %>% 
   filter(day < 1, Group == "Omeprazole") %>% 
   pull(shared_names)
-#groups specification for running dist.shared & pcoa commands on just this subset of samples in mothur"
+#groups specification for running dist.shared & pcoa commands on just this subset of samples in mothur". See README for instructions on analysis using mothur commands.
 # Opos+M1+D0-Opos+M2+D0-Opos+M3+D0-Opos+M4+D0-Opos+M5+D0-Opos+M1+D2-Opos+M2+D2-Opos+M3+D2-Opos+M4+D2-Cpos+M5+D-Opos+M1+D4-Opos+M3+D4-Opos+M4+D4-Opos+M5+D4-Opos+M11+D6-Opos+M2+D6-Opos+M3+D-Opos+M4+D6-Opos+M5+D6-Opos+M1+D7-Opos+M2+D7-Opos+M3+D7-Opos+M4+D7-Opos+M5+D7
[…]

### 2019-08-22 submission/manuscript.Rmd
Fix minor typos
Making changes in response to Pat's edits

@@
 output:
-  word_document: default
   pdf_document:
@@
     keep_tex: yes
+  word_document: default
 ---
@@
 These results suggest omeprazole treatment alone is not sufficient to disrupt microbiota resistance to *C.
-difficle* infection in mice that are normally resistant in the absence of antibiotic treatment.
+difficile* infection in mice that are normally resistant in the absence of antibiotic treatment.
 
@@
 Antibiotics are the primary risk factor for *Clostridium difficile* infections (CDIs), but other factors may also increase a person's risk.
-In epidemiological studies, PPI use has been associated with CDI incidence and recurrence.
-Proton pump inhibitors (PPIs) have also been associated with alterations in the human intestinal microbiota in observational and interventional studies.
+In epidemiological studies, proton pump inhibitor (PPI) use has been associated with CDI incidence and recurrence.
+PPIs have also been associated with alterations in the human intestinal microbiota in observational and interventional studies.
 We evaluated the effects of the PPI omeprazole on the structure of the murine intestinal microbiota and its ability to disrupt colonization resistance to *C.
@@
 PPI-associated microbiota changes have been attributed to the ability of PPIs to increase stomach acid pH which may promote the survival of oral and pathogenic bacteria [@Imhann2017; @Naito2018].
-In human fecal samples, PPI use results in increases in Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae and decreases in Ruminococcaceae [@Imhann2015; @Imhann2017; @Freedberg2015; @maier2018extensive].
+In human fecal samples, PPI use results in increases in *Enterococcaceae*, *Lactobacillaceae*, *Micrococcaceae*, *Staphylococcaceae* and *Streptococcaceae* and decreases in *Ruminococcaceae* [@Imhann2015; @Imhann2017; @Freedberg2015; @maier2018extensive].
 Several of these taxa have also been associated with *C.
 difficile* colonization in humans [@Schuberte01021-14].
-Unfortunately, most of the studies suggesting a link between PPIs and *C.
+Unfortunately, the studies suggesting a link between PPIs and *C.
 difficile* were retrospective and did not evaluate changes in the microbiome [@tariq2017association; @nehra2018proton; @Elias2019].
@@
 
-**Murine fecal microbiomes were minimally affected by omeprazole treatment ** To test whether omeprazole treatment alters the microbiome and promotes susceptibility to CDIs, we gavaged mice with 40 mg/kg of omeprazole for 7 days before *C.
+**Murine fecal microbiomes were minimally affected by omeprazole treatment.
+** To test whether omeprazole treatment alters the microbiome and promotes susceptibility to CDIs, we gavaged mice with 40 mg/kg of omeprazole for 7 days before *C.
 difficile* challenge (Figure 1A).
 A principle coordinates analysis (PCoA) of the Bray-Curtis distances over the initial 7 days of treatment revealed the bacterial communities of omeprazole-treated mice remained relatively unchanged (Figure 1B).
-We observed no significant fluctuations in the relative abundance of those taxa previously shown to respond to PPI treatment throughout the course of the 16-day experiment (Figure 1C-D, S1).
-We also observed no significant changes in relative abundances at the family and genus level over the course of the experiment for the omeprazole-treated mice (data not shown).
+We observed no significant changes in the relative abundance of those taxa previously shown to respond to PPI treatment throughout the course of the 16-day experiment (Figure 1C-D, S1).
+We also observed no significant changes in relative abundances at the family and genus level over the course of the experiment for the omeprazole-treated mice (all corrected P-values > 0.36).
 These results demonstrated that the omeprazole treatment alone had a minimal impact on the murine fecal bacterial community after 7 days of pretreatment.
@@
 **Omeprazole treatment did not promote susceptibility to *C.
-difficile infection* in mice** Next, we examined whether omeprazole treatment altered susceptibility to *C.
+difficile infection* in mice.** Next, we examined whether omeprazole treatment altered susceptibility to *C.
 difficile* infection in mice.
@@
 
-## Conclusions
-We found the PPI omeprazole did not impact the gut microbiota and did not promote *C.
+**Conclusions.** The PPI omeprazole did not meaningfully impact the gut microbiota and did not promote *C.
 difficile* infection in mice.
@@
 The same group demonstrated lansoprazole treatment increased the stomach pH in mice [@Yasutomi2018], which may improve survival of bacteria passing through the stomach.
-We did not find significant changes for the taxa observed to be significantly impacted by PPI use in human studies.
+We did not find significant changes in the relative abundances of the taxa observed to be significantly impacted by PPI use in human studies.
 However, 3 of the human-associated taxa were absent or at low abundance in our mice.
 Although our fecal microbiota findings are comparable to what has been shown in another mouse study [@Yasutomi2018], whether PPI-induced changes in specific bacterial abundances observed in humans play a role in CDIs remains to be determined.
-While several *C.
+Although several *C.
 difficile* mouse model studies have shown that PPIs have an effect on CDIs with or without additional antibiotic treatment [@Kaur2007; @kaur2011effect; @hung2015proton], there were insufficient controls to attribute the effect solely to PPI treatment.
@@
 difficile* in mice not treated with antibiotics [@Kaur2007; @kaur2011effect].
-The other mouse study demonstrated the antibiotic/PPI (esomeprazole)-treated mice developed more severe CDIs compared to antibiotic-treated mice, but did not have a group treated with just esomeprazole for comparison [@hung2015proton].
+The other mouse study demonstrated the antibiotic/PPI esomeprazole-treated mice developed more severe CDIs compared to antibiotic-treated mice, but the researchers did not have a group treated with just esomeprazole for comparison [@hung2015proton].
 We tested the same high 40 mg/kg PPI dose and expanded pre-treatment to 7 days before challenge to test the impact of omeprazole treatment alone on our CDI mouse model.
-Additionally, we've previously demonstrated that mice from our colony were resistant to *C.
+Additionally, we have previously demonstrated that mice from our breeding colony are resistant to *C.
 difficile* 630 colonization without antibiotic treatment [@Jenior2017], ensuring there was not already partial susceptiblity to *C.
@@
 difficile* in mice by incorporating the contribution of the intestinal microbiota.
-We found the PPI omeprazole had no significant impact on bacterial taxa within the murine intestinal microbiota over the 16-day experiment.
-In contrast to previous work, omeprazole did not alter *C.
-difficile* colonization resistance.
+We found omeprazole had no significant impact on bacterial taxa within the murine intestinal microbiota over the 16-day experiment.
+In contrast to previous work with PPIs [@Kaur2007; @kaur2011effect; @hung2015proton], omeprazole did not alter *C.
+difficile* colonization resistance in mice.
 16S rRNA sequencing suggested that *Streptococcus* and *Enterococcus* are rare genera in our C57BL/6 mouse colony and variation in these 2 genera have been observed across other facilities and vendors [@Rausch2016; @Robertson2019].
@@
 difficile* colonization in our CDI mouse model.
-Gastrointestinal physiological differences, particularly the higher stomach pH in mice (pH 3-4) compared to humans (ph 1) [@Hugenholtz2017] could also explain why omeprazole had a limited impact on the murine microbiome.
+Gastrointestinal physiological differences, particularly the higher stomach pH in mice (pH 3-4) compared to humans (pH 1) [@Hugenholtz2017] could also explain why omeprazole had a limited impact on the murine microbiome.
 The microbiota and physiological differences between humans and mice may limit the usefulness of employing mouse models to study the impact of PPIs on the microbiota and CDIs.
@@
 
-**Statistical Analysis.** To test for differences in relative abundances in families and genera across our 3 different treatment groups (Clindamycin, Clindamycin + PPI, and PPI) or within the PPI treatment group across 3 timepoints (Day -7, 0, and 9), we used a Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
+**Statistical Analysis.** To test for differences in relative abundances in families and genera across our 3 different treatment groups at different timepoints (Clindamycin, Clind.
++ Omep., and Omeprazole on Day -7, 0, 2, and 9) or within the Omeprazole treatment group across 3 timepoints (Day -7, 0, and 9), we used a Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
 
-**Code availability.** The code for all sequence processing and analysis step as well as an Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Tomkovich_PPI_mSphere_2019.
+**Code availability.** The code for all sequence processing and analysis steps as well as an Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Tomkovich_PPI_mSphere_2019.
 
@@
 Each circle represents an individual mouse.
-There were no significant differences acoss treatment groups for any of the identified families in the seqeunce data at day -7 and day 0, analyzed by Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
+There were no significant differences acoss treatment groups for any of the identified families in the sequence data at day -7 (all P-values > 0.448) and day 0 (all P-values > 0.137), analyzed by Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
 For C-D, the grey vertical line indicates the limit of detection.
@@
 difficile* CFUs/g stool measured each day post *C.
-difficile* challenge for clindamycin, clindamycin/PPI, and PPI-treated mice.
+difficile* challenge for clindamycin, clindamycin/omeprazole, and omeprazole-treated mice.
 Lines represent the mean CFU/g for each treatment group while points represent CFU/g for individual mice within each group.
@@
 Genera that vary the most across treatment groups for stool samples collected from mice 2 days post-infection.
-Data were analyzed by Kruskal-Wallis test, and no P values were significant after Benjamini-Hochberg correction for multiple comparisons.
+Data were analyzed by Kruskal-Wallis test, and no P-values were significant after Benjamini-Hochberg correction for multiple comparisons (all P-values > 0.092).
 The grey vertical line indicates the limit of detection.
@@
 **Figure S1.
-Families within omeprazole treated mice fluctuate over time with no overall trend in either direction.** Relative abundance over time for Lactobacillaceae (A) and Ruminococcaceae (B), 2 of the PPI-associated familes from human PPI studies across all 3 treatment groups.
+Families within omeprazole treated mice fluctuate over time with no overall trend in either direction.** Relative abundance over time for *Lactobacillaceae* (A) and *Ruminococcaceae* (B), 2 of the PPI-associated familes from human PPI studies across all 3 treatment groups.
 Each point represents the relative abundance for an individual mouse stool sample, while the lines represent the mean relative abundances for each treatment group of mice.

### 2019-08-22 submission/manuscript.tex
Making changes in response to Pat's edits

@@
 exposure. These results suggest omeprazole treatment alone is not
-sufficient to disrupt microbiota resistance to \emph{C. difficle}
+sufficient to disrupt microbiota resistance to \emph{C. difficile}
 infection in mice that are normally resistant in the absence of
@@
 infections (CDIs), but other factors may also increase a person's risk.
-In epidemiological studies, PPI use has been associated with CDI
-incidence and recurrence. Proton pump inhibitors (PPIs) have also been
+In epidemiological studies, proton pump inhibitor (PPI) use has been
+associated with CDI incidence and recurrence. PPIs have also been
 associated with alterations in the human intestinal microbiota in
@@
 survival of oral and pathogenic bacteria (4, 6). In human fecal samples,
-PPI use results in increases in Enterococcaceae, Lactobacillaceae,
-Micrococcaceae, Staphylococcaceae and Streptococcaceae and decreases in
-Ruminococcaceae (6--9). Several of these taxa have also been associated
-with \emph{C. difficile} colonization in humans (10).\\
-Unfortunately, most of the studies suggesting a link between PPIs and
-\emph{C. difficile} were retrospective and did not evaluate changes in
-the microbiome (2, 3, 5). Thus, it is unclear whether the
-gastrointestinal microbiome changes associated with PPI use explain the
-association between PPIs and CDIs. Additionally, epidemiological studies
-have a limited capacity to address potential confounders and
-comorbidities in patients that were on PPIs and developed CDIs or
-recurrent CDIs (2, 5). Here, we evaluated the impact of daily PPI
-treatment with omeprazole on the murine microbiome and susceptibility to
-\emph{C. difficile} colonization in relation to clindamycin, an
-antibiotic that perturbs the microbiome enough to allow \emph{C.
-difficile} to colonize but is mild enough that \emph{C. difficile} is
-cleared within 10 days (11).
+PPI use results in increases in \emph{Enterococcaceae},
+\emph{Lactobacillaceae}, \emph{Micrococcaceae}, \emph{Staphylococcaceae}
+and \emph{Streptococcaceae} and decreases in \emph{Ruminococcaceae}
+(6--9). Several of these taxa have also been associated with \emph{C.
+difficile} colonization in humans (10).\\
+Unfortunately, the studies suggesting a link between PPIs and \emph{C.
+difficile} were retrospective and did not evaluate changes in the
+microbiome (2, 3, 5). Thus, it is unclear whether the gastrointestinal
+microbiome changes associated with PPI use explain the association
+between PPIs and CDIs. Additionally, epidemiological studies have a
+limited capacity to address potential confounders and comorbidities in
+patients that were on PPIs and developed CDIs or recurrent CDIs (2, 5).
+Here, we evaluated the impact of daily PPI treatment with omeprazole on
+the murine microbiome and susceptibility to \emph{C. difficile}
+colonization in relation to clindamycin, an antibiotic that perturbs the
+microbiome enough to allow \emph{C. difficile} to colonize but is mild
+enough that \emph{C. difficile} is cleared within 10 days (11).
 
 \textbf{Murine fecal microbiomes were minimally affected by omeprazole
-treatment } To test whether omeprazole treatment alters the microbiome
+treatment. } To test whether omeprazole treatment alters the microbiome
 and promotes susceptibility to CDIs, we gavaged mice with 40 mg/kg of
@@
 of omeprazole-treated mice remained relatively unchanged (Figure 1B). We
-observed no significant fluctuations in the relative abundance of those
-taxa previously shown to respond to PPI treatment throughout the course
-of the 16-day experiment (Figure 1C-D, S1). We also observed no
-significant changes in relative abundances at the family and genus level
-over the course of the experiment for the omeprazole-treated mice (data
-not shown). These results demonstrated that the omeprazole treatment
-alone had a minimal impact on the murine fecal bacterial community after
-7 days of pretreatment.
+observed no significant changes in the relative abundance of those taxa
+previously shown to respond to PPI treatment throughout the course of
+the 16-day experiment (Figure 1C-D, S1). We also observed no significant
+changes in relative abundances at the family and genus level over the
+course of the experiment for the omeprazole-treated mice (all corrected
+P-values \textgreater{} 0.36). These results demonstrated that the
+omeprazole treatment alone had a minimal impact on the murine fecal
+bacterial community after 7 days of pretreatment.
 
 \textbf{Omeprazole treatment did not promote susceptibility to \emph{C.
-difficile infection} in mice} Next, we examined whether omeprazole
+difficile infection} in mice.} Next, we examined whether omeprazole
 treatment altered susceptibility to \emph{C. difficile} infection in
@@
 
-\subsection{Conclusions}\label{conclusions}
-
-We found the PPI omeprazole did not impact the gut microbiota and did
-not promote \emph{C. difficile} infection in mice. Our findings that
-omeprazole treatment had minimal impact on the fecal microbiome were
-comparable to another PPI mouse study that indicated the PPI
-lansoprazole had more of an effect on the small intestinal microbiota
-compared to the fecal microbiota (12). The same group demonstrated
-lansoprazole treatment increased the stomach pH in mice (12), which may
-improve survival of bacteria passing through the stomach. We did not
-find significant changes for the taxa observed to be significantly
-impacted by PPI use in human studies. However, 3 of the human-associated
-taxa were absent or at low abundance in our mice. Although our fecal
-microbiota findings are comparable to what has been shown in another
-mouse study (12), whether PPI-induced changes in specific bacterial
-abundances observed in humans play a role in CDIs remains to be
-determined.\\
-While several \emph{C. difficile} mouse model studies have shown that
+\textbf{Conclusions.} The PPI omeprazole did not meaningfully impact the
+gut microbiota and did not promote \emph{C. difficile} infection in
+mice. Our findings that omeprazole treatment had minimal impact on the
+fecal microbiome were comparable to another PPI mouse study that
+indicated the PPI lansoprazole had more of an effect on the small
+intestinal microbiota compared to the fecal microbiota (12). The same
+group demonstrated lansoprazole treatment increased the stomach pH in
+mice (12), which may improve survival of bacteria passing through the
+stomach. We did not find significant changes in the relative abundances
+of the taxa observed to be significantly impacted by PPI use in human
+studies. However, 3 of the human-associated taxa were absent or at low
+abundance in our mice. Although our fecal microbiota findings are
+comparable to what has been shown in another mouse study (12), whether
+PPI-induced changes in specific bacterial abundances observed in humans
+play a role in CDIs remains to be determined.\\
+Although several \emph{C. difficile} mouse model studies have shown that
 PPIs have an effect on CDIs with or without additional antibiotic
@@
 with antibiotics (13, 14). The other mouse study demonstrated the
-antibiotic/PPI (esomeprazole)-treated mice developed more severe CDIs
-compared to antibiotic-treated mice, but did not have a group treated
-with just esomeprazole for comparison (15). We tested the same high 40
-mg/kg PPI dose and expanded pre-treatment to 7 days before challenge to
-test the impact of omeprazole treatment alone on our CDI mouse model.
-Additionally, we've previously demonstrated that mice from our colony
-were resistant to \emph{C. difficile} 630 colonization without
-antibiotic treatment (16), ensuring there was not already partial
-susceptiblity to \emph{C. difficile} before treatment. The additional
-controls in our study allowed us to assess the contibution of omeprazole
-alone to \emph{C. difficile} susceptibility in mice.\\
+antibiotic/PPI esomeprazole-treated mice developed more severe CDIs
+compared to antibiotic-treated mice, but the researchers did not have a
+group treated with just esomeprazole for comparison (15). We tested the
+same high 40 mg/kg PPI dose and expanded pre-treatment to 7 days before
+challenge to test the impact of omeprazole treatment alone on our CDI
+mouse model. Additionally, we have previously demonstrated that mice
+from our breeding colony are resistant to \emph{C. difficile} 630
+colonization without antibiotic treatment (16), ensuring there was not
+already partial susceptiblity to \emph{C. difficile} before treatment.
+The additional controls in our study allowed us to assess the
+contibution of omeprazole alone to \emph{C. difficile} susceptibility in
+mice.\\
 Our study also extended previous work examining PPIs and \emph{C.
 difficile} in mice by incorporating the contribution of the intestinal
-microbiota. We found the PPI omeprazole had no significant impact on
-bacterial taxa within the murine intestinal microbiota over the 16-day
-experiment. In contrast to previous work, omeprazole did not alter
-\emph{C. difficile} colonization resistance. 16S rRNA sequencing
-suggested that \emph{Streptococcus} and \emph{Enterococcus} are rare
-genera in our C57BL/6 mouse colony and variation in these 2 genera have
-been observed across other facilities and vendors (17, 18). These genera
-could be important contributors to the associations between PPIs and
-CDIs in humans, and could be a contributing factor to our observation
-that PPI treatment had no effect on \emph{C. difficile} colonization in
-our CDI mouse model. Gastrointestinal physiological differences,
-particularly the higher stomach pH in mice (pH 3-4) compared to humans
-(ph 1) (19) could also explain why omeprazole had a limited impact on
-the murine microbiome. The microbiota and physiological differences
-between humans and mice may limit the usefulness of employing mouse
-models to study the impact of PPIs on the microbiota and CDIs.\\
+microbiota. We found omeprazole had no significant impact on bacterial
+taxa within the murine intestinal microbiota over the 16-day experiment.
+In contrast to previous work with PPIs (13--15), omeprazole did not
+alter \emph{C. difficile} colonization resistance in mice. 16S rRNA
+sequencing suggested that \emph{Streptococcus} and \emph{Enterococcus}
+are rare genera in our C57BL/6 mouse colony and variation in these 2
+genera have been observed across other facilities and vendors (17, 18).
+These genera could be important contributors to the associations between
+PPIs and CDIs in humans, and could be a contributing factor to our
+observation that PPI treatment had no effect on \emph{C. difficile}
+colonization in our CDI mouse model. Gastrointestinal physiological
+differences, particularly the higher stomach pH in mice (pH 3-4)
+compared to humans (pH 1) (19) could also explain why omeprazole had a
+limited impact on the murine microbiome. The microbiota and
+physiological differences between humans and mice may limit the
+usefulness of employing mouse models to study the impact of PPIs on the
+microbiota and CDIs.\\
 Beyond microbiome differences, factors such as age, body mass index,
@@
 abundances in families and genera across our 3 different treatment
-groups (Clindamycin, Clindamycin + PPI, and PPI) or within the PPI
-treatment group across 3 timepoints (Day -7, 0, and 9), we used a
-Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple
-comparisons.
+groups at different timepoints (Clindamycin, Clind. + Omep., and
+Omeprazole on Day -7, 0, 2, and 9) or within the Omeprazole treatment
+group across 3 timepoints (Day -7, 0, and 9), we used a Kruskal-Wallis
+test with a Benjamini-Hochberg correction for multiple comparisons.
 
 \textbf{Code availability.} The code for all sequence processing and
-analysis step as well as an Rmarkdown version of this manuscript is
+analysis steps as well as an Rmarkdown version of this manuscript is
 available at
@@
 There were no significant differences acoss treatment groups for any of
-the identified families in the seqeunce data at day -7 and day 0,
+the identified families in the sequence data at day -7 (all P-values
+\textgreater{} 0.448) and day 0 (all P-values \textgreater{} 0.137),
 analyzed by Kruskal-Wallis test with a Benjamini-Hochberg correction for
@@
 stool measured each day post \emph{C. difficile} challenge for
-clindamycin, clindamycin/PPI, and PPI-treated mice. Lines represent the
-mean CFU/g for each treatment group while points represent CFU/g for
-individual mice within each group. The black dashed line indicates the
-limit of detection. B. PCoA of of Bray-curtis distances from stool
-samples collected after antibiotic treatment (last 9 days of the
-experiment). Transparency of the symbol corresponds to treatment day.
-Symbols represent the \emph{C. difficile} colonization status of the
-mice measured 2 days post-infection. Circles represent resistant mice
-(\emph{C. difficile} was undetectable in stool samples), while X-shapes
-represent mice that were colonized with \emph{C. difficile}, although
-all mice cleared \emph{C. difficile} within 5 days of infection.
-Omeprazole treated fecal samples primarily cluster together throughout
-the experiment. C. Genera that vary the most across treatment groups for
-stool samples collected from mice 2 days post-infection. Data were
-analyzed by Kruskal-Wallis test, and no P values were significant after
-Benjamini-Hochberg correction for multiple comparisons. The grey
-vertical line indicates the limit of detection.
+clindamycin, clindamycin/omeprazole, and omeprazole-treated mice. Lines
+represent the mean CFU/g for each treatment group while points represent
+CFU/g for individual mice within each group. The black dashed line
+indicates the limit of detection. B. PCoA of of Bray-curtis distances
+from stool samples collected after antibiotic treatment (last 9 days of
+the experiment). Transparency of the symbol corresponds to treatment
+day. Symbols represent the \emph{C. difficile} colonization status of
+the mice measured 2 days post-infection. Circles represent resistant
+mice (\emph{C. difficile} was undetectable in stool samples), while
+X-shapes represent mice that were colonized with \emph{C. difficile},
+although all mice cleared \emph{C. difficile} within 5 days of
+infection. Omeprazole treated fecal samples primarily cluster together
+throughout the experiment. C. Genera that vary the most across treatment
+groups for stool samples collected from mice 2 days post-infection. Data
+were analyzed by Kruskal-Wallis test, and no P-values were significant
+after Benjamini-Hochberg correction for multiple comparisons (all
+P-values \textgreater{} 0.092). The grey vertical line indicates the
+limit of detection.
 
@@
 omeprazole treated mice fluctuate over time with no overall trend in
-either direction.} Relative abundance over time for Lactobacillaceae (A)
-and Ruminococcaceae (B), 2 of the PPI-associated familes from human PPI
-studies across all 3 treatment groups. Each point represents the
-relative abundance for an individual mouse stool sample, while the lines
-represent the mean relative abundances for each treatment group of mice.
-The grey horizontal lines indicate the limit of detection.
+either direction.} Relative abundance over time for
+\emph{Lactobacillaceae} (A) and \emph{Ruminococcaceae} (B), 2 of the
+PPI-associated familes from human PPI studies across all 3 treatment
+groups. Each point represents the relative abundance for an individual
+mouse stool sample, while the lines represent the mean relative
+abundances for each treatment group of mice. The grey horizontal lines
+indicate the limit of detection.
 

### 2019-08-22 code/merge_labels.R

@@ -34,9 +34,9 @@ metadata <- read_xlsx('data/raw/PPI_metadata.xlsx') %>%
   # convert names of seq to match the format of the shared samples
   mutate(shared_names = gsub('^_{1,3}','', Plate_label),
          shared_names = gsub('_', '+', shared_names)) %>% 
-  mutate(Group = case_when(Group == 'O+' ~ 'PPI', #make group names more readable
+  mutate(Group = case_when(Group == 'O+' ~ 'Omeprazole', #make group names more readable
                            Group == 'C+' ~ 'Clindamycin',
-                           Group == 'CO+' ~ 'Clindamycin + PPI')) 
+                           Group == 'CO+' ~ 'Clind. + Omep.')) 
   # select only the samples that are present in the shared data
 matching <- inner_join(metadata, shared_sample_names, by = c('shared_names' = 'Group')) 
 

### 2019-08-22 code/ppi_alpha.R

@@ -11,15 +11,16 @@ alpha <- read_tsv(file="data/mothur/ppi.opti_mcc.groups.ave-std.summary",
 
 # Import metadata into data frame
 metadata <- read.table('data/process/ppi_metadata.txt', header = T, sep = '\t', stringsAsFactors = F) %>% 
-  filter(Group != "NA") #Exclude the mock community
+  filter(Group != "NA") %>% #Exclude the mock community
+  mutate(Group=factor(Group, levels=c("Clindamycin", "Clind. + Omep.", "Omeprazole"))) 
+
 
 # Combine metadata with alpha data frame
 meta_alpha <- inner_join(metadata, alpha, by=c('shared_names'='group')) %>% 
-  select(Group, sobs, shannon, day) %>% 
-  mutate(Group = as.factor(Group)) # make sure Group is treated as a factor
+  select(Group, sobs, shannon, day) 
 
 # Define color palette:
-color_scheme <- c("#d95f02", "#1b9e77", "#7570b3")
+color_scheme <- c("#d95f02", "#1b9e77",  "#7570b3")
 color_ppi <-  c("#7570b3") # Use for graphs looking at just the PPI group over time
 
 # Boxplots of shannon index (measure of community diversity) across treatment groups
@@ -82,7 +83,7 @@ b <- meta_alpha %>%
 
 # Linegraph of sobs (observed community richness) over time within the PPI group
 e <- meta_alpha %>% 
-  filter(Group == "PPI") %>% 
+  filter(Group == "Omeprazole") %>% 
   select(Group, day, sobs) %>% 
   ggplot(aes(x=day, y=sobs, group=Group, color=Group))+
   scale_colour_manual(values=color_ppi) +
@@ -94,7 +95,7 @@ e <- meta_alpha %>%
 
 #Test if Shannon diversity and richness is significantly different over D-7, 0, and 9 of the experiment
 ppi_day_alpha <- meta_alpha %>% 
-  filter( Group == "PPI") %>% # select just PPI group 
+  filter( Group == "Omeprazole") %>% # select just PPI group 
   filter(day == -7 | day == 0 | day ==9) %>% # select -7, 0, and -9 timepoints (beginning, middle and end of the experiment)
   mutate(day = as.factor(day))# turn day into factor so we can do post hoc comparisons
 
@@ -140,7 +141,7 @@ pairwise.wilcox.test(g=clind_day_alpha[["day"]], x=clind_day_alpha[["sobs"]], p.
 
[…]

### 2019-08-22 code/ppi_cfu_over_time.R

@@ -3,7 +3,7 @@ library(tidyverse)
 library(cowplot)
 
 # Define color palette:
-color_scheme <- c("#d95f02", "#1b9e77", "#7570b3")
+color_scheme <- c("#d95f02", "#1b9e77",  "#7570b3")
 
 # Import metadata into data frame
 metadata <- read.table('data/process/ppi_metadata.txt', header = T, sep = '\t', stringsAsFactors = F) %>% 
@@ -24,7 +24,7 @@ fancy_scientific <- function(l) {
 #Create data frame of CFU data by selecting group column & columns ending with CFU/g
 ppi_cfu <- select(metadata, Group, ends_with("CFU.g")) %>% 
   gather(Day, CFU, -Group) %>% 
-  mutate(Group=factor(Group, levels=c("Clindamycin", "Clindamycin + PPI", "PPI")),
+  mutate(Group=factor(Group, levels=c("Clindamycin", "Clind. + Omep.", "Omeprazole")),
          Day = as.numeric(str_replace(Day, "^[^.](\\d\\d?)[\\w\\.\\s\\/]+", "\\1")),
          Day = Day - 7, #Modify day notation so that C. difficile challenge day is represented as day 0.
          CFU=as.numeric(CFU) + 1) %>% 
@@ -44,7 +44,10 @@ title <-c(expression(paste(italic("C. difficile"), " colonization over time"))) 
 cfu_time <- ggplot(NULL) + 
   geom_point(ppi_cfu, mapping = aes(x= Day, y = CFU, color=Group, fill=Group), alpha = .04, size = 2, shape = 20, show.legend = FALSE, position = position_dodge(width = 0.6))+
   geom_line(ppi_cfu_summary, mapping = aes(x=Day, y=mean, color=Group))+
-  scale_colour_manual(values=color_scheme) +
+  scale_colour_manual(name=NULL, 
+                      values=color_scheme, 
+                      breaks=c("Clindamycin", "Clind. + Omep.", "Omeprazole"),
+                      labels=c("Clindamycin", "Clind. + Omep.", "Omeprazole"))+
   geom_linerange(show.legend=FALSE)+
   geom_hline(yintercept = 100, linetype=2) +
   labs(x='Days Post-Infection', y='CFU/g Feces')+

### 2019-08-22 code/ppi_pcoa.R

@@ -177,11 +177,8 @@ pcoa_after_challenge <- pcoa %>%
   labs(alpha = "Day")+
   geom_point() +
   theme_classic() +
-  xlim(-0.62, 0.62)+
-  ylim(-0.66, 0.66)+
   labs(x = "PCoA 1",
        y = "PCoA 2") +
   theme(plot.title = element_text(hjust = 0.5))+
-  theme(legend.position = c(0.90, 0.5)) #Move legend position
 save_plot("results/figures/after_abx_C.diff.png", pcoa_after_challenge, base_aspect_ratio = 2)
 

### 2019-08-22 code/ppi_taxa.R

@@ -20,7 +20,7 @@ library(broom)
 library(cowplot)
 
 # Define color palette:
-color_scheme <- c("#d95f02", "#1b9e77", "#7570b3")
+color_scheme <- c("#d95f02", "#1b9e77",  "#7570b3")
 color_ppi <-  c("#7570b3") # Use for graphs looking at just the PPI group over time
 color_cppi <- c("#1b9e77") # Use for graphs looking at just the Clindamycin + PPI group over time
 color_c <- c("#d95f02") # Use for graphs looking at just the Clindamycin group over time
@@ -28,7 +28,8 @@ color_day <- c('#c7c5e0' ,"#7570b3", "#514e7d") # Tints and Shades of PPI color 
 
 # Import metadata into data frame
 metadata <- read.table('data/process/ppi_metadata.txt', header = T, sep = '\t', stringsAsFactors = F) %>% 
-  filter(Group != "NA") #Exclude the mock community
+  filter(Group != "NA") %>% #Exclude the mock community
+  mutate(Group=factor(Group, levels=c("Clindamycin", "Clind. + Omep.", "Omeprazole"))) # make sure Group is treated as a factor
 
   
 # Import taxonomy into data frame and clean up taxonomy names
@@ -53,9 +54,7 @@ rarefy <- read_tsv(file="data/mothur/ppi.opti_mcc.groups.rarefaction") %>%
 
 #Join metadata to rarefy data frame
 metadata_rarefy <- inner_join(metadata, rarefy, by = c("shared_names" = "sample")) %>% 
-  filter(Group != "NA") %>% 
-  mutate(Group = as.factor(Group)) # make sure Group is treated as a factor
-
+  filter(Group != "NA") 
 
 #Plotting----
 #Plot rarefaction curves, the more parallel the curves to the x axis, the more confident you can be in the results
@@ -100,7 +99,10 @@ agg_phylum_data %>%
   filter(phylum %in% top_phyla) %>% 
   mutate(agg_rel_abund = agg_rel_abund + 1/6000) %>% 
   ggplot(aes(x= reorder(phylum, agg_rel_abund), y=agg_rel_abund, color=Group))+
-  scale_colour_manual(values=color_scheme) +
+  scale_colour_manual(name=NULL, 
[…]

### 2019-08-22 code/figure_1.R
Convert 1A to pdf so it will upload to GitHub

@@ -2,7 +2,7 @@ library(tidyverse)
 library(magick)
 library(cowplot)
 
-a <- ggdraw() + draw_image("results/figures/PPI_Exp_Scheme.png")
+a <- ggdraw() + draw_image("results/figures/PPI_Exp_Scheme.pdf")
 b <- ggdraw() + draw_image("results/figures/before_C._diff_challenge.png")
 c <- ggdraw() + draw_image("results/figures/families_prev_assoc_w_PPIs_-7.png")
 d <- ggdraw() + draw_image("results/figures/families_prev_assoc_w_PPIs_0.png")

### 2019-08-22 code/ppi_pcoa.R
Make changes so that ordination in 2B is generated the same way as 1B

@@ -124,23 +124,22 @@ before_plus_day_after_abx <- pcoa %>%	filter(day < 1) %>%
   ggsave("results/figures/before_plus_day_after_abx.png")
 
 # Figure 1B----
-# plot 1st 7 days: Includes abx treatment day. 
+# plot 1st 7 days from just the omeprazole-treated mice: Includes abx treatment day.  
 #Get a list of the shared_names from metadata file that are just the omeprazole treated mice before antibiotic treatment
 ppi_before_challenge_subset <- metadata %>% 
   filter(day < 1, Group == "Omeprazole") %>% 
   pull(shared_names)
 #groups specification for running dist.shared & pcoa commands on just this subset of samples in mothur". See README for instructions on analysis using mothur commands.
 # Opos+M1+D0-Opos+M2+D0-Opos+M3+D0-Opos+M4+D0-Opos+M5+D0-Opos+M1+D2-Opos+M2+D2-Opos+M3+D2-Opos+M4+D2-Cpos+M5+D-Opos+M1+D4-Opos+M3+D4-Opos+M4+D4-Opos+M5+D4-Opos+M11+D6-Opos+M2+D6-Opos+M3+D-Opos+M4+D6-Opos+M5+D6-Opos+M1+D7-Opos+M2+D7-Opos+M3+D7-Opos+M4+D7-Opos+M5+D7
-#Subsetted distance matrices and pcoa files outputted by mothur are located in data/mothur/subset
+#Subsetted distance matrices and pcoa files outputted by mothur are located in data/mothur/subset_1
 
 # Read in PCoA file for subset of PPI samples taken before C. difficile challenge & merge to metadata
-pcoa_subset <- read_tsv('data/mothur/subset/ppi.opti_mcc.braycurtis.0.03.lt.ave.pcoa.axes') %>%
+pcoa_subset1 <- read_tsv('data/mothur/subset_1/ppi.opti_mcc.braycurtis.0.03.lt.ave.pcoa.axes') %>%
   select(group, axis1, axis2) %>% rename(shared_names = group) %>% 
-  right_join(metadata, by = "shared_names") # join with cleaned up metadata for day, mouse, treatment
-# Select O+, C+, and CO+ groups. Convert Group letter labels to more descriptive labels
[…]

### 2019-08-22 code/clean_data.R
Removed old code and files that weren't used

@@ -1,82 +1,0 @@
-#####################
-# 
-# clean up labeling issues with data - Treat mice with abx/PPI and challenge with C difficile
-# 
-# input:
-#	data/mothur/stability.opti_mcc.0.03.subsample.shared
-#	PPI_Exp_Sample_List_2018.xlsx
-#	data/raw/Extraction Template_Plate #1#2#3.xlsx
-#	data/raw/Plate_4_Sarah_Omep2Box1.xlsx
-#	data/raw/SarahPPiNov2018_plateMap.xlsx
-#
-#
-# output:
-#	clean_data.txt
-#		
-#
-#####################
-
-library(tidyverse)
-library(readxl)
-
-# get a list of all the samples names from sequencing
-shared_sample_names <- read.table('data/mothur/stability.opti_mcc.0.03.subsample.shared', 
-		sep = '\t', header = T, stringsAsFactors = F) %>% 
-	select(Group) # select column with sample names of shared file which we will need to match our metadata to
-
-# load the metafile which has all of the original information 
-# with a column of what each sample was listed on the plate as
-sample_id_df <- read_xlsx('data/raw/PPI_Exp_Sample_List_2018.xlsx', sheet = 'Sheet1') %>% 
-	rename(ExpNumber = 'Exp. #', CollectionDay = 'Collection D', CageNumber = 'Cage #', 
-		M_F = 'M/F', OmpDailyDose = 'Omp. Daily Dose', MouseID = 'Mouse ID', TubeDayLabel = 'Tube label__1') %>% 
-	filter(ExpNumber == 2) %>% 
-	mutate(day = as.numeric(gsub('D', '', CollectionDay)),
- 		treatment = gsub('\\+', '', Group),
- 		sample_id = paste(treatment, MouseID, day, sep = '_'),
- 		OmpDailyDose = as.numeric(OmpDailyDose),
- 		mouse = as.numeric(MouseID)) %>% 
-	select(sample_id, day, treatment, mouse, plate_label = `Plate map label`, 
-		sex = M_F, cage = CageNumber, ear_mark = `Ear Mark`)
-
-# data was recorded in wide format, so tidy cdiff infection data so that one column per type of data
-exp_2_data <- read_xlsx('data/raw/08_20_18_Omep._#2_exp.xlsx', 
-		range = 'A1:AE15', sheet = 'cfu_final') %>%
[…]

### 2019-08-22 code/metadata_ncbi_formatting.R

@@ -1,20 +1,0 @@
-#Load packages
-library(tidyverse)
-library(lubridate)
-library(janitor)
-
-metadata <- read.table('data/process/ppi_metadata.txt', header = T, sep = '\t', stringsAsFactors = F) %>% 
-  #Unite 2 tube label columns into a new column "description". This conveys group name, mouse ID, and collection date.
-  unite(sample_title, Tube.label, Tube.label__1) %>% 
-  select(shared_names, sample_title)
-
-# Import ppi.files into data frame
-ncbi_metadata <- read.table('data/raw/formatting_metadata_for_ncbi.txt', header = T, sep = '\t', stringsAsFactors = F)  
-
-# Add metadata columns to mimarks file to finish filling out the information for all samples
-ncbi_metadata_merge <- left_join(metadata, ncbi_metadata, by = c("shared_names" = "shared.names"))
-write_tsv(ncbi_metadata_merge, 'data/raw/ncbi_metadata_merge.tsv', na = " ")
-
-# Paste into appropriate columns of ppi_SRA_metadata file to upload to NCBI
-
-

### 2019-08-22 code/mimarks_ncbi_info.R

@@ -1,27 +1,0 @@
-#Load packages
-library(tidyverse)
-library(lubridate)
-library(janitor)
-
-# Import metadata into data frame
-metadata <- read.table('data/process/ppi_metadata.txt', header = T, sep = '\t', stringsAsFactors = F) %>% 
-  #Unite 2 tube label columns into a new column "description". This conveys group name, mouse ID, and collection date.
-  unite(sample_title, Tube.label, Tube.label__1) %>% 
-  mutate(description = Group) %>% 
-  mutate(collection_date = excel_numeric_to_date(Date.Collected)) %>% #Convert excel numeric values to date values with excel_numeric_to_date function from the janitor package
-  mutate(experiment_day = day, #Reformat names of additional columns since they'll be included in mimarks file so that each sample will pass NCBI's uniqueness check.
-         sex = M.F,
-         exp_cage_number = Exp..cage,
-         mouse_id = Mouse.ID) %>% 
-  select(sample_title, description, collection_date, sex, experiment_day, mouse_id, exp_cage_number, shared_names) #Select columns that are missing from mimarks tsv file
-
-
-# Import MIMARKS ppi.tsv file into data frame
-mimarks <- read_tsv('data/raw/ppi.tsv')
-# Add metadata columns to mimarks file to finish filling out the information for all samples
-mimarks <- left_join(mimarks, metadata, by = c("#This is a tab-delimited file. Additional Documentation can be found at http://www.mothur.org/wiki/MIMarks_Data_Packages." = "shared_names"))
-write_tsv(mimarks, 'data/raw/ppi.tsv', na = " ")
-
-# To prepare for NCBI submission. Open up file in excel and move additional column names to their appropriate location on row 10
-          
-                     

### 2019-08-22 code/motility#2_cfu_over_time.R

@@ -1,60 +1,0 @@
-#Load packages
-library(tidyverse)
-library(dtw)
-library(vegan)
-
-#Define input files
-cfu <- 'data/Motility #2/11_14_18_motility_exp_2.xlsx'
-
-# Define output files
-plot_file <- 'results/figures/motility2_cfu.pdf'
-
-#Load in data
-motility2_data <- readxl::read_excel("data/Motility #2/11_14_18_motility_exp_2.xlsx", sheet = "cfu_final")
-
-#Create data frame of CFU data by selecting group column & columns ending with CFU/g
-motility2_cfu <- select(motility2_data, Group, ends_with("CFU/g")) 
-#make sure Group column is a factor vector and all CFU/g are treated as numeric vectors:
-motility2_cfu$Group <- as.factor(motility2_cfu$Group)
-motility2_cfu$`D0 C. difficile CFU/g` <- as.numeric(motility2_cfu$`D0 C. difficile CFU/g`)
-motility2_cfu$`D3 C. difficile CFU/g` <- as.numeric(motility2_cfu$`D3 C. difficile CFU/g`)
-motility2_cfu$`D5 C. difficile CFU/g` <- as.numeric(motility2_cfu$`D5 C. difficile CFU/g`)
-str(motility2_cfu)
-
-
-#Format CFU over time
-motility2_cfu[,2:ncol(motility2_cfu)] <- log10(motility2_cfu[,2:ncol(motility2_cfu)] + 1)
-cfu_median <- aggregate(motility2_cfu[,2:ncol(motility2_cfu)], by=list(motility2_cfu$Group), FUN=quantile, probs=0.5, na.rm = T)
-rownames(cfu_median) <- cfu_median$Group.1
-cfu_median$Group.1 <- NULL
-cfu_median <- as.data.frame(t(cfu_median))
-cfu_q25 <- aggregate(motility2_cfu[,2:ncol(motility2_cfu)], by=list(motility2_cfu$Group), FUN=quantile, probs=0.25, na.rm = T)
-rownames(cfu_q25) <- cfu_q25$Group.1
-cfu_q25$Group.1 <- NULL
-cfu_q25 <- as.data.frame(t(cfu_q25))
[…]

### 2019-08-22 code/ppi_cfu_stats.R

@@ -1,48 +1,0 @@
-library(tidyverse)
-library(broom)
-
-#Load in metadata file and tidy CFU/g data from each timepoint
-ppi_cfu <- select(metadata, Group, ends_with("CFU.g")) %>% 
-  gather(Day, CFU, -Group) %>% 
-  mutate(Group=factor(Group, levels=c("PPI", "Clindamycin", "Clindamycin + PPI")),
-         Day = as.numeric(str_replace(Day, "^[^.](\\d\\d?)[\\w\\.\\s\\/]+", "\\1")),
-         Day = Day - 7, #Modify day notation so that C. difficile challenge day is represented as day 0.
-         CFU=as.numeric(CFU) + 1) %>% 
-  ungroup
-
-#Analysis
-get_mean_cfu <- function(x){
-      x %>% 
-        group_by(Group) %>% 
-        summarize(mean=mean(CFU)) %>% 
-        spread(key=Group, value=mean)
-}
-
-select(ppi_cfu, Group, Day, CFU) %>% 
-  group_by(Day) %>% 
-  nest() %>% 
-  mutate(model=map(data, ~kruskal.test(x=.x$CFU, g=as.factor(.x$Group)) %>% tidy())) %>% 
-  mutate(mean= map(data, get_mean_cfu)) %>% 
-  unnest(model, mean) %>% 
-  mutate(model=map(data,
-                   ~wilcox.test(x=.x$CFU, g=as.factor(.x$Group), p.adjust.method="BH") %>% 
-                     tidy() %>% 
-                     mutate(compare=paste(group1, group2, sep="-")) %>% 
-##                     select(-group1, -group2) %>% 
-##                     spread(key=compare, value=p.value)
-                   )
-         ) %>% 
-##  unnest(model) %>% 
-##  select(-data, -parameter, -statistic) %>% 
-  write_tsv("data/process/ppi_cfu_stats.tsv")
-
-cfu_tests <- ppi_cfu %>% 
-  group_by(Day) %>% 
-  do(tidy(kruskal.test(CFU~Group, data=.))) %>% 
-  ungroup() %>% 
-  mutate(p.value.adj=p.adjust(p.value, method="BH")) %>% 
-  arrange(p.value.adj)
-
-for (i in ppi_cfu$Day){
-print(pairwise.wilcox.test(ppi_cfu$CFU, ppi_cfu$Group, p.adjust.method = "BH"))
-}

### 2019-08-22 code/ppi_exp2_community.R

@@ -1,156 +1,0 @@
-#####################
-# 
-# analysis of experiment - Treat mice with abx/PPI and challenge with C difficile
-# 
-# input:
-#	data/mothur/ppi.opti_mcc.0.03.subsample.shared
-#	data/mothur/ppi.opti_mcc.thetayc.0.03.lt.ave.nmds.axes
-#	data/process/ppi_metadata.txt
-#
-# output:
-#	plotly html plot
-#		running code will open plot in browser
-#		plot can be modified and saved from browser
-#
-#####################
-
-library(tidyverse)
-library(cowplot)
-
-shared_df <- read.table('data/mothur/ppi.opti_mcc.0.03.subsample.shared', sep = '\t', header = T, 
-	stringsAsFactors = F, row.names = 'Group') # set column with sample name to match
-metadata <- read.table('data/process/ppi_metadata.txt', header = T, sep = '\t', stringsAsFactors = F) %>% 
-	# convert Group letter labels to more descriptive labels
-	mutate(Group = case_when(Group == 'O+' ~ 'PPI', 
-		Group == 'C+' ~ 'Clindamycin',
-		Group == 'CO+' ~ 'Clindamycin + PPI'),
-		day = day - 7) # C difficile challenge on day 7
-
-mouse_label_df <- metadata %>% 
-  select(Mouse.ID, Group) %>% 
-  unique %>% 
-  group_by(Group) %>% 
-  arrange(Mouse.ID) %>% 
-  mutate(mouse_position = 1:length(Mouse.ID)) %>% 
-  ungroup %>% 
-  select(Mouse.ID, mouse_position)
-
-meta_df <- left_join(metadata, mouse_label_df, by = 'Mouse.ID') %>% rename(treatment = Group)
-
-## need to get taxonomy file
-#taxonomy_function <- 'code/sum_otu_by_taxa.R'
-#source(taxonomy_function)
-#taxonomy_df <- read.table(taxonomy_file, sep = '\t', header = T, stringsAsFactors = F)
-
-shared_df <- select(shared_df, -label, -numOtus, )
-
-sample_summed_counts <- apply(shared_df, 1, sum)
-rel_abund <- data.frame(100 * shared_df/sample_summed_counts) %>% 
-  mutate(shared_names = rownames(shared_df))
-
-pre_post_end_df <- meta_df %>% 
-  right_join(select(rel_abund, shared_names), by = "shared_names") %>% 
[…]

### 2019-08-22 code/ppi_nmds.R

@@ -1,122 +1,0 @@
-#####################
-# 
-# analysis of experiment - Treat mice with abx and/or PPI and challenge with C difficile
-# 
-# input:
-#	data/mothur/ppi.opti_mcc.0.03.subsample.shared
-#	data/mothur/ppi.opti_mcc.thetayc.0.03.lt.ave.nmds.axes
-#	data/process/ppi_metadata.txt
-#
-# output:
-#	plotly html plot
-#		running code will open plot in browser
-#		plot can be modified and saved from browser
-#
-#####################
-
-library(tidyverse)
-library(plotly)
-
-shared <- read.table('data/mothur/ppi.opti_mcc.0.03.subsample.shared', sep = '\t', header = T, stringsAsFactors = F) %>% 
-	rename(shared_names = Group) # set column with sample name to matching
-
-metadata <- read.table('data/process/ppi_metadata.txt', header = T, sep = '\t', stringsAsFactors = F)
-
-# read in 2d nmds data
-nmds <- read.table('data/mothur/ppi.opti_mcc.thetayc.0.03.lt.ave.nmds.axes', 
-		sep = '\t', header = T, stringsAsFactors = F) %>% rename(shared_names = group) %>% 
-	right_join(metadata, by = "shared_names") # join with cleaned up metadata for day, mouse, treatment
-	
-
-# plot for before abx
-nmds %>% 
-	# convert Group letter labels to more descriptive labels
-	mutate(Group = case_when(Group == 'O+' ~ 'PPI', 
-		Group == 'C+' ~ 'Clindamycin',
-		Group == 'CO+' ~ 'Clindamycin + PPI'),
-	# modify day number for plotting by size
-		size = case_when(day <= 6 ~ 1,
-		day == 7 ~ 2,
-		day > 7 ~ day - 4)) %>% 
-	# look at days before 7 to only include pre-clinda treatment. Mice were treated with clinda on D6, so stool
-  # sample from D6 is still pre-clinda.
-	filter(day < 7) %>% 
-	plot_ly(x = ~axis1, y = ~axis2, type = 'scatter', mode = 'markers',
-		color = ~Group, size = ~size, 
-		marker = list(symbol = 'circle', sizemode = 'diameter'), sizes = c(10,20)) %>% # sizes determines the range of sizes c(smallest, largest)
-		layout(title = 'Community by treatment')
-
-# plot for before abx
-nmds %>% 
[…]

### 2019-08-22 code/ppi_pcoa.R
Remove code to generate unused pcoa plots

@@ -36,93 +36,6 @@ color_c <- c("#d95f02") # Use for graphs looking at just the Clindamycin group o
 color_mouse <- c("#7fc97f", "#beaed4", "#fdc086", "#386cb0", "#f0027f") # To differentiate between 5 individual mice
 shape_mouse <- c(16, 17, 15, 3, 7)
 
-#plot all samples
-pcoa %>%  
-  ggplot(aes(x=axis1, y=axis2, color=Group, alpha = day)) +
-  scale_colour_manual(name=NULL, 
-                      values=color_scheme, 
-                      breaks=c("Clindamycin", "Clind. + Omep.", "Omeprazole"),
-                      labels=c("Clindamycin", "Clind. + Omep.", "Omeprazole"))+
-  geom_point() +
-#  geom_path() + #Add's lines to plots but looks messy  
-  theme_classic()+
-  labs(title="All groups, all timepoints") +
-  theme(plot.title = element_text(hjust = 0.5))
-
-
-#plot just PPI samples
-pcoa %>% filter(Group == "Omeprazole") %>%  
-  ggplot(aes(x=axis1, y=axis2, color=Group, alpha = day)) +
-  scale_colour_manual(values=color_ppi) +
-  geom_point()+
-#  geom_path() + #Add's lines to plots but looks messy
-  theme_classic() +
-  labs(title="PPI-treated mice") +
-  theme(plot.title = element_text(hjust = 0.5))
-
-#plot just PPI samples over the first 7 days
-pcoa %>% filter(Group == "Omeprazole") %>%  filter(day < 8) %>% 
-  ggplot(aes(x=axis1, y=axis2, color=Group, alpha = day)) +
-  scale_colour_manual(values=color_ppi) +
-  geom_point()+
-#  geom_path() + #Add's lines to plots but looks messy
-  theme_classic() +
-  labs(title="PPI-treated mice over first 7 days") +
-  theme(plot.title = element_text(hjust = 0.5))
-
-#plot just Clindamycin samples
-pcoa %>% filter(Group == "Clindamycin") %>%  
-  ggplot(aes(x=axis1, y=axis2, color=Group, alpha = day)) +
-  scale_colour_manual(values=color_c) +
-  geom_point() +
-#  geom_path() + #Add's lines to plots but looks messy 
-  theme_classic()+
-  labs(title="Clindamycin-treated mice") +
-  theme(plot.title = element_text(hjust = 0.5))
-
-#plot just Clindamycin + PPI samples
[…]

### 2019-08-22 code/ppi_alpha.R
Clean up alpha analysis & plot file

@@ -23,77 +23,7 @@ meta_alpha <- inner_join(metadata, alpha, by=c('shared_names'='group')) %>%
 color_scheme <- c("#d95f02", "#1b9e77",  "#7570b3")
 color_ppi <-  c("#7570b3") # Use for graphs looking at just the PPI group over time
 
-# Boxplots of shannon index (measure of community diversity) across treatment groups
-a <- meta_alpha %>% 
-  ggplot(aes(x= reorder(Group, shannon), y=shannon, colour=Group)) +
-  scale_colour_manual(values=color_scheme) +
-  geom_boxplot(outlier.shape = NA)+
-  geom_jitter(shape=19, size=1, alpha=0.7, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.1)) +
-  labs(title=NULL, 
-       x=NULL,
-       y="Shannon")+
-  ylim(0, 3.5)+
-  theme_classic()+
-  theme(legend.title=element_blank(), legend.position="bottom")
-
-# Test if Shannon and sobs (richness) data are normally distributed using Shapiro-Wilk normality test
-shapiro.test(meta_alpha$shannon) # Test for Shannon diversity
-# P < 0.05, means the data are not normally distributed and we should do non parametric test (Kruskal-Wallis)
-
-#Hypothesis testing between groups, all timepoints. Since data is not normally distributed use the Kruskal-Wallis test
-group_shannon_kruskal <- kruskal.test(shannon~Group, data=meta_alpha)
-# P = 1.132e-05, so do post hoc test to determine which comparisons are significant
-pairwise.wilcox.test(g=meta_alpha[["Group"]], x=meta_alpha[["shannon"]], p.adjust.method="BH")
-## PPI vs clindamycin and PPI vs clindamycin+PPI are significantly different (P = 4.5e-05 and P = 4.5e-05)
-
-# Boxplots of sobs (measure of community richness) across treatment groups
-d <- meta_alpha %>% 
-  ggplot(aes(x= reorder(Group, sobs), y=sobs, color=Group)) +
-  scale_colour_manual(values=color_scheme) +
-  geom_boxplot(outlier.shape = NA)+
-  geom_jitter(shape=19, size=1, alpha=0.7, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.1)) +
-  labs(title=NULL, 
-       x=NULL,
-       y="richness")+
-  ylim(0, 100)+
-  theme_classic()+
-  theme(legend.position="none")
-
[…]

### 2019-08-22 code/ppi_pcoa.R
Clean up code and make sure all figures are still correctly generated

@@ -1,18 +1,5 @@
 #####################
-# 
-# analysis of experiment - Treat mice with abx and/or PPI and challenge with C difficile
-# 
-# input:
-#	data/mothur/ppi.opti_mcc.braycurtis.0.03.lt.ave.pcoa.axes
-# data/mothur/ppi.opti_mcc.braycurtis.0.03.lt.ave.dist
-#	data/process/ppi_metadata.txt
-#
-# output:
-#	pcoa plots
-#		
-#
-#####################
-
+#Setup----
 library(tidyverse)
 library(cowplot)
 

### 2019-08-22 code/ppi_taxa.R

@@ -1,18 +1,4 @@
 #####################
-# 
-# Treat mice with abx and/or PPI and challenge with C difficile and examine how taxa change
-# 
-# input:
-#	data/mothur/ppi.taxonomy
-#	data/process/ppi_metadata.txt
-# data/mothur/ppi.opti_mcc.groups.rarefaction
-# data/mothur/ppi.opti_mcc.0.03.subsample.shared
-#
-# output:
-#	To Be Decided
-#		Plots looking at specific taxa levels between treatment groups
-#		
-#
 #####################
 #Setup----
 library(tidyverse)
@@ -44,30 +30,6 @@ taxonomy <- read_tsv(file="data/mothur/ppi.taxonomy") %>%
 # Separate taxonomic levels into separate columns according to semi-colon.
   separate(taxonomy, into=c("kingdom", "phylum", "class", "order", "family", "genus"), sep=';')
 
-
-#Look at rarefaction curves, import rarefaction data frame
-rarefy <- read_tsv(file="data/mothur/ppi.opti_mcc.groups.rarefaction") %>% 
-  select(-contains("lci-"), -contains("hci-")) %>% 
-  gather(-numsampled, key=sample, value=sobs) %>% #not including numsampled, gather samples intow two columns
-  mutate(sample=str_replace_all(sample, patter="0.03-", replacement="")) %>% #drop prefix 0.03
-  drop_na()
-
-#Join metadata to rarefy data frame
-metadata_rarefy <- inner_join(metadata, rarefy, by = c("shared_names" = "sample")) %>% 
-  filter(Group != "NA") 
-
-#Plotting----
-#Plot rarefaction curves, the more parallel the curves to the x axis, the more confident you can be in the results
-ggplot(metadata_rarefy, aes(x=numsampled, y=sobs, group=shared_names, color=Group))+
-  scale_colour_manual(values=color_scheme) +
-  geom_vline(xintercept=3000, color="gray", size=2) +
-  geom_line()+
-  coord_cartesian(xlim=c(0, 10000), ylim=c(0,200))+
-  labs(subtitle="Vertical line indicates the number of sequences that samples were rarefied to",
-       x="Number of sequences sampled per mouse", 
-       y="Number of OTUs per mouse") +
-  theme_classic()
-
[…]

### 2019-09-08 submission/manuscript.Rmd
Fix typos found while working on poster

@@
 **Omeprazole treatment did not promote susceptibility to *C.
-difficile infection* in mice.** Next, we examined whether omeprazole treatment altered susceptibility to *C.
+difficile* infection in mice.** Next, we examined whether omeprazole treatment altered susceptibility to *C.
 difficile* infection in mice.
@@
 Additionally, we have previously demonstrated that mice from our breeding colony are resistant to *C.
-difficile* 630 colonization without antibiotic treatment [@Jenior2017], ensuring there was not already partial susceptiblity to *C.
+difficile* 630 colonization without antibiotic treatment [@Jenior2017], ensuring there was not already partial susceptibility to *C.
 difficile* before treatment.
-The additional controls in our study allowed us to assess the contibution of omeprazole alone to *C.
+The additional controls in our study allowed us to assess the contribution of omeprazole alone to *C.
 difficile* susceptibility in mice.
@@
 This research was supported by NIH grant U01AI12455.
-We would also like to thank the Unit for Laboratory Animal Medicine at the University of Michigan for maintaining our mouse colony and providing the infrastructure and support for performing our mouse expeirments.
+We would also like to thank the Unit for Laboratory Animal Medicine at the University of Michigan for maintaining our mouse colony and providing the infrastructure and support for performing our mouse experiments.
 The authors are also thankful to members of the Schloss lab for helpful discussions throughout the process of designing the experiment, analyzing the results, crafting the figures, and drafting of the manuscript.
@@
 **Animals.** All mouse experiments were performed with 7- to 12-week-old C57BL/6 male and female mice.
-Each experimental group of mice was spit between 2 cages with 2-3 mice housed per cage and male and female mice housed separately.
+Each experimental group of mice was split between 2 cages with 2-3 mice housed per cage and male and female mice housed separately.
 All animal experiments were approved by the University of Michigan Animal Care and Use Committee (IACUC) under protocol number PRO00006983.
@@
 **16S rRNA gene sequencing.** DNA for 16S rRNA gene sequencing was extracted from 10-50 mg fecal pellet from each mouse using the DNeasy Powersoil HTP 96 Kit (Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf).
-The 16S rRNA sequencing libary was prepared as described previously [@Kozich2013].
+The 16S rRNA sequencing library was prepared as described previously [@Kozich2013].
 In brief, the ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used as a mock community [@Sze2019] and water was used as a negative control.
@@
 
-**Statistical Analysis.** To test for differences in relative abundances in families and genera across our 3 different treatment groups at different timepoints (Clindamycin, Clind.
-+ Omep., and Omeprazole on Day -7, 0, 2, and 9) or within the Omeprazole treatment group across 3 timepoints (Day -7, 0, and 9), we used a Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
+**Statistical Analysis.** To test for differences in relative abundances in families and genera across our 3 different treatment groups at different timepoints (Clindamycin, Clindamycin + Omeprazole, and Omeprazole on Day -7, 0, 2, and 9) or within the Omeprazole treatment group across 3 timepoints (Day -7, 0, and 9), we used a Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
 
-**Code availability.** The code for all sequence processing and analysis steps as well as an Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Tomkovich_PPI_mSphere_2019.
+**Code availability.** The code for all sequence processing and analysis steps as well as a Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Tomkovich_PPI_mSphere_2019.
 
@@
 B.
-Principal Coordinates Analaysis (PCoA) of Bray-curtis distances from stool samples of mice in the omeprazole treatment group during the intial 7 days of the experiment.
+Principal Coordinates Analysis (PCoA) of Bray-Curtis distances from stool samples of mice in the omeprazole treatment group during the initial 7 days of the experiment.
 Each color represents stool samples from the same mouse and lines connect sequentially collected samples.
@@
 Each circle represents an individual mouse.
-There were no significant differences acoss treatment groups for any of the identified families in the sequence data at day -7 (all P-values > 0.448) and day 0 (all P-values > 0.137), analyzed by Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
+There were no significant differences across treatment groups for any of the identified families in the sequence data at day -7 (all P-values > 0.448) and day 0 (all P-values > 0.137), analyzed by Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.
 For C-D, the grey vertical line indicates the limit of detection.
@@
 B.
-PCoA of of Bray-curtis distances from stool samples collected after antibiotic treatment (last 9 days of the experiment).
+PCoA of of Bray-Curtis distances from stool samples collected after antibiotic treatment (last 9 days of the experiment).
 Transparency of the symbol corresponds to treatment day.
@@
 **Figure S1.
-Families within omeprazole treated mice fluctuate over time with no overall trend in either direction.** Relative abundance over time for *Lactobacillaceae* (A) and *Ruminococcaceae* (B), 2 of the PPI-associated familes from human PPI studies across all 3 treatment groups.
+Families within omeprazole treated mice fluctuate over time with no overall trend in either direction.** Relative abundance over time for *Lactobacillaceae* (A) and *Ruminococcaceae* (B), 2 of the PPI-associated families from human PPI studies across all 3 treatment groups.
 Each point represents the relative abundance for an individual mouse stool sample, while the lines represent the mean relative abundances for each treatment group of mice.
@@
 **Figure S2.
-Microbiota diversity and richness decrease with antibiotic treatment but remain relatively constant with omeprazole treatment.** A.
-Boxplots of the Shannon Diversity Index values (A) and number of observed OTUs (B) for each group of mice over 3 timepoints (Day -7, 0, and 9).
+Microbiota diversity and richness decrease with antibiotic treatment but remain relatively constant with omeprazole treatment.** Boxplots of the Shannon Diversity Index values (A) and number of observed OTUs (B) for each group of mice over 3 timepoints (Day -7, 0, and 9).
 Each circle represents the value for a stool sample from an individual mouse.

### 2019-09-09 submission/manuscript.Rmd
Edits from Pat

@@
 difficile* 630 spores.
-While *C.
+Although *C.
 difficile* colonized the clindamycin-treated mice, it did not colonize the omeprazole-treated mice (Figure 2A).
@@
 
-**Conclusions.** The PPI omeprazole did not meaningfully impact the gut microbiota and did not promote *C.
+**Conclusions.** The PPI omeprazole did not meaningfully impact the structure of the gut microbiota and did not promote *C.
 difficile* infection in mice.
@@
 difficile* in mice not treated with antibiotics [@Kaur2007; @kaur2011effect].
-The other mouse study demonstrated the antibiotic/PPI esomeprazole-treated mice developed more severe CDIs compared to antibiotic-treated mice, but the researchers did not have a group treated with just esomeprazole for comparison [@hung2015proton].
+The other mouse study demonstrated antibiotic/esomeprazole-treated mice developed more severe CDIs compared to antibiotic-treated mice, but the researchers did not have a group treated with just esomeprazole for comparison [@hung2015proton].
 We tested the same high 40 mg/kg PPI dose and expanded pre-treatment to 7 days before challenge to test the impact of omeprazole treatment alone on our CDI mouse model.
@@
 difficile* colonization resistance in mice.
-16S rRNA sequencing suggested that *Streptococcus* and *Enterococcus* are rare genera in our C57BL/6 mouse colony and variation in these 2 genera have been observed across other facilities and vendors [@Rausch2016; @Robertson2019].
-These genera could be important contributors to the associations between PPIs and CDIs in humans, and could be a contributing factor to our observation that PPI treatment had no effect on *C.
+16S rRNA sequencing suggested that *Streptococcus* and *Enterococcus* are rare genera in our C57BL/6 mouse colony.
+These two genera could be important contributors to the associations between PPIs and CDIs in humans, and could be a contributing factor to our observation that PPI treatment had no effect on *C.
 difficile* colonization in our CDI mouse model.

### 2019-09-09 submission/manuscript.tex
Edits from Pat

@@
 \textbf{Omeprazole treatment did not promote susceptibility to \emph{C.
-difficile infection} in mice.} Next, we examined whether omeprazole
+difficile} infection in mice.} Next, we examined whether omeprazole
 treatment altered susceptibility to \emph{C. difficile} infection in
@@
 challenged with 10\textsuperscript{3} \emph{C. difficile} 630 spores.
-While \emph{C. difficile} colonized the clindamycin-treated mice, it did
-not colonize the omeprazole-treated mice (Figure 2A). Interestingly,
+Although \emph{C. difficile} colonized the clindamycin-treated mice, it
+did not colonize the omeprazole-treated mice (Figure 2A). Interestingly,
 only 1 cage of mice that received both omeprazole and clindamycin were
@@
 \textbf{Conclusions.} The PPI omeprazole did not meaningfully impact the
-gut microbiota and did not promote \emph{C. difficile} infection in
-mice. Our findings that omeprazole treatment had minimal impact on the
-fecal microbiome were comparable to another PPI mouse study that
-indicated the PPI lansoprazole had more of an effect on the small
-intestinal microbiota compared to the fecal microbiota (12). The same
-group demonstrated lansoprazole treatment increased the stomach pH in
-mice (12), which may improve survival of bacteria passing through the
+structure of the gut microbiota and did not promote \emph{C. difficile}
+infection in mice. Our findings that omeprazole treatment had minimal
+impact on the fecal microbiome were comparable to another PPI mouse
+study that indicated the PPI lansoprazole had more of an effect on the
+small intestinal microbiota compared to the fecal microbiota (12). The
+same group demonstrated lansoprazole treatment increased the stomach pH
+in mice (12), which may improve survival of bacteria passing through the
 stomach. We did not find significant changes in the relative abundances
@@
 however there was detectable \emph{C. difficile} in mice not treated
-with antibiotics (13, 14). The other mouse study demonstrated the
-antibiotic/PPI esomeprazole-treated mice developed more severe CDIs
-compared to antibiotic-treated mice, but the researchers did not have a
-group treated with just esomeprazole for comparison (15). We tested the
-same high 40 mg/kg PPI dose and expanded pre-treatment to 7 days before
+with antibiotics (13, 14). The other mouse study demonstrated
+antibiotic/esomeprazole-treated mice developed more severe CDIs compared
+to antibiotic-treated mice, but the researchers did not have a group
+treated with just esomeprazole for comparison (15). We tested the same
+high 40 mg/kg PPI dose and expanded pre-treatment to 7 days before
 challenge to test the impact of omeprazole treatment alone on our CDI
@@
 colonization without antibiotic treatment (16), ensuring there was not
-already partial susceptiblity to \emph{C. difficile} before treatment.
+already partial susceptibility to \emph{C. difficile} before treatment.
 The additional controls in our study allowed us to assess the
-contibution of omeprazole alone to \emph{C. difficile} susceptibility in
-mice.\\
+contribution of omeprazole alone to \emph{C. difficile} susceptibility
+in mice.\\
 Our study also extended previous work examining PPIs and \emph{C.
@@
 sequencing suggested that \emph{Streptococcus} and \emph{Enterococcus}
-are rare genera in our C57BL/6 mouse colony and variation in these 2
-genera have been observed across other facilities and vendors (17, 18).
-These genera could be important contributors to the associations between
-PPIs and CDIs in humans, and could be a contributing factor to our
-observation that PPI treatment had no effect on \emph{C. difficile}
-colonization in our CDI mouse model. Gastrointestinal physiological
-differences, particularly the higher stomach pH in mice (pH 3-4)
-compared to humans (pH 1) (19) could also explain why omeprazole had a
-limited impact on the murine microbiome. The microbiota and
-physiological differences between humans and mice may limit the
-usefulness of employing mouse models to study the impact of PPIs on the
-microbiota and CDIs.\\
+are rare genera in our C57BL/6 mouse colony. These two genera could be
+important contributors to the associations between PPIs and CDIs in
+humans, and could be a contributing factor to our observation that PPI
+treatment had no effect on \emph{C. difficile} colonization in our CDI
+mouse model. Gastrointestinal physiological differences, particularly
+the higher stomach pH in mice (pH 3-4) compared to humans (pH 1) (17)
+could also explain why omeprazole had a limited impact on the murine
+microbiome. The microbiota and physiological differences between humans
+and mice may limit the usefulness of employing mouse models to study the
+impact of PPIs on the microbiota and CDIs.\\
 Beyond microbiome differences, factors such as age, body mass index,
@@
 determining causality and whether other risk factors play a role
-challenging (20). Future studies are needed to determine whether age,
+challenging (18). Future studies are needed to determine whether age,
 other comorbidities and bacterial strains that are less common in mice
@@
 Michigan for maintaining our mouse colony and providing the
-infrastructure and support for performing our mouse expeirments. The
+infrastructure and support for performing our mouse experiments. The
 authors are also thankful to members of the Schloss lab for helpful
@@
 12-week-old C57BL/6 male and female mice. Each experimental group of
-mice was spit between 2 cages with 2-3 mice housed per cage and male and
-female mice housed separately. All animal experiments were approved by
-the University of Michigan Animal Care and Use Committee (IACUC) under
-protocol number PRO00006983.
+mice was split between 2 cages with 2-3 mice housed per cage and male
+and female mice housed separately. All animal experiments were approved
+by the University of Michigan Animal Care and Use Committee (IACUC)
+under protocol number PRO00006983.
 
@@
 Powersoil HTP 96 Kit (Qiagen) and an EpMotion 5075 automated pipetting
-system (Eppendorf). The 16S rRNA sequencing libary was prepared as
-described previously (21). In brief, the ZymoBIOMICS\textsuperscript{TM}
+system (Eppendorf). The 16S rRNA sequencing library was prepared as
+described previously (19). In brief, the ZymoBIOMICS\textsuperscript{TM}
 Microbial Community DNA Standard (Zymo, CA, USA) was used as a mock
-community (22) and water was used as a negative control. The V4
+community (20) and water was used as a negative control. The V4
 hypervariable region of the 16S rRNA gene was amplified with Accuprime
 Pfx DNA polymerase (Thermo Fisher Scientific) using previously described
-custom barcoded primers (21). The 16S rRNA amplicon library was
+custom barcoded primers (19). The 16S rRNA amplicon library was
 sequenced with the MiSeq (Illumina). Amplicons were cleaned up and
@@
 \textbf{16S rRNA gene sequence analysis.} mothur (v1.40.5) was used for
-all sequence processing steps (23) using a previously published protocol
-(21). In brief, forward and reverse reads for each sample were combined
+all sequence processing steps (21) using a previously published protocol
+(19). In brief, forward and reverse reads for each sample were combined
 and low-quality sequences and chimeras were removed. Duplicate sequences
@@
 abundances in families and genera across our 3 different treatment
-groups at different timepoints (Clindamycin, Clind. + Omep., and
-Omeprazole on Day -7, 0, 2, and 9) or within the Omeprazole treatment
-group across 3 timepoints (Day -7, 0, and 9), we used a Kruskal-Wallis
-test with a Benjamini-Hochberg correction for multiple comparisons.
+groups at different timepoints (Clindamycin, Clindamycin + Omeprazole,
+and Omeprazole on Day -7, 0, 2, and 9) or within the Omeprazole
+treatment group across 3 timepoints (Day -7, 0, and 9), we used a
+Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple
+comparisons.
 
 \textbf{Code availability.} The code for all sequence processing and
-analysis steps as well as an Rmarkdown version of this manuscript is
+analysis steps as well as a Rmarkdown version of this manuscript is
 available at
@@
 
-\hypertarget{ref-Rausch2016}{}
-17. \textbf{Rausch P}, \textbf{Basic M}, \textbf{Batra A},
-\textbf{Bischoff SC}, \textbf{Blaut M}, \textbf{Clavel T},
-\textbf{Gläsner J}, \textbf{Gopalakrishnan S}, \textbf{Grassl GA},
-\textbf{Günther C}, \textbf{Haller D}, \textbf{Hirose M},
-\textbf{Ibrahim S}, \textbf{Loh G}, \textbf{Mattner J}, \textbf{Nagel
-S}, \textbf{Pabst O}, \textbf{Schmidt F}, \textbf{Siegmund B},
-\textbf{Strowig T}, \textbf{Volynets V}, \textbf{Wirtz S},
-\textbf{Zeissig S}, \textbf{Zeissig Y}, \textbf{Bleich A},
-\textbf{Baines JF}. 2016. Analysis of factors contributing to variation
-in the c57bl/6J fecal microbiota across german animal facilities.
-International Journal of Medical Microbiology \textbf{306}:343--355.
-doi:\href{https://doi.org/10.1016/j.ijmm.2016.03.004}{10.1016/j.ijmm.2016.03.004}.
-
-\hypertarget{ref-Robertson2019}{}
-18. \textbf{Robertson SJ}, \textbf{Lemire P}, \textbf{Maughan H},
-\textbf{Goethel A}, \textbf{Turpin W}, \textbf{Bedrani L},
-\textbf{Guttman DS}, \textbf{Croitoru K}, \textbf{Girardin SE},
-\textbf{Philpott DJ}. 2019. Comparison of co-housing and littermate
-methods for microbiota standardization in mouse models. Cell Reports
-\textbf{27}:1910--1919.e2.
-doi:\href{https://doi.org/10.1016/j.celrep.2019.04.023}{10.1016/j.celrep.2019.04.023}.
-
 \hypertarget{ref-Hugenholtz2017}{}
-19. \textbf{Hugenholtz F}, \textbf{Vos WM de}. 2017. Mouse models for
+17. \textbf{Hugenholtz F}, \textbf{Vos WM de}. 2017. Mouse models for
 human intestinal microbiota research: A critical evaluation. Cellular
@@
 \hypertarget{ref-Eze2017}{}
-20. \textbf{Eze P}, \textbf{Balsells E}, \textbf{Kyaw MH}, \textbf{Nair
+18. \textbf{Eze P}, \textbf{Balsells E}, \textbf{Kyaw MH}, \textbf{Nair
 H}. 2017. Risk factors for clostridium difficile infections an overview
@@
 \hypertarget{ref-Kozich2013}{}
-21. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+19. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \hypertarget{ref-Sze2019}{}
-22. \textbf{Sze MA}, \textbf{Schloss PD}. 2019. The impact of DNA
+20. \textbf{Sze MA}, \textbf{Schloss PD}. 2019. The impact of DNA
 polymerase and number of rounds of amplification in PCR on 16S rRNA gene
@@
 \hypertarget{ref-Schloss2009}{}
-23. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+21. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 stool was quantified daily through 6 days post-infection by anaerobic
-culture. B. Principal Coordinates Analaysis (PCoA) of Bray-curtis
+culture. B. Principal Coordinates Analysis (PCoA) of Bray-Curtis
 distances from stool samples of mice in the omeprazole treatment group
-during the intial 7 days of the experiment. Each color represents stool
+during the initial 7 days of the experiment. Each color represents stool
 samples from the same mouse and lines connect sequentially collected
@@
 omeprazole treatment (D). Each circle represents an individual mouse.
-There were no significant differences acoss treatment groups for any of
+There were no significant differences across treatment groups for any of
 the identified families in the sequence data at day -7 (all P-values
@@
 CFU/g for individual mice within each group. The black dashed line
-indicates the limit of detection. B. PCoA of of Bray-curtis distances
+indicates the limit of detection. B. PCoA of of Bray-Curtis distances
 from stool samples collected after antibiotic treatment (last 9 days of
@@
 \emph{Lactobacillaceae} (A) and \emph{Ruminococcaceae} (B), 2 of the
-PPI-associated familes from human PPI studies across all 3 treatment
+PPI-associated families from human PPI studies across all 3 treatment
 groups. Each point represents the relative abundance for an individual
@@
 and richness decrease with antibiotic treatment but remain relatively
-constant with omeprazole treatment.} A. Boxplots of the Shannon
-Diversity Index values (A) and number of observed OTUs (B) for each
-group of mice over 3 timepoints (Day -7, 0, and 9). Each circle
-represents the value for a stool sample from an individual mouse.
+constant with omeprazole treatment.} Boxplots of the Shannon Diversity
+Index values (A) and number of observed OTUs (B) for each group of mice
+over 3 timepoints (Day -7, 0, and 9). Each circle represents the value
+for a stool sample from an individual mouse.
 

### 2019-09-09 code/ppi_pcoa.R

@@ -42,7 +42,6 @@ pcoa_subset1 <- read_tsv('data/mothur/subset_1/ppi.opti_mcc.braycurtis.0.03.lt.a
 pcoa_before_challenge <- pcoa_subset1 %>% 
   filter(day < 1, Group == "Omeprazole", Mouse.ID == as.factor(Mouse.ID)) %>%	
   ggplot(aes(x=axis1, y=axis2, color=Mouse.ID, shape=Mouse.ID)) +
-  labs(color = "Omeprazole Mice")+
   scale_color_manual(name="Omeprazole Mice", values=color_mouse) + 
   scale_shape_manual(name= "Omeprazole Mice", values=shape_mouse)+
   geom_point() +
@@ -53,7 +52,7 @@ pcoa_before_challenge <- pcoa_subset1 %>%
   xlim(-0.4, 0.4)+
   ylim(-0.4, 0.4)+
   theme(plot.title = element_text(hjust = 0.5)) +
-  theme(legend.position = c(0.15, 0.8)) #Move legend position
+  theme(legend.position = "none") #Remove
 save_plot("results/figures/before_C._diff_challenge.png", pcoa_before_challenge, base_aspect_ratio = 2) #Use save_plot over ggsave because it works better with cowplot
 
 # Figure 2B----

### 2019-10-30 submission/manuscript.Rmd
Begin responding to reviewers

@@
 bibliography: references.bib
-csl: mbio.csl
+csl: msphere.csl
 fontsize: 11pt

### 2019-11-01 submission/response_to_reviewers.md
Respond to reviewers' comments

---
output:
  pdf_document: default
  html_document: default
---
### Editor Comments:

**1. Both reviewers commented about the appropriateness of strain that was used in your experiments. Please provide a justification for why this strain was chosen.**

> We have responded to the reviewers' comments below regarding why we chose to test *C. difficile* 630 and have added the following sentence to the discussion to acknowledge the strain choice as a potential limitation: 
"The type of *C. difficile* strain type could also be an important contributing factor, our study was limited in that we only tested *C. difficile* 630 (ribotype 012)."

**2. Both reviewers commented on the low number of mice used in the experiments, perhaps making it hard to pick up any small differences. Please add some additional discussion around this point.**

> We have responded to the reviewers' comments below and added the following sentence to the discussion regarding the low number of mice used:
"One limitation of our study is that there were only 4-5 mice per group, which may have limited our ability to identify PPI-induced changes in specific bacteria genera."

**3. Lastly, from my own reading of the paper, I recall that rodent gut microflora consist mainly of Gram-positive organisms, whereas human gut flora is mainly Gram-negatives and anaerobes. One would presume that any treatment that disrupts the microbiome would interact very differently with these two environments, and that the rodent microbiome may not accurately reflect what happens in the human gut. This caveat is hinted at in lines 106-107 of your manuscript, but I would like you to make this point more clearly.**

> We have added an additional sentence to precede the sentence from lines 106-107 to draw attention to the differences in intestinal microbiome composition between mice and humans: 
"*While the intestinal microbiomes of both humans and mice are dominated by the Bacteroidetes and Firmicutes phyla, there are significant differences in the relative abundances of genera that are present and some genera are unique to each mammal (Hugenholtz & de Vos Cell. Mol. Life Sci. 2018), differences that may partly explain our results.* The microbiota and physiological differences between humans and mice may limit the usefulness of employing mouse models to study the impact of PPIs on the microbiota and CDIs."

### Reviewer #1 (Comments for the Author):

**In their manuscript Tomkovich et al. address the question whether the correlative association of proton pump inhibitors with the occurence of CDI is due to the changes in the microbiome following PPI administration as it is speculated in the field. This correlation drawn mainly from epidemiological studies and typically retrospectively, lacks experimental evaluations. Hence, the authors used the mouse colonization model of C. difficile. They checked the changes in the microbiome and the colonization efficiencies in mice treated either with the PPI, omeprazole or clindamycin or with a commbination of both. They show that efficient colonization is dependent on clindamycin treatment prior to infction and that omeprazole alone has no significant effect on the diversity of the microbiome, and does not result in effective colonzation by C. difficile. As this is the first attempt to experimentally address a question arising from medical practice and observation, the study and its findings are interesting.**
**Neverteheless, the study has some points that need to be addressed at least in the discussion before publication**

**1. Is it correct that the study was perfomed on max. 4 mice/treatment group? This might be too low a number, especially in case of the colonization studies to draw conclusions.**

> The reviewer is correct that the number of mice per group was low. We state the exact number of mice per treatment group in Figure 1A (5 mice each for 2 of the treatment groups and 4 mice for the 3rd treatment group). We agree that this is a low number, but would argue that the low microbiome variation across laboratory mice means less mice are needed compared to human epidemiological studies when examining *C. difficile* colonization. Previous *C. difficile* infection mouse model studies from our group and a close collaborator have used 3-5 mice per treatment group (see Koenigsknecht et al. Infection and Immunity 2015 and Theriot et al. Gut Microbes 2011 ) and were still able to draw meaningful conclusions. We have added the following sentence to the discussion to acknowledge the low number of mice used for these experiments as a limitation:
"One limitation of our study is that there were only 4-5 mice per group, which may have limited our ability to identify PPI-induced changes in specific bacteria genera."

**2. The authors should address the fact that they have used only C. difficile str. 630 for their experiments. This is a ribotype 012 strain that is not predominant in the clinical setting. Which ribotypes are typically observed in PPI associated CDI, is there a tendency towards a certain ribotype? Could this have an impact on the interpretation of the data derived from the model described in this manuscript?**

> We thank the reviewers for pointing out that the type of *C. difficile* strain could be an important factor influencing the association between PPIs and CDIs observed in human epidemiological studies. Unfortunately, the studies to date seldom examine the *C. difficile* strains found in PPI users that develop CDIs. The studies that do discuss *C. difficile* strains merely note the type of *C. difficile* frequently found in the region that the study was conducted (Roughead et al. Expert Opinion on Drug Safety 2016; McDonald et al. JAMA Internal Medicine 2015; Chitnis et al. JAMA Internal Medicine 2013; Vardakas et al. International Journal of Infections Diseases 2012; Dalton et al. Alimentary Pharmacology & Therapeutics 2008). In addition to our lab's prior experience with the *C. difficile* 630 infection mouse model, we also decided to use *C. difficile* 630 because of a recent study that demonstrated three proton pump inhibitors had no effect on *in vitro C. difficile* 630 growth (Maier et al. Nature 2018). We have added the following sentence to the discussion to acknowledge that the type of *C. difficile* strain is another potential contributing factor to the association between PPIs and CDIs seen in human epidemiological studies:
"The type of *C. difficile* strain type could also be an important contributing factor, however our study was limited in that we only tested *C. difficile* 630 (ribotype 012)."

**3. As the manuscript is pretty concise and has in total 4 figures, the supplementary figures can be integrated into the main body.**

> We have submitted this manuscript as an observation, which is limited to 2 figures total. Additionally, in response to Reviewer #2, we have expanded the size of all the panels in our figures, limiting our ability to include additional panels for the 2 main text figures. Because of the changes to the sizes of the figures, we have opted to keep the supplementary figures as supplemental.

**5. There are some studies that looked into the effect of PPIs on microbiomes in cats and infants, which seem to be in accordance with the findings in this manuscript. Hence, the authors should include these in their discussion:**
**- Schmid SM, et al.,2018**
**- Castellani C, et al., 2017**

> Thank you for the suggestion, we have added the following sentence to the discussion referring to these studies to indicate that other groups have also observed minimal changes in the mammalian microbiome after proton pump inhibitor treatment:
"Interestingly, other groups examining fecal microbiota communities before and after PPI administration to healthy cats and infants with gastroesophageal reflux disease, found PPIs have minimal effects on fecal bacterial community structures, although there were a few significant changes in specific genera (Schmid et al. Frontiers in Veterinary Science 2018; Castellani et al. Frontiers in Cellular and Infection Microbiology 2017)."

**6. It is becoming more and more accepted that Clostridium difficile, due to its biology, is a separate genus, resulting in Clostridioides difficile as the accepted name. The authors should give credit to this and at least state it as Clostridium (Clostridioides) difficile in the text the first time they use it.**

> We appreciate the suggestion and have added "*Clostridium* (*Clostridioides*) *difficile*" to the abstract, importance, and introduction sections of our manuscript.

### Reviewer #2 (Comments for the Author):

**The manuscript by Tomkovich et al tests if the PPI omeprazole can influence the fecal microbiome of mice and promote C. difficile infection. The study is based on epidemiological findings in humans that associate increased PPI use with CDI as well as a few mouse studies that implicate PPIs in CDI. The data presented here show no significant impact of omeprazole on the microbiome of mice after 7 days of treatment and the mice did not become susceptible to C. diff colonization. This is a small study but adds to the literature in that it conclusively shows no impact of omeprazole on the microbiome.**

**1. The size of the figures and the fonts are very small and should be improved in a resubmission. It makes it difficult to see individual points in the graphs.**

> We have adjusted the sizes of all figures and increased the sizing of the fonts and points in the graphs.

**2. The choice of CD630 and dose may have a big impact on the results and comparison with previous studies. In ref. 15 the authors used a more virulent strain (VPi) and at a 10,000 fold increased concentration. The authors do a good job of citing other factors that may influence the results (like the increased pH of the mouse stomach) but this should also be noted for readers.**

> We thank the reviewer for pointing out *C. difficile* strain type as an additional factor that could explain the association between proton pump inhibitors and CDIs in human studies, which was also pointed out by Reviewer # 1. In response, we have added the following sentence to the discussion:
"The type of *C. difficile* strain type could also be an important contributing factor, however our study was limited in that we only tested *C. difficile* 630 (ribotype 012)."

**3. Please provide evidence for how the dose of 40mg/kg of OMZ corresponds to what humans get in a dose.**

>  Omeprazole is prescribed to adults in 20 mg or 40 mg doses that are taken once per day. The 40 mg/kg dose administered to the mice in our study is higher than what is administered to humans translating to ~ 3.25 mg/kg (compared to ~ 0.33-0.66 mg/kg for the prescribed human dose) based on a reference body weight of 60 kg for humans and 0.02 kg for mice (converted according to the calculation described in Nair & Jacob J Basic Clin Pharm. 2016). However, previous studies have justified administering omeprazole at higher doses to mice because the drug administered to the mice lacks an enteric coating (Llorente et al. Nature Communications 2017) and is eliminated more quickly compared to humans (plasma elimination half life of 5-15 minutes compared to 1 hour for humans, see Regardh et al. Scandinavian Journal of Gastroenterology 1985). The 40 mg/kg dose chosen for our study is in line with what was used by one of the previous studies that examined the proton pump inhibitor esomeprazole and *C. difficile* in mice (see Hung et al. Journal of Infectious Diseases 2015). We have added the following sentence to the methods section to account for the higher dosing:
"Although the omeprazole dose administered to mice is higher than the recommended dose for humans, omeprazole has a shorter half-life in mice compared to humans (Regardh et al. Scandinavian Journal of Gastroenterology 1985) and lacks an enteric coating (Llorente et al. Nature Communications 2017)."

**4. Although OMZ did not have a major impact on community structure by itself, in figure 2 it seems that it did have an impact on a few groups that were treated with clindamycin. Although the P-values aren't significant, this could be mainly due to the low number of mice that were used in the study as a few of the genera showed 10-fold changes. I can see why it isn't commented on due to the lack of stat significance, but it is pretty striking as presented in the figure.**

> We agree with the reviewer regarding Figure 2C, where the relative abundances for specific taxa in the Clindamycin + omeprazole treated mice were typically in between the clindamycin-treated and omeprazole-treated mice. Beyond the P-values not being significant, we decided not to comment on this because we suspect that the majority of the differences in bacterial relative abundances across groups is driven by the clindamycin treatment. We have added the following sentence to the discussion to acknowledge the low number of mice used as a potential limitation in detecting omeprazole-induced changes in the fecal microbiota:
"One limitation of our study is that there were only 4-5 mice per group, which may have limited our ability to identify PPI-induced changes in bacteria genera."

**5. The final sentence of the importance section is a run-on sentence and needs to be broken up or simplified.**

> To simplify the sentence, we have removed the part that elaborates on what other factors ("composition of the starting bacterial community, comorbidities, and use of additional medications") might be contributing to the correlations between PPIs and CDIs seen in humans.

**6. There are many typos and missed italics in the references.**

> Thank you for pointing these errors out, typos have been corrected and italics have been added to the references.

### 2019-11-01 submission/manuscript.Rmd
Get rid of warnings about packages being built under specific R versions when markdown is knitted
Respond to reviewers' comments

@@
 
-```{r knitr_settings, tidy=TRUE, eval=TRUE, echo=FALSE, cache=FALSE}
+```{r knitr_settings, tidy=TRUE, eval=TRUE, warning = FALSE, echo=FALSE, cache=FALSE}
 options(tidyverse.quiet = TRUE)
@@
 ## Abstract
-Proton pump inhibitor (PPI) use has been associated with microbiota alterations and susceptibility to *Clostridium difficile* infections (CDIs) in humans.
+Proton pump inhibitor (PPI) use has been associated with microbiota alterations and susceptibility to *Clostridium* (*Clostridioides*) *difficile* infections (CDIs) in humans.
 We assessed how PPI treatment alters the fecal microbiota and whether treatment promotes CDIs in a mouse model.
@@
 ## Importance
-Antibiotics are the primary risk factor for *Clostridium difficile* infections (CDIs), but other factors may also increase a person's risk.
+Antibiotics are the primary risk factor for *Clostridium* (*Clostridioides*) *difficile* infections (CDIs), but other factors may also increase a person's risk.
 In epidemiological studies, proton pump inhibitor (PPI) use has been associated with CDI incidence and recurrence.
@@
 difficile* colonization.
-Further studies are needed to determine whether other factors such as the composition of the starting bacterial community, comorbidities, and use of additional medications contribute to the association between PPIs and CDIs seen in humans or whether aspects of murine physiology may limit its utility to test these types of hypotheses.
+Further studies are needed to determine whether other factors contribute to the association between PPIs and CDIs seen in humans or whether aspects of murine physiology may limit its utility to test these types of hypotheses.
 
@@
 
-Antibiotics have a large impact on the intestinal microbiome and are a primary risk factor for developing *Clostridium difficile* infections (CDIs) [@Schubert2015].
+Antibiotics have a large impact on the intestinal microbiome and are a primary risk factor for developing *Clostridium* (*Clostridioides*) *difficile* infections (CDIs) [@Schubert2015].
 It is less clear whether other human medications that impact the microbiota also influence *C.
@@
 However, 3 of the human-associated taxa were absent or at low abundance in our mice.
+Interestingly, other groups examining fecal microbiota communities before and after PPI administration to healthy cats and infants with gastroesophageal reflux disease, found PPIs have minimal effects on fecal bacterial community structures, although there were a few significant changes in specific genera [@Schmid2018; @Castellani2017].
+One limitation of our study is that there were only 4-5 mice per group, which may have limited our ability to identify PPI-induced changes in specific bacteria genera.
 Although our fecal microbiota findings are comparable to what has been shown in another mouse study [@Yasutomi2018], whether PPI-induced changes in specific bacterial abundances observed in humans play a role in CDIs remains to be determined.
@@
 difficile* colonization in our CDI mouse model.
-Gastrointestinal physiological differences, particularly the higher stomach pH in mice (pH 3-4) compared to humans (pH 1) [@Hugenholtz2017] could also explain why omeprazole had a limited impact on the murine microbiome.
+While the intestinal microbiomes of both humans and mice are dominated by the *Bacteroidetes* and *Firmicutes* phyla, there are significant differences in the relative abundances of genera that are present and some genera are unique to each mammal [@Hugenholtz2018], differences that may partly explain our results.
+Gastrointestinal physiological differences, particularly the higher stomach pH in mice (pH 3-4) compared to humans (pH 1) [@Hugenholtz2018] could also explain why omeprazole had a limited impact on the murine microbiome.
 The microbiota and physiological differences between humans and mice may limit the usefulness of employing mouse models to study the impact of PPIs on the microbiota and CDIs.
 Beyond microbiome differences, factors such as age, body mass index, comorbidities, and use of other medications in human studies may also be contributing to the association between PPIs and CDI incidence or recurrence.
+The type of *C.
+difficile* strain type could also be an important contributing factor, however our study was limited in that we only tested *C.
+difficile* 630 (ribotype 012).
 This study addressed the impact of PPIs with or without antibiotics on a murine model of CDI, and found PPIs did not promote *C.
@@
 difficile* challenge (Figure 1A).
+Although the omeprazole dose administered to mice is higher than the recommended dose for humans, omeprazole has a shorter half-life in mice compared to humans [@Regrdh1985] and lacks an enteric coating [@Llorente2017].
 One day prior to *C.

### 2019-11-01 submission/manuscript.tex
Respond to reviewers' comments

@@
 Proton pump inhibitor (PPI) use has been associated with microbiota
-alterations and susceptibility to \emph{Clostridium difficile}
-infections (CDIs) in humans. We assessed how PPI treatment alters the
-fecal microbiota and whether treatment promotes CDIs in a mouse model.
-Mice receiving a PPI treatment were gavaged with 40 mg/kg of omeprazole
-during a 7-day pretreatment phase, the day of \emph{C. difficile}
-challenge, and the following 9 days. We found that mice treated with
-omeprazole were not colonized by \emph{C. difficile}. When omeprazole
-treatment was combined with a single clindamycin treatment, one cage of
-mice remained resistant to \emph{C. difficile} colonization, while the
-other cage was colonized. Treating mice with only clindamycin followed
-by challenge resulted in \emph{C. difficile} colonization. 16S rRNA gene
-sequencing analysis revealed that omeprazole had minimal impact on the
-structure of the murine microbiota throughout the 16 days of omeprazole
-exposure. These results suggest omeprazole treatment alone is not
-sufficient to disrupt microbiota resistance to \emph{C. difficile}
-infection in mice that are normally resistant in the absence of
-antibiotic treatment.
+alterations and susceptibility to \emph{Clostridium}
+(\emph{Clostridioides}) \emph{difficile} infections (CDIs) in humans. We
+assessed how PPI treatment alters the fecal microbiota and whether
+treatment promotes CDIs in a mouse model. Mice receiving a PPI treatment
+were gavaged with 40 mg/kg of omeprazole during a 7-day pretreatment
+phase, the day of \emph{C. difficile} challenge, and the following 9
+days. We found that mice treated with omeprazole were not colonized by
+\emph{C. difficile}. When omeprazole treatment was combined with a
+single clindamycin treatment, one cage of mice remained resistant to
+\emph{C. difficile} colonization, while the other cage was colonized.
+Treating mice with only clindamycin followed by challenge resulted in
+\emph{C. difficile} colonization. 16S rRNA gene sequencing analysis
+revealed that omeprazole had minimal impact on the structure of the
+murine microbiota throughout the 16 days of omeprazole exposure. These
+results suggest omeprazole treatment alone is not sufficient to disrupt
+microbiota resistance to \emph{C. difficile} infection in mice that are
+normally resistant in the absence of antibiotic treatment.
 
@@
 
-Antibiotics are the primary risk factor for \emph{Clostridium difficile}
-infections (CDIs), but other factors may also increase a person's risk.
-In epidemiological studies, proton pump inhibitor (PPI) use has been
-associated with CDI incidence and recurrence. PPIs have also been
-associated with alterations in the human intestinal microbiota in
-observational and interventional studies. We evaluated the effects of
-the PPI omeprazole on the structure of the murine intestinal microbiota
-and its ability to disrupt colonization resistance to \emph{C.
-difficile}. We found omeprazole treatment had minimal impact on the
-murine fecal microbiota and did not promote \emph{C. difficile}
-colonization. Further studies are needed to determine whether other
-factors such as the composition of the starting bacterial community,
-comorbidities, and use of additional medications contribute to the
-association between PPIs and CDIs seen in humans or whether aspects of
-murine physiology may limit its utility to test these types of
-hypotheses.
+Antibiotics are the primary risk factor for \emph{Clostridium}
+(\emph{Clostridioides}) \emph{difficile} infections (CDIs), but other
+factors may also increase a person's risk. In epidemiological studies,
+proton pump inhibitor (PPI) use has been associated with CDI incidence
+and recurrence. PPIs have also been associated with alterations in the
+human intestinal microbiota in observational and interventional studies.
+We evaluated the effects of the PPI omeprazole on the structure of the
+murine intestinal microbiota and its ability to disrupt colonization
+resistance to \emph{C. difficile}. We found omeprazole treatment had
+minimal impact on the murine fecal microbiota and did not promote
+\emph{C. difficile} colonization. Further studies are needed to
+determine whether other factors contribute to the association between
+PPIs and CDIs seen in humans or whether aspects of murine physiology may
+limit its utility to test these types of hypotheses.
 
@@
 Antibiotics have a large impact on the intestinal microbiome and are a
-primary risk factor for developing \emph{Clostridium difficile}
-infections (CDIs) (1). It is less clear whether other human medications
-that impact the microbiota also influence \emph{C. difficile}
-colonization resistance. Multiple epidemiological studies have suggested
-an association between proton pump inhibitor (PPI) use and incidence or
-recurrence of CDIs (2--5). There have also been a number of large cohort
-studies and interventional clinical trials that demonstrated specific
-alterations in the intestinal microbiome were associated with PPI use
-(4, 6). PPI-associated microbiota changes have been attributed to the
-ability of PPIs to increase stomach acid pH which may promote the
-survival of oral and pathogenic bacteria (4, 6). In human fecal samples,
-PPI use results in increases in \emph{Enterococcaceae},
-\emph{Lactobacillaceae}, \emph{Micrococcaceae}, \emph{Staphylococcaceae}
-and \emph{Streptococcaceae} and decreases in \emph{Ruminococcaceae}
-(6--9). Several of these taxa have also been associated with \emph{C.
-difficile} colonization in humans (10).\\
+primary risk factor for developing \emph{Clostridium}
+(\emph{Clostridioides}) \emph{difficile} infections (CDIs) (1). It is
+less clear whether other human medications that impact the microbiota
+also influence \emph{C. difficile} colonization resistance. Multiple
+epidemiological studies have suggested an association between proton
+pump inhibitor (PPI) use and incidence or recurrence of CDIs (2--5).
+There have also been a number of large cohort studies and interventional
+clinical trials that demonstrated specific alterations in the intestinal
+microbiome were associated with PPI use (4, 6). PPI-associated
+microbiota changes have been attributed to the ability of PPIs to
+increase stomach acid pH which may promote the survival of oral and
+pathogenic bacteria (4, 6). In human fecal samples, PPI use results in
+increases in \emph{Enterococcaceae}, \emph{Lactobacillaceae},
+\emph{Micrococcaceae}, \emph{Staphylococcaceae} and
+\emph{Streptococcaceae} and decreases in \emph{Ruminococcaceae} (6--9).
+Several of these taxa have also been associated with \emph{C. difficile}
+colonization in humans (10).\\
 Unfortunately, the studies suggesting a link between PPIs and \emph{C.
@@
 studies. However, 3 of the human-associated taxa were absent or at low
-abundance in our mice. Although our fecal microbiota findings are
+abundance in our mice. Interestingly, other groups examining fecal
+microbiota communities before and after PPI administration to healthy
+cats and infants with gastroesophageal reflux disease, found PPIs have
+minimal effects on fecal bacterial community structures, although there
+were a few significant changes in specific genera (13, 14). One
+limitation of our study is that there were only 4-5 mice per group,
+which may have limited our ability to identify PPI-induced changes in
+specific bacteria genera. Although our fecal microbiota findings are
 comparable to what has been shown in another mouse study (12), whether
@@
 PPIs have an effect on CDIs with or without additional antibiotic
-treatment (13--15), there were insufficient controls to attribute the
+treatment (15--17), there were insufficient controls to attribute the
 effect solely to PPI treatment. One group administered 0.5 mg/kg of the
@@
 however there was detectable \emph{C. difficile} in mice not treated
-with antibiotics (13, 14). The other mouse study demonstrated
+with antibiotics (15, 16). The other mouse study demonstrated
 antibiotic/esomeprazole-treated mice developed more severe CDIs compared
 to antibiotic-treated mice, but the researchers did not have a group
-treated with just esomeprazole for comparison (15). We tested the same
+treated with just esomeprazole for comparison (17). We tested the same
 high 40 mg/kg PPI dose and expanded pre-treatment to 7 days before
@@
 from our breeding colony are resistant to \emph{C. difficile} 630
-colonization without antibiotic treatment (16), ensuring there was not
+colonization without antibiotic treatment (18), ensuring there was not
 already partial susceptibility to \emph{C. difficile} before treatment.
@@
 taxa within the murine intestinal microbiota over the 16-day experiment.
-In contrast to previous work with PPIs (13--15), omeprazole did not
+In contrast to previous work with PPIs (15--17), omeprazole did not
 alter \emph{C. difficile} colonization resistance in mice. 16S rRNA
@@
 treatment had no effect on \emph{C. difficile} colonization in our CDI
-mouse model. Gastrointestinal physiological differences, particularly
-the higher stomach pH in mice (pH 3-4) compared to humans (pH 1) (17)
-could also explain why omeprazole had a limited impact on the murine
-microbiome. The microbiota and physiological differences between humans
-and mice may limit the usefulness of employing mouse models to study the
-impact of PPIs on the microbiota and CDIs.\\
+mouse model. While the intestinal microbiomes of both humans and mice
+are dominated by the \emph{Bacteroidetes} and \emph{Firmicutes} phyla,
+there are significant differences in the relative abundances of genera
+that are present and some genera are unique to each mammal (19),
+differences that may partly explain our results. Gastrointestinal
+physiological differences, particularly the higher stomach pH in mice
+(pH 3-4) compared to humans (pH 1) (19) could also explain why
+omeprazole had a limited impact on the murine microbiome. The microbiota
+and physiological differences between humans and mice may limit the
+usefulness of employing mouse models to study the impact of PPIs on the
+microbiota and CDIs.\\
 Beyond microbiome differences, factors such as age, body mass index,
@@
 contributing to the association between PPIs and CDI incidence or
-recurrence. This study addressed the impact of PPIs with or without
-antibiotics on a murine model of CDI, and found PPIs did not promote
-\emph{C. difficile} colonization. The epidemiological evidence linking
-PPIs to CDIs is primarily from observational studies, which makes
-determining causality and whether other risk factors play a role
-challenging (18). Future studies are needed to determine whether age,
-other comorbidities and bacterial strains that are less common in mice
-can increase the risk of CDIs or recurrent CDIs when combined with PPI
-treatment.
+recurrence. The type of \emph{C. difficile} strain type could also be an
+important contributing factor, however our study was limited in that we
+only tested \emph{C. difficile} 630 (ribotype 012). This study addressed
+the impact of PPIs with or without antibiotics on a murine model of CDI,
+and found PPIs did not promote \emph{C. difficile} colonization. The
+epidemiological evidence linking PPIs to CDIs is primarily from
+observational studies, which makes determining causality and whether
+other risk factors play a role challenging (20). Future studies are
+needed to determine whether age, other comorbidities and bacterial
+strains that are less common in mice can increase the risk of CDIs or
+recurrent CDIs when combined with PPI treatment.
 
@@
 aliquots and diluted to an 8 mg/mL prior to gavage. All mice received 40
-mg/kg omeprazole (a dose previously used in mouse experiments (15)) or
+mg/kg omeprazole (a dose previously used in mouse experiments (17)) or
 vehicle solution once per day through the duration of the experiment
 with treatment starting 7 days before \emph{C. difficile} challenge
-(Figure 1A). One day prior to \emph{C. difficile} challenge, 2 groups of
-mice received an intraperitoneal injection of 10 mg/kg clindamycin or
-sterile saline vehicle (11). All drugs were filter sterilized through a
-0.22 micron syringe filter before administration to animals.
+(Figure 1A). Although the omeprazole dose administered to mice is higher
+than the recommended dose for humans, omeprazole has a shorter half-life
+in mice compared to humans (21) and lacks an enteric coating (22). One
+day prior to \emph{C. difficile} challenge, 2 groups of mice received an
+intraperitoneal injection of 10 mg/kg clindamycin or sterile saline
+vehicle (11). All drugs were filter sterilized through a 0.22 micron
+syringe filter before administration to animals.
 
@@
 system (Eppendorf). The 16S rRNA sequencing library was prepared as
-described previously (19). In brief, the ZymoBIOMICS\textsuperscript{TM}
+described previously (23). In brief, the ZymoBIOMICS\textsuperscript{TM}
 Microbial Community DNA Standard (Zymo, CA, USA) was used as a mock
-community (20) and water was used as a negative control. The V4
+community (24) and water was used as a negative control. The V4
 hypervariable region of the 16S rRNA gene was amplified with Accuprime
 Pfx DNA polymerase (Thermo Fisher Scientific) using previously described
-custom barcoded primers (19). The 16S rRNA amplicon library was
+custom barcoded primers (23). The 16S rRNA amplicon library was
 sequenced with the MiSeq (Illumina). Amplicons were cleaned up and
@@
 \textbf{16S rRNA gene sequence analysis.} mothur (v1.40.5) was used for
-all sequence processing steps (21) using a previously published protocol
-(19). In brief, forward and reverse reads for each sample were combined
+all sequence processing steps (25) using a previously published protocol
+(23). In brief, forward and reverse reads for each sample were combined
 and low-quality sequences and chimeras were removed. Duplicate sequences
@@
 \hypertarget{ref-Schubert2015}{}
-1. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
-Antibiotic-induced alterations of the murine gut microbiota and
-subsequent effects on colonization resistance against clostridium
-difficile. mBio \textbf{6}.
-doi:\href{https://doi.org/10.1128/mbio.00974-15}{10.1128/mbio.00974-15}.
+1. Schubert AM, Sinani H, Schloss PD. 2015. Antibiotic-induced
+alterations of the murine gut microbiota and subsequent effects on
+colonization resistance against \emph{Clostridium difficile}. mBio 6.
 
 \hypertarget{ref-tariq2017association}{}
-2. \textbf{Tariq R}, \textbf{Singh S}, \textbf{Gupta A}, \textbf{Pardi
-DS}, \textbf{Khanna S}. 2017. Association of gastric acid suppression
-with recurrent clostridium difficile infection: A systematic review and
-meta-analysis. JAMA internal medicine \textbf{177}:784--791.
+2. Tariq R, Singh S, Gupta A, Pardi DS, Khanna S. 2017. Association of
+gastric acid suppression with recurrent \emph{Clostridium difficile}
+infection: A systematic review and meta-analysis. JAMA internal medicine
+177:784--791.
 
 \hypertarget{ref-nehra2018proton}{}
-3. \textbf{Nehra AK}, \textbf{Alexander JA}, \textbf{Loftus CG},
-\textbf{Nehra V}. 2018. Proton pump inhibitors: Review of emerging
-concerns, pp. 240--246. \emph{In} Mayo clinic proceedings. Elsevier.
+3. Nehra AK, Alexander JA, Loftus CG, Nehra V. 2018. Proton pump
+inhibitors: Review of emerging concerns, pp. 240--246. \emph{In} Mayo
+clinic proceedings. Elsevier.
 
 \hypertarget{ref-Naito2018}{}
-4. \textbf{Naito Y}, \textbf{Kashiwagi K}, \textbf{Takagi T},
-\textbf{Andoh A}, \textbf{Inoue R}. 2018. Intestinal dysbiosis secondary
-to proton-pump inhibitor use. Digestion \textbf{97}:195--204.
-doi:\href{https://doi.org/10.1159/000481813}{10.1159/000481813}.
+4. Naito Y, Kashiwagi K, Takagi T, Andoh A, Inoue R. 2018. Intestinal
+dysbiosis secondary to proton-pump inhibitor use. Digestion 97:195--204.
 
 \hypertarget{ref-Elias2019}{}
-5. \textbf{Elias E}, \textbf{Targownik LE}. 2019. The clinician's guide
-to proton pump inhibitor related adverse events. Drugs
-\textbf{79}:715--731.
-doi:\href{https://doi.org/10.1007/s40265-019-01110-3}{10.1007/s40265-019-01110-3}.
+5. Elias E, Targownik LE. 2019. The clinician's guide to proton pump
+inhibitor related adverse events. Drugs 79:715--731.
 
 \hypertarget{ref-Imhann2017}{}
-6. \textbf{Imhann F}, \textbf{Vila AV}, \textbf{Bonder MJ},
-\textbf{Manosalva AGL}, \textbf{Koonen DP}, \textbf{Fu J},
-\textbf{Wijmenga C}, \textbf{Zhernakova A}, \textbf{Weersma RK}. 2017.
-The influence of proton pump inhibitors and other commonly used
-medication on the gut microbiota. Gut Microbes \textbf{8}:351--358.
-doi:\href{https://doi.org/10.1080/19490976.2017.1284732}{10.1080/19490976.2017.1284732}.
+6. Imhann F, Vila AV, Bonder MJ, Manosalva AGL, Koonen DP, Fu J,
+Wijmenga C, Zhernakova A, Weersma RK. 2017. The influence of proton pump
+inhibitors and other commonly used medication on the gut microbiota. Gut
+Microbes 8:351--358.
 
 \hypertarget{ref-Imhann2015}{}
-7. \textbf{Imhann F}, \textbf{Bonder MJ}, \textbf{Vila AV}, \textbf{Fu
-J}, \textbf{Mujagic Z}, \textbf{Vork L}, \textbf{Tigchelaar EF},
-\textbf{Jankipersadsing SA}, \textbf{Cenit MC}, \textbf{Harmsen HJM},
-\textbf{Dijkstra G}, \textbf{Franke L}, \textbf{Xavier RJ},
-\textbf{Jonkers D}, \textbf{Wijmenga C}, \textbf{Weersma RK},
-\textbf{Zhernakova A}. 2015. Proton pump inhibitors affect the gut
-microbiome. Gut \textbf{65}:740--748.
-doi:\href{https://doi.org/10.1136/gutjnl-2015-310376}{10.1136/gutjnl-2015-310376}.
+7. Imhann F, Bonder MJ, Vila AV, Fu J, Mujagic Z, Vork L, Tigchelaar EF,
+Jankipersadsing SA, Cenit MC, Harmsen HJM, Dijkstra G, Franke L, Xavier
+RJ, Jonkers D, Wijmenga C, Weersma RK, Zhernakova A. 2015. Proton pump
+inhibitors affect the gut microbiome. Gut 65:740--748.
 
 \hypertarget{ref-Freedberg2015}{}
-8. \textbf{Freedberg DE}, \textbf{Toussaint NC}, \textbf{Chen SP},
-\textbf{Ratner AJ}, \textbf{Whittier S}, \textbf{Wang TC}, \textbf{Wang
-HH}, \textbf{Abrams JA}. 2015. Proton pump inhibitors alter specific
-taxa in the human gastrointestinal microbiome: A crossover trial.
-Gastroenterology \textbf{149}:883--885.e9.
-doi:\href{https://doi.org/10.1053/j.gastro.2015.06.043}{10.1053/j.gastro.2015.06.043}.
+8. Freedberg DE, Toussaint NC, Chen SP, Ratner AJ, Whittier S, Wang TC,
+Wang HH, Abrams JA. 2015. Proton pump inhibitors alter specific taxa in
+the human gastrointestinal microbiome: A crossover trial.
+Gastroenterology 149:883--885.e9.
 
 \hypertarget{ref-maier2018extensive}{}
-9. \textbf{Maier L}, \textbf{Pruteanu M}, \textbf{Kuhn M},
-\textbf{Zeller G}, \textbf{Telzerow A}, \textbf{Anderson EE},
-\textbf{Brochado AR}, \textbf{Fernandez KC}, \textbf{Dose H},
-\textbf{Mori H}, \textbf{others}. 2018. Extensive impact of
-non-antibiotic drugs on human gut bacteria. Nature \textbf{555}:623.
+9. Maier L, Pruteanu M, Kuhn M, Zeller G, Telzerow A, Anderson EE,
+Brochado AR, Fernandez KC, Dose H, Mori H, others. 2018. Extensive
+impact of non-antibiotic drugs on human gut bacteria. Nature 555:623.
 
 \hypertarget{ref-Schuberte01021-14}{}
-10. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
-\textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
-\textbf{Aronoff DM}, \textbf{Schloss PD}. 2014. Microbiome data
-distinguish patients with clostridium difficile infection and non-c.
-difficile-associated diarrhea from healthy controls. mBio \textbf{5}.
-doi:\href{https://doi.org/10.1128/mBio.01021-14}{10.1128/mBio.01021-14}.
+10. Schubert AM, Rogers MAM, Ring C, Mogle J, Petrosino JP, Young VB,
+Aronoff DM, Schloss PD. 2014. Microbiome data distinguish patients with
+\emph{Clostridium difficile} infection and non-\emph{C.
+difficile}-associated diarrhea from healthy controls. mBio 5.
 
 \hypertarget{ref-Jenior2018}{}
-11. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
-\textbf{Schloss PD}. 2018. Clostridium difficile alters the structure
-and metabolism of distinct cecal microbiomes during initial infection to
-promote sustained colonization. mSphere \textbf{3}.
-doi:\href{https://doi.org/10.1128/msphere.00261-18}{10.1128/msphere.00261-18}.
+11. Jenior ML, Leslie JL, Young VB, Schloss PD. 2018. \emph{Clostridium
+difficile} alters the structure and metabolism of distinct cecal
+microbiomes during initial infection to promote sustained colonization.
+mSphere 3.
 
 \hypertarget{ref-Yasutomi2018}{}
-12. \textbf{Yasutomi E}, \textbf{Hoshi N}, \textbf{Adachi S},
-\textbf{Otsuka T}, \textbf{Kong L}, \textbf{Ku Y}, \textbf{Yamairi H},
-\textbf{Inoue J}, \textbf{Ishida T}, \textbf{Watanabe D}, \textbf{Ooi
-M}, \textbf{Yoshida M}, \textbf{Tsukimi T}, \textbf{Fukuda S},
-\textbf{Azuma T}. 2018. Proton pump inhibitors increase the
-susceptibility of mice to oral infection with enteropathogenic bacteria.
-Digestive Diseases and Sciences \textbf{63}:881--889.
-doi:\href{https://doi.org/10.1007/s10620-017-4905-3}{10.1007/s10620-017-4905-3}.
+12. Yasutomi E, Hoshi N, Adachi S, Otsuka T, Kong L, Ku Y, Yamairi H,
+Inoue J, Ishida T, Watanabe D, Ooi M, Yoshida M, Tsukimi T, Fukuda S,
+Azuma T. 2018. Proton pump inhibitors increase the susceptibility of
+mice to oral infection with enteropathogenic bacteria. Digestive
+Diseases and Sciences 63:881--889.
+
+\hypertarget{ref-Schmid2018}{}
+13. Schmid SM, Suchodolski JS, Price JM, Tolbert MK. 2018. Omeprazole
+minimally alters the fecal microbial community in six cats: A pilot
+study. Frontiers in Veterinary Science 5.
+
+\hypertarget{ref-Castellani2017}{}
+14. Castellani C, Singer G, Kashofer K, Huber-Zeyringer A, Flucher C,
+Kaiser M, Till H. 2017. The influence of proton pump inhibitors on the
+fecal microbiome of infants with gastroesophageal refluxA prospective
+longitudinal interventional study. Frontiers in Cellular and Infection
+Microbiology 7.
 
 \hypertarget{ref-Kaur2007}{}
-13. \textbf{Kaur S}, \textbf{Vaishnavi C}, \textbf{Prasad KK},
-\textbf{Ray P}, \textbf{Kochhar R}. 2007. Comparative role of antibiotic
-and proton pump inhibitor in ExperimentalClostridium difficileInfection
-in mice. Microbiology and Immunology \textbf{51}:1209--1214.
-doi:\href{https://doi.org/10.1111/j.1348-0421.2007.tb04016.x}{10.1111/j.1348-0421.2007.tb04016.x}.
+15. Kaur S, Vaishnavi C, Prasad KK, Ray P, Kochhar R. 2007. Comparative
+role of antibiotic and proton pump inhibitor in experimental
+\emph{Clostridium difficile} infection in mice. Microbiology and
+Immunology 51:1209--1214.
 
 \hypertarget{ref-kaur2011effect}{}
-14. \textbf{Kaur S}, \textbf{Vaishnavi C}, \textbf{Prasad KK},
-\textbf{Ray P}, \textbf{Kochhar R}. 2011. Effect of lactobacillus
-acidophilus \& epidermal growth factor on experimentally induced
-clostridium difficile infection. The Indian journal of medical research
-\textbf{133}:434.
+16. Kaur S, Vaishnavi C, Prasad KK, Ray P, Kochhar R. 2011. Effect of
+lactobacillus acidophilus \& epidermal growth factor on experimentally
+induced \emph{Clostridium difficile} infection. The Indian journal of
+medical research 133:434.
 
 \hypertarget{ref-hung2015proton}{}
-15. \textbf{Hung Y-P}, \textbf{Ko W-C}, \textbf{Chou P-H}, \textbf{Chen
-Y-H}, \textbf{Lin H-J}, \textbf{Liu Y-H}, \textbf{Tsai H-W}, \textbf{Lee
-J-C}, \textbf{Tsai P-J}. 2015. Proton-pump inhibitor exposure aggravates
-clostridium difficile--associated colitis: Evidence from a mouse model.
-The Journal of infectious diseases \textbf{212}:654--663.
+17. Hung Y-P, Ko W-C, Chou P-H, Chen Y-H, Lin H-J, Liu Y-H, Tsai H-W,
+Lee J-C, Tsai P-J. 2015. Proton-pump inhibitor exposure aggravates
+\emph{Clostridium difficile}--associated colitis: Evidence from a mouse
+model. The Journal of infectious diseases 212:654--663.
 
 \hypertarget{ref-Jenior2017}{}
-16. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
-\textbf{Schloss PD}. 2017. Clostridium difficile colonizes alternative
-nutrient niches during infection across distinct murine gut microbiomes.
-mSystems \textbf{2}.
-doi:\href{https://doi.org/10.1128/msystems.00063-17}{10.1128/msystems.00063-17}.
+18. Jenior ML, Leslie JL, Young VB, Schloss PD. 2017. \emph{Clostridium
+difficile} colonizes alternative nutrient niches during infection across
+distinct murine gut microbiomes. mSystems 2.
 
-\hypertarget{ref-Hugenholtz2017}{}
-17. \textbf{Hugenholtz F}, \textbf{Vos WM de}. 2017. Mouse models for
-human intestinal microbiota research: A critical evaluation. Cellular
-and Molecular Life Sciences \textbf{75}:149--160.
-doi:\href{https://doi.org/10.1007/s00018-017-2693-8}{10.1007/s00018-017-2693-8}.
+\hypertarget{ref-Hugenholtz2018}{}
+19. Hugenholtz F, Vos WM de. 2018. Mouse models for human intestinal
+microbiota research: A critical evaluation. Cellular and Molecular Life
+Sciences 75:149--160.
 
 \hypertarget{ref-Eze2017}{}
-18. \textbf{Eze P}, \textbf{Balsells E}, \textbf{Kyaw MH}, \textbf{Nair
-H}. 2017. Risk factors for clostridium difficile infections an overview
-of the evidence base and challenges in data synthesis. Journal of Global
-Health \textbf{7}.
-doi:\href{https://doi.org/10.7189/jogh.07.010417}{10.7189/jogh.07.010417}.
+20. Eze P, Balsells E, Kyaw MH, Nair H. 2017. Risk factors for
+\emph{Clostridium difficile} infections -- an overview of the evidence
+base and challenges in data synthesis. Journal of Global Health 7.
+
+\hypertarget{ref-Regrdh1985}{}
+21. Regårdh C-G, Gabrielsson M, Hoffman K-J, Löfberg I, Skånberg I.
+1985. Pharmacokinetics and metabolism of omeprazole in animals and man -
+an overview. Scandinavian Journal of Gastroenterology 20:79--94.
+
+\hypertarget{ref-Llorente2017}{}
+22. Llorente C, Jepsen P, Inamine T, Wang L, Bluemel S, Wang HJ, Loomba
+R, Bajaj JS, Schubert ML, Sikaroodi M, Gillevet PM, Xu J, Kisseleva T,
+Ho SB, DePew J, Du X, Sørensen HT, Vilstrup H, Nelson KE, Brenner DA,
+Fouts DE, Schnabl B. 2017. Gastric acid suppression promotes alcoholic
+liver disease by inducing overgrowth of intestinal enterococcus. Nature
+Communications 8.
 
 \hypertarget{ref-Kozich2013}{}
-19. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
-\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
-dual-index sequencing strategy and curation pipeline for analyzing
-amplicon sequence data on the MiSeq illumina sequencing platform.
-Applied and Environmental Microbiology \textbf{79}:5112--5120.
-doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+23. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD. 2013.
+Development of a dual-index sequencing strategy and curation pipeline
+for analyzing amplicon sequence data on the MiSeq illumina sequencing
+platform. Applied and Environmental Microbiology 79:5112--5120.
 
 \hypertarget{ref-Sze2019}{}
-20. \textbf{Sze MA}, \textbf{Schloss PD}. 2019. The impact of DNA
-polymerase and number of rounds of amplification in PCR on 16S rRNA gene
-sequence data. mSphere \textbf{4}.
-doi:\href{https://doi.org/10.1128/msphere.00163-19}{10.1128/msphere.00163-19}.
+24. Sze MA, Schloss PD. 2019. The impact of DNA polymerase and number of
+rounds of amplification in PCR on 16S rRNA gene sequence data. mSphere
+4.
 
 \hypertarget{ref-Schloss2009}{}
-21. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
-\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
-\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
-\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
-\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
-Introducing mothur: Open-source, platform-independent,
-community-supported software for describing and comparing microbial
-communities. Applied and Environmental Microbiology
-\textbf{75}:7537--7541.
-doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
+25. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+EB, Lesniewski RA, Oakley BB, Parks DH, Robinson CJ, Sahl JW, Stres B,
+Thallinger GG, Horn DJV, Weber CF. 2009. Introducing mothur:
+Open-source, platform-independent, community-supported software for
+describing and comparing microbial communities. Applied and
+Environmental Microbiology 75:7537--7541.
 

### 2019-11-01 code/figure_2.R

@@ -7,6 +7,6 @@ b <- ggdraw() + draw_image("results/figures/after_abx_C.diff.png")
 c <- ggdraw() + draw_image("results/figures/genera_assoc_w_treatment.png")
 
 plot_grid(a, b, c, labels = "AUTO", label_size = 12)+
-  ggsave("results/figures/figure_2.pdf", width=6.5, height=3.8)+
-  ggsave("submission/figure_2.pdf", width=6.5, height=3.8)
+  ggsave("results/figures/figure_2.pdf", width=6.875, height=4.2)+
+  ggsave("submission/figure_2.pdf", width=6.875, height=4.2)
 

### 2019-11-01 code/figure_s1.R

@@ -6,8 +6,8 @@ a <- ggdraw() + draw_image("results/figures/lactobacillaceae_time.png")
 b <- ggdraw() + draw_image("results/figures/ruminococcaceae_time.png")
 
 plot_grid(a, b, labels = "AUTO", label_size = 12)+
-  ggsave("results/figures/figure_s1.pdf", width=6.5, height=1.5)+
-  ggsave("submission/figure_s1.pdf", width=6.5, height=1.5)
+  ggsave("results/figures/figure_s1.pdf", width=6.875, height=2.5)+
+  ggsave("submission/figure_s1.pdf", width=6.875, height=2.5)
 
 
 

### 2019-11-01 code/figure_s2.R

@@ -6,5 +6,5 @@ a <- ggdraw() + draw_image("results/figures/groups_shannon.png")
 b <- ggdraw() + draw_image("results/figures/groups_richness.png")
 
 plot_grid(a, b, labels = "AUTO", label_size = 12)+
-  ggsave("results/figures/figure_s2.pdf", width=6.5, height=3)+
-  ggsave("submission/figure_s2.pdf", width=6.5, height=3.5)
+  ggsave("results/figures/figure_s2.pdf", width=6.875, height=3)+
+  ggsave("submission/figure_s2.pdf", width=6.875, height=3)

### 2019-11-01 code/ppi_alpha.R

@@ -106,14 +106,15 @@ shannon <- groups_day_alpha %>%
                       values=color_scheme, 
                       breaks=c("Clindamycin", "Clind. + Omep.", "Omeprazole"),
                       labels=c("Clindamycin", "Clind. + Omep.", "Omeprazole")) +
-  geom_boxplot(outlier.shape = NA)+
-  geom_jitter(shape=19, size=1, alpha=0.4, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.1), show.legend = FALSE) +
+  geom_boxplot(outlier.shape = NA, size = 1.2)+
+  geom_jitter(shape=19, size=2, alpha=0.6, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.1), show.legend = FALSE) +
   labs(title=NULL, 
        x="Day",
        y="Shannon Diversity Index")+
   ylim(0, 3.5)+
   theme_classic()+
-  theme(legend.position = c(0.8, 0.2))
+  theme(legend.position = c(0.8, 0.2))+
+  theme(text = element_text(size = 16))  # Change font size for entire plot
 save_plot("results/figures/groups_shannon.png", shannon) #Use save_plot instead of ggsave because it works better with cowplot
 
 # Figure S2B----
@@ -124,14 +125,15 @@ richness <- groups_day_alpha %>%
                       values=color_scheme, 
                       breaks=c("Clindamycin", "Clind. + Omep.", "Omeprazole"),
                       labels=c("Clindamycin", "Clind. + Omep.", "Omeprazole")) +
-  geom_boxplot(outlier.shape = NA)+
-  geom_jitter(shape=19, size=1, alpha=0.4, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.1), show.legend = FALSE) +
+  geom_boxplot(outlier.shape = NA, size = 1.2)+
+  geom_jitter(shape=19, size=2, alpha=0.6, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.1), show.legend = FALSE) +
   labs(title=NULL, 
        x="Day",
        y="Number of Observed OTUs")+
   ylim(0, 100)+
   theme_classic()+
-  theme(legend.position = c(0.8, 0.2))
+  theme(legend.position = c(0.8, 0.2))+
+  theme(text = element_text(size = 16))  # Change font size for entire plot
 save_plot("results/figures/groups_richness.png", richness) #Use save_plot instead of ggsave because it works better with cowplot
 
 

### 2019-11-01 code/ppi_taxa.R

@@ -118,8 +118,8 @@ ppi_family_plot_neg7 <- agg_family_data %>%
   ggplot(aes(x= reorder(family, agg_rel_abund), y=agg_rel_abund, color=Group))+
   scale_colour_manual(values=color_scheme) +
   geom_hline(yintercept=1/3000, color="gray")+
-  geom_boxplot(outlier.shape = NA)+
-  geom_jitter(shape=19, size=1, alpha=0.4, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.2)) +
+  geom_boxplot(outlier.shape = NA, size = 1.2)+
+  geom_jitter(shape=19, size=2, alpha=0.6, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.2)) +
   labs(title=NULL, 
        x=NULL,
        y="Relative abundance (%)",
@@ -130,7 +130,8 @@ ppi_family_plot_neg7 <- agg_family_data %>%
   theme(plot.title=element_text(hjust=0.5))+
   theme(axis.text.y = element_text(face = "italic"))+ #Have the families show up as italics
   theme(plot.title=element_text(hjust=0.5))+
-  theme(legend.position = c(0.9, 0.2)) #Move legend position
+  theme(legend.position = c(0.85, 0.2)) + #Move legend position
+  theme(text = element_text(size = 16))  # Change font size for entire plot
 save_plot("results/figures/families_prev_assoc_w_PPIs_-7.png", ppi_family_plot_neg7, base_aspect_ratio = 2)
 
 # Figure 1D----
@@ -142,8 +143,8 @@ ppi_family_plot_day0 <- agg_family_data %>%
   ggplot(aes(x= reorder(family, agg_rel_abund), y=agg_rel_abund, color=Group))+
   scale_colour_manual(values=color_scheme) +
   geom_hline(yintercept=1/3000, color="gray")+
-  geom_boxplot(outlier.shape = NA)+
-  geom_jitter(shape=19, size=1, alpha=0.4, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.2)) +
+  geom_boxplot(outlier.shape = NA, size = 1.2)+
+  geom_jitter(shape=19, size=2, alpha=0.6, position=position_jitterdodge(dodge.width=0.7, jitter.width=0.2)) +
   labs(title=NULL, 
        x=NULL,
        y="Relative abundance (%)",
[…]

### 2019-11-04 submission/manuscript.tex
Incorporating Pat's edits for the response to reviewers

@@
 \section{\texorpdfstring{The proton pump inhibitor omeprazole does not
-promote \emph{Clostridium difficile} colonization in a murine
-model}{The proton pump inhibitor omeprazole does not promote Clostridium difficile colonization in a murine model}}\label{the-proton-pump-inhibitor-omeprazole-does-not-promote-clostridium-difficile-colonization-in-a-murine-model}
+promote \emph{Clostridioides difficile} colonization in a murine
+model}{The proton pump inhibitor omeprazole does not promote Clostridioides difficile colonization in a murine model}}\label{the-proton-pump-inhibitor-omeprazole-does-not-promote-clostridioides-difficile-colonization-in-a-murine-model}
 
@@
 Proton pump inhibitor (PPI) use has been associated with microbiota
-alterations and susceptibility to \emph{Clostridium}
-(\emph{Clostridioides}) \emph{difficile} infections (CDIs) in humans. We
-assessed how PPI treatment alters the fecal microbiota and whether
-treatment promotes CDIs in a mouse model. Mice receiving a PPI treatment
-were gavaged with 40 mg/kg of omeprazole during a 7-day pretreatment
-phase, the day of \emph{C. difficile} challenge, and the following 9
-days. We found that mice treated with omeprazole were not colonized by
-\emph{C. difficile}. When omeprazole treatment was combined with a
-single clindamycin treatment, one cage of mice remained resistant to
-\emph{C. difficile} colonization, while the other cage was colonized.
-Treating mice with only clindamycin followed by challenge resulted in
-\emph{C. difficile} colonization. 16S rRNA gene sequencing analysis
-revealed that omeprazole had minimal impact on the structure of the
-murine microbiota throughout the 16 days of omeprazole exposure. These
-results suggest omeprazole treatment alone is not sufficient to disrupt
-microbiota resistance to \emph{C. difficile} infection in mice that are
-normally resistant in the absence of antibiotic treatment.
+alterations and susceptibility to \emph{Clostridioides difficile}
+infections (CDIs) in humans. We assessed how PPI treatment alters the
+fecal microbiota and whether treatment promotes CDIs in a mouse model.
+Mice receiving a PPI treatment were gavaged with 40 mg/kg of omeprazole
+during a 7-day pretreatment phase, the day of \emph{C. difficile}
+challenge, and the following 9 days. We found that mice treated with
+omeprazole were not colonized by \emph{C. difficile}. When omeprazole
+treatment was combined with a single clindamycin treatment, one cage of
+mice remained resistant to \emph{C. difficile} colonization, while the
+other cage was colonized. Treating mice with only clindamycin followed
+by challenge resulted in \emph{C. difficile} colonization. 16S rRNA gene
+sequencing analysis revealed that omeprazole had minimal impact on the
+structure of the murine microbiota throughout the 16 days of omeprazole
+exposure. These results suggest omeprazole treatment alone is not
+sufficient to disrupt microbiota resistance to \emph{C. difficile}
+infection in mice that are normally resistant in the absence of
+antibiotic treatment.
 
@@
 
-Antibiotics are the primary risk factor for \emph{Clostridium}
-(\emph{Clostridioides}) \emph{difficile} infections (CDIs), but other
-factors may also increase a person's risk. In epidemiological studies,
-proton pump inhibitor (PPI) use has been associated with CDI incidence
-and recurrence. PPIs have also been associated with alterations in the
-human intestinal microbiota in observational and interventional studies.
-We evaluated the effects of the PPI omeprazole on the structure of the
-murine intestinal microbiota and its ability to disrupt colonization
-resistance to \emph{C. difficile}. We found omeprazole treatment had
-minimal impact on the murine fecal microbiota and did not promote
-\emph{C. difficile} colonization. Further studies are needed to
-determine whether other factors contribute to the association between
-PPIs and CDIs seen in humans or whether aspects of murine physiology may
-limit its utility to test these types of hypotheses.
+Antibiotics are the primary risk factor for \emph{Clostridioides
+difficile} infections (CDIs), but other factors may also increase a
+person's risk. In epidemiological studies, proton pump inhibitor (PPI)
+use has been associated with CDI incidence and recurrence. PPIs have
+also been associated with alterations in the human intestinal microbiota
+in observational and interventional studies. We evaluated the effects of
+the PPI omeprazole on the structure of the murine intestinal microbiota
+and its ability to disrupt colonization resistance to \emph{C.
+difficile}. We found omeprazole treatment had minimal impact on the
+murine fecal microbiota and did not promote \emph{C. difficile}
+colonization. Further studies are needed to determine whether other
+factors contribute to the association between PPIs and CDIs seen in
+humans or whether aspects of murine physiology may limit its utility to
+test these types of hypotheses.
 
@@
 Antibiotics have a large impact on the intestinal microbiome and are a
-primary risk factor for developing \emph{Clostridium}
-(\emph{Clostridioides}) \emph{difficile} infections (CDIs) (1). It is
-less clear whether other human medications that impact the microbiota
-also influence \emph{C. difficile} colonization resistance. Multiple
-epidemiological studies have suggested an association between proton
-pump inhibitor (PPI) use and incidence or recurrence of CDIs (2--5).
-There have also been a number of large cohort studies and interventional
-clinical trials that demonstrated specific alterations in the intestinal
-microbiome were associated with PPI use (4, 6). PPI-associated
-microbiota changes have been attributed to the ability of PPIs to
-increase stomach acid pH which may promote the survival of oral and
-pathogenic bacteria (4, 6). In human fecal samples, PPI use results in
-increases in \emph{Enterococcaceae}, \emph{Lactobacillaceae},
-\emph{Micrococcaceae}, \emph{Staphylococcaceae} and
-\emph{Streptococcaceae} and decreases in \emph{Ruminococcaceae} (6--9).
-Several of these taxa have also been associated with \emph{C. difficile}
-colonization in humans (10).\\
+primary risk factor for developing \emph{Clostridioides difficile}
+infections (CDIs) (1). It is less clear whether other human medications
+that impact the microbiota also influence \emph{C. difficile}
+colonization resistance. Multiple epidemiological studies have suggested
+an association between proton pump inhibitor (PPI) use and incidence or
+recurrence of CDIs (2--5). There have also been a number of large cohort
+studies and interventional clinical trials that demonstrated specific
+alterations in the intestinal microbiome were associated with PPI use
+(4, 6). PPI-associated microbiota changes have been attributed to the
+ability of PPIs to increase stomach acid pH which may promote the
+survival of oral and pathogenic bacteria (4, 6). In human fecal samples,
+PPI use results in increases in \emph{Enterococcaceae},
+\emph{Lactobacillaceae}, \emph{Micrococcaceae}, \emph{Staphylococcaceae}
+and \emph{Streptococcaceae} and decreases in \emph{Ruminococcaceae}
+(6--9). Several of these taxa have also been associated with \emph{C.
+difficile} colonization in humans (10).\\
 Unfortunately, the studies suggesting a link between PPIs and \emph{C.

### 2019-11-04 submission/response_to_reviewers.md

@@
 Alimentary Pharmacology & Therapeutics 2008).
+Our lab prefers using *C.
+difficile* 630 for experiments because we are interested in colonization and clearance.
+*C.
+difficile* VPI 10463 administration to mice results in a more agressive disease that leads to mortality.
+Using *C.
+difficile* 630 after clindamycin administration results in an intermediate phenotype, where the mice are colonized but able to clear infection within 10 days.
+The intermediate phenotype with *C.
+difficile* 630 allows us to examine whether the addition of a drug exacerbates or prevents infection and track the mice for a longer period of time.
 In addition to our lab's prior experience with the *C.
@@
 
-> We have submitted this manuscript as an observation, which is limited to 2 figures total.
+> We have submitted this manuscript as an Observation, which is limited to 2 figures total.
 Additionally, in response to Reviewer #2, we have expanded the size of all the panels in our figures, limiting our ability to include additional panels for the 2 main text figures.
@@
 
-> We appreciate the suggestion and have added "*Clostridium* (*Clostridioides*) *difficile*" to the abstract, importance, and introduction sections of our manuscript.
+> We appreciate the suggestion and have changed "*Clostridium*" to "*Clostridioides*" in the title, abstract, importance, and introduction sections of our manuscript.
 

### 2019-11-04 submission/manuscript.Rmd

@@
 
-# The proton pump inhibitor omeprazole does not promote *Clostridium difficile* colonization in a murine model
+# The proton pump inhibitor omeprazole does not promote *Clostridioides difficile* colonization in a murine model
 
@@
 ## Abstract
-Proton pump inhibitor (PPI) use has been associated with microbiota alterations and susceptibility to *Clostridium* (*Clostridioides*) *difficile* infections (CDIs) in humans.
+Proton pump inhibitor (PPI) use has been associated with microbiota alterations and susceptibility to *Clostridioides difficile* infections (CDIs) in humans.
 We assessed how PPI treatment alters the fecal microbiota and whether treatment promotes CDIs in a mouse model.
@@
 ## Importance
-Antibiotics are the primary risk factor for *Clostridium* (*Clostridioides*) *difficile* infections (CDIs), but other factors may also increase a person's risk.
+Antibiotics are the primary risk factor for *Clostridioides difficile* infections (CDIs), but other factors may also increase a person's risk.
 In epidemiological studies, proton pump inhibitor (PPI) use has been associated with CDI incidence and recurrence.
@@
 
-Antibiotics have a large impact on the intestinal microbiome and are a primary risk factor for developing *Clostridium* (*Clostridioides*) *difficile* infections (CDIs) [@Schubert2015].
+Antibiotics have a large impact on the intestinal microbiome and are a primary risk factor for developing *Clostridioides difficile* infections (CDIs) [@Schubert2015].
 It is less clear whether other human medications that impact the microbiota also influence *C.
